### 2022-03-07 README.Rmd
Initial commit

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Research Compendium Template

This is a research compendium template geared towards producing a manuscript output as a Word document reproducibly using the `targets` package for workflow management.
See the [`targets` documentation](https://docs.ropensci.org/targets/) for more information on how to program analyses in this style.
I've found that writing `paper.Rmd` using RStudio's [visual markdown editor](https://rstudio.github.io/visual-markdown-editing/) and automatic text wrapping (with each sentence as a new line) works best for version control and collaboration with git and GitHub.
The visual editor also integrates seamlessly with [Zotero](https://www.zotero.org/) for adding in-text citations and bibliographies.
The [`renv` package](https://rstudio.github.io/renv/articles/renv.html) my also be useful for collaboration to keep package version synced across collaborators.
Before submitting the manuscript, I like to [archive my GitHub repository with Zenodo](https://guides.github.com/activities/citable-code/) and add a DOI badge to the README, and a link to the Zenodo archive to the Data Availability Statement.
In the `review` directory, you'll find my preferred way of structuring response to reviewers complete with the ability to cross-reference responses to avoid repeating yourself.

**GithHub Actions** This template also uses [GitHub actions](https://github.com/features/actions) to automatically render a .html version of the manuscript every time an edit to docs/paper.Rmd is made.
If [GitHub pages](https://pages.github.com/) are active, then the rendered draft is viewable at \<username\>.github.io/\<repository name\>/paper.html.
The workflow for this lives in .github/workflows/build-manuscript.yaml and can be disabled on GitHub in the "Actions" tab.

Readme text for *your* repository below:

------------------------------------------------------------------------

<!-- badges: start -->

<!-- DOI badges go here -->

<!-- badges: end -->

This repository contains data and code for a manuscript by \_\_\_\_\_\_ written for submission to \_\_\_\_\_\_ tentatively titled: \_\_\_\_\_.

The most recent draft of the manuscript is available [here](https://aariq.github.io/paper-template/paper.html).

# Reproducibility

This research compendium has been developed using the statistical programming language R.
To work with the compendium, you will need installed on your computer the [R software](https://cloud.r-project.org/) itself and [RStudio Desktop](https://rstudio.com/products/rstudio/download/).

You can download the compendium as a zip file from from this URL: [master.zip](/archive/master.zip).

To run the compendium and reproduce all outputs:

-   Open the project in RStudio by double-clicking the `.Rproj` file.
-   Install packages listed in `packages.R` (if this compendium uses `renv`, this may happen automatically or with `renv::restore()`).
-   Run scripts in /R/ in numerical order.
-   Knit /doc/paper.Rmd to produce output

Additional instructions if project uses `targets`:

-   Install the [`targets` package](https://docs.ropensci.org/targets/).
-   Run `targets::tar_make()` or `targets::tar_make_clustermq()` from the R console to run all code and produce all outputs.

**NOTE**: the GitHub actions in .github/workflows/build-manuscript.yaml will not work with a `targets` project.
Consider `targets::tar_github_actions()` as an alternative

### 2022-03-07 docs/paper.Rmd
Initial commit

---
title: "Full Title"
shorttitle: "Running Title"
author:
  - Eric R. Scott:
      email: scottericr@gmail.com
      institute: [UF_WEC]
      correspondence: true
  - Co-author two:
      email: secondperson@university.edu
      institute: [affiliation2]
      correspondence: false
  - Co-author three:
      email: thirdperson@university.edu
      institute: [UF_WEC, affiliation2]
      correspondence: false

institute:
  - UF_WEC: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
  - affiliation2: Department of Ecology, Evolution and Environmental Biology, Example University 123 Center Avenue, New York, New York 10027 USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      linenumbers: yes
      number_sections: no
      fig_caption: yes
      reference_docx: "templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=templates/scholarly-metadata.lua
      - --lua-filter=templates/author-info-blocks.lua
      #- -Fpandoc-crossref #required for cross-ref equations
linkcolor: blue
bibliography: references.bib
csl: "templates/global-change-biology.csl" # Insert path for the bib-style
keywords: |
  keyword1; keyword2; keyword3
highlights: |
  These are the highlights

editor_options:
  markdown:
    wrap: sentence
    canonical: true
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Running title: `r rmarkdown::metadata$shorttitle`

<!-- Highlights: `r rmarkdown::metadata$highlights` -->

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.path = "figures/",
	dev = "png",
	dpi = 300,
	message = FALSE,
	warning = FALSE,
	collapse = TRUE,
	comment = "#>"
)
library(targets)
```

<!-- The actual document text starts here: -->

\pagebreak

# Abstract

Abstract text.

\pagebreak

# Introduction

Introduction with an in-text citation [@scottInteractiveEffectsDrought2019].

# Methods

Methods text.

# Results

A reference to Table \@ref(tab:example-table).
A reference to Figure \@ref(fig:example-figure).

# Discussion

# Acknowledgments

We thank **,** \_, \_\_\_ and \_\_\_ anonymous reviewers for helpful discussions and comments on the manuscript.
Financial support was provided by the U.S.
National Science Foundation (awards \_\_\_\_, and \_\_\_\_).
The authors delcare no conflicts of interest.

# Data Availability Statement

Data and R code used in this study are archived with Zenodo at <doi url>.

\pagebreak

# Tables

```{r}
library(pander)
```

```{r example-table, results='asis'}
pandoc.table(
  head(iris),
  justify = c("right", "right", "right", "right", "left"),
  full_width = FALSE,
  caption = "(\\#tab:example-table) A table must include the table tag in order to cross-reference it correctly."
)
```

# Figures

```{r}
#| example-figure,
#| fig.cap = "A caption for an example figure. It can be long and span mutiple
#|   lines like this."
plot(cars)
```

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

### 2022-03-07 docs/templates/template_maker.Rmd
Initial commit

---
title: "My title"
author:
- author 1
- other author
date: '`r Sys.Date()`'
output: word_document
abstract: "My abstract \n"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Knit this doc, and edit the styles in the resulting docx file. Then rename as 'template' and use this in the yml:

```
output:
  word_document:
    reference_docx: templates/template.docx
```

# Heading 1

## Heading 2

### Heading 3

#### Heading 4

\pagebreak

```{r fig.cap="This is a figure caption. *italics*, **bold**, ~subscript~, ^superscript^ and $y = mx + \\beta$"}
plot(cars)
```

\pagebreak

```{r results='asis'}
library(pander)
pandoc.table(head(iris),
             caption = "This is a table caption for a markdown table made with `pander`. *italics*, **bold**, ~subscript~, ^superscript^ and $y = mx + \\beta$")
```

To edit table style in the template, click on the table, choose "Table Design" from the ribbon,
click the little tab to pull down the table style drawer, click "Modify Table Style..." and modify the style called "Table".  You may also want to modify the text style used (default "Compact") for the table.

### 2022-03-07 R/1-wrangle-data.R

@@ -1,0 +1,1 @@
+#code for wrangling data

### 2022-03-07 R/2-analysis.R

@@ -1,0 +1,3 @@
+# code for analysis, figures and tables
+# if figures and table code is big, maybe split into a third R script.
+# separate multiple analyses into multiple R scripts

### 2022-03-07 README.Rmd
customize templates

@@
 
-# Research Compendium Template
-
-This is a research compendium template geared towards producing a manuscript output as a Word document reproducibly using the `targets` package for workflow management.
-See the [`targets` documentation](https://docs.ropensci.org/targets/) for more information on how to program analyses in this style.
-I've found that writing `paper.Rmd` using RStudio's [visual markdown editor](https://rstudio.github.io/visual-markdown-editing/) and automatic text wrapping (with each sentence as a new line) works best for version control and collaboration with git and GitHub.
-The visual editor also integrates seamlessly with [Zotero](https://www.zotero.org/) for adding in-text citations and bibliographies.
-The [`renv` package](https://rstudio.github.io/renv/articles/renv.html) my also be useful for collaboration to keep package version synced across collaborators.
-Before submitting the manuscript, I like to [archive my GitHub repository with Zenodo](https://guides.github.com/activities/citable-code/) and add a DOI badge to the README, and a link to the Zenodo archive to the Data Availability Statement.
-In the `review` directory, you'll find my preferred way of structuring response to reviewers complete with the ability to cross-reference responses to avoid repeating yourself.
-
-**GithHub Actions** This template also uses [GitHub actions](https://github.com/features/actions) to automatically render a .html version of the manuscript every time an edit to docs/paper.Rmd is made.
-If [GitHub pages](https://pages.github.com/) are active, then the rendered draft is viewable at \<username\>.github.io/\<repository name\>/paper.html.
-The workflow for this lives in .github/workflows/build-manuscript.yaml and can be disabled on GitHub in the "Actions" tab.
-
-Readme text for *your* repository below:
-
-------------------------------------------------------------------------
-
 <!-- badges: start -->
@@
 
-This repository contains data and code for a manuscript by \_\_\_\_\_\_ written for submission to \_\_\_\_\_\_ tentatively titled: \_\_\_\_\_.
+This repository contains data and code for a manuscript by Eric Scott and Emilio Bruna written for submission to The American Naturalist tentatively titled: \_\_\_\_\_.
 
-The most recent draft of the manuscript is available [here](https://aariq.github.io/paper-template/paper.html).
+The most recent draft of the manuscript is available [here](https://brunalab.github.io/lagged-ipms/paper.html).
 
@@
 
-Additional instructions if project uses `targets`:
+<!-- Additional instructions if project uses `targets`: -->
 
--   Install the [`targets` package](https://docs.ropensci.org/targets/).
--   Run `targets::tar_make()` or `targets::tar_make_clustermq()` from the R console to run all code and produce all outputs.
+<!-- -   Install the [`targets` package](https://docs.ropensci.org/targets/). -->
 
-**NOTE**: the GitHub actions in .github/workflows/build-manuscript.yaml will not work with a `targets` project.
-Consider `targets::tar_github_actions()` as an alternative
+<!-- -   Run `targets::tar_make()` or `targets::tar_make_clustermq()` from the R console to run all code and produce all outputs. -->
+
+<!-- **NOTE**: the GitHub actions in .github/workflows/build-manuscript.yaml will not work with a `targets` project. -->
+
+<!-- Consider `targets::tar_github_actions()` as an alternative -->

### 2022-03-07 docs/paper.Rmd
customize templates

@@
       correspondence: true
-  - Co-author two:
-      email: secondperson@university.edu
-      institute: [affiliation2]
-      correspondence: false
-  - Co-author three:
-      email: thirdperson@university.edu
-      institute: [UF_WEC, affiliation2]
+  - Emilio M. Bruna:
+      email: embruna@ufl.edu
+      institute: [UF_WEC, UF_LAS, BDFFP]
       correspondence: false
@@
   - UF_WEC: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
-  - affiliation2: Department of Ecology, Evolution and Environmental Biology, Example University 123 Center Avenue, New York, New York 10027 USA
+  - UF_LAS: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530  
+  - BDFFP: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 bibliography: references.bib
-csl: "templates/global-change-biology.csl" # Insert path for the bib-style
+csl: "templates/the-american-naturalist.csl" # Insert path for the bib-style
 keywords: |

### 2022-03-07 docs/paper.Rmd
transfer notes from meeting with Emilio

@@
 ---
-title: "Full Title"
-shorttitle: "Running Title"
+title: "Context-dependent consequences of including lagged effects in demographic models"
+shorttitle: "Lagged Effects IPMs"
 author:
@@
 
-Introduction with an in-text citation [@scottInteractiveEffectsDrought2019].
+1.  Demographic models are widely used for all kinds of stuff
+
+2.  It has long been recognized that there is the potential for lagged effects.
+    Now there's evidence for that [@scottInteractiveEffectsDrought2019].
+
+3.  Lagged effects could be different in different habitat types (and this could be a big reason for differences between habitats)
 

### 2022-03-07 docs/paper.Rmd
rough out some text

@@
 2.  It has long been recognized that there is the potential for lagged effects.
-    Now there's evidence for that [@scottInteractiveEffectsDrought2019].
+    Now there's evidence for that [@eversLaggedDormantSeason2021; @scottDelayedEffectsClimate2022].
 
@@
 
-Methods text.
+## Modeling vital rates
+
+Used 10 yr demographic dataset for modeling post-seedling survival, growth, and flowering; seedling survival and growth.
+For post-seedlings, all vital rates were modeled as a smooth function of log(size_t) using a penalized spline in generalized additive models (GAMs) using the `mgcv` package [@mgcv].
+
+-   Where the rest of the fecundity data comes from
+
+For stochastic IPMs, we took two approaches.
+In the first approach, we added a random intercept for year to all post-seedling vital rates GAMs, when supported by the data (i.e.
+p-value for random effect \< 0.05) <!--# something about how you can actually use p-values for random effects fit with mgcv -->.
+Then, we used kernel-resampled ipms to generate kernels for each year of data and iterate the IPM by drawing from these randomly.
+Equivalent to "matrix shuffling" approach in matrix population models (CITATION).
+In the second approach, we modeled the impacts of the environment on vital rates explicitly.
+We calculated the standardized precipitation evapotranspiraion index (SPEI).
+We modeled effects of SPEI on post-seedling survival, growth, and flowering probability using distributed lag non-linear models fit as two dimensional tensor product splines [@scottDelayedEffectsClimate2022].
+Then these models were incorporated into a parameter-resampled IPM.
+To iterate such an IPM, a random sequence of SPEI values is created by re-sampling the observed monthly SPEI data.
+Then, 36 month lags are calculated for each year starting in February (the month of the demographic census).
+These values are then used to iterate the IPM.
+
+IPMs built and iterated with the @ipmr package.
+
+Use fitted_samples() for bootsrapping CIs around lambda?
 
 # Results
+
+-   table of vital rate models.
+    Which vital rate, which IPM, ∆AIC, model edf.
+
+-   table of lambdas.
+    2 habitats x 3 (4?) IPMs
+
+-   figure of PVA??
+    Conceptual diagram instead?
 
@@
 We thank **,** \_, \_\_\_ and \_\_\_ anonymous reviewers for helpful discussions and comments on the manuscript.
+We thank Sam Levin for his help with the `ipmr` package.
 Financial support was provided by the U.S.
 National Science Foundation (awards \_\_\_\_, and \_\_\_\_).
-The authors delcare no conflicts of interest.
+The authors declare no conflicts of interest.
 

### 2022-03-07 R/00-functions.R
copy over data and wrangling code

@@ -1,0 +1,37 @@
+#' Convert demographic survey data into binary survival
+#' 
+#' This function might only make sense to use if you continue surveying for
+#' plants long after you plan to count them as dead.  It is convenient if you
+#' have many NA entries after a plant has died (or is assumed dead) and you want
+#' to know which years it was alive and which it was dead.  You want to ignore
+#' NAs that are not trailing and you want to ignore trailing NAs < n
+#'
+#' @param x any ordered numeric vector such as height, size, or number of shoots, that is collected every observation unless plants are not detected.
+#' @param n number of successive, trailing missing survey points required to consider an organism dead.
+#'
+#' @return a logical vector with TRUE = alive and FALSE = assumed dead
+#' @examples
+#' #Assume plants dead with after not detected for 2 years
+#' dead  <- c(1, 3, NA, NA, 1, 2, 1, NA, NA)
+#' alive <- c(1, 3, NA, NA, 1, 2, 1, 3, NA)
+#' as_living(dead)
+#' as_living(alive)
+as_living <- function(x, n = 2) {
+  
+  #if at least last 2 values are NA
+  trail <- (length(x) - n + 1):length(x)
+  #exclude negative indexes
+  trail <- trail[trail >= 0]
+  
+  if(all(is.na(x))){
+    alive <- rep(NA, length(x))
+  } else if (all(is.na(x[trail]))) {
+    #find the last non-NA value
+    y <- cumsum(!is.na(x))
+    last_real <- which.max(y == max(y))
+    alive <- c(rep(TRUE, last_real), rep(FALSE, length(x) - last_real))
+  } else {
+    alive <- rep(TRUE, length(x))
+  }
+  return(alive)
+}

### 2022-03-07 R/1-wrangle-data.R

@@ -1,1 +1,118 @@
-#code for wrangling data
+library(tidyverse)
+library(janitor)
+library(here)
+library(SPEI)
+library(tsibble)
+library(lubridate)
+library(tsModel)
+source(here("R", "00-functions.R"))
+
+
+# wrangle demography data -------------------------------------------------
+
+ha_raw <- read_csv(here::here("data", "Ha_survey_pre_submission.csv"), 
+                   col_types = cols(notes = col_character()))
+#clean colnames
+ha <- 
+  ha_raw %>% 
+  clean_names() %>% 
+  #remove unused cols
+  select(-tag_number, -row, -column, -x_09, -y_09, -duplicate_tag, -plot_id, -survey_status, -treefall_status, -adult_recruit) %>% 
+  #create logical flowering column
+  mutate(flwr = !is.na(infl) & infl > 0, .after = infl) %>% 
+  #create logical seedling column
+  mutate(sdlg = !is.na(sdlg_status), .before = code) %>% 
+  select(-sdlg_status) %>% 
+  #remove plants with 0 shoots or height (categorically different than small plant)
+  filter(ht > 0 | is.na(ht), shts > 0 | is.na(shts)) %>% 
+  #calculate log(size)
+  mutate(log_size = log(shts * ht), .after = ht) %>% 
+  #try to separate true NAs for flwr from what should be 0s
+  mutate(flwr = if_else(is.na(log_size), NA, flwr)) %>% 
+  #create column for survival
+  group_by(ha_id_number) %>% 
+  mutate(surv = as_living(ht, n = 3), .after = sdlg) %>% 
+  mutate(surv = if_else(!is.na(code) & str_detect(code, "dead"), FALSE, surv)) %>% 
+  filter(surv | lag(surv)) %>%  #remove entries after last year alive.
+  filter(pmin(cumsum(!is.na(shts))) != 0) %>%  #remove leading NAs.
+  ungroup() %>% 
+  
+  #create some columns with info about the previous timestep
+  group_by(ha_id_number) %>% 
+  mutate(log_size_prev = lag(log_size), .after = log_size) %>% 
+  mutate(sdlg_prev = case_when(
+    lag(sdlg) ~ TRUE, #if seedling prev timestep, use TRUE
+    sdlg ~ NA, #if current seedling, use NA
+    !lag(sdlg) ~ FALSE #otehrwise use FALSE
[…]

### 2022-03-07 R/2-analysis.R
copy over vital rates functions

@@ -1,3 +1,0 @@
-# code for analysis, figures and tables
-# if figures and table code is big, maybe split into a third R script.
-# separate multiple analyses into multiple R scripts

### 2022-03-07 R/2-vital-rates.R

@@ -1,0 +1,380 @@
+#' Here I'll fit the vital rates models for deterministic, kernel resampled, and parameter resampled IPM
+#' 
+#' 
+library(tidyverse)
+library(mgcv)
+library(gratia)
+library(broom)
+
+
+# read and separate data --------------------------------------------------
+
+ha <- read_rds("data/data_cleaned.rds")
+ha_ff <- ha %>% filter(sdlg_prev == FALSE, habitat == "1-ha") 
+ha_cf <- ha %>% filter(sdlg_prev == FALSE, habitat == "CF")
+ha_sdlg_ff <- ha %>% filter(sdlg_prev == TRUE, habitat == "1-ha")
+ha_sdlg_cf <- ha %>% filter(sdlg_prev == TRUE, habitat == "CF")
+
+
+# Simple -----------------------------------------------------------
+
+## Survival --------------------------------------------------------------
+
+
+surv_det_ff <- gam(
+  surv ~ s(log_size_prev, bs = "cr", k = 15),
+  family = binomial,
+  data = ha_ff,
+  method = "REML"
+)
+# appraise(surv_det_ff, method = "simulate")
+# gam.check(surv_det_ff, k.rep = 999)
+
+surv_det_cf <- gam(
+  surv ~ s(log_size_prev, bs = "cr", k = 15),
+  family = binomial,
+  data = ha_cf,
+  method = "REML"
+)
+# appraise(surv_det_cf, method = "simulate")
+# gam.check(surv_det_cf, k.rep = 999)
+
+
+## Size ------------------------------------------------------------------
+
+size_det_ff <- gam(
+  log_size ~ s(log_size_prev, bs = "cr", k = 20),
+  family = scat,
+  data = ha_ff %>% filter(surv), #only surviving plants get to grow
+  method = "REML"
+)
+# appraise(size_det_ff)
+# gam.check(size_det_ff, k.rep = 999) #no diff between k = 20 and k = 25
+
+size_det_cf <- gam(
+  log_size ~ s(log_size_prev, bs = "cr", k = 20),
+  family = scat,
+  data = ha_cf %>% filter(surv), #only surviving plants get to grow
+  method = "REML"
+)
+# appraise(size_det_cf)
[…]

### 2022-03-08 README.Rmd
switch to using targets

@@
 -   Install packages listed in `packages.R` (if this compendium uses `renv`, this may happen automatically or with `renv::restore()`).
--   Run scripts in /R/ in numerical order.
--   Knit /doc/paper.Rmd to produce output
+-   Install the [`targets` package](https://docs.ropensci.org/targets/).
+-   Run `targets::tar_make()` or `targets::tar_make_clustermq()` from the R console to run all code and produce all outputs.
 
-<!-- Additional instructions if project uses `targets`: -->
-
-<!-- -   Install the [`targets` package](https://docs.ropensci.org/targets/). -->
-
-<!-- -   Run `targets::tar_make()` or `targets::tar_make_clustermq()` from the R console to run all code and produce all outputs. -->
-
-<!-- **NOTE**: the GitHub actions in .github/workflows/build-manuscript.yaml will not work with a `targets` project. -->
-
-<!-- Consider `targets::tar_github_actions()` as an alternative -->
+<!-- Consider using `targets::tar_github_actions()` if targets aren't too big -->

### 2022-03-08 R/read_wrangle_clim.R

@@ -1,0 +1,32 @@
+read_wrangle_clim <- function(path, scale = 3, maxlag = 36) {
+  xa_raw <- read_csv(here("data", "xavier_daily_0.25x0.25.csv"))
+  
+  # Calculate SPEI
+  xa_raw %>% 
+    unite(latlon, lat, lon) %>% 
+    mutate(year = year(date),
+           month = month(date)) %>% 
+    group_by(latlon, year, month) %>% 
+    #convert to monthly
+    summarize(precip = sum(precip), eto = sum(eto)) %>% 
+    mutate(date = make_date(year = year, month = month, day = 15),
+           .after = year) %>% 
+    group_by(latlon) %>% 
+    mutate(spei = as.numeric(
+      spei(
+        ts(precip - eto, freq = 12, start = c(year(min(date)), month(min(date)))),
+        scale = scale
+      )$fitted
+    )) %>% 
+    
+    #create lags
+    
+    group_by(latlon) %>% 
+    mutate(spei_history = tsModel::Lag(spei, 0:maxlag),
+           L = matrix(0:maxlag, nrow = n(), ncol = maxlag+1, byrow = TRUE)) %>% 
+    separate(latlon, into = c("lat", "lon"), sep = "_") %>% 
+    #clean up 
+    select(-month) %>% 
+    ungroup()
+  
+}

### 2022-03-08 R/read_wrangle_demog.R

@@ -1,0 +1,43 @@
+read_wrangle_demog <- function(path) {
+  ha_raw <- read_csv(here::here("data", "Ha_survey_pre_submission.csv"), 
+                     col_types = cols(notes = col_character()))
+  #clean colnames
+  ha_raw %>% 
+    clean_names() %>% 
+    #remove unused cols
+    select(-tag_number, -row, -column, -x_09, -y_09, -duplicate_tag, -plot_id, -survey_status, -treefall_status, -adult_recruit) %>% 
+    #create logical flowering column
+    mutate(flwr = !is.na(infl) & infl > 0, .after = infl) %>% 
+    #create logical seedling column
+    mutate(sdlg = !is.na(sdlg_status), .before = code) %>% 
+    select(-sdlg_status) %>% 
+    #remove plants with 0 shoots or height (categorically different than small plant)
+    filter(ht > 0 | is.na(ht), shts > 0 | is.na(shts)) %>% 
+    #calculate log(size)
+    mutate(log_size = log(shts * ht), .after = ht) %>% 
+    #try to separate true NAs for flwr from what should be 0s
+    mutate(flwr = if_else(is.na(log_size), NA, flwr)) %>% 
+    #create column for survival
+    group_by(ha_id_number) %>% 
+    mutate(surv = as_living(ht, n = 3), .after = sdlg) %>% 
+    mutate(surv = if_else(!is.na(code) & str_detect(code, "dead"), FALSE, surv)) %>% 
+    filter(surv | lag(surv)) %>%  #remove entries after last year alive.
+    filter(pmin(cumsum(!is.na(shts))) != 0) %>%  #remove leading NAs.
+    ungroup() %>% 
+    
+    #create some columns with info about the previous timestep
+    group_by(ha_id_number) %>% 
+    mutate(log_size_prev = lag(log_size), .after = log_size) %>% 
+    mutate(sdlg_prev = case_when(
+      lag(sdlg) ~ TRUE, #if seedling prev timestep, use TRUE
+      sdlg ~ NA, #if current seedling, use NA
+      !lag(sdlg) ~ FALSE #otehrwise use FALSE
+    ), .after = sdlg) %>% 
+    ungroup() %>% 
+    #set column types
+    mutate(across(c("habitat", "ranch", "plot", "ha_id_number"), factor)) %>% 
+    mutate(year_fct = factor(year), .after = year) %>% 
+    #not using 10-ha data
+    filter(habitat %in% c("CF", "1-ha"))
+  # head(ha)
+}

### 2022-03-08 _targets.R

@@ -1,0 +1,16 @@
+## Load your packages, e.g. library(targets).
+source("./packages.R")
+
+## Load your R files
+lapply(list.files("./R", full.names = TRUE), source)
+
+## tar_plan supports drake-style targets and also tar_target()
+tar_plan(
+
+  tar_file(demog_file, here("data", "Ha_survey_pre_submission.csv")),
+  tar_file(clim_file, here("data", "xavier_daily_0.25x0.25.csv")),
+  demog = read_wrangle_demog(demog_file),
+  clim = read_wrangle_clim(clim_file),
+  data_full = join_data(demog, clim)
+
+) 

### 2022-03-08 packages.R

@@ -1,0 +1,18 @@
+## library() calls go here
+library(conflicted)
+library(dotenv)
+library(targets)
+library(tarchetypes)
+library(tidyverse)
+library(janitor)
+library(here)
+library(SPEI)
+library(tsibble)
+library(lubridate)
+library(tsModel)
+library(mgcv)
+library(gratia)
+library(broom)
+
+conflict_prefer("filter", "dplyr")
+conflict_prefer("lag", "dplyr")

### 2022-03-08 R/1-wrangle-data.R
use targets

@@ -1,118 +1,0 @@
-library(tidyverse)
-library(janitor)
-library(here)
-library(SPEI)
-library(tsibble)
-library(lubridate)
-library(tsModel)
-source(here("R", "00-functions.R"))
-
-
-# wrangle demography data -------------------------------------------------
-
-ha_raw <- read_csv(here::here("data", "Ha_survey_pre_submission.csv"), 
-                   col_types = cols(notes = col_character()))
-#clean colnames
-ha <- 
-  ha_raw %>% 
-  clean_names() %>% 
-  #remove unused cols
-  select(-tag_number, -row, -column, -x_09, -y_09, -duplicate_tag, -plot_id, -survey_status, -treefall_status, -adult_recruit) %>% 
-  #create logical flowering column
-  mutate(flwr = !is.na(infl) & infl > 0, .after = infl) %>% 
-  #create logical seedling column
-  mutate(sdlg = !is.na(sdlg_status), .before = code) %>% 
-  select(-sdlg_status) %>% 
-  #remove plants with 0 shoots or height (categorically different than small plant)
-  filter(ht > 0 | is.na(ht), shts > 0 | is.na(shts)) %>% 
-  #calculate log(size)
-  mutate(log_size = log(shts * ht), .after = ht) %>% 
-  #try to separate true NAs for flwr from what should be 0s
-  mutate(flwr = if_else(is.na(log_size), NA, flwr)) %>% 
-  #create column for survival
-  group_by(ha_id_number) %>% 
-  mutate(surv = as_living(ht, n = 3), .after = sdlg) %>% 
-  mutate(surv = if_else(!is.na(code) & str_detect(code, "dead"), FALSE, surv)) %>% 
-  filter(surv | lag(surv)) %>%  #remove entries after last year alive.
-  filter(pmin(cumsum(!is.na(shts))) != 0) %>%  #remove leading NAs.
-  ungroup() %>% 
-  
-  #create some columns with info about the previous timestep
-  group_by(ha_id_number) %>% 
-  mutate(log_size_prev = lag(log_size), .after = log_size) %>% 
-  mutate(sdlg_prev = case_when(
-    lag(sdlg) ~ TRUE, #if seedling prev timestep, use TRUE
-    sdlg ~ NA, #if current seedling, use NA
-    !lag(sdlg) ~ FALSE #otehrwise use FALSE
-    ), .after = sdlg) %>% 
[…]

### 2022-03-08 R/join_data.R

@@ -1,0 +1,23 @@
+join_data <- function(demog, clim) {
+  demog <- demog %>% 
+    mutate(lon = case_when(
+      plot %in% c("Florestal-CF", "5752", "5751", "5750", "5756", "CaboFrio-CF") ~ "-59.875",
+      plot %in% c("2206", "2108", "2107", "Dimona-CF", "5753", "PortoAlegre-CF") ~ "-60.125",
+      TRUE ~ NA_character_
+    ), .before = plot)
+  
+  #filter climate data so only months with census (february)
+  clim_sub <- 
+    clim %>% 
+    filter(month(date) == 2) %>% 
+    #remove burnin for SPEI
+    rowwise() %>% 
+    filter(!any(is.na(spei_history)))
+  
+  #join and clean up
+  df_full <-
+    left_join(demog, clim_sub, by = c("lon", "year")) %>% 
+    select(-lon, -lat)
+  #return:
+  df_full
+}

### 2022-03-08 R/2-vital-rates.R
merge

@@ -1,380 +1,380 @@
-#' Here I'll fit the vital rates models for deterministic, kernel resampled, and parameter resampled IPM
+#' #' Here I'll fit the vital rates models for deterministic, kernel resampled, and parameter resampled IPM
+#' #' 
+#' #' 
+#' library(tidyverse)
+#' library(mgcv)
+#' library(gratia)
+#' library(broom)
 #' 
 #' 
-library(tidyverse)
-library(mgcv)
-library(gratia)
-library(broom)
-
-
-# read and separate data --------------------------------------------------
-
-ha <- read_rds("data/data_cleaned.rds")
-ha_ff <- ha %>% filter(sdlg_prev == FALSE, habitat == "1-ha") 
-ha_cf <- ha %>% filter(sdlg_prev == FALSE, habitat == "CF")
-ha_sdlg_ff <- ha %>% filter(sdlg_prev == TRUE, habitat == "1-ha")
-ha_sdlg_cf <- ha %>% filter(sdlg_prev == TRUE, habitat == "CF")
-
-
-# Simple -----------------------------------------------------------
-
-## Survival --------------------------------------------------------------
-
-
-surv_det_ff <- gam(
-  surv ~ s(log_size_prev, bs = "cr", k = 15),
-  family = binomial,
-  data = ha_ff,
-  method = "REML"
-)
-# appraise(surv_det_ff, method = "simulate")
-# gam.check(surv_det_ff, k.rep = 999)
-
-surv_det_cf <- gam(
-  surv ~ s(log_size_prev, bs = "cr", k = 15),
-  family = binomial,
-  data = ha_cf,
-  method = "REML"
-)
-# appraise(surv_det_cf, method = "simulate")
-# gam.check(surv_det_cf, k.rep = 999)
-
-
-## Size ------------------------------------------------------------------
-
-size_det_ff <- gam(
-  log_size ~ s(log_size_prev, bs = "cr", k = 20),
-  family = scat,
-  data = ha_ff %>% filter(surv), #only surviving plants get to grow
-  method = "REML"
-)
-# appraise(size_det_ff)
-# gam.check(size_det_ff, k.rep = 999) #no diff between k = 20 and k = 25
-
-size_det_cf <- gam(
[…]

### 2022-03-08 R/vital_rates_dlnm.R

@@ -1,0 +1,80 @@
+#for checking knots, load data and run k.check() for the models
+# library(mgcv)
+# library(tidyverse)
+# data <- tar_read(data_full)
+
+surv_dlnm <- function(data, sdlg = FALSE, habitat = c("CF", "1-ha")) {
+  
+  habitat_choice <- match.arg(habitat)
+  
+  df <- data %>%
+    filter(sdlg_prev == sdlg, habitat == habitat_choice)
+  
+  gam(
+    surv ~ s(log_size_prev, bs = "cr", k = 10) +
+      te(spei_history, L, bs = "cr", k = c(15,15)),
+    family = binomial,
+    data = df,
+    method = "REML"
+  )
+}
+# k.check(surv_dlnm(data))
+# k.check(surv_dlnm(data, habitat = "1-ha"))
+
+
+size_dlnm <- function(data, sdlg = FALSE, habitat = c("CF", "1-ha")) {
+  
+  habitat_choice <- match.arg(habitat)
+  
+  df <- data %>%
+    dplyr::filter(sdlg_prev == sdlg, habitat == habitat_choice) %>% 
+    #only plants that survived get to grow
+    dplyr::filter(surv == 1, !is.na(log_size))
+  
+  gam(
+    log_size ~ s(log_size_prev, bs = "cr", k = 25) +
+      te(spei_history, L, bs = "cr", k = c(5,15)),
+    family = scat,
+    data = df,
+    method = "REML"
+  )
+}
+# k.check(size_dlnm(data))
+# k.check(size_dlnm(data, habitat = "1-ha"))
+
+
+size_sdlg_dlnm <- function(data, sdlg = TRUE, habitat = c("CF", "1-ha")) {
+  
+  habitat_choice <- match.arg(habitat)
+  
+  df <- data %>%
+    dplyr::filter(sdlg_prev == sdlg, habitat == habitat_choice) %>% 
+    dplyr::filter(surv == 1, !is.na(log_size))
+  
+  gam(
+    log_size ~ 1 + te(spei_history, L, bs = "cr", k = c(10, 15)),
+    family = scat,
+    data = df,
+    method = "REML"
+  )
+}
+#TODO: double-check k for this one
+# k.check(size_sdlg_dlnm(data))
+
+flwr_dlnm <- function(data, habitat = c("CF", "1-ha")) {
+  habitat_choice <- match.arg(habitat)
+  df <- data %>% 
[…]

### 2022-03-08 R/vital_rates_simple.R

@@ -1,0 +1,76 @@
+#for checking knots, load data and run k.check() for the models
+# library(mgcv)
+# library(tidyverse)
+# data <- tar_read(data_full)
+
+surv_det <- function(data, sdlg = FALSE, habitat = c("CF", "1-ha")) {
+
+  habitat_choice <- match.arg(habitat)
+  
+  df <- data %>%
+    filter(sdlg_prev == sdlg, habitat == habitat_choice)
+  
+  gam(
+    surv ~ s(log_size_prev, bs = "cr", k = 20),
+    family = binomial,
+    data = df,
+    method = "REML"
+  )
+}
+# k.check(surv_det(data))
+# k.check(surv_det(data, habitat = "1-ha"))
+
+
+size_det <- function(data, sdlg = FALSE, habitat = c("CF", "1-ha")) {
+  
+  habitat_choice <- match.arg(habitat)
+  
+  df <- data %>%
+    dplyr::filter(sdlg_prev == sdlg, habitat == habitat_choice) %>% 
+    #only plants that survived get to grow
+    dplyr::filter(surv == 1, !is.na(log_size))
+  
+  gam(
+    log_size ~ s(log_size_prev, bs = "cr", k = 20),
+    family = scat,
+    data = df,
+    method = "REML"
+  )
+}
+# k.check(size_det(data))
+# k.check(size_det(data, habitat = "1-ha")) #no diff between k = 20 and k = 25
+
+
+size_sdlg_det <- function(data, sdlg = TRUE, habitat = c("CF", "1-ha")) {
+  
+  habitat_choice <- match.arg(habitat)
+  
+  df <- data %>%
+    dplyr::filter(sdlg_prev == sdlg, habitat == habitat_choice) %>% 
+    dplyr::filter(surv == 1, !is.na(log_size))
+  
+  gam(
+    log_size ~ 1,
+    family = scat,
+    data = df,
+    method = "REML"
+  )
+}
+
+
+flwr_det <- function(data, habitat = c("CF", "1-ha")) {
+  habitat_choice <- match.arg(habitat)
+  df <- data %>% 
+    dplyr::filter(habitat == habitat_choice, sdlg == FALSE) %>% 
+    #current seedlings are excluded, but not plants that were seedlings in the previous year
+    dplyr::filter(surv == 1, !is.na(log_size))
+  
+  gam(
+    flwr ~ s(log_size_prev, bs = "cr", k = 15),
[…]

### 2022-03-08 R/vital_rates_stoch.R

@@ -1,0 +1,79 @@
+#for checking knots, load data and run k.check() for the models
+# library(mgcv)
+# library(tidyverse)
+# data <- tar_read(data_full)
+
+surv_raneff <- function(data, sdlg = FALSE, habitat = c("CF", "1-ha")) {
+  
+  habitat_choice <- match.arg(habitat)
+  
+  df <- data %>%
+    filter(sdlg_prev == sdlg, habitat == habitat_choice)
+  
+  gam(
+    surv ~ s(log_size_prev, bs = "cr", k = 20) +
+      s(year_fct, bs = "re"),
+    family = binomial,
+    data = df,
+    method = "REML"
+  )
+}
+# k.check(surv_raneff(data))
+# k.check(surv_raneff(data, habitat = "1-ha"))
+
+
+size_raneff <- function(data, sdlg = FALSE, habitat = c("CF", "1-ha")) {
+  
+  habitat_choice <- match.arg(habitat)
+  
+  df <- data %>%
+    dplyr::filter(sdlg_prev == sdlg, habitat == habitat_choice) %>% 
+    #only plants that survived get to grow
+    dplyr::filter(surv == 1, !is.na(log_size))
+  
+  gam(
+    log_size ~ s(log_size_prev, bs = "cr", k = 20) +
+      s(year_fct, bs = "re"),
+    family = scat,
+    data = df,
+    method = "REML"
+  )
+}
+# k.check(size_raneff(data))
+# k.check(size_raneff(data, habitat = "1-ha")) #no diff between k = 20 and k = 25
+
+
+size_sdlg_raneff <- function(data, sdlg = TRUE, habitat = c("CF", "1-ha")) {
+  
+  habitat_choice <- match.arg(habitat)
+  
+  df <- data %>%
+    dplyr::filter(sdlg_prev == sdlg, habitat == habitat_choice) %>% 
+    dplyr::filter(surv == 1, !is.na(log_size))
+  
+  gam(
+    log_size ~ 1 + s(year_fct, bs = "re"),
+    family = scat,
+    data = df,
+    method = "REML"
+  )
+}
+
+
+flwr_raneff <- function(data, habitat = c("CF", "1-ha")) {
+  habitat_choice <- match.arg(habitat)
+  df <- data %>% 
+    dplyr::filter(habitat == habitat_choice, sdlg == FALSE) %>% 
[…]

### 2022-03-08 _targets.R

@@ -11,6 +11,8 @@ tar_plan(
   tar_file(clim_file, here("data", "xavier_daily_0.25x0.25.csv")),
   demog = read_wrangle_demog(demog_file),
   clim = read_wrangle_clim(clim_file),
-  data_full = join_data(demog, clim)
-
+  data_full = join_data(demog, clim),
+  vit_list_det_ff = list(
+    surv = surv_det(data_full, habitat = "1-ha")
+  )
 ) 

### 2022-03-08 _targets.R
just checking

@@ -13,6 +13,7 @@ tar_plan(
   clim = read_wrangle_clim(clim_file),
   data_full = join_data(demog, clim),
   vit_list_det_ff = list(
-    surv = surv_det(data_full, habitat = "1-ha")
+    surv = surv_det(data_full, habitat = "1-ha"),
+    size = size_det(data_full, habitat = "1-ha")
   )
 ) 

### 2022-03-08 _targets.R
get targets to render manuscript

@@ -15,5 +15,6 @@ tar_plan(
   vit_list_det_ff = list(
     surv = surv_det(data_full, habitat = "1-ha"),
     size = size_det(data_full, habitat = "1-ha")
-  )
+  ),
+  tar_render(paper, here("docs", "paper.Rmd"), packages = "bookdown")
 ) 

### 2022-03-08 packages.R

@@ -13,6 +13,7 @@ library(tsModel)
 library(mgcv)
 library(gratia)
 library(broom)
+library(bookdown)
 
 conflict_prefer("filter", "dplyr")
 conflict_prefer("lag", "dplyr")

### 2022-04-15 _targets.R
finish code for bootstrapped lambdas

@@ -172,10 +172,8 @@ tar_plan(
     vit_list_stoch_ff
   ),
 
-# Lambdas -----------------------------------------------------------------
+# Bootstrapped Lambdas -----------------------------------------------------------------
 
-  # bootstrapping
-#TODO: add tar_rep() for stochastic and dlnm models
   tar_rep(
     lambda_bt_det_ff,
     ipm_boot_det(data_full, vit_other = vit_other_ff, habitat = "1-ha"),
@@ -204,15 +202,35 @@ tar_plan(
     reps = 100
   ),
 
-#TODO: edit this function to use `calc_ci()` to get estimate, lower, upper
-  lambda_tbl_df = make_lambda_tbl(
-    ipm_det_cf,
-    ipm_det_ff,
-    ipm_stoch_cf,
-    ipm_stoch_ff,
-    ipm_dlnm_cf,
-    ipm_dlnm_ff
+  tar_rep(
+    lambda_bt_dlnm_ff,
+    ipm_boot_dlnm(data_full, vit_other = vit_other_ff, habitat = "1-ha", clim = clim),
+    batches = 5,
+    reps = 100
   ),
+  
+  tar_rep(
+    lambda_bt_dlnm_cf,
+    ipm_boot_dlnm(data_full, vit_other = vit_other_cf, habitat = "CF", clim = clim),
+    batches = 5,
+    reps = 100
+  ),
+
+  tar_target(
+    lambda_df,
+    list(
+      det_ff = lambda_bt_det_ff,
+      det_cf = lambda_bt_det_cf,
+      stoch_ff = lambda_bt_stoch_ff,
+      stoch_cf = lambda_bt_stoch_cf,
+      dlnm_ff = lambda_bt_dlnm_ff,
+      dlnm_cf = lambda_bt_dlnm_cf
+    ) %>% 
+      map_df(calc_ci, .id = "model") %>% 
+      separate(model, into = c("ipm", "habitat")),
+    deployment = "main"
+  ),
+
   # Manuscript --------------------------------------------------------------
   tar_render(paper, here("docs", "paper.Rmd"), packages = "bookdown")
 ) %>% 

### 2022-04-15 notes/bootstrapping lambda.Rmd
consolidated notes

---
title: "bootstrapping lambda"
author: "Eric Scott"
date: '2022-04-14'
output: 
  html_document: 
    df_print: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(targets)
library(tidyverse)
library(progress)
library(tictoc)
library(ipmr)
library(metRology)
library(here)
library(rsample)
source(here("packages.R"))
lapply(list.files(here("R"), full.names = TRUE), source)
conflict_prefer("filter", "dplyr")
```

```{r}
tar_load(data_full)
tar_load(data_1998)
tar_load(data_2008)
```

# Bootstrapping individuals

Bootstrapping should mirror sampling methods.
So we want to know how the data would be different if different individuals were sampled.
We probably also want to stratify by plot?

So, within each plot, get the unique `ha_id_number`s, then sample from those with replacement and use to filter the full dataset.

```{r}
# View(head(data_full))
```

```{r}
set.seed(123)
boot_ids <-
  data_full %>% 
  group_by(habitat) %>% 
  summarize(ha_id_number = unique(ha_id_number)) %>% 
  group_by(habitat) %>% 
  sample_n(n(), replace = TRUE) %>% 
  #for validation:
  mutate(unique_id = paste(ha_id_number, row_number(), sep = "-"))
boot_ids %>% arrange(ha_id_number)

boot <- inner_join(data_full, boot_ids)
length(unique(data_full$ha_id_number)) == length(unique(boot$unique_id))
```

Nice!
That works.

So now just need to map an IPM mega function that fits the vital rate models, builds the IPMs, and outputs lambda.

```{r}
ipm_det <- function(boot, vit_other, habitat) {
   vit_list_det <- c(list(
    vit_surv = surv_det(boot, habitat = habitat),
    vit_size = size_det(boot, habitat = habitat),
    vit_flwr = flwr_det(boot, habitat = habitat),
    vit_size_sdlg = size_sdlg_det(boot, habitat = habitat),
    vit_surv_sdlg = surv_sdlg_det(boot, habitat = habitat)
  ), vit_other)

  #make IPM
  make_proto_ipm_det(vit_list_det) %>% 
    make_ipm(iterations = 100,  #only needs 100 to converge
             normalize_pop_size = TRUE,
             usr_funs = list(get_scat_params = get_scat_params)
    ) %>% 
    #calculate lambda
    ipmr::lambda(log = FALSE)
}
```

```{r}
tar_load(vit_other_ff)
```

```{r}
ipm_det(boot, vit_other_ff, habitat = "1-ha")
```

Nice!

### 2022-04-15 notes/bootstrapping_sketch.R

@@ -1,56 +1,0 @@
-library(targets)
-library(tidyverse)
-library(progress)
-library(tictoc)
-library(ipmr)
-library(metRology)
-source("packages.R")
-lapply(list.files("./R", full.names = TRUE), source)
-conflict_prefer("filter", "dplyr")
-
-tar_load(data_full)
-tar_load(data_1998)
-tar_load(data_2008)
-
-#start with just 1-ha data
-ha <- data_full %>% filter(habitat == "1-ha")
-
-# using rsample package ---------------------------------------------------
-library(rsample)
-
-#possibly works best with targets since it is a dataframe that we can iterate over with dynamic branching.
-
-ha_bt <- ha %>% bootstraps(10, apparent = TRUE) 
-ha_bt
-
-# IPM mega-function -------------------------------------------------------
-ipm_det <- function(boots, habitat, ...) {
-  splits <- boots[["splits"]] %>% map(analysis)
-  map_dbl(splits, ~{
-    
-    vit_list_det_ff <- list(
-      vit_surv = surv_det(.x, habitat = habitat),
-      vit_size = size_det(.x, habitat = habitat),
-      vit_flwr = flwr_det(.x, habitat = habitat),
-      vit_fruits = fruits_gam(data_1998),
-      vit_seeds = seeds_gam(data_1998, data_2008),
-      vit_germ_est = 0.018921527, #germination and establishment
-      vit_size_sdlg = size_sdlg_det(.x, habitat = habitat),
-      vit_surv_sdlg = surv_sdlg_det(.x, habitat = habitat)
-    )
-    
-    make_proto_ipm_det(vit_list_det_ff) %>% 
-      make_ipm(iterations = 100,  #only needs 100 to converge
-               normalize_pop_size = FALSE, # to run as PVA
-               usr_funs = list(get_scat_params = get_scat_params)) %>% 
-      ipmr::lambda(log = FALSE)
-  })
-}
-
-lambda_bt <- ipm_det(ha_bt, habitat = "1-ha")
-
-
-quantile(lambda_bt, c(0.025, 0.975))
-mean(lambda_bt)
-
-

### 2022-04-20 docs/paper.Rmd
get targets to render manuscript
add reminder to add CRediT statement
worked on methods section

--- 2022-03-07 docs/paper.Rmd
@@
 
-Methods text.
+## Demographic Data
+
+## Modeling vital rates
+
+Ultimately, we built and iterated three types of IPMs which each required different underlying vital rate model structure.
+For all models, we used the long term demographic dataset for modeling survival, growth, and flowering.
+For established plants, these three vital rates were modeled as a smooth function of size in the previous census using generalized additive models (GAMs) fit with the `mgcv` package[@mgcv] in R [@rcoreteam2020].
+Seedlings were considered a separate discrete class not structured by size.
+For consistency, seedling survival and growth were also modeled using GAMs, but without size in the previous census as a predictor.
+For growth models, a scaled t family distribution provided a better fit to the data than a gaussian fit as the residuals were leptokurtic with a simple gaussian model.
+
+To estimate reproduction we drew on additional data sources to estimate the number of fruits per flowering plant as a function of plant size, the number of seeds per fruit, and germination and establishment rates in continuous forest and forest fragments.
+<!--# Emilio: add more info here -->
+
+These models were used to build a general, density-independent, deterministic IPM with four sub-kernels: growth and survival, reproduction, probability of staying a seedling (always = 0), and seedling survival and establishment.
+<!--# include equation --> The seedling bank, or probability of staying a seedling, was always equal to zero, but was required <!--# for reasons? -->.
+
+For the second approach, we included environmental stochasticity by adding a random intercept for year to all vital rate models built with the long term demographic dataset.
+Then, we used these models to build general, density independent, stochastic, kernel-resampled IPMs.
+The kernel-resampling approach is to generate kernels for each year of data, using the random intercepts for year, and to iterate the IPM by drawing from these randomly.
+This is equivalent to "matrix shuffling" approach in matrix population models [@caswell2001] <!--# I'm not actually sure that Caswell calls it "matrix shuffling" -->.
+
+For the third method, we modeled the impacts of drought on vital rates explicitly.
+We calculated the standardized precipitation evapotranspiraton index (SPEI) for our site using a published gridded dataset based on ground measurements [@xavierDailyGriddedMeteorological2016].
+We modeled delayed effects of SPEI on survival, growth, and flowering probability using distributed lag non-linear models fit as two dimensional tensor product splines, with a maximum lag of 36 months[@scottDelayedEffectsClimate2022].
+Then these models were incorporated into a general, density independent, stochastic, parameter-resampled IPM.
+To iterate such an IPM, a random sequence of SPEI values is created by re-sampling the observed monthly SPEI data.
+Then, 36 month lags are calculated for each year starting in February (the month of the demographic census).
+These values are then used to predict fitted values from the vital rates models, generating different kernels at each iteration of the IPM.
+With this method, these kernels are not entirely independent because the SPEI from the previous two iterations is included in the vital rates underlying each kernel.
+
+All IPMs were constructed and iterated with the `ipmr` package in R [@ipmr].
+We iterated all IPMs for 1000 iterations and checked for convergence of the population growth rate, lambda.
+Stochastic growth rates ($\lambda_s$) were calculated as the mean of $ln(\lambda)$ from each iteration and back-transformed to be on the same scale as deterministic lambdas for comparison <!--# could replace entirely with an equation -->.
+To calculate 95% confidence intervals around the growth rates, we bootstrapped plants in the 10 year demographic dataset within each habitat type 1000 times.
+For each bootstrap, we then re-fit vital rates models (all except germination and establishment rate, fruits per flowering plant, and seeds per fruit, which were estimated using different datasets), constructed IPMs, and iterated them 1000 times to calculate lambda.
+
+This workflow was managed using the `targets` R package [@targets] which also allowed us to track computational time spent on each IPM for comparison.
 
 # Results
+
+For all vital rates estimated using the long term demographic dataset, the DLNM model fit the best.
+
+-   table of vital rate models.
+    Which vital rate, which IPM, ∆AIC, model edf.
+
+-   table of lambdas.
+    2 habitats x 3 (4?) IPMs
+
+-   figure of PVA??
+    Conceptual diagram instead?
 
@@
 We thank **,** \_, \_\_\_ and \_\_\_ anonymous reviewers for helpful discussions and comments on the manuscript.
+We thank Sam Levin for his help with the `ipmr` package.
 Financial support was provided by the U.S.
 National Science Foundation (awards \_\_\_\_, and \_\_\_\_).
-The authors delcare no conflicts of interest.
+The authors declare no conflicts of interest.
+
+# CRediT Statement
+
+<!--# https://casrai.org/credit/ -->
 

### 2022-04-20 docs/paper.Rmd
inserted AIC table and referenced in results

@@
 We calculated the standardized precipitation evapotranspiraton index (SPEI) for our site using a published gridded dataset based on ground measurements [@xavierDailyGriddedMeteorological2016].
-We modeled delayed effects of SPEI on survival, growth, and flowering probability using distributed lag non-linear models fit as two dimensional tensor product splines, with a maximum lag of 36 months[@scottDelayedEffectsClimate2022].
+We modeled delayed effects of SPEI on survival, growth, and flowering probability using distributed lag non-linear models fit as two dimensional tensor product splines, with a maximum lag of 36 months [@scottDelayedEffectsClimate2022].
 Then these models were incorporated into a general, density independent, stochastic, parameter-resampled IPM.
@@
 
-All IPMs were constructed and iterated with the `ipmr` package in R [@ipmr].
+All IPMs were constructed and iterated with the `ipmr` package in R [@levinIpmrFlexibleImplementation2021].
 We iterated all IPMs for 1000 iterations and checked for convergence of the population growth rate, lambda.
@@
 
-For all vital rates estimated using the long term demographic dataset, the DLNM model fit the best.
+For all vital rates estimated using the long term demographic dataset, the DLNM model fit the best (dAIC = 0) followed by the model with a random effect of year, followed by the deterministic model (Table \@ref(tab:aic))
 
@@
     Conceptual diagram instead?
-
-A reference to Table \@ref(tab:example-table).
-A reference to Figure \@ref(fig:example-figure).
 
@@
 
-```{r example-table, results='asis'}
+```{r aic, results='asis'}
 pandoc.table(
-  head(iris),
-  justify = c("right", "right", "right", "right", "left"),
+  tar_read(aic_tbl_df),
+  justify = c("left", "left", "left", "right", "right", "right"),
   full_width = FALSE,
-  caption = "(\\#tab:example-table) A table must include the table tag in order to cross-reference it correctly."
+  caption = "(\\#tab:aic) Comparison of vital rate models for deterministic, stochastic kernel-resampled, and stochastic parameter-resampled IPMs.
+'edf' is the estimated degrees of freedom of the penalized GAM."
 )

### 2022-04-20 R/vital_rates_simple.R
make everything possible identical between deterministic and stochastic models

@@ -49,7 +49,7 @@ flwr_det <- function(data, habitat = c("CF", "1-ha")) {
     dplyr::filter(surv == 1, !is.na(log_size))
   
   bam(
-    flwr ~ s(log_size_prev, bs = "cr", k = 15),
+    flwr ~ s(log_size_prev, bs = "cr", k = 20),
     family = binomial, 
     data = df,
     method = "fREML"

### 2022-04-20 R/vital_rates_stoch.R

@@ -69,11 +69,11 @@ size_sdlg_raneff <- function(data, sdlg = TRUE, habitat = c("CF", "1-ha")) {
     dplyr::filter(sdlg_prev == sdlg, habitat == habitat_choice) %>% 
     dplyr::filter(surv == 1, !is.na(log_size))
   
-  bam(
+  gam(
     log_size ~ 1 + s(year_fct, bs = "re"),
     family = scat,
     data = df,
-    method = "fREML"
+    method = "REML"
   )
 }
 
@@ -84,11 +84,11 @@ surv_sdlg_raneff <- function(data, sdlg = TRUE, habitat = c("CF", "1-ha")) {
   df <- data %>%
     filter(sdlg_prev == sdlg, habitat == habitat_choice)
   
-  bam(
+  gam(
     surv ~ 1 + s(year_fct, bs = "re"),
     family = binomial,
     data = df,
-    method = "fREML"
+    method = "REML"
   )
 }
 

### 2022-04-20 R/calc_ci.R
table of lambdas

@@ -1,5 +1,0 @@
-calc_ci <- function(x, alpha = 0.05) {
-  est <- mean(x)
-  ci <- quantile(x, c(0 + (alpha/2), 1 - (alpha/2)))
-  tibble(estimate = est, lower = ci[1], upper = ci[2])
-}

### 2022-04-20 R/make_lambda_table.R

@@ -1,0 +1,39 @@
+make_lambda_table <- function(ipm_list, bt_list) {
+
+#not very reusable, but that's fine.
+#ipm_list should look like this:
+  # ipm_list <- list(
+  #   det_ff = ipm_det_ff,
+  #   det_cf = ipm_det_cf,
+  #   stoch_ff = ipm_stoch_ff,
+  #   stoch_cf = ipm_stoch_cf
+  # )
+#bt_list should look like this:
+  # bt_list <- list(
+  #   det_ff = lambda_bt_det_ff,
+  #   det_cf = lambda_bt_det_cf,
+  #   stoch_ff = lambda_bt_stoch_ff,
+  #   stoch_cf = lambda_bt_stoch_cf
+  #   # dlnm_ff = lambda_bt_dlnm_ff,
+  #   # dlnm_cf = lambda_bt_dlnm_cf
+  # )
+
+  
+ lambdas <- 
+   ipm_list %>% 
+   map_df(~lambda(.x, log = FALSE) %>% set_names("lambda"), .id = "model")
+  
+  calc_ci <- function(x, alpha = 0.05) {
+    mu <- mean(x)
+    ci <- quantile(x, c(0 + (alpha/2), 1 - (alpha/2)))
+    tibble(mean = mu, lower = ci[1], upper = ci[2])
+  }
+  
+lambdas_bt <-
+  bt_list %>%
+  map_df(calc_ci, .id = "model") #TODO move calc_ci here
+
+full_join(lambdas, lambdas_bt, by = "model") %>%
+  separate(model, into = c("ipm", "habitat"))
+
+}

### 2022-04-20 _targets.R

@@ -9,7 +9,7 @@ options(
   clustermq.scheduler = "ssh",
   clustermq.template = "ssh_clustermq.tmpl", #custom SSH template to use R 4.0
   clustermq.ssh.host = "hpg", #set up in ~/.ssh/config.
-  clustermq.ssh.timeout = 30, #longer timeout helps with 2FA
+  clustermq.ssh.timeout = 60, #longer timeout helps with 2FA
   clustermq.ssh.log = "~/cmq_ssh.log" # log for easier debugging
 )
 
@@ -137,8 +137,7 @@ tar_plan(
   proto_ipm_dlnm_ff = make_proto_ipm_dlnm(vit_list_dlnm_ff),
   ipm_dlnm_ff = proto_ipm_dlnm_ff %>%
     make_dlnm_ipm(clim, seed = 1234, iterations = 1000,
-                  return_sub_kernels = FALSE, # don't save every iteration
-                  normalize_pop_size = FALSE, # to run as PVA
+                  return_main_env = FALSE, # don't save iteration mesh and other stuff
                   usr_funs = list(get_scat_params = get_scat_params)),
   
   ## continuous forest
@@ -155,9 +154,7 @@ tar_plan(
   proto_ipm_dlnm_cf = make_proto_ipm_dlnm(vit_list_dlnm_cf),
   ipm_dlnm_cf = proto_ipm_dlnm_cf %>%
     make_dlnm_ipm(clim, seed = 1234, iterations = 1000,
-                  return_sub_kernels = FALSE, # don't save every iteration
-                  normalize_pop_size = FALSE, # to run as PVA
-                  report_progress = TRUE,
+                  return_main_env = FALSE, # don't save iteration mesh and other stuff
                   usr_funs = list(get_scat_params = get_scat_params)),
   
   
@@ -204,33 +201,41 @@ tar_plan(
 
 
 # for these I use more batches, fewer reps because each rep is like an hour.  That way I can make incremental progress easier.
-  tar_rep(
-    lambda_bt_dlnm_ff,
-    ipm_boot_dlnm(data_full, vit_other = vit_other_ff, habitat = "1-ha", clim = clim),
-    batches = 500,
-    reps = 1
-  ),
-
-  tar_rep(
-    lambda_bt_dlnm_cf,
[…]

### 2022-04-20 docs/paper.Rmd
draft CRediT statement
add table of lambdas and reference to results text

@@
 
-For all vital rates estimated using the long term demographic dataset, the DLNM model fit the best (dAIC = 0) followed by the model with a random effect of year, followed by the deterministic model (Table \@ref(tab:aic))
+For all vital rates estimated using the long term demographic dataset, the DLNM model fit the best (dAIC = 0) followed by the model with a random effect of year, followed by the deterministic model (Table \@ref(tab:aic)).
 
--   table of vital rate models.
-    Which vital rate, which IPM, ∆AIC, model edf.
-
--   table of lambdas.
-    2 habitats x 3 (4?) IPMs
-
--   figure of PVA??
-    Conceptual diagram instead?
+Population growth rates were consistently higher in continuous forest compared to forest fragments across IPM types (Table \@ref(tab:lambdas)).
 
@@
 <!--# https://casrai.org/credit/ -->
+
+ERS contributed to the conceptualization, methodology, formal analysis, and led the writing of the original draft.
+EMB contributed to the conceptualization, methodology, and writing and also acquired funding.
 
@@
 
+```{r lambdas, results='asis'}
+pandoc.table(
+  tar_read(lambda_table),
+  digits = 5,
+  keep.trailing.zeros = TRUE,
+  justify = c("left", "left", "right", "right", "right", "right"),
+  full_width = FALSE,
+  caption = "(\\#tab:lambda) Population growth rates for continuous forest (cf) and forest fragments (ff) under different kinds of IPMs with bootstrapped 95% confidence intervals."
+)
+
+```
+
 # Figures
@@
 
+Some kind of conceptual diagram of the methods?
+The life cycle diagram with equations on it?
+
 # References

### 2022-04-20 docs/paper.Rmd
revisions to methods

@@
 For all models, we used the long term demographic dataset for modeling survival, growth, and flowering.
-For established plants, these three vital rates were modeled as a smooth function of size in the previous census using generalized additive models (GAMs) fit with the `mgcv` package[@mgcv] in R [@rcoreteam2020].
+For established plants, these three vital rates were modeled as a smooth function of size in the previous census using generalized additive models (GAMs) fit with the `mgcv` package[@mgcv] in `r version$version.string` [@rcoreteam2020].
 Seedlings were considered a separate discrete class not structured by size.
@@
 The kernel-resampling approach is to generate kernels for each year of data, using the random intercepts for year, and to iterate the IPM by drawing from these randomly.
-This is equivalent to "matrix shuffling" approach in matrix population models [@caswell2001] <!--# I'm not actually sure that Caswell calls it "matrix shuffling" -->.
+This is equivalent to the matrix selection approach for matrix population models described by @caswell2001.
 
@@
 All IPMs were constructed and iterated with the `ipmr` package in R [@levinIpmrFlexibleImplementation2021].
-We iterated all IPMs for 1000 iterations and checked for convergence of the population growth rate, lambda.
-Stochastic growth rates ($\lambda_s$) were calculated as the mean of $ln(\lambda)$ from each iteration and back-transformed to be on the same scale as deterministic lambdas for comparison <!--# could replace entirely with an equation -->.
-To calculate 95% confidence intervals around the growth rates, we bootstrapped plants in the 10 year demographic dataset within each habitat type 1000 times.
-For each bootstrap, we then re-fit vital rates models (all except germination and establishment rate, fruits per flowering plant, and seeds per fruit, which were estimated using different datasets), constructed IPMs, and iterated them 1000 times to calculate lambda.
+For each type of IPM we iterated the model for 1000 time steps, discarding the first 100 time steps to omit transient effects.
+Stochastic growth rates ($\lambda_s$) were calculated as the average $ln(\lambda)$ from each time step [@caswell2001] and back-transformed to be on the same scale as deterministic lambdas for comparison.
+<!--# should  mention starting population vector here.
+Observed population could be a reasonable starting point.
+Emphasize that choice of starting vector should only impact transient dynamics, which we're not interested in here.
+-->
 
+To calculate 95% confidence intervals around the per-capita growth rates (lambdas), we created 1000 bootstraps of the 10 year demographic dataset by sampling individual plants with replacement within each habitat.
+For each bootstrap, we then re-fit vital rates models (all except germination and establishment rate, fruits per flowering plant, and seeds per fruit, which were estimated using different datasets), constructed IPMs, and calculated a value for lambda as described above.
 This workflow was managed using the `targets` R package [@targets] which also allowed us to track computational time spent on each IPM for comparison.
@@
 Population growth rates were consistently higher in continuous forest compared to forest fragments across IPM types (Table \@ref(tab:lambdas)).
+Similar to @kayeEffectStochasticTechnique2003 finding that method effected stochastic lambda, but relative ranking of populations was consistent.
 

### 2022-04-21 notes/optimizing meshpoints.Rmd
work on optimizing number of meshpoints (#6)

---
title: "Meshpoint optimization"
author: "Eric R. Scott"
date: "2022-04-21"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    number_sections: yes
    highlight: kate
    theme:
      version: 4
      bootswatch: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(metRology)
library(targets)
library(ipmr)

library(conflicted)
source(here("R", "make_proto_ipm_det.R"))
source(here("R", "make_proto_ipm_stoch.R"))
source(here("R", "get_scat_params.R"))

conflict_prefer("lag", "dplyr")
conflict_prefer("filter", "dplyr")
```

*Last compiled: `r Sys.Date()`*

# Purpose

To figure out the minimum number of meshpoints needed to accurately (precisely?) model population growth.

# Load Data

```{r data, echo=TRUE}
tar_load(vit_list_det_ff)
tar_load(vit_list_stoch_ff)
```

# Make proto IPM

```{r}
proto_det <- make_proto_ipm_det(vit_list_det_ff)
```

# Alter meshpoints

Ok, so let's alter the number of meshpoints systematically and see where lambda converges

```{r}
proto_det_list <- 
  seq(20, 150, 5) %>% 
  map(~{
    proto_det %>% 
    define_domains(log_size = c(0, 8.018296, .x)) %>%  
    #arbitrary starting population state
    define_pop_state(n_log_size = runif(.x), n_sdlg = .1)
  })
lambdas_det <- proto_det_list %>% 
  map_dbl(~make_ipm(.x, usr_funs = list(get_scat_params = get_scat_params)) %>% lambda())

plot(x =seq(20, 150, 5), y = lambdas_det, ylab = "lambda", xlab = "meshpoints")
```

Looks like 100 is probably the best.
Could *maybe* do 80.

# Stochastic IPM

Let's try with stochastic IPM.

```{r}
proto_stoch <- make_proto_ipm_stoch(vit_list_stoch_ff)
```

```{r}
year_seq <- sample(2000:2009, 100, replace = TRUE)

proto_stoch_list <- 
  seq(60, 120, 10) %>%
  map(~{
    proto_stoch %>% 
    define_domains(log_size = c(0, 8.018296, .x)) %>%  
    #arbitrary starting population state
    define_pop_state(n_log_size = runif(.x), n_sdlg = .1)
  })

# make_ipm(proto_stoch_list[[1]],
#          iterations = 100,
#          kernel_seq = year_seq,
#          usr_funs = list(get_scat_params = get_scat_params))
```

###Note:

This currently isn't working because of a bug in `ipmr` that causes `make_ipm()` to hang when `kernel_seq` is supplied <https://github.com/levisc8/ipmr/issues/58>

```{r}
lambdas_stoch <-
  proto_stoch_list %>%
  map_dbl( ~ make_ipm(
    .x,
    iterations = 100,
    kernel_seq = year_seq,
    usr_funs = list(get_scat_params = get_scat_params, coef = coef)
  ) %>% lambda(log = FALSE))

plot(x = seq(60, 120, 10), y = lambdas_stoch, ylab = "stochastic lambda", xlab = "meshpoints")
```

### 2022-04-21 debug_ipmr.R

@@ -1,0 +1,164 @@
+library(targets)
+library(ipmr)
+library(purrr)
+library(metRology)
+library(here)
+library(tibble)
+source(here("R", "make_proto_ipm_stoch.R"))
+source(here("R", "get_scat_params.R"))
+
+tar_load(vit_list_stoch_ff)
+proto_stoch <- make_proto_ipm_stoch(vit_list_stoch_ff)
+
+make_ipm(proto_stoch,
+         iterations = 10,
+         kernel_seq = sample(2000:2009, 10, replace = TRUE),
+         usr_funs = list(get_scat_params = get_scat_params))
+
+
+
+# use ipmr vignette -------------------------------------------------------
+library(ipmr)
+
+# Define some fixed parameters
+
+fixed_list <- list(
+  s_int     = 1.03,   # fixef(my_surv_mod)[1] - uses fixef because we now have a model with random effects
+  s_slope   = 2.2,    # fixef(my_surv_mod)[2]
+  g_int     = 3.7,    # fixef(my_grow_mod)[1]
+  g_slope   = 0.92,   # fixef(my_grow_mod)[2]
+  sd_g      = 0.9,    # sd(resid(my_grow_mod))
+  r_r_int   = 0.09,   # coef(my_repro_mod)[1] - uses coef because there are no random effects in this model
+  r_r_slope = 0.05,   # coef(my_repro_mod)[2]
+  r_s_int   = 0.1,    # fixef(my_flower_mod)[1]
+  r_s_slope = 0.005,  # fixef(my_flower_mod)[2]
+  mu_fd     = 9,      # mean(my_recr_data$size_2, na.rm = TRUE)
+  sd_fd     = 2       # sd(my_recr_data$size_2, na.rm = TRUE)
+)
+# Now, simulate some random intercepts for growth (g_), survival (s_), 
+# and offspring production (r_s_). This part is for the purpose of the example.
+
+# First, we create vector of values corresponding to 
+
+g_r_int   <- rnorm(5, 0, 0.3) # unlist(ranef(my_grow_mod)) for an lme4 output
+s_r_int   <- rnorm(5, 0, 0.7) # unlist(ranef(my_surv_mod)) for an lme4 output
+r_s_r_int <- rnorm(5, 0, 0.2) # unlist(ranef(my_flower_mod)) for an lme4 output
+
+nms <- paste("r_", 1:5, sep = "")
+
[…]

### 2022-04-21 _targets.R
run dlnm IPMs locally because they keep timing out on hipergator

@@ -9,7 +9,7 @@ options(
   clustermq.scheduler = "ssh",
   clustermq.template = "ssh_clustermq.tmpl", #custom SSH template to use R 4.0
   clustermq.ssh.host = "hpg", #set up in ~/.ssh/config.
-  clustermq.ssh.timeout = 60, #longer timeout helps with 2FA
+  clustermq.ssh.timeout = 80, #longer timeout helps with 2FA
   clustermq.ssh.log = "~/cmq_ssh.log" # log for easier debugging
 )
 
@@ -134,11 +134,24 @@ tar_plan(
     ),
     vit_other_ff),
   
+  # Unfortunately, I think the dlnm IPMs need to be run locally (deployment =
+  # "main") because otherwise it times out when trying to pass the resulting
+  # object back to the local session from the HPC.
+  
   proto_ipm_dlnm_ff = make_proto_ipm_dlnm(vit_list_dlnm_ff),
-  ipm_dlnm_ff = proto_ipm_dlnm_ff %>%
-    make_dlnm_ipm(clim, seed = 1234, iterations = 1000,
-                  return_main_env = FALSE, # don't save iteration mesh and other stuff
-                  usr_funs = list(get_scat_params = get_scat_params)),
+  tar_target(
+    ipm_dlnm_ff,
+    proto_ipm_dlnm_ff %>%
+      make_dlnm_ipm(
+        clim,
+        seed = 1234,
+        iterations = 1000,
+        return_main_env = FALSE,
+        # don't save iteration mesh and other stuff
+        usr_funs = list(get_scat_params = get_scat_params)
+      ),
+    deployment = "main"
+  ), 
   
   ## continuous forest
   vit_list_dlnm_cf = c(
@@ -152,10 +165,19 @@ tar_plan(
     vit_other_cf),
   
   proto_ipm_dlnm_cf = make_proto_ipm_dlnm(vit_list_dlnm_cf),
-  ipm_dlnm_cf = proto_ipm_dlnm_cf %>%
-    make_dlnm_ipm(clim, seed = 1234, iterations = 1000,
-                  return_main_env = FALSE, # don't save iteration mesh and other stuff
-                  usr_funs = list(get_scat_params = get_scat_params)),
+  tar_target(
+    ipm_dlnm_cf,
+    proto_ipm_dlnm_cf %>%
+      make_dlnm_ipm(
+        clim,
+        seed = 1234,
+        iterations = 1000,
[…]

### 2022-04-21 notes/environmental stochasticity.Rmd
add notes about how to model environmental stochasticity

---
title: "Modeling Environmental Stochasticity"
author: "Eric R. Scott"
date: "2022-04-21"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    number_sections: yes
    highlight: kate
    theme:
      version: 4
      bootswatch: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(conflicted)
library(targets)
library(mgcv)
library(gratia)
library(bbmle)
library(tictoc)

conflict_prefer("lag", "dplyr")
conflict_prefer("filter", "dplyr")
```

*Last compiled: `r Sys.Date()`*

# Purpose

Try out different ways of doing random effects from these sources:

-   <https://fromthebottomoftheheap.net/2021/02/02/random-effects-in-gams/>

-   Pedersen, Eric J., David L. Miller, Gavin L. Simpson, and Noam Ross.
    "Hierarchical Generalized Additive Models in Ecology: An Introduction with Mgcv." PeerJ 7 (May 27, 2019).
    <https://doi.org/10.7717/peerj.6876>.

# Load Data

```{r data, echo=TRUE}
tar_load(data_full)
```

```{r}
df <- data_full %>%
  filter(sdlg_prev == FALSE, habitat == "1-ha") %>%
  filter(!is.na(log_size_prev)) %>% 
  select(-L, -spei_history)
df
```

# Fit GAMs

With no random effect

```{r}
tic()
m0 <- bam(
  surv ~ s(log_size_prev, bs = "cr", k = 20),
  family = binomial,
    data = df,
    method = "fREML"
)
toc()
draw(m0)
df <- df %>% mutate(fitted0 = predict(m0, type = "response"))
```

With random intercepts for each year

```{r}
tic()
m1 <- bam(
  surv ~ s(log_size_prev, bs = "cr", k = 20) +
    s(year_fct, bs = "re"),
  family = binomial,
    data = df,
    method = "fREML"
)
toc()
draw(m1)
df <- df %>% mutate(fitted1 = predict(m1, type = "response"))
```

With random intercept and "slope" for each year.
Not clear to me what slope means exactly in this context.

```{r}
tic()
m2 <- bam(
  surv ~ s(log_size_prev, bs = "cr", k = 20) +
    s(year_fct, log_size_prev, bs = "re"),
  family = binomial,
    data = df,
    method = "fREML"
)
toc()
draw(m2)
df <- df %>% mutate(fitted2 = predict(m2, type = "response"))
```

With a global smoother and a different shapes for each year (but not different wiggliness).

```{r}
tic()
m3 <- bam(
  surv ~ s(log_size_prev, bs = "cr", k = 20, m = 2) +
    s(log_size_prev, year_fct, bs = "fs",  m = 2),
  family = binomial,
    data = df,
    method = "fREML"
)
toc()
draw(m3)
df <- df %>% mutate(fitted3 = predict(m3, type = "response"))
```

```{r}
AICtab(m0, m1, m2, m3)
```

The last one fits the data best, but also takes the longest to fit.
m2 is not far off with a dAIC of 5.

This is what the fitted values look like

```{r}
p0 <- ggplot(df, aes(x = log_size_prev)) +
  geom_line(aes(y = fitted0)) +
  labs(title = "no random effect")
p1 <- ggplot(df, aes(x = log_size_prev)) +
  geom_line(aes(y = fitted1, color = year_fct)) +
  labs(title = "random intercept")
p2 <- ggplot(df, aes(x = log_size_prev)) +
  geom_line(aes(y = fitted2, color = year_fct)) +
  labs(title = "random intercept and slope")
p3 <- ggplot(df, aes(x = log_size_prev)) +
  geom_line(aes(y = fitted3, color = year_fct)) +
  labs(title = "random shape")

library(patchwork)
p0 + p1 + p2 + p3 & labs(y = ".fitted") & theme(legend.position = "none") & plot_annotation(title=  "survival probability")
```

### 2022-04-21 notes/environmental stochasticity.Rmd
format notes

@@
 ```{r data, echo=TRUE}
-tar_load(data_full)
+withr::with_dir(here(), tar_load(data_full))
 ```
@@
 library(patchwork)
-p0 + p1 + p2 + p3 & labs(y = ".fitted") & theme(legend.position = "none") & plot_annotation(title=  "survival probability")
+p0 + p1 + p2 + p3 &
+  labs(y = ".fitted") &
+  theme(legend.position = "none") &
+  plot_annotation(title = "survival probability")
 ```

### 2022-04-21 R/vital_rates_stoch.R
change random effect of year to allow for different functional forms each year.

@@ -11,8 +11,8 @@ surv_raneff <- function(data, sdlg = FALSE, habitat = c("CF", "1-ha")) {
     filter(sdlg_prev == sdlg, habitat == habitat_choice)
   
   bam(
-    surv ~ s(log_size_prev, bs = "cr", k = 20) +
-      s(year_fct, bs = "re"),
+    surv ~ s(log_size_prev, bs = "cr", k = 20, m = 2) +
+      s(log_size_prev, year_fct, bs = "fs", m = 2),
     family = binomial,
     data = df,
     method = "fREML"
@@ -32,8 +32,8 @@ size_raneff <- function(data, sdlg = FALSE, habitat = c("CF", "1-ha")) {
     dplyr::filter(surv == 1, !is.na(log_size))
   
   bam(
-    log_size ~ s(log_size_prev, bs = "cr", k = 20) +
-      s(year_fct, bs = "re"),
+    log_size ~ s(log_size_prev, bs = "cr", k = 20, m = 2) +
+      s(log_size_prev, year_fct, bs = "fs", m = 2),
     family = scat,
     data = df,
     method = "fREML"
@@ -51,8 +51,8 @@ flwr_raneff <- function(data, habitat = c("CF", "1-ha")) {
     dplyr::filter(surv == 1, !is.na(log_size))
   
   bam(
-    flwr ~ s(log_size_prev, bs = "cr", k = 20) +
-      s(year_fct, bs = "re"),
+    flwr ~ s(log_size_prev, bs = "cr", k = 20, m = 2) +
+      s(log_size_prev, year_fct, bs = "fs", m = 2),
     family = binomial, 
     data = df,
     method = "fREML"

### 2022-04-21 _targets.R
comment out targets I don't need for this experiment

@@ -124,69 +124,69 @@ tar_plan(
   # Stochastic parameter resampled IPM --------------------------------------
   
   ## forest fragments
-  vit_list_dlnm_ff = c(
-    list(
-      vit_surv = surv_dlnm(data_full, habitat = "1-ha"),
-      vit_size = size_dlnm(data_full, habitat = "1-ha"),
-      vit_flwr = flwr_dlnm(data_full, habitat = "1-ha"),
-      vit_size_sdlg = size_sdlg_dlnm(data_full, habitat = "1-ha"),
-      vit_surv_sdlg = surv_sdlg_dlnm(data_full, habitat = "1-ha")
-    ),
-    vit_other_ff),
-  
+  # vit_list_dlnm_ff = c(
+  #   list(
+  #     vit_surv = surv_dlnm(data_full, habitat = "1-ha"),
+  #     vit_size = size_dlnm(data_full, habitat = "1-ha"),
+  #     vit_flwr = flwr_dlnm(data_full, habitat = "1-ha"),
+  #     vit_size_sdlg = size_sdlg_dlnm(data_full, habitat = "1-ha"),
+  #     vit_surv_sdlg = surv_sdlg_dlnm(data_full, habitat = "1-ha")
+  #   ),
+  #   vit_other_ff),
+  # 
   # Unfortunately, I think the dlnm IPMs need to be run locally (deployment =
   # "main") because otherwise it times out when trying to pass the resulting
   # object back to the local session from the HPC.
   
-  proto_ipm_dlnm_ff = make_proto_ipm_dlnm(vit_list_dlnm_ff),
-  tar_target(
-    ipm_dlnm_ff,
-    proto_ipm_dlnm_ff %>%
-      make_dlnm_ipm(
-        clim,
-        seed = 1234,
-        iterations = 1000,
-        return_main_env = FALSE,
-        # don't save iteration mesh and other stuff
-        usr_funs = list(get_scat_params = get_scat_params)
-      ),
-    deployment = "main"
-  ), 
-  
+  # proto_ipm_dlnm_ff = make_proto_ipm_dlnm(vit_list_dlnm_ff),
+  # tar_target(
+  #   ipm_dlnm_ff,
+  #   proto_ipm_dlnm_ff %>%
+  #     make_dlnm_ipm(
+  #       clim,
+  #       seed = 1234,
+  #       iterations = 1000,
[…]

### 2022-04-21 notes/environmental stochasticity.Rmd
fix random intercept + slope to include intercept.

@@
   surv ~ s(log_size_prev, bs = "cr", k = 20) +
+    s(year_fct, bs = "re") +
     s(year_fct, log_size_prev, bs = "re"),
@@
 The last one fits the data best, but also takes the longest to fit.
-m2 is not far off with a dAIC of 5.
+m2 is not far off with a dAIC of 6.8.
 
@@
 ```
+
+With newdata to better reflect what the IPM is doing under the hood:
+
+```{r}
+newdata <- expand.grid(
+  log_size_prev = seq(min(df$log_size_prev), max(df$log_size_prev), length.out = 50),
+  year_fct = unique(df$year_fct)
+)
+
+fitted_newdf<- list(m0 = m0, m1 = m1, m2 = m2, m3 = m3) %>% 
+  map_dfc(~predict(.x, newdata = newdata, type = "response")) %>% 
+  bind_cols(newdata)
+
+p0 <- ggplot(fitted_newdf, aes(x = log_size_prev)) +
+  geom_line(aes(y = m0)) +
+  labs(title = "no random effect")
+p1 <- ggplot(fitted_newdf, aes(x = log_size_prev)) +
+  geom_line(aes(y = m1, color = year_fct)) +
+  labs(title = "random intercept")
+p2 <- ggplot(fitted_newdf, aes(x = log_size_prev)) +
+  geom_line(aes(y = m2, color = year_fct)) +
+  labs(title = "random intercept and slope")
+p3 <- ggplot(fitted_newdf, aes(x = log_size_prev)) +
+  geom_line(aes(y = m3, color = year_fct)) +
+  labs(title = "random shape")
+```
+
+```{r}
+p0 + p1 + p2 + p3 &
+  labs(y = ".fitted") &
+  theme(legend.position = "none") &
+  plot_annotation(title = "survival probability") & ylim(0,1)
+```

### 2022-04-21 notes/understanding stochastic lambda.Rmd
figuring out what's up with IPMs

---
title: "Why is Stochastic Lambda > Deterministic?"
author: "Eric R. Scott"
date: "2022-04-21"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    number_sections: yes
    highlight: kate
    theme:
      version: 4
      bootswatch: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(conflicted)

conflict_prefer("lag", "dplyr")
conflict_prefer("filter", "dplyr")
```

*Last compiled: `r Sys.Date()`*

# Purpose

Figure out what the heck is going on.

# Load Data

```{r include=FALSE}
## Load your packages, e.g. library(targets).
source(here("packages.R"))

## Load your R files
lapply(list.files(here("R"), full.names = TRUE), source)

withr::with_dir(here(),{
  tar_load(data_full)
  tar_load(vit_other_ff)
})
```

# Pick two years, worst and best

Fit determinsitic IPMs to all years separately to identify a bad year and a good year.
Then subset data to only contain those years

```{r warning=FALSE}
lambdas <- data_full %>%
  filter(year>1998) %>% 
  group_by(year_fct) %>% 
  group_split() %>% 
  map_dbl(~{
    vit_list_det_ff = c(
      list(
        vit_surv = surv_det(.x, habitat = "1-ha"),
        vit_size = size_det(.x, habitat = "1-ha"),
        vit_flwr = flwr_det(.x, habitat = "1-ha"),
        vit_size_sdlg = size_sdlg_det(.x, habitat = "1-ha"),
        vit_surv_sdlg = surv_sdlg_det(.x, habitat = "1-ha")
      ),
      vit_other_ff)

    make_proto_ipm_det(vit_list_det_ff) %>% 
      make_ipm(iterations = 1000,  #only needs 100 to converge
               normalize_pop_size = FALSE, # to run as PVA
               usr_funs = list(get_scat_params = get_scat_params)) %>% lambda()
  })
worst_year <- c(1999:max(data_full$year))[which.min(lambdas)]; min(lambdas)
best_year <- c(1999:max(data_full$year))[which.max(lambdas)]; max(lambdas)

two_yrs <- data_full %>% filter(year %in% c(best_year, worst_year))
```

# Fit IPMs

## Deterministic

Fit a deterministic IPM with just those two years of data

```{r}
vit_list_det_ff = c(
  list(
    vit_surv = surv_det(two_yrs, habitat = "1-ha"),
    vit_size = size_det(two_yrs, habitat = "1-ha"),
    vit_flwr = flwr_det(two_yrs, habitat = "1-ha"),
    vit_size_sdlg = size_sdlg_det(two_yrs, habitat = "1-ha"),
    vit_surv_sdlg = surv_sdlg_det(two_yrs, habitat = "1-ha")
  ),
  vit_other_ff)

ipm_det_ff = make_proto_ipm_det(vit_list_det_ff) %>% 
  make_ipm(iterations = 1000,  #only needs 100 to converge
           normalize_pop_size = FALSE, # to run as PVA
           usr_funs = list(get_scat_params = get_scat_params))
lambda(ipm_det_ff)
```

## Stochastic

The stochastic IPM requires re-writing the proto_ipm function because it's now using different parameter sets for year.

```{r}
## forets fragments
vit_list_stoch_ff = c(
  list(
    vit_surv = surv_raneff(two_yrs, habitat = "1-ha"),
    vit_size = size_raneff(two_yrs, habitat = "1-ha"),
    vit_flwr = flwr_raneff(two_yrs, habitat = "1-ha"),
    vit_size_sdlg = size_sdlg_raneff(two_yrs, habitat = "1-ha"),
    vit_surv_sdlg = surv_sdlg_raneff(two_yrs, habitat = "1-ha")
  ), 
  vit_other_ff)

proto_ipm_stoch_ff <- 
  init_ipm(
    sim_gen = "general",
    di_dd = "di",
    det_stoch = "stoch",
    kern_param = "kern"
  ) %>%

  # Define growth and survival kernel
  define_kernel(
    "P_yr", #_yr makes a different P kernel for each value of `yr` from par_set_indices
    formula = s_z * G_z1z * d_log_size,
    family  = "CC",
    # Probability of survival
    s_z = predict(vit_surv,
                  newdata = tibble(log_size_prev = log_size_1, year_fct = yr),
                  type = 'response'),
    # Growth
    G_z1z = dt.scaled(
      log_size_2,
      df   = size_nu,
      mean = size_mu,
      sd   = size_sd
    ),
    size_mu = predict(vit_size,
                      newdata = tibble(log_size_prev = log_size_1, year_fct = yr),
                      type = 'response'),
    size_nu = get_scat_params(vit_size)["nu"],
    size_sd = get_scat_params(vit_size)["sd"],

    data_list = vit_list_stoch_ff,
    states          = list(c('log_size')),
    uses_par_sets   = TRUE,
    par_set_indices = list(yr = c(2001, 2003)),
    evict_cor       = TRUE,
    evict_fun       = truncated_distributions("t.scaled", "G_z1z")
  ) %>%

  # Define fecundity and recruitment kernel
  define_kernel(
    "go_sdlg_yr",
    formula = p_f * fruits * seeds * g_e * d_log_size,
    family  = "CD",
    # probability of flowering
    p_f = predict(vit_flwr,
                  newdata = tibble(log_size_prev = log_size_1, year_fct = yr),
                  type = 'response'),
    fruits = predict(vit_fruits,
                     newdata = tibble(log_size_prev = log_size_1),
                     type = 'response'), 
    seeds = exp(coef(vit_seeds)[1]), #mean seeds per fruit
    g_e = vit_germ_est,

    data_list     = vit_list_stoch_ff,
    states        = list(c('log_size', 'sdlg')),
    uses_par_sets = TRUE,
    par_set_indices = list(yr = c(2001, 2003))
  ) %>%

  define_kernel(
    name = "stay_sdlg",
    formula = 0,
    family = "DD",
    states = list(c("sdlg")),
    evict_cor = FALSE
  ) %>%

  # Enter into post-seedling
  define_kernel(
    name = "leave_sdlg_yr",
    formula = s_sdlg * G_z2_sdlg,
    family = "DC",
    s_sdlg = predict(vit_surv_sdlg,
                     newdata = tibble(year_fct = yr),
                     type = 'response'), #intercept from intercept-only glm
    G_z2_sdlg = dt.scaled(
      x    = log_size_2,
      df   = size_sdlg_nu,
      mean = size_sdlg_mu,
      sd   = size_sdlg_sd
    ),
    size_sdlg_mu = coef(vit_size_sdlg)[1], #intercept from intercept-only glm
    size_sdlg_nu = get_scat_params(vit_size_sdlg)["nu"],
    size_sdlg_sd = get_scat_params(vit_size_sdlg)["sd"],

    data_list = vit_list_stoch_ff,
    states = list(c("sdlg", "log_size")),
    uses_par_sets = TRUE,
    par_set_indices = list(yr = c(2001, 2003)),
    evict_cor = TRUE,
    evict_fun = truncated_distributions("t.scaled", "G_z2_sdlg")
  ) %>%

  # define implementation with midpoint rule
  define_impl(make_impl_args_list(
    kernel_names = c("P_yr",        "go_sdlg_yr", "stay_sdlg", "leave_sdlg_yr"),
    int_rule = "midpoint",
    state_start  = c("log_size", "log_size", "sdlg",     "sdlg"),
    state_end    = c("log_size", "sdlg",     "sdlg",     "log_size")
  )) %>% 
  #define lower bound, upper bound, and number of meshpoints for log_size
  define_domains(log_size = c(0, 8.018296, 100)) %>%  
  #arbitrary starting population state
  define_pop_state(n_log_size = runif(100), n_sdlg = .1)
```

Iterate IPM

```{r}
ipm_stoch_ff <- 
  proto_ipm_stoch_ff %>%
  make_ipm(iterations = 1000,
           normalize_pop_size = FALSE, # to run as PVA
           usr_funs = list(get_scat_params = get_scat_params))
```

# Investigation

First, with just the worst and best years, is the stochastic lambda still higher than the deterministic?

```{r}
lambda(ipm_det_ff)
lambda(ipm_stoch_ff, log = FALSE)
```

Yes, stochastic lambda is still higher!

What about the averages for each year?
Are they consistent with a bad and good year?

```{r}
all_lams <- tibble(year = ipm_stoch_ff$env_seq,
       lambda = lambda(ipm_stoch_ff, type_lambda = "all")[,1])
all_lams %>% group_by(year) %>% summarize(mean(lambda)) %>% knitr::kable()
```

Yes, but both the good and bad years mean lambdas are *greater* than the deterministic lambda.

How do they compare at the level of vital rates?

```{r}
newdata <- expand.grid(
  log_size_prev = seq(min(two_yrs$log_size_prev, na.rm = TRUE), max(two_yrs$log_size_prev, na.rm = TRUE), length.out = 50),
  year_fct = as.character(c(2001, 2003))
)

safe_predict <- possibly(predict, NA)

det_fitted <- 
  vit_list_det_ff %>% 
  map_df(~safe_predict(.x, newdata = newdata, type = "response")) %>% 
  bind_cols(newdata) %>%
  select(-vit_fruits, -vit_seeds, -vit_germ_est) %>% 
  pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_") %>% 
  mutate(year_fct = "deterministic")

stoch_fitted <- 
  vit_list_stoch_ff %>%
  map_df(~safe_predict(.x, newdata= newdata, type = "response")) %>%
  bind_cols(newdata) %>%
  select(-vit_fruits, -vit_seeds, -vit_germ_est) %>% 
  pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_")

bind_rows(stoch_fitted, det_fitted) %>% 
  ggplot(aes(x = log_size_prev, y = .fitted, color = year_fct)) +
  geom_line() +
  facet_wrap(~vital_rate, scales = "free")

```

Hold on.
Those numbers for size_sdlg don't make sense!
The mean size a seedling establishes at should be more like 2

```{r}
two_yrs %>% filter(sdlg_prev == TRUE, !is.na(log_size), habitat== "1-ha") %>% select(-L, -spei_history) %>% pull(log_size) %>% hist()
data_full %>% filter(ha_id_number == "2") %>% View()
data_full %>% filter(sdlg_prev == TRUE, !is.na(log_size), habitat == "1-ha") %>% View()
```

### 2022-04-21 R/vital_rates_stoch.R
fix a tidyeval related bug that was not filtering seedlings correctly.

@@ -3,12 +3,12 @@
 # library(tidyverse)
 # data <- tar_read(data_full)
 
-surv_raneff <- function(data, sdlg = FALSE, habitat = c("CF", "1-ha")) {
+surv_raneff <- function(data, habitat = c("CF", "1-ha")) {
   
   habitat_choice <- match.arg(habitat)
   
   df <- data %>%
-    filter(sdlg_prev == sdlg, habitat == habitat_choice)
+    filter(sdlg_prev == FALSE, habitat == habitat_choice)
   
   bam(
     surv ~ s(log_size_prev, bs = "cr", k = 20) +
@@ -22,12 +22,12 @@ surv_raneff <- function(data, sdlg = FALSE, habitat = c("CF", "1-ha")) {
 # k.check(surv_raneff(data, habitat = "1-ha"))
 
 
-size_raneff <- function(data, sdlg = FALSE, habitat = c("CF", "1-ha")) {
+size_raneff <- function(data, habitat = c("CF", "1-ha")) {
   
   habitat_choice <- match.arg(habitat)
   
   df <- data %>%
-    dplyr::filter(sdlg_prev == sdlg, habitat == habitat_choice) %>% 
+    dplyr::filter(sdlg_prev == FALSE, habitat == habitat_choice) %>% 
     #only plants that survived get to grow
     dplyr::filter(surv == 1, !is.na(log_size))
   
@@ -44,7 +44,9 @@ size_raneff <- function(data, sdlg = FALSE, habitat = c("CF", "1-ha")) {
 
 
 flwr_raneff <- function(data, habitat = c("CF", "1-ha")) {
+  
   habitat_choice <- match.arg(habitat)
+  
   df <- data %>% 
     dplyr::filter(habitat == habitat_choice, sdlg == FALSE) %>% 
     #current seedlings are excluded, but not plants that were seedlings in the previous year
@@ -61,12 +63,13 @@ flwr_raneff <- function(data, habitat = c("CF", "1-ha")) {
 # k.check(flwr_raneff(data))
 # k.check(flwr_raneff(data, habitat = "1-ha"))
 
-size_sdlg_raneff <- function(data, sdlg = TRUE, habitat = c("CF", "1-ha")) {
+size_sdlg_raneff <- function(data, habitat = c("CF", "1-ha")) {
   
   habitat_choice <- match.arg(habitat)
+  sdlg_choice <- sdlg
   
[…]

### 2022-04-22 notes/understanding stochastic lambda.Rmd
tweak and render notes

@@
     highlight: kate
+    code_folding: show
     theme:
@@
 conflict_prefer("filter", "dplyr")
+set.seed(123)
 ```
@@
 
-```{r warning=FALSE}
-lambdas <- data_full %>%
-  filter(year>1998) %>% 
+```{r warning=FALSE, class.source = 'fold-hide'}
+single_years <-   
+  data_full %>%
+  filter(year>1999) %>% 
+  select(-L, -spei_history) %>% 
   group_by(year_fct) %>% 
-  group_split() %>% 
-  map_dbl(~{
+  group_split()
+
+vit_list_yr <- 
+  single_years %>% 
+  map(~{
     vit_list_det_ff = c(
@@
       vit_other_ff)
+    })
 
-    make_proto_ipm_det(vit_list_det_ff) %>% 
+ipms_single_yr <-
+  map(vit_list_yr, ~{
+    make_proto_ipm_det(.x) %>% 
       make_ipm(iterations = 1000,  #only needs 100 to converge
                normalize_pop_size = FALSE, # to run as PVA
-               usr_funs = list(get_scat_params = get_scat_params)) %>% lambda()
+               usr_funs = list(get_scat_params = get_scat_params))
   })
-worst_year <- c(1999:max(data_full$year))[which.min(lambdas)]; min(lambdas)
-best_year <- c(1999:max(data_full$year))[which.max(lambdas)]; max(lambdas)
+
+lambdas <- ipms_single_yr %>% map_dbl(lambda)
+```
+
+```{r warning=FALSE}
+#worst year
+(worst_year <- c(2000:max(data_full$year))[which.min(lambdas)]); min(lambdas)
+
+#best year
+(best_year <- c(2000:max(data_full$year))[which.max(lambdas)]); max(lambdas)
 
@@
 
-```{r}
+```{r warning=FALSE, class.source = 'fold-hide'}
 vit_list_det_ff = c(
@@
 
-The stochastic IPM requires re-writing the proto_ipm function because it's now using different parameter sets for year.
+The stochastic IPM requires re-writing the `proto_ipm` because it's now using different parameter sets for year.
+This now uses vital rates models where there is a random effect of year fit such that each year can have a different shape smoother (but not different wiggliness).
 
-```{r}
+```{r warning=FALSE, class.source = 'fold-hide'}
 ## forets fragments
@@
 
-Iterate IPM
+Iterate IPM with 1000 iterations
 
-```{r}
+```{r warning=FALSE}
 ipm_stoch_ff <- 
@@
 lambda(ipm_det_ff) 
-lambda(ipm_stoch_ff, log = FALSE)
+lambda(ipm_stoch_ff, log = FALSE, burn_in = 0.2)
 ```
 
-Yes, stochastic lambda is still higher!
+Yes, stochastic lambda is still a bit higher!
 
@@
 ```{r}
-all_lams <- tibble(year = ipm_stoch_ff$env_seq,
+all_lams <- 
+  tibble(year = ipm_stoch_ff$env_seq,
          lambda = lambda(ipm_stoch_ff, type_lambda = "all")[,1])
-all_lams %>% group_by(year) %>% summarize(mean(lambda)) %>% knitr::kable()
+all_lams %>%
+  group_by(year) %>%
+  summarize(mean(lambda))
 ```
@@
 
+How do the mean lambdas from the good and bad year compare to the deterministic lambdas from individual years?
+
+```{r}
+max(lambdas)
+min(lambdas)
+```
+
+The deterministic lambda for 2001 is a lot higher than the mean of the 2001 lambdas from the stochastic simulation and the deterministic lambda for 2003 is a lot lower than the mean of the 2003 lambdas from the stochastic simulation.
+This makes sense to me, as fitting year as a random effect rather than a fixed effect will shrink estimates toward a global mean.
+
 How do they compare at the level of vital rates?
 
-```{r}
+```{r class.source = 'fold-hide'}
 newdata <- expand.grid(
@@
 
+#deterministic, just 2001
+det_fitted_2001 <- 
+  vit_list_yr[[which.max(lambdas)]] %>% 
+  map_df(~safe_predict(.x, newdata = newdata, type = "response")) %>% 
+  bind_cols(newdata) %>% 
+  select(-vit_fruits, -vit_seeds, -vit_germ_est) %>% 
+  pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_") %>% 
+  mutate(year_fct = "det. 2001")
+
+#deterministic, just 2003
+det_fitted_2003 <- 
+  vit_list_yr[[which.min(lambdas)]] %>% 
+  map_df(~safe_predict(.x, newdata = newdata, type = "response")) %>% 
+  bind_cols(newdata) %>% 
+  select(-vit_fruits, -vit_seeds, -vit_germ_est) %>% 
+  pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_") %>% 
+  mutate(year_fct = "det. 2003")
+
+#deterministic with both years combined
 det_fitted <- 
@@
   pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_") %>% 
-  mutate(year_fct = "deterministic")
+  mutate(year_fct = "det. both years")
 
+#stochastic
 stoch_fitted <- 
@@
   select(-vit_fruits, -vit_seeds, -vit_germ_est) %>% 
-  pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_")
+  pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_") %>% 
+  mutate(year_fct = paste("stoch.", year_fct))
 
-bind_rows(stoch_fitted, det_fitted) %>% 
-  ggplot(aes(x = log_size_prev, y = .fitted, color = year_fct)) +
-  geom_line() +
+bind_rows(stoch_fitted, det_fitted, det_fitted_2001, det_fitted_2003) %>% 
+  ggplot(aes(x = log_size_prev, y = .fitted, color = year_fct, linetype = year_fct)) +
+  geom_line(alpha = 0.9) +
   facet_wrap(~vital_rate, scales = "free")
@@
 
-Hold on.
-Those numbers for size_sdlg don't make sense!
-The mean size a seedling establishes at should be more like 2
-
-```{r}
-two_yrs %>% filter(sdlg_prev == TRUE, !is.na(log_size), habitat== "1-ha") %>% select(-L, -spei_history) %>% pull(log_size) %>% hist()
-data_full %>% filter(ha_id_number == "2") %>% View()
-data_full %>% filter(sdlg_prev == TRUE, !is.na(log_size), habitat == "1-ha") %>% View()
-```
+For the stochastic models, the two years are nearly indistinguishable for seedlings.
+I believe this is because of shrinkage---there is little information for either year so means are shrunk toward global average.
+You can see this more clearly for flowering probability where the order of lines at the larger end is 2001 deterministic (essentially a fixed effect), 2001 stochastic (random effect), both years combined, 2003 stochastic (random effect), and 2003 deterministic (fixed effect).

### 2022-04-22 notes/environmental stochasticity.Rmd

@@
   theme(legend.position = "none") &
-  plot_annotation(title = "survival probability")
+  plot_annotation(title = "survival probability") &
+  ylim(0,1)
 ```
@@
   theme(legend.position = "none") &
-  plot_annotation(title = "survival probability") & ylim(0,1)
+  plot_annotation(title = "survival probability") & 
+  ylim(0,1)
 ```

### 2022-04-22 notes/index.Rmd
add more notes for github pages

---
output: html_document
---

# Notes:

```{r eval=FALSE, include=FALSE, results='asis'}
base_url <- "https://brunalab.github.io/lagged-ipms/notes/"
# paste(base_url, list.files(here::here("notes"), pattern = "[^index].html"), sep = "")
```

[Exploring different ways of fitting random effects GAMs](notes/environmental-stochasticity.html)

[Why is stochastic lambda \> deterministic?](notes/understanding-stochastic-lambda.html)

[Optimizing the number of meshpoints](notes/optimizing-meshpoints.html)

### 2022-04-22 notes/optimizing meshpoints.Rmd
add more notes for github pages

@@
 ```{r data, echo=TRUE}
+withr::with_dir(here::here(), {
   tar_load(vit_list_det_ff)
   tar_load(vit_list_stoch_ff)
+})
 ```
@@
 
-###Note:
+### Note:
 
@@
 
-```{r}
+```{r eval=FALSE}
 lambdas_stoch <-

### 2022-04-22 notes/index.Rmd
link to github *pages*

@@
 
-[Exploring different ways of fitting random effects GAMs](notes/environmental-stochasticity.html)
+[Exploring different ways of fitting random effects GAMs](https://brunalab.github.io/lagged-ipms/notes/environmental-stochasticity.html)
 
-[Why is stochastic lambda \> deterministic?](notes/understanding-stochastic-lambda.html)
+[Why is stochastic lambda \> deterministic?](https://brunalab.github.io/lagged-ipms/notes/understanding-stochastic-lambda.html)
 
-[Optimizing the number of meshpoints](notes/optimizing-meshpoints.html)
+[Optimizing the number of meshpoints](https://brunalab.github.io/lagged-ipms/notes/optimizing-meshpoints.html)

### 2022-04-25 R/ipm_boot_det.R
define population vector using data (#7)

@@ -33,8 +33,11 @@ ipm_boot_det <- function(data, vit_other, habitat = c("1-ha", "CF")) {
     vit_surv_sdlg = surv_sdlg_det(boot, habitat = habitat)
   ), vit_other)
   
+  #make starting population vector
+  pop_vec <- make_pop_vec(boot, n_mesh = 100)
+  
   #make IPM
-  make_proto_ipm_det(vit_list_det) %>% 
+  make_proto_ipm_det(vit_list_det, pop_vec) %>% 
     make_ipm(iterations = 100,  #only needs 100 to converge
              normalize_pop_size = TRUE,
              usr_funs = list(get_scat_params = get_scat_params)

### 2022-04-25 docs/paper.Rmd
notes for methods
revise methods and results

@@
 keywords: |
-  keyword1; keyword2; keyword3
+  integral projection models; environmental stochasticity; lagged effects
 highlights: |
@@
 
+\<!-
+-# TODO:
+
+-   Mention number of meshpoints
+
+-   Add info about starting pop vector for IPMs
+
+-   Add equations / figures
+
+-   Add info about additional datasets (Emilio)
+
+--\>
+
 Ultimately, we built and iterated three types of IPMs which each required different underlying vital rate model structure.
@@
 Seedlings were considered a separate discrete class not structured by size.
-For consistency, seedling survival and growth were also modeled using GAMs, but without size in the previous census as a predictor.
+For consistency, seedling survival and growth were also modeled using GAMs, but without size in the previous census as a predictor (i.e. intercept only models).
 For growth models, a scaled t family distribution provided a better fit to the data than a gaussian fit as the residuals were leptokurtic with a simple gaussian model.
@@
 These models were used to build a general, density-independent, deterministic IPM with four sub-kernels: growth and survival, reproduction, probability of staying a seedling (always = 0), and seedling survival and establishment.
-<!--# include equation --> The seedling bank, or probability of staying a seedling, was always equal to zero, but was required <!--# for reasons? -->.
+<!--# include equation --> The seedling bank, or probability of staying a seedling, was always equal to zero, since our definition of seedlings was first year plants only.
 
-For the second approach, we included environmental stochasticity by adding a random intercept for year to all vital rate models built with the long term demographic dataset.
+For the second approach, we included environmental stochasticity by adding a random intercept <!--# or intercept and slope or factor-smooth interaction -->for year to all vital rate models built with the long term demographic dataset.
 Then, we used these models to build general, density independent, stochastic, kernel-resampled IPMs.
-The kernel-resampling approach is to generate kernels for each year of data, using the random intercepts for year, and to iterate the IPM by drawing from these randomly.
+The kernel-resampling approach is to generate kernels corresponding to each transition year in the demographic dataset using the random intercepts for year, and to iterate the IPM by drawing from these randomly.
 This is equivalent to the matrix selection approach for matrix population models described by @caswell2001.
@@
 For the third method, we modeled the impacts of drought on vital rates explicitly.
-We calculated the standardized precipitation evapotranspiraton index (SPEI) for our site using a published gridded dataset based on ground measurements [@xavierDailyGriddedMeteorological2016].
+We calculated the standardized precipitation evapotranspiraton index (SPEI) for our site using a published gridded dataset based on ground measurements [@xavierDailyGriddedMeteorological2016] as described in @scottDelayedEffectsClimate2022.
 We modeled delayed effects of SPEI on survival, growth, and flowering probability using distributed lag non-linear models fit as two dimensional tensor product splines, with a maximum lag of 36 months [@scottDelayedEffectsClimate2022].
@@
 Stochastic growth rates ($\lambda_s$) were calculated as the average $ln(\lambda)$ from each time step [@caswell2001] and back-transformed to be on the same scale as deterministic lambdas for comparison.
-<!--# should  mention starting population vector here.
-Observed population could be a reasonable starting point.
-Emphasize that choice of starting vector should only impact transient dynamics, which we're not interested in here.
--->
+We used the distribution of established plant sizes and proportion of seedlings from the full dataset as a starting population vector.
+While other starting population vectors were possible, the choice is of little importance as it will only impact transient dynamics, which we aren't interested in for this study.
 
-To calculate 95% confidence intervals around the per-capita growth rates (lambdas), we created 1000 bootstraps of the 10 year demographic dataset by sampling individual plants with replacement within each habitat.
+To calculate 95% confidence intervals around the per-capita growth rates (lambdas), we created 1000 bootstraps of the demographic dataset by sampling individual plants with replacement within each habitat.
 For each bootstrap, we then re-fit vital rates models (all except germination and establishment rate, fruits per flowering plant, and seeds per fruit, which were estimated using different datasets), constructed IPMs, and calculated a value for lambda as described above.
+
 This workflow was managed using the `targets` R package [@targets] which also allowed us to track computational time spent on each IPM for comparison.
@@
 For all vital rates estimated using the long term demographic dataset, the DLNM model fit the best (dAIC = 0) followed by the model with a random effect of year, followed by the deterministic model (Table \@ref(tab:aic)).
+<!--# No longer true if random effect = different smooth for each year -->
 
 Population growth rates were consistently higher in continuous forest compared to forest fragments across IPM types (Table \@ref(tab:lambdas)).
-Similar to @kayeEffectStochasticTechnique2003 finding that method effected stochastic lambda, but relative ranking of populations was consistent.
 
 # Discussion
+
+-   Our finding that the choice of IPM didn't change the relative ranking of CF and FF is consistent with @kayeEffectStochasticTechnique2003 finding that method effected stochastic lambda, but relative ranking of populations was consistent.
 

### 2022-04-25 R/ipm_boot_det.R
initially split data by habitat instead of having `habitat` arg in every function

@@ -1,23 +1,20 @@
 #' Deterministic IPM mega-function for bootstrapping lambda
 #'
-#' This samples individuals within habitats with replacement, fits vital rate
+#' This samples individuals with replacement, fits vital rate
 #' models, builds an IPM, iterates it, then returns lambda. No intermediate
 #' steps are saved, so this is intended for use in bootstrapping confidence
 #' intervals around lambda.
 #' 
-#' @param data the data_full target
+#' @param data demographic data.  E.g. the data_ff or data_cf targets
 #' @param vit_other a list containing vit_fruits, vit_seeds, and vit_germ_est
-#' @param habitat character, either "1-ha" or "CF"
 #'
 #' @return lambda
 #' 
-ipm_boot_det <- function(data, vit_other, habitat = c("1-ha", "CF")) {
+ipm_boot_det <- function(data, vit_other) {
   #sample ha_id_numbers with replacement within plots
   boot_ids <-
     data %>% 
-    group_by(habitat) %>% 
     summarize(ha_id_number = unique(ha_id_number)) %>% 
-    group_by(habitat) %>% 
     sample_n(n(), replace = TRUE) %>% 
     #for validation:
     mutate(unique_id = paste(ha_id_number, row_number(), sep = "-"))
@@ -26,11 +23,11 @@ ipm_boot_det <- function(data, vit_other, habitat = c("1-ha", "CF")) {
   
   #fit vital rates
   vit_list_det <- c(list(
-    vit_surv = surv_det(boot, habitat = habitat),
-    vit_size = size_det(boot, habitat = habitat),
-    vit_flwr = flwr_det(boot, habitat = habitat),
-    vit_size_sdlg = size_sdlg_det(boot, habitat = habitat),
-    vit_surv_sdlg = surv_sdlg_det(boot, habitat = habitat)
+    vit_surv = surv_det(boot),
+    vit_size = size_det(boot),
+    vit_flwr = flwr_det(boot),
+    vit_size_sdlg = size_sdlg_det(boot),
+    vit_surv_sdlg = surv_sdlg_det(boot)
   ), vit_other)
   
   #make starting population vector

### 2022-04-25 R/ipm_boot_dlnm.R

@@ -1,24 +1,21 @@
 #' Stochastic DLNM IPM mega-function for bootstrapping lambda
 #'
-#' This samples individuals within habitats with replacement, fits vital rate
+#' This samples individuals with replacement, fits vital rate
 #' models as a function of SPEI history, builds an IPM, iterates it, then
 #' returns lambda. No intermediate steps are saved, so this is intended for use
 #' in bootstrapping confidence intervals around lambda.
 #' 
-#' @param data the data_full target
+#' @param data demographic data.  E.g. the data_ff or data_cf targets
 #' @param vit_other a list containing vit_fruits, vit_seeds, and vit_germ_est
-#' @param habitat character, either "1-ha" or "CF"
 #' @param clim the clim target.  Full climate timeseries (no lags calculated).
 #'
 #' @return lambda
 #' 
-ipm_boot_dlnm_raw <- function(data, vit_other, habitat = c("1-ha", "CF"), clim) {
+ipm_boot_dlnm_raw <- function(data, vit_other, clim) {
   #sample ha_id_numbers with replacement within plots
   boot_ids <-
     data %>% 
-    group_by(habitat) %>% 
     summarize(ha_id_number = unique(ha_id_number)) %>% 
-    group_by(habitat) %>% 
     sample_n(n(), replace = TRUE) %>% 
     #for validation:
     mutate(unique_id = paste(ha_id_number, row_number(), sep = "-"))
@@ -27,11 +24,11 @@ ipm_boot_dlnm_raw <- function(data, vit_other, habitat = c("1-ha", "CF"), clim) 
   
   #fit vital rates
   vit_list_dlnm <- c(list(
-    vit_surv = surv_dlnm(boot, habitat = habitat),
-    vit_size = size_dlnm(boot, habitat = habitat),
-    vit_flwr = flwr_dlnm(boot, habitat = habitat),
-    vit_size_sdlg = size_sdlg_dlnm(boot, habitat = habitat),
-    vit_surv_sdlg = surv_sdlg_dlnm(boot, habitat = habitat)
+    vit_surv = surv_dlnm(boot),
+    vit_size = size_dlnm(boot),
+    vit_flwr = flwr_dlnm(boot),
+    vit_size_sdlg = size_sdlg_dlnm(boot),
+    vit_surv_sdlg = surv_sdlg_dlnm(boot)
   ), vit_other)
   
   #make starting population vector

### 2022-04-25 R/ipm_boot_stoch.R

@@ -1,23 +1,20 @@
 #' Stochastic IPM mega-function for bootstrapping lambda
 #'
-#' This samples individuals within habitats with replacement, fits vital rate
+#' This samples individuals with replacement, fits vital rate
 #' models with a random effect of `year`, builds an IPM, iterates it, then
 #' returns lambda. No intermediate steps are saved, so this is intended for use
 #' in bootstrapping confidence intervals around lambda.
 #' 
-#' @param data the data_full target
+#' @param data demographic data.  E.g. the data_ff or data_cf targets
 #' @param vit_other a list containing vit_fruits, vit_seeds, and vit_germ_est
-#' @param habitat character, either "1-ha" or "CF"
 #'
 #' @return lambda
 #' 
-ipm_boot_stoch <- function(data, vit_other, habitat = c("1-ha", "CF")) {
+ipm_boot_stoch <- function(data, vit_other) {
   #sample ha_id_numbers with replacement within plots
   boot_ids <-
     data %>% 
-    group_by(habitat) %>% 
     summarize(ha_id_number = unique(ha_id_number)) %>% 
-    group_by(habitat) %>% 
     sample_n(n(), replace = TRUE) %>% 
     #for validation:
     mutate(unique_id = paste(ha_id_number, row_number(), sep = "-"))
@@ -26,11 +23,11 @@ ipm_boot_stoch <- function(data, vit_other, habitat = c("1-ha", "CF")) {
 
   #fit vital rates
   vit_list_stoch <- c(list(
-    vit_surv = surv_raneff(boot, habitat = habitat),
-    vit_size = size_raneff(boot, habitat = habitat),
-    vit_flwr = flwr_raneff(boot, habitat = habitat),
-    vit_size_sdlg = size_sdlg_raneff(boot, habitat = habitat),
-    vit_surv_sdlg = surv_sdlg_raneff(boot, habitat = habitat)
+    vit_surv = surv_raneff(boot),
+    vit_size = size_raneff(boot),
+    vit_flwr = flwr_raneff(boot),
+    vit_size_sdlg = size_sdlg_raneff(boot),
+    vit_surv_sdlg = surv_sdlg_raneff(boot)
   ), vit_other)
   
   #make starting population vector

### 2022-04-25 R/vital_rates_dlnm.R

@@ -1,14 +1,13 @@
 #for checking knots, load data and run k.check() for the models
 # library(mgcv)
 # library(tidyverse)
-# data <- tar_read(data_full)
+# tar_load(data_ff)
+# tar_load(data_cf)
 
-surv_dlnm <- function(data, habitat = c("CF", "1-ha")) {
-  
-  habitat_choice <- match.arg(habitat)
+surv_dlnm <- function(data) {
   
   df <- data %>%
-    dplyr::filter(sdlg_prev == FALSE, habitat == habitat_choice)
+    dplyr::filter(sdlg_prev == FALSE)
   
   bam(
     surv ~ s(log_size_prev, bs = "cr", k = 10) +
@@ -18,16 +17,14 @@ surv_dlnm <- function(data, habitat = c("CF", "1-ha")) {
     method = "fREML"
   )
 }
-# k.check(surv_dlnm(data))
-# k.check(surv_dlnm(data, habitat = "1-ha"))
+# k.check(surv_dlnm(data_ff))
+# k.check(surv_dlnm(data_cf))
 
 
-size_dlnm <- function(data, habitat = c("CF", "1-ha")) {
-  
-  habitat_choice <- match.arg(habitat)
+size_dlnm <- function(data) {
   
   df <- data %>%
-    dplyr::filter(sdlg_prev == FALSE, habitat == habitat_choice) %>% 
+    dplyr::filter(sdlg_prev == FALSE) %>% 
     #only plants that survived get to grow
     dplyr::filter(surv == 1, !is.na(log_size))
   
@@ -39,14 +36,13 @@ size_dlnm <- function(data, habitat = c("CF", "1-ha")) {
     method = "fREML"
   )
 }
-# k.check(size_dlnm(data))
-# k.check(size_dlnm(data, habitat = "1-ha"))
+# k.check(size_dlnm(data_ff))
+# k.check(size_dlnm(data_cf))
 
 
-flwr_dlnm <- function(data, habitat = c("CF", "1-ha")) {
-  habitat_choice <- match.arg(habitat)
+flwr_dlnm <- function(data) {
   df <- data %>% 
-    dplyr::filter(habitat == habitat_choice, sdlg == FALSE) %>% 
+    dplyr::filter(sdlg == FALSE) %>% 
     #current seedlings are excluded, but not plants that were seedlings in the previous year
[…]

### 2022-04-25 R/vital_rates_stoch.R

@@ -1,14 +1,13 @@
 #for checking knots, load data and run k.check() for the models
 # library(mgcv)
 # library(tidyverse)
-# data <- tar_read(data_full)
+# tar_load(data_ff)
+# tar_load(data_cf)
 
-surv_raneff <- function(data, habitat = c("CF", "1-ha")) {
-  
-  habitat_choice <- match.arg(habitat)
+surv_raneff <- function(data) {
   
   df <- data %>%
-    filter(sdlg_prev == FALSE, habitat == habitat_choice)
+    filter(sdlg_prev == FALSE)
   
   bam(
     surv ~ s(log_size_prev, bs = "cr", k = 20, m = 2) +
@@ -18,16 +17,14 @@ surv_raneff <- function(data, habitat = c("CF", "1-ha")) {
     method = "fREML"
   )
 }
-# k.check(surv_raneff(data))
-# k.check(surv_raneff(data, habitat = "1-ha"))
+# k.check(surv_raneff(data_ff))
+# k.check(surv_raneff(data_cf))
 
 
-size_raneff <- function(data, habitat = c("CF", "1-ha")) {
-  
-  habitat_choice <- match.arg(habitat)
+size_raneff <- function(data) {
   
   df <- data %>%
-    dplyr::filter(sdlg_prev == FALSE, habitat == habitat_choice) %>% 
+    dplyr::filter(sdlg_prev == FALSE) %>% 
     #only plants that survived get to grow
     dplyr::filter(surv == 1, !is.na(log_size))
   
@@ -39,16 +36,14 @@ size_raneff <- function(data, habitat = c("CF", "1-ha")) {
     method = "fREML"
   )
 }
-# k.check(size_raneff(data))
-# k.check(size_raneff(data, habitat = "1-ha")) #no diff between k = 20 and k = 25
+# k.check(size_raneff(data_ff))#no diff between k = 20 and k = 25
+# k.check(size_raneff(data_cf)) 
 
 
-flwr_raneff <- function(data, habitat = c("CF", "1-ha")) {
-  
-  habitat_choice <- match.arg(habitat)
+flwr_raneff <- function(data) {
   
   df <- data %>% 
-    dplyr::filter(habitat == habitat_choice, sdlg == FALSE) %>% 
+    dplyr::filter(sdlg == FALSE) %>% 
[…]

### 2022-04-25 _targets.R

@@ -32,6 +32,8 @@ tar_plan(
   demog = read_wrangle_demog(file_demog),
   clim = read_wrangle_clim(file_clim),
   data_full = join_data(demog, clim),
+  data_cf = data_full %>% filter(habitat == "CF"),
+  data_ff = data_full %>% filter(habitat == "1-ha"),
   
   data_1998 = read_wrangle_1998(file_1998),
   data_2008 = read.csv(file_2008),
@@ -54,18 +56,18 @@ tar_plan(
 
   # Starting population for simulations -------------------------------------
 
-  pop_vec_ff = make_pop_vec(data_full %>% filter(habitat == "1-ha"), n_mesh = 100),
-  pop_vec_cf = make_pop_vec(data_full %>% filter(habitat == "CF"),   n_mesh = 100),
+  pop_vec_ff = make_pop_vec(data_ff, n_mesh = 100),
+  pop_vec_cf = make_pop_vec(data_cf, n_mesh = 100),
   
   # Deterministic IPM -------------------------------------------------------
   ## forests fragments
   vit_list_det_ff = c(
     list(
-      vit_surv = surv_det(data_full, habitat = "1-ha"),
-      vit_size = size_det(data_full, habitat = "1-ha"),
-      vit_flwr = flwr_det(data_full, habitat = "1-ha"),
-      vit_size_sdlg = size_sdlg_det(data_full, habitat = "1-ha"),
-      vit_surv_sdlg = surv_sdlg_det(data_full, habitat = "1-ha")
+      vit_surv = surv_det(data_ff),
+      vit_size = size_det(data_ff),
+      vit_flwr = flwr_det(data_ff),
+      vit_size_sdlg = size_sdlg_det(data_ff),
+      vit_surv_sdlg = surv_sdlg_det(data_ff)
     ),
     vit_other_ff),
   
@@ -77,11 +79,11 @@ tar_plan(
   ## continuous forest
   vit_list_det_cf = c(
     list(
-      vit_surv = surv_det(data_full, habitat = "CF"),
-      vit_size = size_det(data_full, habitat = "CF"),
-      vit_flwr = flwr_det(data_full, habitat = "CF"),
-      vit_size_sdlg = size_sdlg_det(data_full, habitat = "CF"),
[…]

### 2022-04-26 notes/understanding stochastic lambda.Rmd

@@
   single_years %>% 
+  filter(habitat == "1-ha") %>% 
   map(~{
@@
       list(
-        vit_surv = surv_det(.x, habitat = "1-ha"),
-        vit_size = size_det(.x, habitat = "1-ha"),
-        vit_flwr = flwr_det(.x, habitat = "1-ha"),
-        vit_size_sdlg = size_sdlg_det(.x, habitat = "1-ha"),
-        vit_surv_sdlg = surv_sdlg_det(.x, habitat = "1-ha")
+        vit_surv = surv_det(.x),
+        vit_size = size_det(.x),
+        vit_flwr = flwr_det(.x),
+        vit_size_sdlg = size_sdlg_det(.x),
+        vit_surv_sdlg = surv_sdlg_det(.x)
       ),
@@
   list(
-    vit_surv = surv_det(two_yrs, habitat = "1-ha"),
-    vit_size = size_det(two_yrs, habitat = "1-ha"),
-    vit_flwr = flwr_det(two_yrs, habitat = "1-ha"),
-    vit_size_sdlg = size_sdlg_det(two_yrs, habitat = "1-ha"),
-    vit_surv_sdlg = surv_sdlg_det(two_yrs, habitat = "1-ha")
+    vit_surv = surv_det(two_yrs),
+    vit_size = size_det(two_yrs),
+    vit_flwr = flwr_det(two_yrs),
+    vit_size_sdlg = size_sdlg_det(two_yrs),
+    vit_surv_sdlg = surv_sdlg_det(two_yrs)
   ),
@@
   list(
-    vit_surv = surv_raneff(two_yrs, habitat = "1-ha"),
-    vit_size = size_raneff(two_yrs, habitat = "1-ha"),
-    vit_flwr = flwr_raneff(two_yrs, habitat = "1-ha"),
-    vit_size_sdlg = size_sdlg_raneff(two_yrs, habitat = "1-ha"),
-    vit_surv_sdlg = surv_sdlg_raneff(two_yrs, habitat = "1-ha")
+    vit_surv = surv_raneff(two_yrs),
+    vit_size = size_raneff(two_yrs),
+    vit_flwr = flwr_raneff(two_yrs),
+    vit_size_sdlg = size_sdlg_raneff(two_yrs),
+    vit_surv_sdlg = surv_sdlg_raneff(two_yrs)
   ), 

### 2022-04-26 R/bcpi.R
calculate bias-corrected percentile CIs

@@ -1,0 +1,28 @@
+#' Bias-correction for bootstrap percentile intervals
+#'
+#' Method described in Caswell (2008) ch. 12. (Eqs 12.19--12.22).
+#'
+#' @param t0 the parameter estimate
+#' @param t a vector of bootstrapped parameter estimates
+#' @param alpha confidence level of the interval
+#'
+#' @return a tibble with the original estimate, `t0`, and the lower and upper
+#'   CIs
+#'
+#' @references Caswell, Hal. Matrix Population Models: Construction, Analysis,
+#'   and Interpretation. 2. ed., Sunderland, Mass: Sinauer Associates, 2008.
+
+#' 
+#'
+#' @examples
+#' tar_load(c(lambda_bt_det_ff, ipm_det_ff))
+#' bcpi(t0 = lambda(ipm_det_ff), t = lambda_bt_det_ff)
+bcpi <- function(t0, t, alpha = 0.05) {
+  B <- length(t)
+  z0 <- qnorm(sum(t < t0)/B)
+  a1 <- pnorm(2 * z0 + qnorm(alpha/2))
+  a2 <- pnorm(2 * z0 + qnorm(1 - alpha/2))
+  c1 <- quantile(t, a1)
+  c2 <- quantile(t, a2)
+  return(tibble::tibble(est = t0, lower = c1, upper = c2))
+}

### 2022-04-26 R/make_lambda_table.R

@@ -1,4 +1,4 @@
-make_lambda_table <- function(ipm_list, bt_list) {
+make_lambda_table <- function(ipm_list, bt_list, alpha = 0.05) {
 
 #not very reusable, but that's fine.
 #ipm_list should look like this:
@@ -17,23 +17,12 @@ make_lambda_table <- function(ipm_list, bt_list) {
   #   # dlnm_ff = lambda_bt_dlnm_ff,
   #   # dlnm_cf = lambda_bt_dlnm_cf
   # )
-
-  
  lambdas <- 
    ipm_list %>% 
-   map_df(~lambda(.x, log = FALSE) %>% set_names("lambda"), .id = "model")
+   map(~lambda(.x, log = FALSE))
   
-  calc_ci <- function(x, alpha = 0.05) {
-    mu <- mean(x)
-    ci <- quantile(x, c(0 + (alpha/2), 1 - (alpha/2)))
-    tibble(mean = mu, lower = ci[1], upper = ci[2])
-  }
-  
-lambdas_bt <-
-  bt_list %>%
-  map_df(calc_ci, .id = "model") #TODO move calc_ci here
-
-full_join(lambdas, lambdas_bt, by = "model") %>%
+ #calc bias-corrected percentile CIs
+  map2_df(lambdas, bt_list, ~bcpi(t0 = .x, t = .y, alpha = alpha), .id = "model") %>%
   separate(model, into = c("ipm", "habitat"))
 
 }

### 2022-04-26 R/make_lambda_tbl.R

@@ -1,13 +1,0 @@
-make_lambda_tbl <- function(...) {
-  all_ipms <- rlang::list2(...)
-  ipm_names <- rlang::ensyms(...)
-  names(all_ipms) <- ipm_names
-  
-  all_ipms %>% 
-    map_df(~ipmr::lambda(.x, log = FALSE), .id = "ipm") %>% 
-    pivot_longer(everything(), names_to = "ipm", names_prefix = "ipm_", values_to = "lambda") %>% 
-    separate(ipm, into = c("IPM", "habitat")) %>% 
-    mutate(habitat = toupper(habitat)) %>% 
-    pivot_wider(names_from = habitat, values_from = lambda) 
-  
-}

### 2022-04-26 notes/optimizing bootstrapping.Rmd
add notes on optimizing bootstrapping

---
title: "Optimizing Bootstrapping"
author: "Eric R. Scott"
date: "2022-04-26"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    number_sections: yes
    highlight: kate
    theme:
      version: 4
      bootswatch: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(conflicted)
library(targets)

source(here("R", "bcpi.R"))

conflict_prefer("lag", "dplyr")
conflict_prefer("filter", "dplyr")
```

*Last compiled: `r Sys.Date()`*

# Purpose

How many bootstraps is enough to estimate CIs around lambda?

# Load Data

```{r data, echo=TRUE}
withr::with_dir(here::here(), tar_load(c(
  ipm_det_cf,
  ipm_det_ff,
  ipm_stoch_cf,
  ipm_stoch_ff,
  lambda_bt_det_cf,
  lambda_bt_det_ff,
  lambda_bt_stoch_cf,
  lambda_bt_stoch_ff
)))

```

```{r}
est <- lambda(ipm_det_cf)
df <- seq_along(lambda_bt_det_cf) %>% 
  map_df(~bcpi(t0 = est, t = lambda_bt_det_cf[1:.x]), .id = "B") %>% 
  mutate(B = as.numeric(B))

ggplot(df,aes(x = B)) +
  geom_line(aes(y = lower), color = "red") +
  geom_line(aes(y = upper), color = "blue")
```

Looks like it does converge after like 200 bootstraps

```{r}
est <- lambda(ipm_det_ff)
df <- seq_along(lambda_bt_det_ff) %>% 
  map_df(~bcpi(t0 = est, t = lambda_bt_det_ff[1:.x]), .id = "B") %>% 
  mutate(B = as.numeric(B))

ggplot(df,aes(x = B)) +
  geom_line(aes(y = lower), color = "red") +
  geom_line(aes(y = upper), color = "blue")
```

```{r}
est <- lambda(ipm_stoch_cf, log = FALSE)
df <- seq_along(lambda_bt_stoch_cf) %>% 
  map_df(~bcpi(t0 = est, t = lambda_bt_stoch_cf[1:.x]), .id = "B") %>% 
  mutate(B = as.numeric(B))

ggplot(df,aes(x = B)) +
  geom_line(aes(y = lower), color = "red") +
  geom_line(aes(y = upper), color = "blue")
```

For stochastic it does take longer.
Why the sudden leaps for the upper limit?

```{r}
est <- lambda(ipm_stoch_ff, log = FALSE)
df <- seq_along(lambda_bt_stoch_ff) %>% 
  map_df(~bcpi(t0 = est, t = lambda_bt_stoch_ff[1:.x]), .id = "B") %>% 
  mutate(B = as.numeric(B))

ggplot(df,aes(x = B)) +
  geom_line(aes(y = lower), color = "red") +
  geom_line(aes(y = upper), color = "blue")
```

### 2022-04-26 R/make_lambda_table.R
update lambda table

@@ -1,28 +1,29 @@
-make_lambda_table <- function(ipm_list, bt_list, alpha = 0.05) {
-
 #not very reusable, but that's fine.
 #ipm_list should look like this:
-  # ipm_list <- list(
-  #   det_ff = ipm_det_ff,
-  #   det_cf = ipm_det_cf,
-  #   stoch_ff = ipm_stoch_ff,
-  #   stoch_cf = ipm_stoch_cf
-  # )
+# ipm_list <- list(
+#   det_ff = ipm_det_ff,
+#   det_cf = ipm_det_cf,
+#   stoch_ff = ipm_stoch_ff,
+#   stoch_cf = ipm_stoch_cf
+# )
 #bt_list should look like this:
-  # bt_list <- list(
-  #   det_ff = lambda_bt_det_ff,
-  #   det_cf = lambda_bt_det_cf,
-  #   stoch_ff = lambda_bt_stoch_ff,
-  #   stoch_cf = lambda_bt_stoch_cf
-  #   # dlnm_ff = lambda_bt_dlnm_ff,
-  #   # dlnm_cf = lambda_bt_dlnm_cf
-  # )
+# bt_list <- list(
+#   det_ff = lambda_bt_det_ff,
+#   det_cf = lambda_bt_det_cf,
+#   stoch_ff = lambda_bt_stoch_ff,
+#   stoch_cf = lambda_bt_stoch_cf
+#   # dlnm_ff = lambda_bt_dlnm_ff,
+#   # dlnm_cf = lambda_bt_dlnm_cf
+# )
+
+make_lambda_table <- function(ipm_list, bt_list, alpha = 0.05) {
+
  lambdas <- 
    ipm_list %>% 
    map(~lambda(.x, log = FALSE))
   
  #calc bias-corrected percentile CIs
-  map2_df(lambdas, bt_list, ~bcpi(t0 = .x, t = .y, alpha = alpha), .id = "model") %>%
-  separate(model, into = c("ipm", "habitat"))
+ map2_df(lambdas, bt_list, ~bcpi(t0 = .x, t = .y, alpha = alpha), .id = "model") %>%
+   separate(model, into = c("ipm", "habitat"))
 
 }

### 2022-04-26 _targets.R

@@ -251,9 +251,9 @@ tar_plan(
         det_ff = ipm_det_ff,
         det_cf = ipm_det_cf,
         stoch_ff = ipm_stoch_ff,
-        stoch_cf = ipm_stoch_cf,
-        dlnm_ff = ipm_dlnm_ff,
-        dlnm_cf = ipm_dlnm_cf
+        stoch_cf = ipm_stoch_cf#,
+        # dlnm_ff = ipm_dlnm_ff,
+        # dlnm_cf = ipm_dlnm_cf
       ),
       bt_list = list(
         det_ff = lambda_bt_det_ff,

### 2022-04-26 notes/understanding stochastic lambda.Rmd
update notes on stochastic lambda

@@
 
+# TODO
+
+-   Get code to run
+
+-   Instead of deterministic IPM using different vital rates models, use the random effects models, but plug in `newdata = (log_size_prev = log_size, year = "newlevel")` to get mean fitted values.
+
+-   add analysis with CF
+
+-   Don't normalize population sizes and plot both CF and FF as PVA for both models
+
+-   Try year as fixed effect
+
+-   Read Rees & Ellner 2009.
+    Consider either fixed effect of year + kernel sampling OR random effect of year + parameter sampling.
+    Maybe no added value for random effect of year + kernel sampling??
+
+Rees, Mark, and Stephen P. Ellner.
+"Integral Projection Models for Populations in Temporally Varying Environments." Ecological Monographs 79, no. 4 (2009): 575--94.
+<https://doi.org/10.1890/08-1474.1>.
+
 *Last compiled: `r Sys.Date()`*
@@
 withr::with_dir(here(),{
-  tar_load(data_full)
-  tar_load(vit_other_ff)
+  tar_load(c(
+    data_full,
+    pop_vec_ff,
+    pop_vec_cf,
+    vit_other_ff,
+    vit_other_cf
+  ))
 })
+
+#don't need matrix columns for this
+data_full <- data_full %>% select(-L, -spei_history)
 ```
@@
   filter(year>1999) %>% 
-  select(-L, -spei_history) %>% 
   group_by(year_fct) %>% 
@@
 
-vit_list_yr <- 
+#fit vital rates for FF  
+vit_list_yr_ff <- 
   single_years %>% 
-  filter(habitat == "1-ha") %>% 
+  map(~dplyr::filter(.x, habitat == "1-ha")) %>% 
   map(~{
-    vit_list_det_ff = c(
+    vit_list_det = c(
       list(
@@
     })
-
-ipms_single_yr <-
-  map(vit_list_yr, ~{
-    make_proto_ipm_det(.x) %>% 
+# make IPMs for FF
+ipms_yr_ff <-
+  map(vit_list_yr_ff, ~{
+    make_proto_ipm_det(.x, pop_vec = pop_vec_ff) %>% 
       make_ipm(iterations = 1000,  #only needs 100 to converge
@@
 
-lambdas <- ipms_single_yr %>% map_dbl(lambda)
+# fit vital rates for CF
+vit_list_yr_cf <- 
+  single_years %>% 
+  map(~dplyr::filter(.x, habitat == "CF")) %>% 
+  map(~{
+    vit_list_det = c(
+      list(
+        vit_surv = surv_det(.x),
+        vit_size = size_det(.x),
+        vit_flwr = flwr_det(.x),
+        vit_size_sdlg = size_sdlg_det(.x),
+        vit_surv_sdlg = surv_sdlg_det(.x)
+      ),
+      vit_other_cf)
+    })
+
+#make IPMs for CF
+ipms_yr_cf <-
+  map(vit_list_yr_cf, ~{
+    make_proto_ipm_det(.x, pop_vec = pop_vec_cf) %>% 
+      make_ipm(iterations = 1000,  #only needs 100 to converge
+               normalize_pop_size = FALSE, # to run as PVA
+               usr_funs = list(get_scat_params = get_scat_params))
+  })
+
+#lambdas
+lambdas_ff <- ipms_yr_ff %>% map_dbl(lambda)
+lambdas_cf <- ipms_yr_cf %>% map_dbl(lambda)
 ```
@@
 #worst year
-(worst_year <- c(2000:max(data_full$year))[which.min(lambdas)]); min(lambdas)
+(worst_year_ff <- c(2000:max(data_full$year))[which.min(lambdas_ff)]); min(lambdas_ff)
+(worst_year_cf <- c(2000:max(data_full$year))[which.min(lambdas_cf)]); min(lambdas_cf)
 
 #best year
-(best_year <- c(2000:max(data_full$year))[which.max(lambdas)]); max(lambdas)
+(best_year_ff <- c(2000:max(data_full$year))[which.max(lambdas_ff)]); max(lambdas_ff)
+(best_year_cf <- c(2000:max(data_full$year))[which.max(lambdas_cf)]); max(lambdas_cf)
 
-two_yrs <- data_full %>% filter(year %in% c(best_year, worst_year))
+plot(2000:max(data_full$year), lambdas_cf, type = "l", col = "red"); lines(x =2000:max(data_full$year), y = lambdas_ff, col = "blue", type = "l")
+```
+
+2003 was the worst year in both habitats.
+2001 was a good year in both habitats (but not the best in CF)
+
+```{r warning=FALSE}
+#combine good and bad year
+two_yrs <- data_full %>% filter(year %in% c(2003, 2001))
+two_yrs_ff <- two_yrs %>% filter(habitat == "1-ha")
+two_yrs_cf <- two_yrs %>% filter(habitat == "CF")
 ```
@@
 
+Fragments:
+
 ```{r warning=FALSE, class.source = 'fold-hide'}
@@
   list(
-    vit_surv = surv_det(two_yrs),
-    vit_size = size_det(two_yrs),
-    vit_flwr = flwr_det(two_yrs),
-    vit_size_sdlg = size_sdlg_det(two_yrs),
-    vit_surv_sdlg = surv_sdlg_det(two_yrs)
+    vit_surv = surv_det(two_yrs_ff),
+    vit_size = size_det(two_yrs_ff),
+    vit_flwr = flwr_det(two_yrs_ff),
+    vit_size_sdlg = size_sdlg_det(two_yrs_ff),
+    vit_surv_sdlg = surv_sdlg_det(two_yrs_ff)
   ),
@@
 
-ipm_det_ff = make_proto_ipm_det(vit_list_det_ff) %>% 
+ipm_det_ff = make_proto_ipm_det(vit_list_det_ff, pop_vec = pop_vec_ff) %>% 
   make_ipm(iterations = 1000,  #only needs 100 to converge
@@
            usr_funs = list(get_scat_params = get_scat_params))
+# FF
 lambda(ipm_det_ff)
+```
+
+CF:
+
+```{r warning=FALSE, class.source = 'fold-hide'}
+vit_list_det_cf = c(
+  list(
+    vit_surv = surv_det(two_yrs_cf),
+    vit_size = size_det(two_yrs_cf),
+    vit_flwr = flwr_det(two_yrs_cf),
+    vit_size_sdlg = size_sdlg_det(two_yrs_cf),
+    vit_surv_sdlg = surv_sdlg_det(two_yrs_cf)
+  ),
+  vit_other_cf)
+
+ipm_det_cf = make_proto_ipm_det(vit_list_det_cf, pop_vec = pop_vec_cf) %>% 
+  make_ipm(iterations = 1000,  #only needs 100 to converge
+           normalize_pop_size = FALSE, # to run as PVA
+           usr_funs = list(get_scat_params = get_scat_params))
+lambda(ipm_det_cf)
 ```
@@
 ```{r warning=FALSE, class.source = 'fold-hide'}
-## forets fragments
+## forests fragments
 vit_list_stoch_ff = c(
   list(
-    vit_surv = surv_raneff(two_yrs),
-    vit_size = size_raneff(two_yrs),
-    vit_flwr = flwr_raneff(two_yrs),
-    vit_size_sdlg = size_sdlg_raneff(two_yrs),
-    vit_surv_sdlg = surv_sdlg_raneff(two_yrs)
+    vit_surv = surv_raneff(two_yrs_ff),
+    vit_size = size_raneff(two_yrs_ff),
+    vit_flwr = flwr_raneff(two_yrs_ff),
+    vit_size_sdlg = size_sdlg_raneff(two_yrs_ff),
+    vit_surv_sdlg = surv_sdlg_raneff(two_yrs_ff)
   ), 
@@
   #arbitrary starting population state
-  define_pop_state(n_log_size = runif(100), n_sdlg = .1)
+  define_pop_state(pop_vectors = pop_vec_ff)
+```
+
+```{r warning=FALSE, class.source = 'fold-hide'}
+## forests fragments
+vit_list_stoch_cf = c(
+  list(
+    vit_surv = surv_raneff(two_yrs_cf),
+    vit_size = size_raneff(two_yrs_cf),
+    vit_flwr = flwr_raneff(two_yrs_cf),
+    vit_size_sdlg = size_sdlg_raneff(two_yrs_cf),
+    vit_surv_sdlg = surv_sdlg_raneff(two_yrs_cf)
+  ), 
+  vit_other_cf)
+
+proto_ipm_stoch_cf <- 
+  init_ipm(
+    sim_gen = "general",
+    di_dd = "di",
+    det_stoch = "stoch",
+    kern_param = "kern"
+  ) %>%
+
+  # Define growth and survival kernel
+  define_kernel(
+    "P_yr", #_yr makes a different P kernel for each value of `yr` from par_set_indices
+    formula = s_z * G_z1z * d_log_size,
+    family  = "CC",
+    # Probability of survival
+    s_z = predict(vit_surv,
+                  newdata = tibble(log_size_prev = log_size_1, year_fct = yr),
+                  type = 'response'),
+    # Growth
+    G_z1z = dt.scaled(
+      log_size_2,
+      df   = size_nu,
+      mean = size_mu,
+      sd   = size_sd
+    ),
+    size_mu = predict(vit_size,
+                      newdata = tibble(log_size_prev = log_size_1, year_fct = yr),
+                      type = 'response'),
+    size_nu = get_scat_params(vit_size)["nu"],
+    size_sd = get_scat_params(vit_size)["sd"],
+
+    data_list = vit_list_stoch_cf,
+    states          = list(c('log_size')),
+    uses_par_sets   = TRUE,
+    par_set_indices = list(yr = c(2001, 2003)),
+    evict_cor       = TRUE,
+    evict_fun       = truncated_distributions("t.scaled", "G_z1z")
+  ) %>%
+
+  # Define fecundity and recruitment kernel
+  define_kernel(
+    "go_sdlg_yr",
+    formula = p_f * fruits * seeds * g_e * d_log_size,
+    family  = "CD",
+    # probability of flowering
+    p_f = predict(vit_flwr,
+                  newdata = tibble(log_size_prev = log_size_1, year_fct = yr),
+                  type = 'response'),
+    fruits = predict(vit_fruits,
+                     newdata = tibble(log_size_prev = log_size_1),
+                     type = 'response'), 
+    seeds = exp(coef(vit_seeds)[1]), #mean seeds per fruit
+    g_e = vit_germ_est,
+
+    data_list     = vit_list_stoch_cf,
+    states        = list(c('log_size', 'sdlg')),
+    uses_par_sets = TRUE,
+    par_set_indices = list(yr = c(2001, 2003))
+  ) %>%
+
+  define_kernel(
+    name = "stay_sdlg",
+    formula = 0,
+    family = "DD",
+    states = list(c("sdlg")),
+    evict_cor = FALSE
+  ) %>%
+
+  # Enter into post-seedling
+  define_kernel(
+    name = "leave_sdlg_yr",
+    formula = s_sdlg * G_z2_sdlg,
+    family = "DC",
+    s_sdlg = predict(vit_surv_sdlg,
+                     newdata = tibble(year_fct = yr),
+                     type = 'response'), #intercept from intercept-only glm
+    G_z2_sdlg = dt.scaled(
+      x    = log_size_2,
+      df   = size_sdlg_nu,
+      mean = size_sdlg_mu,
+      sd   = size_sdlg_sd
+    ),
+    size_sdlg_mu = coef(vit_size_sdlg)[1], #intercept from intercept-only glm
+    size_sdlg_nu = get_scat_params(vit_size_sdlg)["nu"],
+    size_sdlg_sd = get_scat_params(vit_size_sdlg)["sd"],
+
+    data_list = vit_list_stoch_cf,
+    states = list(c("sdlg", "log_size")),
+    uses_par_sets = TRUE,
+    par_set_indices = list(yr = c(2001, 2003)),
+    evict_cor = TRUE,
+    evict_fun = truncated_distributions("t.scaled", "G_z2_sdlg")
+  ) %>%
+
+  # define implementation with midpoint rule
+  define_impl(make_impl_args_list(
+    kernel_names = c("P_yr",        "go_sdlg_yr", "stay_sdlg", "leave_sdlg_yr"),
+    int_rule = "midpoint",
+    state_start  = c("log_size", "log_size", "sdlg",     "sdlg"),
+    state_end    = c("log_size", "sdlg",     "sdlg",     "log_size")
+  )) %>% 
+  #define lower bound, upper bound, and number of meshpoints for log_size
+  define_domains(log_size = c(0, 8.018296, 100)) %>%  
+  #arbitrary starting population state
+  define_pop_state(pop_vectors = pop_vec_cf)
 ```
@@
            usr_funs = list(get_scat_params = get_scat_params))
+
+ipm_stoch_cf <- 
+  proto_ipm_stoch_cf %>%
+  make_ipm(iterations = 1000,
+           normalize_pop_size = FALSE, # to run as PVA
+           usr_funs = list(get_scat_params = get_scat_params))
 ```
@@
 ```{r}
-lambda(ipm_det_ff) 
-lambda(ipm_stoch_ff, log = FALSE, burn_in = 0.2)
+all_lams <-
+  list(ff = ipm_stoch_ff, cf = ipm_stoch_cf) %>% 
+  map_df(
+    ~tibble(year = .x$env_seq,
+         lambda = lambda(.x, type_lambda = "all")[,1])%>% 
+  #remove burnin
+  slice(101:n()),
+  .id = "habitat"
+  )
+
 ```
 
-Yes, stochastic lambda is still a bit higher!
-
-What about the averages for each year?
-Are they consistent with a bad and good year?
-
-```{r}
-all_lams <- 
-  tibble(year = ipm_stoch_ff$env_seq,
-         lambda = lambda(ipm_stoch_ff, type_lambda = "all")[,1])
+```{r lambdas}
+tribble(~model, ~habitat, ~year, ~lambda,
+        "det", "ff", "both combined", lambda(ipm_det_ff),
+        "stoch", "ff", "both combined", lambda(ipm_stoch_ff, log = FALSE),
+        "det", "cf", "both combined", lambda(ipm_det_cf),
+        "stoch", "cf", "both combined", lambda(ipm_stoch_ff, log = FALSE),
+        "det", "ff", "2001", lambdas_ff[2],
+        "det", "ff", "2003", lambdas_ff[4],
+        "det", "cf", "2001", lambdas_cf[2],
+        "det", "cf", "2003", lambdas_cf[4]
+) %>% 
+  bind_rows(
     all_lams %>%
-  group_by(year) %>%
-  summarize(mean(lambda))
+      group_by(habitat, year) %>%
+      summarize(lambda = mean(lambda)) %>% 
+      mutate(year = paste(year, "mean"),
+             model = "stoch")
+  ) %>% arrange(model, habitat)
 ```
 
-Yes, but both the good and bad years mean lambdas are *greater* than the deterministic lambda.
+Stochastic lambda is still slightly higher than deterministic
 
-How do the mean lambdas from the good and bad year compare to the deterministic lambdas from individual years?
-
-```{r}
-max(lambdas)
-min(lambdas)
-```
+Within stochastic IPMs, the mean lambdas for individual transition years are consistent with good and bad years, but both are *higher* than deterministic lambda with both years combined!
 
@@
 #deterministic, just 2001
-det_fitted_2001 <- 
-  vit_list_yr[[which.max(lambdas)]] %>% 
+det_fitted_2001_ff <- 
+  vit_list_yr_ff[[2]] %>% 
   map_df(~safe_predict(.x, newdata = newdata, type = "response")) %>% 
@@
 #deterministic, just 2003
-det_fitted_2003 <- 
-  vit_list_yr[[which.min(lambdas)]] %>% 
+det_fitted_2003_ff <- 
+  vit_list_yr_ff[[4]] %>% 
   map_df(~safe_predict(.x, newdata = newdata, type = "response")) %>% 
@@
 #deterministic with both years combined
-det_fitted <- 
+det_fitted_ff <- 
   vit_list_det_ff %>% 
@@
 #stochastic
-stoch_fitted <- 
+stoch_fitted_ff <- 
   vit_list_stoch_ff %>%
@@
 
-bind_rows(stoch_fitted, det_fitted, det_fitted_2001, det_fitted_2003) %>% 
+bind_rows(stoch_fitted_ff, det_fitted_ff, det_fitted_2001_ff, det_fitted_2003_ff) %>% 
   ggplot(aes(x = log_size_prev, y = .fitted, color = year_fct, linetype = year_fct)) +
   geom_line(alpha = 0.9) +
-  facet_wrap(~vital_rate, scales = "free")
+  facet_wrap(~vital_rate, scales = "free") +
+  labs(title = "Forest Fragments")
+
+```
+
+```{r class.source = 'fold-hide'}
+newdata <- expand.grid(
+  log_size_prev = seq(min(two_yrs$log_size_prev, na.rm = TRUE), max(two_yrs$log_size_prev, na.rm = TRUE), length.out = 50),
+  year_fct = as.character(c(2001, 2003))
+)
+
+safe_predict <- possibly(predict, NA)
+
+#deterministic, just 2001
+det_fitted_2001_cf <- 
+  vit_list_yr_cf[[2]] %>% 
+  map_df(~safe_predict(.x, newdata = newdata, type = "response")) %>% 
+  bind_cols(newdata) %>% 
+  select(-vit_fruits, -vit_seeds, -vit_germ_est) %>% 
+  pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_") %>% 
+  mutate(year_fct = "det. 2001")
+
+#deterministic, just 2003
+det_fitted_2003_cf <- 
+  vit_list_yr_cf[[4]] %>% 
+  map_df(~safe_predict(.x, newdata = newdata, type = "response")) %>% 
+  bind_cols(newdata) %>% 
+  select(-vit_fruits, -vit_seeds, -vit_germ_est) %>% 
+  pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_") %>% 
+  mutate(year_fct = "det. 2003")
+
+#deterministic with both years combined
+det_fitted_cf <- 
+  vit_list_det_cf %>% 
+  map_df(~safe_predict(.x, newdata = newdata, type = "response")) %>% 
+  bind_cols(newdata) %>%
+  select(-vit_fruits, -vit_seeds, -vit_germ_est) %>% 
+  pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_") %>% 
+  mutate(year_fct = "det. both years")
+
+#stochastic
+stoch_fitted_cf <- 
+  vit_list_stoch_cf %>%
+  map_df(~safe_predict(.x, newdata= newdata, type = "response")) %>%
+  bind_cols(newdata) %>%
+  select(-vit_fruits, -vit_seeds, -vit_germ_est) %>% 
+  pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_") %>% 
+  mutate(year_fct = paste("stoch.", year_fct))
+
+bind_rows(stoch_fitted_cf, det_fitted_cf, det_fitted_2001_cf, det_fitted_2003_cf) %>% 
+  ggplot(aes(x = log_size_prev, y = .fitted, color = year_fct, linetype = year_fct)) +
+  geom_line(alpha = 0.9) +
+  facet_wrap(~vital_rate, scales = "free") +
+  labs(title = "Continuous Forest")
 

### 2022-04-26 notes/understanding stochastic lambda.Rmd
updated notes

@@
 
--   Get code to run
-
 -   Instead of deterministic IPM using different vital rates models, use the random effects models, but plug in `newdata = (log_size_prev = log_size, year = "newlevel")` to get mean fitted values.
 
--   add analysis with CF
-
--   Don't normalize population sizes and plot both CF and FF as PVA for both models
-
--   Try year as fixed effect
+-   Plot population trajectories for all models
 
@@
 
-```{r}
+```{r class.source = 'fold-hide'}
 all_lams <-
@@
 
-```{r lambdas}
+```{r lambdas, class.source = 'fold-hide'}
 tribble(~model, ~habitat, ~year, ~lambda,

### 2022-05-03 README.Rmd
add more notes for github pages
link to github *pages*
update README with project description

@@
 
+The idea here is to investigate the consequences of including lagged effects (such as those modeled by distributed lag non-linear models) into population models (integral projection models, IPMs).
+We'll compare population growth rates (lambda) with deterministic, stochastic (matrix shuffling), and lagged effects IPMs.
+
 The most recent draft of the manuscript is available [here](https://brunalab.github.io/lagged-ipms/paper.html).
+
+[Notes for collaborators](https://brunalab.github.io/lagged-ipms/notes/index.html)
 
@@
 -   Run `targets::tar_make()` or `targets::tar_make_clustermq()` from the R console to run all code and produce all outputs.
-
-<!-- Consider using `targets::tar_github_actions()` if targets aren't too big -->

### 2022-05-03 notes/understanding stochastic lambda.Rmd
finished notes after updating ipmr

@@
 
+```{r}
+year_seq = sample(c(2001, 2003), size = 1000, replace = TRUE)
+```
+
 ```{r warning=FALSE}
@@
            normalize_pop_size = FALSE, # to run as PVA
+           kernel_seq = as.character(year_seq),
            usr_funs = list(get_scat_params = get_scat_params))

### 2022-05-04 _targets.R
rearranged _targets.R to make the outline view look nice

@@ -39,7 +39,9 @@ tar_plan(
   data_2008 = read.csv(file_2008),
   
 
-  # Shared vital rates ------------------------------------------------------
+  # Vital rate models -------------------------------------------------------
+
+  ## ├Shared vital rates ----------------------------------------------------
   # Vital rates from other datasets besides the 10 year demographic experiment
   vit_other_ff = list(
     vit_fruits = fruits_gam(data_1998),
@@ -53,13 +55,8 @@ tar_plan(
     vit_germ_est = 0.057098765 #germination and establishment
   ),
 
-
-  # Starting population for simulations -------------------------------------
-
-  pop_vec_ff = make_pop_vec(data_ff, n_mesh = 100),
-  pop_vec_cf = make_pop_vec(data_cf, n_mesh = 100),
+  ## ├Deterministic ---------------------------------------------------------
   
-  # Deterministic IPM -------------------------------------------------------
   ## forests fragments
   vit_list_det_ff = c(
     list(
@@ -70,11 +67,6 @@ tar_plan(
       vit_surv_sdlg = surv_sdlg_det(data_ff)
     ),
     vit_other_ff),
-  
-  ipm_det_ff = make_proto_ipm_det(vit_list_det_ff, pop_vec_ff) %>% 
-    make_ipm(iterations = 1000,  #only needs 100 to converge
-             normalize_pop_size = FALSE, # to run as PVA
-             usr_funs = list(get_scat_params = get_scat_params)),
   
   ## continuous forest
   vit_list_det_cf = c(
@@ -87,15 +79,9 @@ tar_plan(
     ),
     vit_other_cf),
   
-  ipm_det_cf = make_proto_ipm_det(vit_list_det_cf, pop_vec_cf) %>% 
-    make_ipm(iterations = 1000, #only needs 100 to converge
-             normalize_pop_size = FALSE, # to run as PVA
-             usr_funs = list(get_scat_params = get_scat_params)),
+  ## ├Stochastic (random effect of year) ------------------------------------
   
-  
-  # Stochastic kernel resampled IPM -----------------------------------------
-  
-  ## forets fragments
+  ## forest fragments
   vit_list_stoch_ff = c(
     list(
       vit_surv = surv_raneff(data_ff),
@@ -105,11 +91,6 @@ tar_plan(
[…]

### 2022-05-04 _targets.R
document targets

@@ -177,31 +177,29 @@ tar_plan(
              usr_funs = list(get_scat_params = get_scat_params)),
   
   ## ├Stochastic, DLNM ------------------------------------------------------
+  # These are set up a little differently since to iterate the IPM I need to
+  # first create the sequence of environments (lagged SPEI history) from year
+  # sequence of years and the climate data.  `make_dlnm_ipm()` does this, then
+  # add the environmental states to the proto IPM, then iterates the IPM.
+  
   #TODO: make this use year_seq
   proto_ipm_dlnm_ff = make_proto_ipm_dlnm(vit_list_dlnm_ff, pop_vec_ff),
-  tar_target(
-    ipm_dlnm_ff,
-    proto_ipm_dlnm_ff %>%
-      make_dlnm_ipm(
-        clim,
-        seed = 1234,
-        iterations = 1000,
-        usr_funs = list(get_scat_params = get_scat_params)
-      )
-  ),
+  ipm_dlnm_ff = proto_ipm_dlnm_ff %>%
+    make_dlnm_ipm(
+      clim,
+      seed = 1234,
+      iterations = 1000,
+      usr_funs = list(get_scat_params = get_scat_params)
+    ),
  
   proto_ipm_dlnm_cf = make_proto_ipm_dlnm(vit_list_dlnm_cf, pop_vec_cf),
-  tar_target(
-    ipm_dlnm_cf,
-    proto_ipm_dlnm_cf %>%
-      make_dlnm_ipm(
-        clim,
-        seed = 1234,
-        iterations = 1000,
-        usr_funs = list(get_scat_params = get_scat_params)
-      )
-  ),
-
+  ipm_dlnm_cf = proto_ipm_dlnm_cf %>%
+    make_dlnm_ipm(
+      clim,
+      seed = 1234,
+      iterations = 1000,
+      usr_funs = list(get_scat_params = get_scat_params)
+    ),
  
   # Bootstrapped Lambdas -----------------------------------------------------------------
   # uses tar_rep() target factory to create dynamic branches to do 500
@@ -224,6 +222,7 @@ tar_plan(
   ),
   
   ## ├Stochastic, matrix sampling --------------------------------------------
+  #TODO: also make these use year_seq??
   tar_rep(
     lambda_bt_stoch_ff,
     ipm_boot_stoch(data_ff, vit_other = vit_other_ff),
@@ -242,6 +241,7 @@ tar_plan(
[…]

### 2022-05-06 R/make_dlnm_ipm.R
make DLNM ipms able to use specific sequence of years (#8)

@@ -1,12 +1,53 @@
-make_dlnm_ipm <- function(proto_ipm, clim, seed, iterations, ...) {
+#' Add lagged SPEI history and iterate a DLNM-based IPM
+#' 
+#' Takes a `proto_ipm` object and the `clim` dataset, calculates lagged SPEI
+#' history, adds those environmental states to the `proto_ipm`, then iterates
+#' it. Takes *either* a specific sequence of years found in the `clim` dataset
+#' (`year_seq`) *or* a number of `iterations`.
+#'
+#' @param proto_ipm a proto_ipm built using DLNM vital rates models
+#' @param clim the `clim` target climate dataset
+#' @param year_seq a vector of years passed to `make_env_states()`
+#' @param seed random seed passed to `make_env_states()`
+#' @param iterations number of iterations passed to `make_env_states()`
+#' @param ... additional arguments passed to `ipmr::make_ipm()`
+#'
+#' @details When `year_seq` is specified, that exact sequence of years is drawn
+#'   from the climate dataset.  However, because lags are calculated, there is
+#'   incomplete data for the first few years.  Therefore, a number of years are
+#'   added to the beginning of the sequence so that lagged SPEI can be
+#'   calculated, but then trimmed from the final dataset so that the total
+#'   number of iterations is the same as the length of `year_seq`.  These few
+#'   "burn-in" years are chosen randomly, and when `year_seq` is specified, the
+#'   behavior of `seed` is to make these "burn-in" years reproducible.  In the
+#'   case when `year_seq` is not specified, years are chosen at random
+#'   (reproducibly if `seed` is set), with a "burn-in" so the total number of
+#'   environmental states is `iterations`.
+#'   
+#' @return an iterated ipm
+#' 
+make_dlnm_ipm <- function(proto_ipm, clim, year_seq = NULL, seed = NULL, iterations = NULL, ...) {
   
+  if(is.null(year_seq) & is.null(iterations)) {
+    stop("One of `year_seq` or `iterations` must be specified")
+  }
+  
+  if(!is.null(year_seq) & !is.null(iterations)) {
+    stop("Only `year_seq` or `iterations` can be specified, not both")
+  }
+  
+  # if iterations are specified, draw random years with make_env_states()
+  env_states <- make_env_states(clim, seed = seed, maxlag = 36, iterations = iterations, year_seq = year_seq)
+  n_it <- nrow(env_states) #should be the same as iterations, but just in case.
+  
[…]

### 2022-05-06 R/make_env_states.R

@@ -1,4 +1,4 @@
-make_env_states <- function(clim, seed = NULL, maxlag = 36, iterations = NULL) {
+make_env_states <- function(clim, seed = NULL, maxlag = 36, iterations = NULL, year_seq = NULL) {
   env_df <- 
     clim %>% 
     #remove leading NAs and any Inf or -Inf values that might exist in spei
@@ -23,15 +23,26 @@ make_env_states <- function(clim, seed = NULL, maxlag = 36, iterations = NULL) {
   
   #pick random years
   yrs <- unique(env_df$year)
-  if (is.null(iterations)) {
-    iterations <- length(yrs)
+  
+  if(!is.null(year_seq)) {
+  #check that years in year_seq are in env_df
+    if(any(!unique(year_seq) %in% yrs)) {
+      stop("Not all values of year_seq are in the clim dataset")
+    }
+    iterations <- length(year_seq)
+    #add burnin years to the beginning
+    year_seq <- c(sample(yrs, burnin, replace = TRUE), year_seq)
+  } else {
+    #samples n_years years plus necessary "burn-in" years to calculate lags.  These will be removed later.
+    if (is.null(iterations)) {
+      iterations <- length(yrs)
+    }
+    year_seq <- sample(yrs, iterations + burnin, replace = TRUE)
   }
   
-  #samples n_years years plus necessary "burn-in" years to calculate lags.  These will be removed later.
-  yrs_samp <- sample(yrs, iterations + burnin, replace = TRUE)
-  # pull those random years from the dataset to construct a new dataset in the random order
+  # pull those random years from the dataset to construct a new dataset in the order of year_seq
   env_states <- 
-    map_dfr(yrs_samp, ~ env_df %>% 
+    map_dfr(year_seq, ~ env_df %>% 
                dplyr::filter(year == .x)) %>% 
     # then calculate lagged `spei_history` starting in February of each year going back `maxlag` months
     mutate(spei_history = tsModel::Lag(spei, 0:maxlag)) %>% 
@@ -39,7 +50,7 @@ make_env_states <- function(clim, seed = NULL, maxlag = 36, iterations = NULL) {
     slice(-c(1:burnin)) %>% 
     mutate(t = 1:n()) %>% #add row index
     #return just what's needed to iterate
-    select(t, spei_history)
+    select(year, t, spei_history)
   
[…]

### 2022-05-06 R/ipm_boot_det.R
add year_seq argument to bootstrapping functions also (#8)

@@ -7,10 +7,11 @@
 #' 
 #' @param data demographic data.  E.g. the data_ff or data_cf targets
 #' @param vit_other a list containing vit_fruits, vit_seeds, and vit_germ_est
-#'
+#' @param ... other arguments passed to `ipmr::make_ipm()`
+#' 
 #' @return lambda
 #' 
-ipm_boot_det <- function(data, vit_other) {
+ipm_boot_det <- function(data, vit_other, ...) {
   #sample ha_id_numbers with replacement within plots
   boot_ids <-
     data %>% 
@@ -37,7 +38,8 @@ ipm_boot_det <- function(data, vit_other) {
   make_proto_ipm_det(vit_list_det, pop_vec) %>% 
     make_ipm(iterations = 100,  #only needs 100 to converge
              normalize_pop_size = TRUE,
-             usr_funs = list(get_scat_params = get_scat_params)
+             usr_funs = list(get_scat_params = get_scat_params),
+             ...
     ) %>% 
     #calculate lambda
     ipmr::lambda(log = FALSE)

### 2022-05-06 R/ipm_boot_dlnm.R

@@ -8,10 +8,11 @@
 #' @param data demographic data.  E.g. the data_ff or data_cf targets
 #' @param vit_other a list containing vit_fruits, vit_seeds, and vit_germ_est
 #' @param clim the clim target.  Full climate timeseries (no lags calculated).
+#' @param ... other arguments passed to `ipmr::make_ipm()`
 #'
 #' @return lambda
 #' 
-ipm_boot_dlnm_raw <- function(data, vit_other, clim) {
+ipm_boot_dlnm_raw <- function(data, vit_other, clim, year_seq = NULL, ...) {
   #sample ha_id_numbers with replacement within plots
   boot_ids <-
     data %>% 
@@ -39,6 +40,7 @@ ipm_boot_dlnm_raw <- function(data, vit_other, clim) {
     make_dlnm_ipm(clim, seed = 1234, iterations = 1000,
                   return_sub_kernels = FALSE, # don't save every iteration
                   normalize_pop_size = TRUE,
+                  year_seq = year_seq,
                   usr_funs = list(get_scat_params = get_scat_params)) %>% 
     #calculate lambda
     ipmr::lambda(log = FALSE)

### 2022-05-06 R/ipm_boot_stoch.R

@@ -7,10 +7,11 @@
 #' 
 #' @param data demographic data.  E.g. the data_ff or data_cf targets
 #' @param vit_other a list containing vit_fruits, vit_seeds, and vit_germ_est
-#'
+#' @param ... other arguments passed to `ipmr::make_ipm()`
+#' 
 #' @return lambda
 #' 
-ipm_boot_stoch <- function(data, vit_other) {
+ipm_boot_stoch <- function(data, vit_other, year_seq = NULL, ...) {
   #sample ha_id_numbers with replacement within plots
   boot_ids <-
     data %>% 
@@ -37,7 +38,9 @@ ipm_boot_stoch <- function(data, vit_other) {
   make_proto_ipm_stoch(vit_list_stoch, pop_vec) %>% 
     make_ipm(iterations = 1000,  
              normalize_pop_size = TRUE,
-             usr_funs = list(get_scat_params = get_scat_params)
+             kernel_seq = year_seq,
+             usr_funs = list(get_scat_params = get_scat_params),
+             ...
     ) %>% 
     #calculate lambda
     ipmr::lambda(log = FALSE)

### 2022-05-06 R/make_pop_vec.R
allow initial population vector to be scaled to any total population size.

@@ -7,10 +7,11 @@
 #'
 #' @param data the data_full target
 #' @param n_mesh number of meshpoints to use when discretizing log_size.  Passed to `ipmr::discretize_pop_vector()`.
+#' @param n total individuals. The sum of `n_log_size` and `n_sdlg` will equal this.
 #'
 #' @return a list of two elements, `n_log_size` (length = n_mesh) and `n_sdlg` (length 1)
 #' 
-make_pop_vec <- function(data, n_mesh = 100) {
+make_pop_vec <- function(data, n_mesh = 100, n = 1) {
   
   #filter data to only include years with seedlings
   pop_data <- data %>% filter(year>1999, !is.na(log_size))
@@ -36,5 +37,5 @@ make_pop_vec <- function(data, n_mesh = 100) {
   # sum(prop_established) + prop_sdlg == 1
   
   #return:
-  list(n_log_size = prop_established, n_sdlg = prop_sdlg)
+  list(n_log_size = prop_established * n, n_sdlg = prop_sdlg * n)
 }

### 2022-05-06 _targets.R
implement changes in year_seq and pop_vec into _targets

@@ -143,8 +143,11 @@ tar_plan(
   
   ## ├Simulation Parameters -------------------------------------------------
   # Starting population for simulations
-  pop_vec_ff = make_pop_vec(data_ff, n_mesh = 100),
-  pop_vec_cf = make_pop_vec(data_cf, n_mesh = 100),
+  
+  # Start with 1000 individuals distributed in size classes in a way that
+  # matches observed data
+  pop_vec_ff = make_pop_vec(data_ff, n_mesh = 100, n = 1000),
+  pop_vec_cf = make_pop_vec(data_cf, n_mesh = 100, n = 1000),
   
   # Sequence of years to use for stochastic simulations
   year_seq = sample(2000:2009, 1000, replace = TRUE),
@@ -182,13 +185,12 @@ tar_plan(
   # sequence of years and the climate data.  `make_dlnm_ipm()` does this, then
   # add the environmental states to the proto IPM, then iterates the IPM.
   
-  #TODO: make this use year_seq
   proto_ipm_dlnm_ff = make_proto_ipm_dlnm(vit_list_dlnm_ff, pop_vec_ff),
   ipm_dlnm_ff = proto_ipm_dlnm_ff %>%
     make_dlnm_ipm(
       clim,
       seed = 1234,
-      iterations = 1000,
+      year_seq = year_seq,
       usr_funs = list(get_scat_params = get_scat_params)
     ),
  
@@ -197,7 +199,7 @@ tar_plan(
     make_dlnm_ipm(
       clim,
       seed = 1234,
-      iterations = 1000,
+      year_seq = year_seq,
       usr_funs = list(get_scat_params = get_scat_params)
     ),
  
@@ -225,14 +227,14 @@ tar_plan(
   #TODO: also make these use year_seq??
   tar_rep(
     lambda_bt_stoch_ff,
-    ipm_boot_stoch(data_ff, vit_other = vit_other_ff),
+    ipm_boot_stoch(data_ff, vit_other = vit_other_ff, year_seq = year_seq),
     batches = 5,
     reps = 100
   ),
 
   tar_rep(
     lambda_bt_stoch_cf,
-    ipm_boot_stoch(data_cf, vit_other = vit_other_cf),
+    ipm_boot_stoch(data_cf, vit_other = vit_other_cf, year_seq = year_seq),
     batches = 5,
     reps = 100
   ),
@@ -241,17 +243,16 @@ tar_plan(
[…]

### 2022-05-06 docs/paper.Rmd
one line per sentence
sketch out some plots

@@
   keep.trailing.zeros = TRUE,
-  justify = c("left", "left", "right", "right", "right", "right"),
+  justify = c("left", "left", "right", "right", "right"),
   full_width = FALSE,
@@
 ```{r}
-#| example-figure,
-#| fig.cap = "A caption for an example figure. It can be long and span mutiple
-#|   lines like this."
-plot(cars)
+#| pva,
+#| fig.cap = "Population trajectories for all three IPMs in both habitats"
+tar_load(c(
+  ipm_det_cf,
+  ipm_det_ff,
+  ipm_stoch_cf,
+  ipm_stoch_ff,
+  ipm_dlnm_cf,
+  ipm_dlnm_ff
+))
+
+ipm_list <- 
+  list(det_cf = ipm_det_cf,
+       det_ff = ipm_det_ff,
+       stoch_cf = ipm_stoch_cf,
+       stoch_ff = ipm_stoch_ff,
+       dlnm_cf = ipm_dlnm_cf,
+       dlnm_ff = ipm_dlnm_ff)
+
+plot_df <- 
+  ipm_list %>% map_df(~{
+  pop_states <- pop_state(.x)
+
+    lapply(pop_states, colSums) %>%
+    as_tibble() %>% 
+    mutate(n = n_log_size + n_sdlg,
+           t = 1:n())
+}, .id = "model") %>% 
+  separate(model, into = c("ipm", "habitat"))
+
+ggplot(plot_df, aes(x = t, y = n, color = habitat)) +
+  geom_line() +
+  facet_wrap(~ipm,ncol = 1)
 ```

### 2022-05-10 R/fruits_seeds.R
document all functions! (closes #16)

@@ -35,6 +35,16 @@
 # summary(m3)
 
 
+#' Fit model for number of fruits per plant as a function of size
+#'
+#' This uses data from the 1998 experiment to estimate fruit production as a
+#' function of plant size.  This made more sense than trying to calculate
+#' fruits/flower and flowers/infloresence and infloresence/flowering plant.
+#'
+#' @param data_1998 data from an experiment in 1998 (Emilio, add detail here if
+#'   you see this)
+#'
+#' @return a fitted gam
 fruits_gam <- function(data_1998) {
   gam(ripe_fruits ~ s(log_size_prev, bs = "cr"),
       family = nb,
@@ -42,6 +52,17 @@ fruits_gam <- function(data_1998) {
       data = data_1998)
 }
 
+#' Fit model for number of seeds per fruit
+#' 
+#' Uses an intercept only poisson GAM (really a GLM, because there is no
+#' smoother involved) with an offest of log(#fruits) to estimate seeds/fruit.
+#' This uses two datasets with data on numbers of seeds and fruits of individual
+#' plants.
+#'
+#' @param data_1998 data from 1998 experiment
+#' @param data_2008 data from 2008 experiment
+#'
+#' @return a fitted gam
 seeds_gam <- function(data_1998, data_2008) {
   #combine data on seed number, retaining number of fruits
   seeds_df <- 
@@ -56,8 +77,9 @@ seeds_gam <- function(data_1998, data_2008) {
     ) %>% 
     filter(!is.na(seeds)) %>% 
     filter(!(fruits==0 & seeds>0)) #these are obvious data entry errors
-  #model as intercept-only with offset of log(fruits) to get mean # seeds/fruit with predict()
-  #note, seeds/fruits probably closer to uniform distribution than poisson, but straight mean is not that different.
+  #model as intercept-only with offset of log(fruits) to get mean # seeds/fruit
+  #with predict(). NOTE: seeds/fruits probably closer to uniform distribution
+  #than poisson, but mean is not that different.
   gam(seeds ~ 1,
       offset = log(fruits),
       family = poisson,

### 2022-05-10 R/get_scat_params.R

@@ -1,3 +1,13 @@
+#' Retrieve parameters for scaled t distribution family GAMs
+#' 
+#' A convenience function to extract nu and sd from GAMs with 
+#' `family = scat`.  These parameters can be plugged into the `_t.scaled()`
+#' functions from the `metRology` package.
+#'
+#' @param model 
+#'
+#' @return a named vector
+#' 
 get_scat_params <- function(model) {
   thetas <- model$family$getTheta()
   

### 2022-05-10 R/join_data.R

@@ -1,14 +1,24 @@
+#' Combine demographic and climate data
+#'
+#' @param demog the 10 year demographic dataset
+#' @param clim the `clim` target (data from Xavier et al. with SPEI calculated already)
+#'
+#' @return a tibble
+#' 
 join_data <- function(demog, clim) {
   demog <- demog %>% 
+    # manually align plots with grid cells.  I did this by eyeballing the map of BDFFP.
     mutate(lon = case_when(
+      #old names:
       # plot %in% c("Florestal-CF", "5752", "5751", "5750", "5756", "CaboFrio-CF") ~ "-59.875",
       # plot %in% c("2206", "2108", "2107", "Dimona-CF", "5753", "PortoAlegre-CF") ~ "-60.125",
+      #new names:
       plot %in% c("CF-1", "FF-3", "CF-2", "CF-3", "CF-6") ~ "-59.875",
       plot %in% c("FF-5", "FF-2", "FF-1", "CF-4", "FF-4", "CF-5") ~ "-60.125",
       TRUE ~ NA_character_
     ), .before = plot)
   
-  #filter climate data so only months with census (february)
+  #filter climate data so only months with census (February)
   clim_sub <- 
     clim %>% 
     filter(month == 2) %>% 

### 2022-05-10 R/make_AIC_tbl.R

@@ -1,3 +1,11 @@
+#' Make model competition table data frame
+#' 
+#' Takes lists of vital rates (e.g. vit_list_det_ff, vit_list_dlnm_cf, etc.), extracts the vital rates models contained within, and returns a table of AIC values.
+#'
+#' @param ... targets following the pattern `vit_list_<det/stoch/dlnm>_<ff/cf>`
+#'
+#' @return a tibble
+#' 
 make_AIC_tbl <- function(...) {
   all_mods <- rlang::list2(...)
   mod_names <- rlang::ensyms(...) %>% as.character() %>% str_remove_all("vit_list_")

### 2022-05-10 R/make_env_states.R

@@ -1,3 +1,19 @@
+#' Create a data frame of environmental states for use in stochastic IPMs
+#'
+#' Takes the `clim` target and either a specific sequence of years (`year_seq`)
+#' *or* a number of `iterations` and produces a tibble with everything required
+#' to be used for the stochastic, parameter sampled IPMs (the DLNM IPMs).  Used
+#' by `make_dlnm_ipm()`.
+#'
+#' @param clim the `clim` target
+#' @param seed a random seed for reproducibility, optional.
+#' @param maxlag the maximum lag time.  Should be 36 months.
+#' @param iterations a number of iterations if you want years chosen randomly
+#' @param year_seq a vector of years to use to sample from `clim`
+#'
+#' @return a tibble with columns `t` (iteration number, used internally by
+#'   `ipmr`), and `spei_history` (a matrix column with 36 columns)
+#' 
 make_env_states <- function(clim, seed = NULL, maxlag = 36, iterations = NULL, year_seq = NULL) {
   env_df <- 
     clim %>% 
@@ -33,18 +49,21 @@ make_env_states <- function(clim, seed = NULL, maxlag = 36, iterations = NULL, y
     #add burnin years to the beginning
     year_seq <- c(sample(yrs, burnin, replace = TRUE), year_seq)
   } else {
-    #samples n_years years plus necessary "burn-in" years to calculate lags.  These will be removed later.
+    #samples n_years years plus necessary "burn-in" years to calculate lags.
+    #These will be removed later.
     if (is.null(iterations)) {
       iterations <- length(yrs)
     }
     year_seq <- sample(yrs, iterations + burnin, replace = TRUE)
   }
   
-  # pull those random years from the dataset to construct a new dataset in the order of year_seq
+  # pull those random years from the dataset to construct a new dataset in the
+  # order of year_seq
   env_states <- 
     map_dfr(year_seq, ~ env_df %>% 
                dplyr::filter(year == .x)) %>% 
-    # then calculate lagged `spei_history` starting in February of each year going back `maxlag` months
+    # then calculate lagged `spei_history` starting in February of each year
+    # going back `maxlag` months
     mutate(spei_history = tsModel::Lag(spei, 0:maxlag)) %>% 
     dplyr::filter(month == 2) %>%
     slice(-c(1:burnin)) %>% 
[…]

### 2022-05-10 R/make_lambda_table.R

@@ -1,21 +1,32 @@
-#not very reusable, but that's fine.
-#ipm_list should look like this:
-# ipm_list <- list(
-#   det_ff = ipm_det_ff,
-#   det_cf = ipm_det_cf,
-#   stoch_ff = ipm_stoch_ff,
-#   stoch_cf = ipm_stoch_cf
-# )
-#bt_list should look like this:
-# bt_list <- list(
-#   det_ff = lambda_bt_det_ff,
-#   det_cf = lambda_bt_det_cf,
-#   stoch_ff = lambda_bt_stoch_ff,
-#   stoch_cf = lambda_bt_stoch_cf
-#   # dlnm_ff = lambda_bt_dlnm_ff,
-#   # dlnm_cf = lambda_bt_dlnm_cf
-# )
-
+#' Make a table of lambdas and lower and upper confidence limits
+#'
+#' @param ipm_list a named list (see example)
+#' @param bt_list a named list (see example)
+#' @param alpha confidence level of the interval
+#'
+#' @return a tibble
+#' 
+#' @example 
+#' 
+#' #ipm_list should look like this:
+#' ipm_list <- list(
+#'   det_ff = ipm_det_ff,
+#'   det_cf = ipm_det_cf,
+#'   stoch_ff = ipm_stoch_ff,
+#'   stoch_cf = ipm_stoch_cf
+#' )
+#' 
+#' #bt_list should look like this:
+#' bt_list <- list(
+#'   det_ff = lambda_bt_det_ff,
+#'   det_cf = lambda_bt_det_cf,
+#'   stoch_ff = lambda_bt_stoch_ff,
+#'   stoch_cf = lambda_bt_stoch_cf
+#'   # dlnm_ff = lambda_bt_dlnm_ff,
+#'   # dlnm_cf = lambda_bt_dlnm_cf
+#' )
+#' 
+#' make_lambda_table(ipm_list, bt_list, alpha = 0.05)
 make_lambda_table <- function(ipm_list, bt_list, alpha = 0.05) {
 
  lambdas <- 
@@ -23,7 +34,8 @@ make_lambda_table <- function(ipm_list, bt_list, alpha = 0.05) {
    map(~lambda(.x, log = FALSE))
   
  #calc bias-corrected percentile CIs
- map2_df(lambdas, bt_list, ~bcpi(t0 = .x, t = .y, alpha = alpha), .id = "model") %>%
+ map2_df(lambdas, bt_list,
[…]

### 2022-05-10 R/make_pop_vec.R

@@ -6,11 +6,14 @@
 #' Make starting population vector for IPMs
 #'
 #' @param data the data_full target
-#' @param n_mesh number of meshpoints to use when discretizing log_size.  Passed to `ipmr::discretize_pop_vector()`.
-#' @param n total individuals. The sum of `n_log_size` and `n_sdlg` will equal this.
+#' @param n_mesh number of meshpoints to use when discretizing log_size.  Passed
+#'   to `ipmr::discretize_pop_vector()`.
+#' @param n total individuals. The sum of `n_log_size` and `n_sdlg` will equal
+#'   this.
 #'
-#' @return a list of two elements, `n_log_size` (length = n_mesh) and `n_sdlg` (length 1)
-#' 
+#' @return a list of two elements, `n_log_size` (length = n_mesh) and `n_sdlg`
+#'   (length 1)
+#'   
 make_pop_vec <- function(data, n_mesh = 100, n = 1) {
   
   #filter data to only include years with seedlings
@@ -20,8 +23,13 @@ make_pop_vec <- function(data, n_mesh = 100, n = 1) {
   log_size <- pop_data %>% filter(!sdlg) %>% pull(log_size)
   n_established <- length(log_size)
   
-  #use ipmr function to discretize log_size.  No padding because we are assuming we've captured the entire range of plant sizes
-  established_vec <- ipmr::discretize_pop_vector(log_size, n_mesh = n_mesh, pad_low = 1, pad_high = 1)
+  #use ipmr function to discretize log_size.  No padding because we are assuming
+  #we've captured the entire range of plant sizes
+  established_vec <-
+    ipmr::discretize_pop_vector(log_size,
+                                n_mesh = n_mesh,
+                                pad_low = 1,
+                                pad_high = 1)
   
   #count seedlings
   n_sdlg <- pop_data %>% filter(sdlg) %>% nrow()
@@ -31,7 +39,8 @@ make_pop_vec <- function(data, n_mesh = 100, n = 1) {
   
   # sum(established_vec$n_log_size * n_established) == n_established
   
-  prop_established <- (established_vec$n_log_size * n_established) / n_total
+  prop_established <-
+    (established_vec$n_log_size * n_established) / n_total
   prop_sdlg <- n_sdlg / n_total
   
   # sum(prop_established) + prop_sdlg == 1

### 2022-05-10 R/make_proto_ipm_det.R

@@ -1,4 +1,14 @@
-# library(ipmr)
+#' Build general, deterministic, density-independent proto-IPM for Heliconia
+#'
+#' This wraps all the `ipmr` functions used to build a general, deterministic,
+#' density-independent IPM.  See the documentation for `ipmr` for more details.
+#'
+#' @param data_list one of the `vit_list_det_*` targets containing all vital
+#'   rates referenced in the IPM
+#' @param pop_vec an initial population vector such as the `pop_vec_*` targets.
+#'
+#' @return a proto_ipm object
+#' 
 make_proto_ipm_det <- function(data_list, pop_vec) {
   
   init_ipm(
@@ -87,7 +97,8 @@ make_proto_ipm_det <- function(data_list, pop_vec) {
       evict_fun = truncated_distributions("t.scaled", "G_z2_sdlg")
     ) %>% 
     
-    # define implementation with midpoint rule
+    # define implementation with midpoint rule (only available integration rule
+    # currently)
     define_impl(make_impl_args_list(
       kernel_names = c("P",        "go_sdlg",  "stay_sdlg", "leave_sdlg"),
       int_rule     = c("midpoint", "midpoint", "midpoint",  "midpoint"),
@@ -96,7 +107,6 @@ make_proto_ipm_det <- function(data_list, pop_vec) {
     )) %>% 
     #define lower bound, upper bound, and number of meshpoints for log_size
     define_domains(log_size = c(0, 8.018296, 100)) %>%  
-    #arbitrary starting population state
     define_pop_state(pop_vectors = pop_vec)
   
 }

### 2022-05-10 R/make_proto_ipm_dlnm.R

@@ -1,4 +1,18 @@
-library(ipmr)
+#' Build general, stochastic, parameter sampled, density-independent proto-IPM
+#' for Heliconia
+#'
+#' This wraps all the `ipmr` functions used to build a general, stochastic,
+#' parameter sampled, density-independent IPM. The vital rates functions are
+#' GAMs that include lagged effects of SPEI. The returned `proto_ipm` contains
+#' everything needed to iterate the IPM except for the environmental states. See
+#' the documentation for `ipmr` for more details.
+#'
+#' @param data_list one of the `vit_list_dlnm_*` targets containing all vital
+#'   rates referenced in the IPM
+#' @param pop_vec an initial population vector such as the `pop_vec_*` targets.
+#'
+#' @return a proto_ipm object
+#' 
 make_proto_ipm_dlnm <- function(data_list, pop_vec) {
   
   init_ipm(

### 2022-05-10 R/make_proto_ipm_stoch.R

@@ -1,4 +1,17 @@
-library(ipmr)
+#' Build general, stochastic, kernel sampled, density-independent proto-IPM
+#' for Heliconia
+#'
+#' This wraps all the `ipmr` functions used to build a general, stochastic,
+#' kernel sampled, density-independent IPM. The vital rates functions are GAMs
+#' that include random effects of year. See the documentation for `ipmr` for
+#' more details.
+#' 
+#' @param data_list one of the `vit_list_stoch_*` targets containing all vital
+#'   rates referenced in the IPM
+#' @param pop_vec an initial population vector such as the `pop_vec_*` targets.
+#'
+#' @return a `proto_ipm` object
+#' 
 make_proto_ipm_stoch <- function(data_list, pop_vec) {
   
   init_ipm(

### 2022-05-10 R/plot_pop_states.R

@@ -1,13 +1,14 @@
 #' Plot population stage structure over time
-#' 
-#' For a stochastic Heliconia IPM, plots size distribution changes over iterations as a binned area plot.
+#'
+#' For a stochastic Heliconia IPM, plots size distribution changes over
+#' iterations as a binned area plot.
 #'
 #' @param ipm an iterated ipmr ipm
-#' @param bins number of bins to group continuous log_size into (not including seedlings)
-#' @param normalize logical; TRUE plots proportions while FALSE (default) plots actual numbers
-#' 
-#' 
-#' TODO: look at using collapse_pop_state instead.  Might be useful.
+#' @param bins number of bins to group continuous log_size into (not including
+#'   seedlings)
+#' @param normalize logical; TRUE plots proportions while FALSE (default) plots
+#'   actual numbers
+#'
 plot_pop_states <- function(ipm, bins = 6, normalize = FALSE) {
   meshpts <- int_mesh(ipm, full_mesh = FALSE)$log_size_1
   
@@ -15,7 +16,9 @@ plot_pop_states <- function(ipm, bins = 6, normalize = FALSE) {
   rownames(pop_states$n_log_size) <-  meshpts
   
   post_sdlg_df <-
-    as_tibble(pop_states$n_log_size, .name_repair = "unique", rownames = "log_size") %>% 
+    as_tibble(pop_states$n_log_size,
+              .name_repair = "unique",
+              rownames = "log_size") %>%
     mutate(log_size = as.numeric(log_size)) %>% 
     pivot_longer(
       -log_size,
@@ -45,7 +48,7 @@ plot_pop_states <- function(ipm, bins = 6, normalize = FALSE) {
   
   pop_state_tidy <-
     bind_rows(post_sdlg_binned, sdlg_df) %>% 
-    mutate(log_size_bin = fct_relevel(log_size_bin, "seedling", after = 0)) %>% 
+    mutate(log_size_bin = fct_relevel(log_size_bin, "seedling", after = 0)) %>%
     mutate(log_size_bin = fct_rev(log_size_bin)) %>% 
     #calculate proportions
     group_by(iteration) %>% 
@@ -58,7 +61,8 @@ plot_pop_states <- function(ipm, bins = 6, normalize = FALSE) {
   }
   
   p <- 
-    ggplot(pop_state_tidy, aes_string(x = "iteration", y = yvar, fill = "log_size_bin")) +
+    ggplot(pop_state_tidy,
[…]

### 2022-05-10 R/read_wrangle_1998.R

@@ -1,3 +1,9 @@
+#' Read and wrangle data from 1998 experiment
+#'
+#' @param path path to dataset
+#'
+#' @return a tibble
+#' 
 read_wrangle_1998 <- function(path){
   df_raw <- read_csv(path)
   df_raw %>% 

### 2022-05-10 R/read_wrangle_clim.R

@@ -1,3 +1,27 @@
+#' Read and wrangle data from Xavier et al. 2016
+#'
+#' The .csv file can be found in the data from Scott et al. 2022 where it has
+#' already been wrangled from spatial .nc files.  This function calculates SPEI
+#' and calculates monthly lags for each year.
+#'
+#' @param path file path
+#' @param scale passed to `SPEI::spei()`
+#' @param maxlag the number of months to lag SPEI by.  Should be 36 for this
+#'   analysis.
+#'
+#' @return a tibble with two matrix columns: `spei_history`, containing the
+#'   lagged SPEI values and `L`, a matrix indexing the lags.
+#' 
+#' @references 
+#' Scott, Eric R., María Uriarte, and Emilio M. Bruna. “Delayed Effects of
+#' Climate on Vital Rates Lead to Demographic Divergence in Amazonian Forest
+#' Fragments.” Global Change Biology 28, no. 2 (2022): 463–79.
+#' https://doi.org/10.1111/gcb.15900.
+#'
+#' Xavier, Alexandre C., Carey W. King, and Bridget R. Scanlon. “Daily Gridded
+#' Meteorological Variables in Brazil (1980–2013).” International Journal of
+#' Climatology 36, no. 6 (2016): 2644–59. https://doi.org/10.1002/joc.4518.
+#' 
 read_wrangle_clim <- function(path, scale = 3, maxlag = 36) {
   xa_raw <- read_csv(here("data", "xavier_daily_0.25x0.25.csv"))
   

### 2022-05-10 R/read_wrangle_demog.R

@@ -1,3 +1,17 @@
+#' Reand and wrangle Heliconia acuminata demographic dataset
+#' 
+#' Reads in the Dryad-archived demographic dataset and does some wrangling: 1)
+#' creates a new binary column for flowering (from nubmer of infloresences), 2)
+#' creates `log_size` variable (`log(ht*shts)`), 3) creates a column for
+#' survival assuming that plants that haven't been found for 3 years were dead
+#' the last time they were seen, 4) does some filtering and selecting to get
+#' just the data we need.
+#'
+#' @param file_demog the demographic dataset
+#' @param file_plots the dataset with plot information
+#'
+#' @return a tibble
+#' 
 read_wrangle_demog <- function(file_demog, file_plots) {
   ha_raw <- read_csv(file_demog)
   ha_plots <- read_csv(file_plots)

### 2022-05-10 R/vital_rates_dlnm.R

@@ -1,3 +1,11 @@
+#' Distributed lag non-linear models for Heliconia vital rates
+#' 
+#' These functions fit models for survival, growth, flowering, seedling
+#' survival, and seedling recruitment using GAMs.  I've erred on the side of too
+#' many knots so that models have enough knots for both CF and FF data and I
+#' don't have to write separate functions.
+
+
 #for checking knots, load data and run k.check() for the models
 # library(mgcv)
 # library(tidyverse)

### 2022-05-10 R/vital_rates_simple.R

@@ -1,3 +1,11 @@
+#' Simple deterministic models for Heliconia vital rates
+#' 
+#' These functions fit models for survival, growth, flowering, seedling
+#' survival, and seedling recruitment using GAMs.  I've erred on the side of too
+#' many knots so that models have enough knots for both CF and FF data and I
+#' don't have to write separate functions.
+
+
 #for checking knots, load data and run k.check() for the models
 # library(mgcv)
 # library(tidyverse)

### 2022-05-10 R/vital_rates_stoch.R

@@ -1,3 +1,14 @@
+#' Models for Heliconia vital rates including environmental stochasticity as a
+#' random effect of year
+#'
+#' These functions fit models for survival, growth, flowering, seedling
+#' survival, and seedling recruitment using GAMs.  I've erred on the side of too
+#' many knots so that models have enough knots for both CF and FF data and I
+#' don't have to write separate functions. The random effect of year can be
+#' modeled several ways with GAMs.  These are explored in
+#' notes/environmental-stochasticity.Rmd
+
+
 #for checking knots, load data and run k.check() for the models
 # library(mgcv)
 # library(tidyverse)

### 2022-05-10 run.R

@@ -1,1 +1,2 @@
+#to run as a background process in RStudio use the "Source" button and choose "Source as Local Job ..."
 targets::tar_make_clustermq(workers = 10L)

### 2022-05-10 _targets.R
add documentation to _targets.R (closes #17)

@@ -4,7 +4,8 @@
 # )
 
 
-# # Setup SSH connector
+# Or use these options to setup the SSH connector to use hipergator (or another
+# HPC)
 options(
   clustermq.scheduler = "ssh",
   clustermq.template = "ssh_clustermq.tmpl", #custom SSH template to use R 4.0
@@ -23,40 +24,53 @@ lapply(list.files("./R", full.names = TRUE), source)
 tar_plan(
   
   # Load and wrangle data ---------------------------------------------------
-  
+  # All of these targets are run locally (deployment = "main")
+  # Heliconia demographic datset
   tar_file(file_demog, here("data", "HDP_1997_2009.csv"), deployment = "main"),
   tar_file(file_plots, here("data", "HDP_plots.csv"), deployment = "main"),
+  
+  # Climate data from Xavier et al. 2016
   tar_file(file_clim, here("data", "xavier_daily_0.25x0.25.csv"), deployment = "main"),
+  
+  # Other unpublished Heliconia datasets for fruit and seed vital rates
   tar_file(file_1998, here("data", "ha_size_data_1998_cor.csv"), deployment = "main"),
-  tar_file(file_2008, here("data", "Heliconia_acuminata_seedset_2008.csv"), deployment = "main"),
+  tar_file(file_2008, here("data", "Heliconia_acuminata_seedset_2008.csv"),
+           deployment = "main"),
+  tar_target(data_1998, read_wrangle_1998(file_1998), deployment = "main"),
+  tar_target(data_2008, read.csv(file_2008), deployment = "main"),
+  
+  # Wrangle and combine data
   tar_target(demog, read_wrangle_demog(file_demog, file_plots), deployment = "main"),
   tar_target(clim, read_wrangle_clim(file_clim), deployment = "main"),
   tar_target(data_full, join_data(demog, clim), deployment = "main"),
   
+  # Subset data since separate IPMs are fit for continuous forest (cf) and forest
+  # fragments (ff)
   tar_target(data_cf, data_full %>% filter(habitat == "CF"), deployment = "main"),
   tar_target(data_ff, data_full %>% filter(habitat == "1-ha"), deployment = "main"),
-  
-  tar_target(data_1998, read_wrangle_1998(file_1998), deployment = "main"),
-  tar_target(data_2008, read.csv(file_2008), deployment = "main"),
   
 
   # Vital rate models -------------------------------------------------------
 
   ## ├Shared vital rates ----------------------------------------------------
[…]

### 2022-05-10 run.R
nevermind.  10 workers hits the QOS limit.  Guess we only have 9?

@@ -1,2 +1,2 @@
 #to run as a background process in RStudio use the "Source" button and choose "Source as Local Job ..."
-targets::tar_make_clustermq(workers = 10L)
+targets::tar_make_clustermq(workers = 9L)

### 2022-05-10 README.Rmd
update README with roadmap (#18)
add some minimal instructions for setting up hipergator access (closes #20)

@@
 
+# Roadmap
+
+Before June 3:
+
+-   [x] Finish building all IPMs
+-   [x] Document functions and \_targets.R
+-   [ ] Calculate bootstrapped 95% CIs around all lambdas
+-   [ ] Finish writing methods section
+-   [ ] Draft tables & figures
+-   [ ] Archive draft of code in Zenodo
+
+Future:
+
+-   [ ] Finish writing and revising
+-   [ ] Submit
+
 # Reproducibility
@@
 -   Run `targets::tar_make()` or `targets::tar_make_clustermq()` from the R console to run all code and produce all outputs.
+
+### NOTES:
+
+-   The targets `lambda_bt_dlnm_ff` and `lambda_bt_dlnm_cf` will take a very long time to run (days), so you may want to comment those out before attempting to reproduce results.
+-   If you want to make use of HPC, follow instructions [here](https://github.com/BrunaLab/hipergator-targets-ssh) to customize templates and options.

### 2022-05-13 notes/optimizing-meshpoints.Rmd
rename files to remove spaces

--- 2022-04-21 notes/optimizing meshpoints.Rmd
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
 library(tidyverse)
@@
 ```{r data, echo=TRUE}
+withr::with_dir(here::here(), {
   tar_load(vit_list_det_ff)
   tar_load(vit_list_stoch_ff)
+  tar_load(pop_vec_ff)
+})
 ```
@@
 ```{r}
-proto_det <- make_proto_ipm_det(vit_list_det_ff)
+proto_det <- make_proto_ipm_det(vit_list_det_ff, pop_vec = pop_vec_ff)
 ```
@@
 plot(x =seq(20, 150, 5), y = lambdas_det, ylab = "lambda", xlab = "meshpoints")
+plot(x = seq(20, 150, 5), y = lambdas_det - lag(lambdas_det), ylab = "∆lambda", xlab = "meshpoints")
 ```
@@
 ```{r}
-proto_stoch <- make_proto_ipm_stoch(vit_list_stoch_ff)
+proto_stoch <- make_proto_ipm_stoch(vit_list_stoch_ff, pop_vec = pop_vec_ff)
 ```
@@
 ```{r}
-year_seq <- sample(2000:2009, 100, replace = TRUE)
-
 proto_stoch_list <- 
-  seq(60, 120, 10) %>%
+  seq(20, 140, 10) %>%
   map(~{
@@
     define_domains(log_size = c(0, 8.018296, .x)) %>%  
-    #arbitrary starting population state
-    define_pop_state(n_log_size = runif(.x), n_sdlg = .1)
+    #replace starting pop vector with uniform one with correct number of meshpts
+    define_pop_state(n_log_size = rep(1/(.x + 1), .x), n_sdlg = 1/.x)
   })
-
-# make_ipm(proto_stoch_list[[1]],
-#          iterations = 100,
-#          kernel_seq = year_seq,
-#          usr_funs = list(get_scat_params = get_scat_params))
 ```
 
-###Note:
+```{r}
+year_seq <- sample(2000:2009, 1000, replace = TRUE)
 
-This currently isn't working because of a bug in `ipmr` that causes `make_ipm()` to hang when `kernel_seq` is supplied <https://github.com/levisc8/ipmr/issues/58>
+ipm_list <-
+  proto_stoch_list %>%
+  map( ~ make_ipm(
+    .x,
+    iterations = 1000,
+    kernel_seq = year_seq,
+    usr_funs = list(get_scat_params = get_scat_params)
+  ))
 
-```{r}
-lambdas_stoch <-
-  proto_stoch_list %>%
-  map_dbl( ~ make_ipm(
-    .x,
-    iterations = 100,
-    kernel_seq = year_seq,
-    usr_funs = list(get_scat_params = get_scat_params, coef = coef)
-  ) %>% lambda(log = FALSE))
+lambdas_stoch <- ipm_list %>% map_dbl(~lambda(.x, log = FALSE))
 
-plot(x = seq(60, 120, 10), y = lambdas_stoch, ylab = "stochastic lambda", xlab = "meshpoints")
+plot(x = seq(20, 140, 10), y = lambdas_stoch, ylab = "stochastic lambda", xlab = "meshpoints")
+plot(x = seq(20, 140, 10), y = lambdas_stoch - lag(lambdas_stoch), ylab = "∆lambda", xlab = "meshpoints")
 ```
+
+Ok, looks about the same for the stochastic model, which makes sense.

### 2022-05-13 docs/paper.Rmd
update methods

@@
 
--   Mention number of meshpoints
-
--   Add info about starting pop vector for IPMs
-
 -   Add equations / figures
@@
 
-These models were used to build a general, density-independent, deterministic IPM with four sub-kernels: growth and survival, reproduction, probability of staying a seedling (always = 0), and seedling survival and establishment.
-<!--# include equation --> The seedling bank, or probability of staying a seedling, was always equal to zero, since our definition of seedlings was first year plants only.
+These models were used to build a general, density-independent, deterministic IPM with four sub-kernels: growth and survival, reproduction, probability of staying a seedling, and seedling survival and establishment.
+<!--# include equation --> The probability of staying a seedling, was always equal to zero, since our definition of seedlings was first year plants only.
 
-For the second approach, we included environmental stochasticity by adding a random intercept <!--# or intercept and slope or factor-smooth interaction -->for year to all vital rate models built with the long term demographic dataset.
+For the second approach, we included environmental stochasticity by adding a random effect of year to all vital rate models built with the long term demographic dataset.
+The random effect of year was fit using a factor--smooth interaction which allowed the relationship between plant size and vital rates to vary in functional form among transition years.
 Then, we used these models to build general, density independent, stochastic, kernel-resampled IPMs.
-The kernel-resampling approach is to generate kernels corresponding to each transition year in the demographic dataset using the random intercepts for year, and to iterate the IPM by drawing from these randomly.
+The kernel-resampling approach is to generate kernels corresponding to each transition year in the demographic dataset using the random smooths for year, and to iterate the IPM by drawing from these randomly.
 This is equivalent to the matrix selection approach for matrix population models described by @caswell2001.
@@
 We modeled delayed effects of SPEI on survival, growth, and flowering probability using distributed lag non-linear models fit as two dimensional tensor product splines, with a maximum lag of 36 months [@scottDelayedEffectsClimate2022].
-Then these models were incorporated into a general, density independent, stochastic, parameter-resampled IPM.
+Then these models were incorporated into a general, density independent, stochastic, parameter-resampled IPM (*sensu* @metcalfStatisticalModellingAnnual2015).
 To iterate such an IPM, a random sequence of SPEI values is created by re-sampling the observed monthly SPEI data.
@@
 
-All IPMs were constructed and iterated with the `ipmr` package in R [@levinIpmrFlexibleImplementation2021].
+All IPMs were constructed and iterated with 100 meshpoints <!--# optimized in notes/optimizing-meshpoints.Rmd which could be turned into supplemental -->using the `ipmr` package in R [@levinIpmrFlexibleImplementation2021].
 For each type of IPM we iterated the model for 1000 time steps, discarding the first 100 time steps to omit transient effects.
@@
 
-To calculate 95% confidence intervals around the per-capita growth rates (lambdas), we created 1000 bootstraps of the demographic dataset by sampling individual plants with replacement within each habitat.
+To calculate 95% confidence intervals around the per-capita growth rates (lambdas), we created 500 bootstraps of the demographic dataset by sampling individual plants with replacement within each habitat.
 For each bootstrap, we then re-fit vital rates models (all except germination and establishment rate, fruits per flowering plant, and seeds per fruit, which were estimated using different datasets), constructed IPMs, and calculated a value for lambda as described above.

### 2022-05-13 R/make_proto_ipm_stoch.R
fix mistake where seedling size wasn't stochastic

@@ -87,14 +87,16 @@ make_proto_ipm_stoch <- function(data_list, pop_vec) {
       family = "DC",
       s_sdlg = predict(vit_surv_sdlg,
                        newdata = tibble(year_fct = yr),
-                       type = 'response'), #intercept from intercept-only glm
+                       type = 'response'),
       G_z2_sdlg = dt.scaled(
         x    = log_size_2,
         df   = size_sdlg_nu,
         mean = size_sdlg_mu,
         sd   = size_sdlg_sd
       ),
-      size_sdlg_mu = coef(vit_size_sdlg)[1], #intercept from intercept-only glm
+      size_sdlg_mu = predict(vit_size_sdlg,
+                             newdata = tibble(year_fct = yr),
+                             type = 'response'),
       size_sdlg_nu = get_scat_params(vit_size_sdlg)["nu"],
       size_sdlg_sd = get_scat_params(vit_size_sdlg)["sd"],
       

### 2022-05-13 docs/paper.Rmd
update methods

@@
 Then these models were incorporated into a general, density independent, stochastic, parameter-resampled IPM (*sensu* @metcalfStatisticalModellingAnnual2015).
-To iterate such an IPM, a random sequence of SPEI values is created by re-sampling the observed monthly SPEI data.
+To iterate such an IPM, a random sequence of SPEI values is created by sampling years of the observed monthly SPEI data.
 Then, 36 month lags are calculated for each year starting in February (the month of the demographic census).
 These values are then used to predict fitted values from the vital rates models, generating different kernels at each iteration of the IPM.
-With this method, these kernels are not entirely independent because the SPEI from the previous two iterations is included in the vital rates underlying each kernel.
+With this method, the kernels of successive iterations are not entirely independent because the SPEI values used in calculating vital rates include values used in the previous two iterations.
 
-All IPMs were constructed and iterated with 100 meshpoints <!--# optimized in notes/optimizing-meshpoints.Rmd which could be turned into supplemental -->using the `ipmr` package in R [@levinIpmrFlexibleImplementation2021].
+All IPMs were constructed and iterated <!--# optimized in notes/optimizing-meshpoints.Rmd which could be turned into supplemental -->using the `ipmr` package in R [@levinIpmrFlexibleImplementation2021].
+The IPMs used 100 meshpoints and the midpoint rule for calculating kernels <!--# citation? -->.
 For each type of IPM we iterated the model for 1000 time steps, discarding the first 100 time steps to omit transient effects.

### 2022-05-16 docs/paper.Rmd
add kernel equations with english translations

@@
 library(targets)
+library(tidyverse)
 ```
@@
 
--   Add equations / figures
+-   Add figures
+
+-   Add paragraph about why we used IPMs or why a General IPM---depends on what goes in introduction maybe?
 
@@
 --\>
+
+Why general IPM necessary?
+
+-   Seedlings physiologically different from 1-shoot small plants because of lack of underground reserves (of carbs and meristems)
+
+-   Seedling survival not well predicted when included with mature plants.
+
+------------------------------------------------------------------------
+
+### Kernel equation and description
+
+Below is a description of the *basic*, deterministic IPM.
+This maybe goes in a supplement?
+These vital rates get swapped out for ones that vary by year (for the stochastic, kernel resampled IPMs) or ones that include lagged effects of SPEI (for the stochastic, parameter resampled IPMs).
+See Figure \@ref(fig:lifecycle).
+
+For mature plants:
+
+$$
+n(z^{\prime},t+1) = s_s G_s(z^\prime)n_s(t) + \int_L^U s(z) G(z^{\prime}, z) n(z,t)\;dz
+$$
+
+The number and size of mature plants in the next census is determined by seedlings entering the mature plant population (i.e.
+recruitment <!--# *Is* this recruitment?
+-->) and survival and growth (or regression) of mature plants.
+Seedlings survive ($s_s$) and grow into mature plants of a particular size ($G_s(z^\prime)$).
+Mature plants survive as a function of size ($s(z)$), and grow (or regress) to a new size as a function of their previous size ($G(z^\prime, z)$).
+
+For seedlings:
+
+$$
+n_s(t+1) = \int_L^U p_f(z)f(z)g n(z,t) \; dz
+$$
+
+Mature plants flower with a probability that is a function of size ($p_f(z)$) and produce a number of seeds as a function of size ($f(z)$).
+Those seeds then germinate and establish as seedlings with probability $g$.
+
+------------------------------------------------------------------------
 
@@
 For each bootstrap, we then re-fit vital rates models (all except germination and establishment rate, fruits per flowering plant, and seeds per fruit, which were estimated using different datasets), constructed IPMs, and calculated a value for lambda as described above.
+<!--# then calculated bias corrected 95% CIs -->
 
@@
 
+<!--# Note: The old version of this table where "random effect of year" only meant a random intercept has DLNM *always* winning as best fit model.
+Should other ways of fitting random effects be included for this table (even if not in the IPMs)?
+Could add simple GLMMs, for example.
+-->
+
 ```{r aic, results='asis'}
@@
   full_width = FALSE,
-  caption = "(\\#tab:lambda) Population growth rates for continuous forest (cf) and forest fragments (ff) under different kinds of IPMs with bootstrapped 95% confidence intervals."
+  caption = "(\\#tab:lambda) Population growth rates for continuous forest (cf) and forest fragments (ff) under different kinds of IPMs with bootstrapped, bias-corrected, 95% confidence intervals."
 )
@@
 # Figures
+
+```{r}
+#| lifecycle
+#| fig.cap = "Diagram of IPMs with equations"
+
+# Draft of this figure is in docs/figures/Heliconia lifecycle DAG.key
+```
+
+Just playing with something here.
+This is the population trajectories from the iterated IPMs with `normalize_pop_size=FALSE`.
+I thought maybe I'd be able to see some (lagged) similarities between the kernel sampled and DLNM IPMs because they are run with the same sequence of years.
+Nothing obvious pops out.
+Not sure this figure is worth including.
 
@@
 
-ggplot(plot_df, aes(x = t, y = n, color = habitat)) +
+p <- ggplot(plot_df, aes(x = t, y = n, color = ipm)) +
   geom_line() +
-  facet_wrap(~ipm,ncol = 1)
+  facet_wrap(~habitat, ncol = 1)
+
+p + xlim(10, 250)
+
+p+ scale_y_log10() + labs(y = "log10(n)")
+
 ```
-
-Some kind of conceptual diagram of the methods?
-The life cycle diagram with equations on it?
 

### 2022-05-16 docs/paper.Rmd
add lifecycle diagram figure

@@
 
--   Add figures
-
 -   Add paragraph about why we used IPMs or why a General IPM---depends on what goes in introduction maybe?
@@
 
-Ultimately, we built and iterated three types of IPMs which each required different underlying vital rate model structure.
+Ultimately, we built and iterated three types of IPMs which each required different underlying vital rate model structure (Figure \@ref(fig:lifecycle)).
 For all models, we used the long term demographic dataset for modeling survival, growth, and flowering.
@@
 # Draft of this figure is in docs/figures/Heliconia lifecycle DAG.key
+knitr::include_graphics(here::here("docs", "figures", "lifecycle.png"))
 ```
 
+Combine above figure with this table:
+
+| Description                                  | Deterministic    | Stochastic, kernel-resampled | Stochastic, parameter-resampled (DLNM) |
+|----------------------------------------------|------------------|------------------------------|----------------------------------------|
+| Mature plant survival                        | $s(z)$           | $s_y(z)$                     | $s(z;\theta_{0-36})$                   |
+| Seedling survival                            | $s_0$            | $s_{0_y}$                    | $s_0(\theta_{0-36})$                   |
+| Mature plant growth or regression            | $G(z^\prime, z)$ | $G_y(z^\prime,z)$            | $G(z^\prime,z;\theta_{0-36})$          |
+| Seedling growth into mature plant            | $G_0(z^\prime)$  | $G_{0_y}(z^\prime)$          | $G(z^\prime;\theta_{0-36})$            |
+| Probability of flowering                     | $p_f(z)$         | $p_{f_y}(z)$                 | $p_f(z;\theta_{0-36})$                 |
+| Size-specific fecundity                      | $f(z)$           | $f(z)$                       | $f(z)$                                 |
+| Probability of germination and establishment | $g$              | $g$                          | $g$                                    |
+
+<!--# probably can't make the figure with the table in it in R easily. -->\
+\
 Just playing with something here.

### 2022-05-17 README.Rmd
update renv instructions and roadmap

@@
 
-This repository contains data and code for a manuscript by Eric Scott and Emilio Bruna written for submission to The American Naturalist tentatively titled: \_\_\_\_\_.
+This repository contains data and code for a manuscript by Eric Scott and Emilio Bruna written for submission to The American Naturalist tentatively titled: *Context-dependent consequences of including lagged effects in demographic models*.
 
@@
 -   [ ] Calculate bootstrapped 95% CIs around all lambdas
--   [ ] Finish writing methods section
--   [ ] Draft tables & figures
+-   [x] Finish draft of methods section
+-   [x] Draft tables & figures
 -   [ ] Archive draft of code in Zenodo
@@
 -   Open the project in RStudio by double-clicking the `.Rproj` file.
--   Install packages listed in `packages.R` (if this compendium uses `renv`, this may happen automatically or with `renv::restore()`).
--   Install the [`targets` package](https://docs.ropensci.org/targets/).
+-   This project uses `renv` for package management, which should be installed automatically when opening the project. Run `renv::restore()` to install all required packages.
 -   Run `targets::tar_make()` or `targets::tar_make_clustermq()` from the R console to run all code and produce all outputs.

### 2022-05-23 _targets.R
keep sub-kernels for possible future analyses

@@ -212,22 +212,26 @@ tar_plan(
   # add the environmental states to the proto IPM, then iterates the IPM.
   
   proto_ipm_dlnm_ff = make_proto_ipm_dlnm(vit_list_dlnm_ff, pop_vec_ff),
+  
   ipm_dlnm_ff = proto_ipm_dlnm_ff %>%
     make_dlnm_ipm(
       clim,
       seed = 1234,
       year_seq = year_seq,
       normalize_pop_size = FALSE,
+      return_sub_kernels = TRUE,
       usr_funs = list(get_scat_params = get_scat_params)
     ),
  
   proto_ipm_dlnm_cf = make_proto_ipm_dlnm(vit_list_dlnm_cf, pop_vec_cf),
+  
   ipm_dlnm_cf = proto_ipm_dlnm_cf %>%
     make_dlnm_ipm(
       clim,
       seed = 1234,
       year_seq = year_seq,
       normalize_pop_size = FALSE,
+      return_sub_kernels = TRUE,
       usr_funs = list(get_scat_params = get_scat_params)
     ),
  
@@ -292,17 +296,17 @@ tar_plan(
         det_ff = ipm_det_ff,
         det_cf = ipm_det_cf,
         stoch_ff = ipm_stoch_ff,
-        stoch_cf = ipm_stoch_cf#,
-        # dlnm_ff = ipm_dlnm_ff,
-        # dlnm_cf = ipm_dlnm_cf
+        stoch_cf = ipm_stoch_cf,
+        dlnm_ff = ipm_dlnm_ff,
+        dlnm_cf = ipm_dlnm_cf
       ),
       bt_list = list(
         det_ff = lambda_bt_det_ff,
         det_cf = lambda_bt_det_cf,
         stoch_ff = lambda_bt_stoch_ff,
-        stoch_cf = lambda_bt_stoch_cf#,
-        # dlnm_ff = lambda_bt_dlnm_ff,
-        # dlnm_cf = lambda_bt_dlnm_cf
+        stoch_cf = lambda_bt_stoch_cf,
+        dlnm_ff = lambda_bt_dlnm_ff,
+        dlnm_cf = lambda_bt_dlnm_cf
       )
     ),
     deployment = "main" #data limited, not computation limited

### 2022-05-23 docs/paper.Rmd
try to make a table of figures to go with lifecycle diagram
don't print/save experimental figures yet.

--- 2022-05-16 docs/paper.Rmd
@@
 # Draft of this figure is in docs/figures/Heliconia lifecycle DAG.key
+knitr::include_graphics(here::here("docs", "figures", "lifecycle.png"))
 ```
 
+Combine above figure with this table:
+
+| Description                                  | Deterministic    | Stochastic, kernel-resampled | Stochastic, parameter-resampled (DLNM) |
+|----------------------|----------------|----------------|-------------------|
+| Mature plant survival                        | $s(z)$           | $s_y(z)$                     | $s(z;\theta_{0-36})$                   |
+| Seedling survival                            | $s_0$            | $s_{0_y}$                    | $s_0(\theta_{0-36})$                   |
+| Mature plant growth or regression            | $G(z^\prime, z)$ | $G_y(z^\prime,z)$            | $G(z^\prime,z;\theta_{0-36})$          |
+| Seedling growth into mature plant            | $G_0(z^\prime)$  | $G_{0_y}(z^\prime)$          | $G(z^\prime;\theta_{0-36})$            |
+| Probability of flowering                     | $p_f(z)$         | $p_{f_y}(z)$                 | $p_f(z;\theta_{0-36})$                 |
+| Size-specific fecundity                      | $f(z)$           | $f(z)$                       | $f(z)$                                 |
+| Probability of germination and establishment | $g$              | $g$                          | $g$                                    |
+
+<!--# probably can't make the figure with the table in it in R easily. -->
+
+\
+\
 Just playing with something here.
@@
 
-```{r}
+```{r eval=FALSE}
 #| pva,

### 2022-05-24 docs/paper.Rmd
run to get working

--- 2022-05-16 docs/paper.Rmd
@@
 
-<!--# probably can't make the figure with the table in it in R easily. -->\
+<!--# probably can't make the figure with the table in it in R easily. -->
+
+\
 \

### 2022-05-24 _targets.R

@@ -273,20 +273,20 @@ tar_plan(
   ## ├Stochastic, DLNM ------------------------------------------------------
   # For these I use more batches, fewer reps because each rep takes about 90 min.
   # That way I can make incremental progress easier.
-  
-  tar_rep(
-    lambda_bt_dlnm_ff,
-    ipm_boot_dlnm(data_ff, vit_other = vit_other_ff, clim = clim, year_seq = year_seq),
-    batches = 500,
-    reps = 1
-  ),
-
-  tar_rep(
-    lambda_bt_dlnm_cf,
-    ipm_boot_dlnm(data_cf, vit_other = vit_other_cf, clim = clim, year_seq = year_seq),
-    batches = 500,
-    reps = 1
-  ),
+  # 
+  # tar_rep(
+  #   lambda_bt_dlnm_ff,
+  #   ipm_boot_dlnm(data_ff, vit_other = vit_other_ff, clim = clim, year_seq = year_seq),
+  #   batches = 500,
+  #   reps = 1
+  # ),
+  # 
+  # tar_rep(
+  #   lambda_bt_dlnm_cf,
+  #   ipm_boot_dlnm(data_cf, vit_other = vit_other_cf, clim = clim, year_seq = year_seq),
+  #   batches = 500,
+  #   reps = 1
+  # ),
   
   ## ├Lambda table ------------------------------------------------------
   tar_target(
@@ -296,17 +296,17 @@ tar_plan(
         det_ff = ipm_det_ff,
         det_cf = ipm_det_cf,
         stoch_ff = ipm_stoch_ff,
-        stoch_cf = ipm_stoch_cf,
-        dlnm_ff = ipm_dlnm_ff,
-        dlnm_cf = ipm_dlnm_cf
+        stoch_cf = ipm_stoch_cf#,
+        # dlnm_ff = ipm_dlnm_ff,
+        # dlnm_cf = ipm_dlnm_cf
       ),
       bt_list = list(
         det_ff = lambda_bt_det_ff,
         det_cf = lambda_bt_det_cf,
         stoch_ff = lambda_bt_stoch_ff,
-        stoch_cf = lambda_bt_stoch_cf,
-        dlnm_ff = lambda_bt_dlnm_ff,
-        dlnm_cf = lambda_bt_dlnm_cf
+        stoch_cf = lambda_bt_stoch_cf#,
+        # dlnm_ff = lambda_bt_dlnm_ff,
+        # dlnm_cf = lambda_bt_dlnm_cf
[…]

### 2022-05-24 docs/paper.Rmd
update html version of manuscript

--- 2022-05-23 docs/paper.Rmd
@@
       #- -Fpandoc-crossref #required for cross-ref equations
+    bookdown::html_document2: 
+      linenumbers: yes
+      number_sections: no
+      fig_caption: yes
+      pandoc_args:
+      - --lua-filter=templates/scholarly-metadata.lua
+      - --lua-filter=templates/author-info-blocks.lua
 linkcolor: blue

### 2022-05-24 _targets.R

@@ -274,19 +274,19 @@ tar_plan(
   # For these I use more batches, fewer reps because each rep takes about 90 min.
   # That way I can make incremental progress easier.
   # 
-  # tar_rep(
-  #   lambda_bt_dlnm_ff,
-  #   ipm_boot_dlnm(data_ff, vit_other = vit_other_ff, clim = clim, year_seq = year_seq),
-  #   batches = 500,
-  #   reps = 1
-  # ),
-  # 
-  # tar_rep(
-  #   lambda_bt_dlnm_cf,
-  #   ipm_boot_dlnm(data_cf, vit_other = vit_other_cf, clim = clim, year_seq = year_seq),
-  #   batches = 500,
-  #   reps = 1
-  # ),
+  tar_rep(
+    lambda_bt_dlnm_ff,
+    ipm_boot_dlnm(data_ff, vit_other = vit_other_ff, clim = clim, year_seq = year_seq),
+    batches = 500,
+    reps = 1
+  ),
+
+  tar_rep(
+    lambda_bt_dlnm_cf,
+    ipm_boot_dlnm(data_cf, vit_other = vit_other_cf, clim = clim, year_seq = year_seq),
+    batches = 500,
+    reps = 1
+  ),
   
   ## ├Lambda table ------------------------------------------------------
   tar_target(
@@ -314,7 +314,7 @@ tar_plan(
 
 
   # Manuscript --------------------------------------------------------------
-  tar_render(paper, here("docs", "paper.Rmd"), packages = "bookdown")
+  tar_render(paper, here("docs", "paper.Rmd"), packages = "bookdown", output_format = "all")
   
   
 ) %>% 

### 2022-05-24 docs/paper.Rmd
make html look better

--- 2022-05-23 docs/paper.Rmd
@@
       #- -Fpandoc-crossref #required for cross-ref equations
+    bookdown::html_document2: 
+      theme: 
+        version: 4
+        bootswatch: journal
+      number_sections: no
+      fig_caption: yes
+      pandoc_args:
+      - --lua-filter=templates/scholarly-metadata.lua
+      - --lua-filter=templates/author-info-blocks.lua
 linkcolor: blue

### 2022-05-24 R/make_proto_ipm_stoch.R
modify code to fit IPM with just low-density continuous forest plots

@@ -12,7 +12,7 @@
 #'
 #' @return a `proto_ipm` object
 #' 
-make_proto_ipm_stoch <- function(data_list, pop_vec) {
+make_proto_ipm_stoch <- function(data_list, pop_vec, years = 2000:2009) {
   
   init_ipm(
     sim_gen = "general",
@@ -46,7 +46,7 @@ make_proto_ipm_stoch <- function(data_list, pop_vec) {
       data_list = data_list,
       states          = list(c('log_size')),
       uses_par_sets   = TRUE,
-      par_set_indices = list(yr = 2000:2009),
+      par_set_indices = list(yr = years),
       evict_cor       = TRUE,
       evict_fun       = truncated_distributions("t.scaled", "G_z1z")
     ) %>% 
@@ -69,7 +69,7 @@ make_proto_ipm_stoch <- function(data_list, pop_vec) {
       data_list     = data_list,
       states        = list(c('log_size', 'sdlg')),
       uses_par_sets = TRUE,
-      par_set_indices = list(yr = 2000:2009)
+      par_set_indices = list(yr = years)
     ) %>% 
     
     define_kernel(
@@ -103,7 +103,7 @@ make_proto_ipm_stoch <- function(data_list, pop_vec) {
       data_list = data_list,
       states = list(c("sdlg", "log_size")),
       uses_par_sets = TRUE,
-      par_set_indices = list(yr = 2000:2009),
+      par_set_indices = list(yr = years),
       evict_cor = TRUE,
       evict_fun = truncated_distributions("t.scaled", "G_z2_sdlg")
     ) %>% 

### 2022-05-24 _targets.R

@@ -48,6 +48,8 @@ tar_plan(
   # fragments (ff)
   tar_target(data_cf, data_full %>% filter(habitat == "CF"), deployment = "main"),
   tar_target(data_ff, data_full %>% filter(habitat == "1-ha"), deployment = "main"),
+  tar_target(data_cf_low, data_full %>% filter(plot %in% c("CF-4", "CF-5", "CF-6")),
+             deployment = "main"),
   
 
   # Vital rate models -------------------------------------------------------
@@ -94,6 +96,17 @@ tar_plan(
     ),
     vit_other_cf),
   
+  ## continuous forest
+  vit_list_det_cf_low = c(
+    list(
+      vit_surv = surv_det(data_cf_low),
+      vit_size = size_det(data_cf_low),
+      vit_flwr = flwr_det(data_cf_low),
+      vit_size_sdlg = size_sdlg_det(data_cf_low),
+      vit_surv_sdlg = surv_sdlg_det(data_cf_low)
+    ),
+    vit_other_cf),
+  
   ## ├Stochastic (random effect of year) ------------------------------------
   
   # For more on how environmental stochasticity was modeled, see
@@ -118,6 +131,17 @@ tar_plan(
       vit_flwr = flwr_raneff(data_cf),
       vit_size_sdlg = size_sdlg_raneff(data_cf),
       vit_surv_sdlg = surv_sdlg_raneff(data_cf)
+    ),
+    vit_other_cf),
+  
+  ## continuous forest low density plots only
+  vit_list_stoch_cf_low = c(
+    list(
+      vit_surv = surv_raneff(data_cf_low),
+      vit_size = size_raneff(data_cf_low),
+      vit_flwr = flwr_raneff(data_cf_low),
+      vit_size_sdlg = size_sdlg_raneff(data_cf_low),
+      vit_surv_sdlg = surv_sdlg_raneff(data_cf_low)
     ),
     vit_other_cf),
   
@@ -146,6 +170,17 @@ tar_plan(
       vit_flwr = flwr_dlnm(data_cf),
       vit_size_sdlg = size_sdlg_dlnm(data_cf),
       vit_surv_sdlg = surv_sdlg_dlnm(data_cf)
+    ),
+    vit_other_cf),
+  
+  # continuous forest low density plots only
+  vit_list_dlnm_cf_low = c(
+    list(
+      vit_surv = surv_dlnm(data_cf_low),
[…]

### 2022-05-24 notes/low-density-continuous.Rmd
do some analysis to see if low and high density CF plots differ

---
title: "Is plot density important?"
author: "Eric R. Scott"
date: "2022-05-24"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    number_sections: yes
    highlight: kate
    theme:
      version: 4
      bootswatch: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(conflicted)
library(targets)
library(mgcv)
library(gratia)
library(ipmr)
library(marginaleffects)

conflict_prefer("lag", "dplyr")
conflict_prefer("filter", "dplyr")
```

*Last compiled: `r Sys.Date()`*

# Purpose

The continuous forest plots are not sampled randomly.
The first three plots (CF-1--3) have a much higher density of plants than the 1-ha fragments.
Emilio then purposefully saught out low-density continuous forest plots with numbers of plants comparable to the plots in 1-ha fragments (CF-4--6).
Some preliminary investigation leads me to believe the high and low density plots may be different in terms of observed population growth, vital rates, and lambda.
I've fit the vital rates models and ran the IPMs using only low-density CF plots for comparison.

# Load Data

```{r data, echo=TRUE}
withr::with_dir(here::here(), {
  tar_load(c(
    lambda_table,
    ipm_det_cf_low,
    ipm_stoch_cf_low,
    ipm_dlnm_cf_low,
    data_full
  ))
})
```

# Vital rates

One simple way to look for differences is to fit vital rates models with different smooths for 1-ha, CF-low density, and cf-high density.
The `by = hab_dens` argument in `s()` generates a different smooth for each level of habitat/density

#### NOTE:

One important confounding factor here is that there is missing data for the year 2000 for all low-density CF plots and missing data for 2003 for one of the plots.
So differences between low and high density or between low density and 1-ha could be due to the year 2000.
To be "fair" I'm going to filter out 2000 (and 2001, since we need the size in the previous year) for all plots.

```{r}
# data_full$habitat %>% unique()
df <-
  data_full %>% 
  select(-L, -spei_history) %>% 
  filter(year != 2000) %>% 
  mutate(hab_dens = case_when(
    habitat == "1-ha" ~ "1-ha",
    plot %in% c("CF-1", "CF-2", "CF-3") ~ "CF-high",
    plot %in% c("CF-4", "CF-5", "CF-6") ~ "CF-low",
    TRUE ~ NA_character_
  ), .after = habitat) %>% 
  mutate(hab_dens = as.factor(hab_dens))

```

## Survival

Fit the model with different intercepts and smooths for the three habitat:density combos

```{r}
m_surv <- bam(
    surv ~ hab_dens + s(log_size_prev, bs = "cr", k = 20, by = hab_dens),
    family = binomial,
    data = df %>% filter(sdlg_prev == FALSE),
    method = "fREML"
)
draw(m_surv)
```

Look at the pairwise differences in these smooths:

```{r}
diff_surv <- difference_smooths(m_surv, "s(log_size_prev)")
ggplot(diff_surv, aes(x = log_size_prev, y = diff)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.4) +
  geom_line() +
  facet_wrap(level_1 ~ level_2) +
  labs(title = "Difference smooths for survival")
```

Calculate contrast, the mean difference between adjusted predictions (See [vignette](https://vincentarelbundock.github.io/marginaleffects/articles/mfx02_contrasts.html)).
The important thing to get about how these work is they evaluate the model at values in the actual dataset, then average the differences.

```{r}
cmp_surv <- comparisons(m_surv, variables = list(hab_dens = "pairwise"))
summary(cmp_surv)
```

The mean difference in survival between low and high density CF is significant, and the mean difference between low density CF and 1-ha is not significant.

## Size

Fit the model with different intercepts and smooths for the three habitat:density combos

```{r}
m_size <- bam(
    log_size ~ hab_dens + s(log_size_prev, bs = "cr", k = 20, by = hab_dens),
    family = scat,
    data = df %>% filter(sdlg_prev == FALSE, surv == 1, !is.na(log_size)),
    method = "fREML"
)
draw(m_size)
```

Look at the pairwise differences in these smooths:

```{r}
diff_size <- difference_smooths(m_size, "s(log_size_prev)")
ggplot(diff_size, aes(x = log_size_prev, y = diff)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.4) +
  geom_line() +
  facet_wrap(level_1 ~ level_2) +
  labs(title = "Difference smooths for size")
```

differences are very minor

Contrasts:

```{r}
cmp_size <- comparisons(m_size, variables = list(hab_dens = "pairwise"))
summary(cmp_size)
```

Here the CF high and 1-ha are not significantly different but other comparisons are.
The smooths for CF-high and 1-ha look really different for small plants, but because there are few small plants the comparison above shows a small effect.

## Flowering

```{r}
m_flwr <- bam(
    flwr ~ hab_dens + s(log_size_prev, bs = "cr", k = 20, by = hab_dens),
    family = binomial, 
    data = df %>% filter(sdlg == FALSE, surv == 1, !is.na(log_size)),
    method = "fREML"
  )
draw(m_flwr)
```

The smooth for high density CF looks differen than the other two.

Look at the pairwise differences in these smooths:

```{r}
diff_flwr <- difference_smooths(m_flwr, "s(log_size_prev)")
ggplot(diff_flwr, aes(x = log_size_prev, y = diff)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.4) +
  geom_line() +
  facet_wrap(level_1 ~ level_2) +
  labs(title = "Difference smooths for size")
```

Contrasts:

```{r}
cmp_flwr <- comparisons(m_flwr, variables = list(hab_dens = "pairwise"))
summary(cmp_flwr)
```

all pairwise comparisons are significantly different and the biggest difference is actually between CF-low and 1-ha

#IPMs

We can also look at the level of the IPMs

```{r}
lambda_table %>% 
  add_row(ipm = "det", habitat = "cf_low", est = lambda(ipm_det_cf_low)) %>% 
  add_row(ipm = "stoch", habitat = "cf_low", est = lambda(ipm_stoch_cf_low, log = FALSE)) %>% 
  add_row(ipm = "dlnm", habitat = "cf_low", est = lambda(ipm_dlnm_cf_low, log = FALSE)) %>% 
  add_row(ipm = "dlnm", habitat = "ff", est = lambda(tar_read(ipm_dlnm_ff), log = FALSE)) %>% 
  add_row(ipm = "dlnm", habitat = "cf", est = lambda(tar_read(ipm_dlnm_cf), log = FALSE)) %>% 
  arrange(ipm, habitat)
```

For the deterministic and stochastic, kernel-resampled (random effect of year) models, CF low density plots have the lowest lambda.
For the DLNM IPM, CF and CF low density are comparable and FF has the lowest lambda.
The caveat is that the stochastic, kernel-resampled IPM for cf_low is iterated using a different sequence of years since it doesn't have data for 2000 or 2001.

### 2022-05-24 docs/paper.Rmd
run everything

@@
 | Description                                  | Deterministic    | Stochastic, kernel-resampled | Stochastic, parameter-resampled (DLNM) |
-|----------------------|----------------|----------------|-------------------|
+|----------------------------------------------|------------------|------------------------------|----------------------------------------|
 | Mature plant survival                        | $s(z)$           | $s_y(z)$                     | $s(z;\theta_{0-36})$                   |

### 2022-05-24 _targets.R

@@ -296,17 +296,17 @@ tar_plan(
         det_ff = ipm_det_ff,
         det_cf = ipm_det_cf,
         stoch_ff = ipm_stoch_ff,
-        stoch_cf = ipm_stoch_cf#,
-        # dlnm_ff = ipm_dlnm_ff,
-        # dlnm_cf = ipm_dlnm_cf
+        stoch_cf = ipm_stoch_cf,
+        dlnm_ff = ipm_dlnm_ff,
+        dlnm_cf = ipm_dlnm_cf
       ),
       bt_list = list(
         det_ff = lambda_bt_det_ff,
         det_cf = lambda_bt_det_cf,
         stoch_ff = lambda_bt_stoch_ff,
-        stoch_cf = lambda_bt_stoch_cf#,
-        # dlnm_ff = lambda_bt_dlnm_ff,
-        # dlnm_cf = lambda_bt_dlnm_cf
+        stoch_cf = lambda_bt_stoch_cf,
+        dlnm_ff = lambda_bt_dlnm_ff,
+        dlnm_cf = lambda_bt_dlnm_cf
       )
     ),
     deployment = "main" #data limited, not computation limited

### 2022-05-25 docs/paper.Rmd
add a paragraph and some results for the time it takes to iterate the IPMs (#1)

--- 2022-05-24 docs/paper.Rmd
@@
 
+The time to iterate the DLNM models is much higher than than deterministic and kernel-resampled.
+The greater use of computational resources is likely a result of `predict()` being much slower for GAMs with 2D smooths because the number of knots is much higher compared to the GAMs used for the vital rates models in the determinsitic and kernel-resampled IPMs.
+<!--# also note that this can't really be paralellized because each iteration depends on the previous one.
+There's *maybe* some potential speed gains in using multiple cores with predict.bam(), but unlikely to be substantial.
+-->
+
+```{r}
+meta_df <- 
+  tar_meta() %>% 
+  select(name, size, bytes, time, seconds) %>% 
+  filter(str_detect(name, "^ipm_\\w+_[cf]{2}$")) %>% 
+  mutate(minutes = seconds / 60,
+         hours = minutes / 60)
+
+meta_df %>% 
+  separate(name, into = c("trash", "IPM", "Habitat")) %>% 
+  select(-trash) %>% 
+  select(IPM, Habitat, minutes) %>% 
+  group_by(IPM) %>% 
+  summarize(mean_time_min = round(mean(minutes), 2))
+```
+
 # Discussion

### 2022-05-25 docs/paper.Rmd
restructured equations and IPM description section and added info on "why general IPMs?"

@@
       - --lua-filter=templates/author-info-blocks.lua
-      #- -Fpandoc-crossref #required for cross-ref equations
+      - -Fpandoc-crossref #required for cross-ref equations
     bookdown::html_document2: 
@@
       - --lua-filter=templates/author-info-blocks.lua
+      - -Fpandoc-crossref #required for cross-ref equations
 linkcolor: blue
@@
 
-## Modeling vital rates
+<!--# Emilio will fill this section -->
+
+## IPMs
 
@@
 
-Why general IPM necessary?
+In preliminary investigation we found that the survival and growth of plants was better explained by treating seedlings and mature plants separately.
+Seedlings are physiologically different from small plants because they necessarily lack the underground reserves (of carbohydrates and meristems) that a small, mature plant may have.
+Therefore, we used general IPMs to model population dynamics with seedlings treated as a separate discreet class not structured by size.
+General IPMs allow for combinations of continuous and discrete states and transitions between them [@ellnerDatadrivenModellingStructured2016].
 
--   Seedlings physiologically different from 1-shoot small plants because of lack of underground reserves (of carbs and meristems)
-
--   Seedling survival not well predicted when included with mature plants.
-
-------------------------------------------------------------------------
-
-### Kernel equation and description
-
-Below is a description of the *basic*, deterministic IPM.
-This maybe goes in a supplement?
-These vital rates get swapped out for ones that vary by year (for the stochastic, kernel resampled IPMs) or ones that include lagged effects of SPEI (for the stochastic, parameter resampled IPMs).
-See Figure \@ref(fig:lifecycle).
-
-For mature plants:
+We built three classes of IPMs for comparison which each required different functional forms of their underlying vital rates models.
+The simplest IPM was a general, density-independent, deterministic IPM with four sub-kernels: growth and survival ($P$, @eq:P), fecundity ($F$, i.e.
+production of new seedlings, @eq:F), probability of staying a seedling (always 0), and recruitment ($R$, i.e.
+seedling survival and establishment, @eq:R) ([@eq:mature; @eq:sdlg], Figure \@ref(fig:lifecycle)).
+The probability of staying a seedling, was always equal to zero, since our definition of seedlings was first year plants only.
 
 $$
-n(z^{\prime},t+1) = s_s G_s(z^\prime)n_s(t) + \int_L^U s(z) G(z^{\prime}, z) n(z,t)\;dz
+n(z^{\prime},t+1) = R(z^\prime)n_s(t) + \int_L^U P(z^\prime,z) n(z,t)\;dz
+$${#eq:mature}
+
 $$
+n_s(t+1) = \int_L^U F(z) n(z,t) \; dz
+$${#eq:sdlg}
+
+$$
+R(z^\prime) = s_sG_s(z^\prime)
+$${#eq:R}
+
+$$
+P(z^\prime, z) = s(z)G(z^\prime, z)
+$${#eq:P}
+
+$$
+F(z) = p_f(z)f(z)g
+$${#eq:F}
 
@@
 recruitment <!--# *Is* this recruitment?
--->) and survival and growth (or regression) of mature plants.
-Seedlings survive ($s_s$) and grow into mature plants of a particular size ($G_s(z^\prime)$).
-Mature plants survive as a function of size ($s(z)$), and grow (or regress) to a new size as a function of their previous size ($G(z^\prime, z)$).
+-->) and survival and growth (or regression) of mature plants [@eq:mature].
+Seedlings survive ($s_s$) and grow into mature plants of a particular size ($G_s(z^\prime)$) [@eq:R].
+Mature plants survive as a function of size ($s(z)$), and grow (or regress) to a new size as a function of their previous size ($G(z^\prime, z)$) [@eq:P].
+Mature plants flower with a probability that is a function of size ($p_f(z)$) and produce a number of seeds as a function of size ($f(z)$), which germinate and establish as seedlings with probability $g$ [@eq:F].
 
-For seedlings:
+Vital rate models for growth ($G_s(z^\prime)$ and $G(z^\prime, z)$) , survival ($s_s$ and $s(z)$), and flowering ($p_f(z)$) were fit using the long term demographic dataset.
+For established plants, these three vital rates were modeled as a smooth function of size in the previous census using generalized additive models (GAMs) fit with the `mgcv` package[@mgcv] in `r version$version.string` [@rcoreteam2020] <!--# probably include a supplement with detail on things like basis, number of knots, etc.
+-->.
+For consistency, seedling survival and growth were also modeled using GAMs, but without size in the previous census as a predictor (i.e. intercept only models).
+For growth models ($G_s(z^\prime)$ and $G(z^\prime, z)$) a scaled t family distribution provided a better fit to the data than a gaussian fit as the residuals were leptokurtic with a simple gaussian model.
 
-$$
-n_s(t+1) = \int_L^U p_f(z)f(z)g n(z,t) \; dz
-$$
-
-Mature plants flower with a probability that is a function of size ($p_f(z)$) and produce a number of seeds as a function of size ($f(z)$).
-Those seeds then germinate and establish as seedlings with probability $g$.
-
-------------------------------------------------------------------------
-
-Ultimately, we built and iterated three types of IPMs which each required different underlying vital rate model structure (Figure \@ref(fig:lifecycle)).
-For all models, we used the long term demographic dataset for modeling survival, growth, and flowering.
-For established plants, these three vital rates were modeled as a smooth function of size in the previous census using generalized additive models (GAMs) fit with the `mgcv` package[@mgcv] in `r version$version.string` [@rcoreteam2020].
-Seedlings were considered a separate discrete class not structured by size.
-For consistency, seedling survival and growth were also modeled using GAMs, but without size in the previous census as a predictor (i.e. intercept only models).
-For growth models, a scaled t family distribution provided a better fit to the data than a gaussian fit as the residuals were leptokurtic with a simple gaussian model.
-
-To estimate reproduction we drew on additional data sources to estimate the number of fruits per flowering plant as a function of plant size, the number of seeds per fruit, and germination and establishment rates in continuous forest and forest fragments.
+To estimate reproduction we drew on additional data sources to estimate the number of fruits per flowering plant as a function of plant size and the number of seeds per fruit (together $f(z)$).
+Germination and establishment rates in continuous forest and forest fragments were estimated using data from ....
 <!--# Emilio: add more info here -->
 
-These models were used to build a general, density-independent, deterministic IPM with four sub-kernels: growth and survival, reproduction, probability of staying a seedling, and seedling survival and establishment.
-<!--# include equation --> The probability of staying a seedling, was always equal to zero, since our definition of seedlings was first year plants only.
-
-For the second approach, we included environmental stochasticity by adding a random effect of year to all vital rate models built with the long term demographic dataset.
-The random effect of year was fit using a factor--smooth interaction which allowed the relationship between plant size and vital rates to vary in functional form among transition years.
-Then, we used these models to build general, density independent, stochastic, kernel-resampled IPMs.
+To build the general, density-independent, stochastic, kernel-resampled IPMs, we included environmental stochasticity in all vital rate models built using the long term demographic datset by adding a random effect of year (Figure \@ref(fig:lifecycle)).
+The random effect of year was included using a factor--smooth interaction which allowed the relationship between plant size and vital rates to vary in functional form among transition years.
 The kernel-resampling approach is to generate kernels corresponding to each transition year in the demographic dataset using the random smooths for year, and to iterate the IPM by drawing from these randomly.
@@
 
-For the third method, we modeled the impacts of drought on vital rates explicitly.
+For the third method, we modeled the impacts of drought on vital rates explicitly and created general, density-independent, stochastic, parameter-resampled IPMs (*sensu* @metcalfStatisticalModellingAnnual2015).
 We calculated the standardized precipitation evapotranspiraton index (SPEI) for our site using a published gridded dataset based on ground measurements [@xavierDailyGriddedMeteorological2016] as described in @scottDelayedEffectsClimate2022.
-We modeled delayed effects of SPEI on survival, growth, and flowering probability using distributed lag non-linear models fit as two dimensional tensor product splines, with a maximum lag of 36 months [@scottDelayedEffectsClimate2022].
-Then these models were incorporated into a general, density independent, stochastic, parameter-resampled IPM (*sensu* @metcalfStatisticalModellingAnnual2015).
-To iterate such an IPM, a random sequence of SPEI values is created by sampling years of the observed monthly SPEI data.
+For all vital rate models fit using the long term demographic dataset, we modeled delayed effects of SPEI using distributed lag non-linear models with a maximum lag of 36 months[@scottDelayedEffectsClimate2022] (Figure \@ref(fig:lifecycle)).
+To iterate these parameter-resampled IPMs, a random sequence of SPEI values was created by sampling years of the observed monthly SPEI data.
 Then, 36 month lags are calculated for each year starting in February (the month of the demographic census).
 These values are then used to predict fitted values from the vital rates models, generating different kernels at each iteration of the IPM.
-With this method, the kernels of successive iterations are not entirely independent because the SPEI values used in calculating vital rates include values used in the previous two iterations.
+With this method, the kernels of successive iterations are not entirely independent because the SPEI values used in calculating vital rates include values used in the previous two iterations, but they are ergodic.
+<!--# ergodicity is somethign discussed in the Caswell book.
+Basically, kernels eventually "forget" previous states.
+I don't know if this last sentence really needs to be included, or if we are just inviting criticism from reviewers.
+-->
 
@@
 
-To calculate 95% confidence intervals around the per-capita growth rates (lambdas), we created 500 bootstraps of the demographic dataset by sampling individual plants with replacement within each habitat.
+To estimate uncertainty around the per-capita growth rates (lambdas), we created 500 bootstraps of the demographic dataset by sampling individual plants with replacement within each habitat.
 For each bootstrap, we then re-fit vital rates models (all except germination and establishment rate, fruits per flowering plant, and seeds per fruit, which were estimated using different datasets), constructed IPMs, and calculated a value for lambda as described above.
-<!--# then calculated bias corrected 95% CIs -->
+We then used these bootstraped estimates of lambda to calculate bias corrected 95% confidence intervals [@ellnerDatadrivenModellingStructured2016].
 
@@
 
-```{r}
-#| lifecycle
+```{r lifecycle}
 #| fig.cap = "Diagram of IPMs with equations"
@@
 | Description                                  | Deterministic    | Stochastic, kernel-resampled | Stochastic, parameter-resampled (DLNM) |
-|----------------------|----------------|----------------|-------------------|
+|--------------------|------------------|------------------|------------------|
 | Mature plant survival                        | $s(z)$           | $s_y(z)$                     | $s(z;\theta_{0-36})$                   |

### 2022-05-26 docs/paper.Rmd
fixed up lifecycle diagram to include rendered equations

@@
 pandoc.table(
-  tar_read(aic_tbl_df),
+  tar_read(aic_tbl_df) %>% rename("edf" = "df"),
   justify = c("left", "left", "left", "right", "right", "right"),
@@
 ```{r lifecycle}
-#| fig.cap = "Diagram of IPMs with equations"
+#| fig.cap = "Lifecycle diagram of IPMs with equations.  Diagram shows equations for the deterministic IPM and the table shows how the equations differ in the other two IPM types."
 
 # Draft of this figure is in docs/figures/Heliconia lifecycle DAG.key
-knitr::include_graphics(here::here("docs", "figures", "lifecycle.png"))
+knitr::include_graphics(here::here("docs", "figures", "lifecycle.pdf"))
 ```
 
-Combine above figure with this table:
-
-| Description                                  | Deterministic    | Stochastic, kernel-resampled | Stochastic, parameter-resampled (DLNM) |
-|--------------------|------------------|------------------|------------------|
-| Mature plant survival                        | $s(z)$           | $s_y(z)$                     | $s(z;\theta_{0-36})$                   |
-| Seedling survival                            | $s_0$            | $s_{0_y}$                    | $s_0(\theta_{0-36})$                   |
-| Mature plant growth or regression            | $G(z^\prime, z)$ | $G_y(z^\prime,z)$            | $G(z^\prime,z;\theta_{0-36})$          |
-| Seedling growth into mature plant            | $G_0(z^\prime)$  | $G_{0_y}(z^\prime)$          | $G(z^\prime;\theta_{0-36})$            |
-| Probability of flowering                     | $p_f(z)$         | $p_{f_y}(z)$                 | $p_f(z;\theta_{0-36})$                 |
-| Size-specific fecundity                      | $f(z)$           | $f(z)$                       | $f(z)$                                 |
-| Probability of germination and establishment | $g$              | $g$                          | $g$                                    |
-
-<!--# probably can't make the figure with the table in it in R easily. -->
-
-\
-\
-Just playing with something here.
+<!-- Just playing with something here.
 This is the population trajectories from the iterated IPMs with `normalize_pop_size=FALSE`.
@@
 Not sure this figure is worth including.
+-->
 

### 2022-05-27 docs/paper.Rmd
Working on tables

@@
 
-\<!-
--# TODO:
+```{=html}
+<!-- TODO:
 
@@
 -   Add info about additional datasets (Emilio)
-
---\>
-
+-->
+```
 In preliminary investigation we found that the survival and growth of plants was better explained by treating seedlings and mature plants separately.
@@
 library(pander)
+library(glue)
+library(snakecase)
 ```
 
-<!--# Note: The old version of this table where "random effect of year" only meant a random intercept has DLNM *always* winning as best fit model.
-Should other ways of fitting random effects be included for this table (even if not in the IPMs)?
-Could add simple GLMMs, for example.
--->
+```{r aic, results='asis'}
+#TODO: only show first instance of habitat and vital rate columns, not on every row.  Alternatively do something fancy with flextable or gt.
+tar_read(aic_tbl_df) %>%
+  group_by(habitat, model) %>% 
+  # mutate(rank = row_number()) %>% 
+  ungroup() %>% 
+  mutate(habitat = toupper(habitat),
+         env_effect_incl = ifelse(
+           env_effect_incl != "DLNM",
+           to_sentence_case(as.character(env_effect_incl)),
+           as.character(env_effect_incl)
+         )
+  )%>% 
+  mutate(
+    model = fct_relevel(
+      model,
+      c("surv", "size", "flwr", "surv_sdlg", "size_sdlg")
+    ),
+    model = fct_recode(
+      model,
+      "Survival"          = "surv",
+      "Growth"            = "size",
+      "Flowering"         = "flwr",
+      "Seedling survival" = "surv_sdlg",
+      "Seedling growth"   = "size_sdlg"
+    )
+  ) %>%
 
-```{r aic, results='asis'}
+   #another option is to use psuedo model notation
+  mutate(RHS = str_replace_all(
+    env_effect_incl,
+    c(
+      "None" = "~s(z)",
+      "Random effect of year" = "~s(z, by = 'year')",
+      "DLNM" = "~s(z) + te(spei, lag)"
+    )
+  )) %>%
+
+  arrange(habitat, model) %>% 
+  select(
+    "Habitat" = "habitat",
+    "Vital Rate" = "model",
+   "Effect of Environment" = "env_effect_incl",
+   #"Model" = "RHS",
+    "edf" = "df",
+    "∆AIC" = "dAIC"
+  ) %>%
+
 pandoc.table(
-  tar_read(aic_tbl_df) %>% rename("edf" = "df"),
-  justify = c("left", "left", "left", "right", "right", "right"),
+  justify = c("left", "left", "left", "right",  "right"),
   full_width = FALSE,
-  caption = "(\\#tab:aic) Comparison of vital rate models for deterministic, stochastic kernel-resampled, and stochastic parameter-resampled IPMs.
-'edf' is the estimated degrees of freedom of the penalized GAM."
+  caption = "(\\#tab:aic) Comparison of vital rate models used to build IPM.
+The 'Effect of Environment' column describes how environmental effects were included in models.
+Those with 'none' were used to build deterministic IPMs; those with a random effect of year were used to build stochastic, kernel-resampled IPMs; and those with a distributed lag non-linear model (DLNM) were used to build stochastic, parameter-resampled IPMs.
+'edf' is the estimated degrees of freedom of the penalized GAM.
+∆AIC is calculated within each habitat and vital rate combination.
+∆AIC within 2 indicates models are equivalent."
 )
@@
 ```{r lambdas, results='asis'}
+tar_read(lambda_table) %>% 
+  mutate(habitat = toupper(habitat)) %>% 
+  mutate(across(c(est, lower, upper), ~format(round(.,4)))) %>% 
+  mutate(lambda = glue::glue("{est} ({lower}, {upper})")) %>%
+  mutate(ipm = str_replace_all(
+    ipm,
+    c(
+      "det" = "Deterministic",
+      "stoch" = "Stochastic, kernel-resampled",
+      "DLNM" = "Stochastic, parameter-resampled"
+    )
+  )) %>% 
+  select(IPM = ipm, Habitat = habitat, "$\\lambda$" = lambda) %>% 
 pandoc.table(
-  tar_read(lambda_table),
   digits = 5,
   keep.trailing.zeros = TRUE,
-  justify = c("left", "left", "right", "right", "right"),
+  justify = c("left", "left", "right"),
   full_width = FALSE,
@@
 ```{r lifecycle}
-#| fig.cap = "Lifecycle diagram of IPMs with equations.  Diagram shows equations for the deterministic IPM and the table shows how the equations differ in the other two IPM types."
+#| fig.cap = "Lifecycle diagram of *Heliconia acuminata*.
+Each transition is associated with an equation for a vital rate function.
+The functions shown on the diagram correspond to those used to construct a general, density-independent, deterministic IPM.
+The table below shows the equivalent equations for stochastic, kernel-resampled IPMs and stochastic, parameter-resampled IPMs."
 
+#TODO: update table on figure once you figure out multi-line tables in lifecycle-eq-table.Rmd
 # Draft of this figure is in docs/figures/Heliconia lifecycle DAG.key

### 2022-05-27 docs/paper.Rmd
finish fixing up tables

@@
 #TODO: only show first instance of habitat and vital rate columns, not on every row.  Alternatively do something fancy with flextable or gt.
+
 tar_read(aic_tbl_df) %>%
@@
 
-   #another option is to use psuedo model notation
+   #another option is to use psuedo model notation.  Can't fit both unless I remove the "habitat" column or something.
   mutate(RHS = str_replace_all(
@@
 pandoc.table(
-  justify = c("left", "left", "left", "right",  "right"),
+  justify = c( "left", "left","left", "right",  "right"),
   full_width = FALSE,
@@
   full_width = FALSE,
-  caption = "(\\#tab:lambda) Population growth rates for continuous forest (cf) and forest fragments (ff) under different kinds of IPMs with bootstrapped, bias-corrected, 95% confidence intervals."
+  caption = "(\\#tab:lambda) Population growth rates for continuous forest (CF) and forest fragments (FF) under different kinds of IPMs with bootstrapped, bias-corrected, 95% confidence intervals."
 )

### 2022-05-27 R/plot_pop_states.R
plot population states as a facet_grid

@@ -1,31 +1,92 @@
+# library(targets)
+# library(ipmr)
+# tar_load(c(
+#   ipm_det_cf,
+#   ipm_det_ff,
+#   ipm_stoch_cf,
+#   ipm_stoch_ff,
+#   ipm_dlnm_cf,
+#   ipm_dlnm_ff
+# ))
+
+
+
 #' Plot population stage structure over time
 #'
-#' For a stochastic Heliconia IPM, plots size distribution changes over
-#' iterations as a binned area plot.
+#' @param ipm_list a named list of `ipmr` IPM objects.  See example
+#' @param bins number of bins for log_size
+#' @param save_path a filepath (including filename and extension) to save the plot.  Passed to `ggsave()`
+#' @param ... other arguments passed to `ggsave()`
 #'
-#' @param ipm an iterated ipmr ipm
-#' @param bins number of bins to group continuous log_size into (not including
-#'   seedlings)
-#' @param normalize logical; TRUE plots proportions while FALSE (default) plots
-#'   actual numbers
+#' @return a ggplot object
 #'
-plot_pop_states <- function(ipm, bins = 6, normalize = FALSE) {
+#' @examples
+#' list(
+#'   det_cf = ipm_det_cf,
+#'   det_ff = ipm_det_ff,
+#'   stoch_cf = ipm_stoch_cf,
+#'   stoch_ff = ipm_stoch_ff,
+#'   dlnm_cf = ipm_dlnm_cf,
+#'   dlnm_ff = ipm_dlnm_ff
+#' ) %>%
+#'   plot_pop_states(bins = 6, save_path = "test.png", height = 5)
+#'   
+plot_pop_states <- function(ipm_list, bins = 7, save_path = NULL, ...) {
+  df <-
+    ipm_list %>%
+    #extract and bin population states from all IPMs
+    map_df(~bin_pop_states(.x, bins), .id = "model") %>% 
+    #wrangle into a pretty dataframe
+    separate(model, into = c("ipm", "habitat")) %>% 
+    mutate(
+      ipm = str_replace_all(
+        ipm,
+        c(
+          "det" = "Deterministic",
+          "stoch" = "Kernel-resampled",
+          "dlnm" = "Parameter-resampled"
+        )
+      ),
+      habitat = toupper(habitat)
+    ) %>% 
+    rename(IPM = ipm, Habitat = habitat)
+  
+  #plot with facet_grid
+  
+  p <- 
[…]

### 2022-05-27 _targets.R

@@ -235,6 +235,22 @@ tar_plan(
       usr_funs = list(get_scat_params = get_scat_params)
     ),
  
+
+  ## ├Plot population states --------------------------------------------------
+
+  tar_target(
+    "plot_pop_states",
+    list(
+      det_cf = ipm_det_cf,
+      det_ff = ipm_det_ff,
+      stoch_cf = ipm_stoch_cf,
+      stoch_ff = ipm_stoch_ff,
+      dlnm_cf = ipm_dlnm_cf,
+      dlnm_ff = ipm_dlnm_ff
+    ) %>%
+      plot_pop_states(bins = 7, save_path = "docs/figures/pop_states.png")
+  ),
+  
   # Bootstrapped Lambdas -----------------------------------------------------------------
   # uses tar_rep() target factory to create dynamic branches to do 500
   # bootstraps of data, then use all-in-one functions to fit vital rate models,

### 2022-06-01 README.Rmd
update html version of manuscript
add mermaid graph to README

@@
 
-The most recent draft of the manuscript is available [here](https://brunalab.github.io/lagged-ipms/paper.html).
+The most recent draft of the manuscript is available [here](https://brunalab.github.io/lagged-ipms/docs/paper.html).
 
@@
 -   This project uses `renv` for package management, which should be installed automatically when opening the project. Run `renv::restore()` to install all required packages.
--   Run `targets::tar_make()` or `targets::tar_make_clustermq()` from the R console to run all code and produce all outputs.
+-   This project uses `targets` for workflow management.
+The functions used in the project are defined in `R/` and the steps are defined in `_targets.R` and summarized in the schematic below.
+Run `targets::tar_make()` or `targets::tar_make_clustermq()` from the R console to run all code and produce all outputs.
+
+```{r, echo=FALSE, message=FALSE, results='asis'}
+cat(
+  "```mermaid",
+  targets::tar_mermaid(targets_only = TRUE),
+  "```",
+  sep = "\n"
+)
+#this should display correctly on GitHub, or code can be pasted into https://mermaid.live
+```
 

### 2022-06-01 _targets.R

@@ -331,8 +331,9 @@ tar_plan(
   ),
 
 
-  # Manuscript --------------------------------------------------------------
-  tar_render(paper, here("docs", "paper.Rmd"), packages = "bookdown", output_format = "all")
+  # Outputs --------------------------------------------------------------
+  tar_render(paper, here("docs", "paper.Rmd"), packages = "bookdown", output_format = "all"),
+  tar_render(readme, "README.Rmd")
   
   
 ) %>% 

### 2022-06-01 README.Rmd
suppress package startup messages

@@
 
-```{r, echo=FALSE, message=FALSE, results='asis'}
+```{r echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
 cat(

### 2022-06-01 packages.R

@@ -1,18 +1,20 @@
-## library() calls go here
-library(conflicted)
-library(dotenv)
-library(targets)
-library(tarchetypes)
-library(tidyverse)
-library(here)
-library(SPEI)
-library(tsibble)
-library(lubridate)
-library(tsModel)
-library(mgcv)
-library(gratia)
-library(broom)
-library(metRology)
-library(clustermq)
-library(ipmr)
+suppressPackageStartupMessages({
+  ## library() calls go here
+  library(conflicted)
+  library(dotenv)
+  library(targets)
+  library(tarchetypes)
+  library(tidyverse)
+  library(here)
+  library(SPEI)
+  library(tsibble)
+  library(lubridate)
+  library(tsModel)
+  library(mgcv)
+  library(gratia)
+  library(broom)
+  library(metRology)
+  library(clustermq)
+  library(ipmr)
+})
 

### 2022-06-01 R/plot_pop_states.R
plot figure of poulation states

@@ -12,13 +12,24 @@
 
 
 #' Plot population stage structure over time
+#' 
+#' Uses custom size bins relating to changes in vital rates.  "seedling" is the
+#' discrete class for seedlings, "pre-reproductive 1" are plants with low
+#' survival (~ < 0.8) and a near 0 probability of flowering, "pre-reproductive
+#' 2" are plants with high survival (~ >0.8) and a near 0 probability of
+#' flowering, "reproductive 1" are plants with high survival probability and low
+#' flowering probability (~ < 0.3), "reproductive 2" are plants with high
+#' survival probability and high flowering probability (~ > 0.3).
 #'
 #' @param ipm_list a named list of `ipmr` IPM objects.  See example
-#' @param bins number of bins for log_size
-#' @param save_path a filepath (including filename and extension) to save the plot.  Passed to `ggsave()`
+#' @param xlim a length 2 vector passed to `coord_cartesian()`.  By default only
+#'   the first 250 iterations are plotted for clarity.
+#' @param save_path a filepath (including filename and extension) to save the
+#'   plot.  Passed to `ggsave()`
 #' @param ... other arguments passed to `ggsave()`
-#'
-#' @return a ggplot object
+#'   
+#' @return if `save_path` is not specified, then a ggplot object is returned.
+#'   Otherwise `save_path` is returned.
 #'
 #' @examples
 #' list(
@@ -29,13 +40,13 @@
 #'   dlnm_cf = ipm_dlnm_cf,
 #'   dlnm_ff = ipm_dlnm_ff
 #' ) %>%
-#'   plot_pop_states(bins = 6, save_path = "test.png", height = 5)
+#'   plot_pop_states(save_path = "test.png", height = 5)
 #'   
-plot_pop_states <- function(ipm_list, bins = 7, save_path = NULL, ...) {
+plot_pop_states <- function(ipm_list, xlim = c(0, 250), save_path = NULL, ...) {
   df <-
     ipm_list %>%
     #extract and bin population states from all IPMs
-    map_df(~bin_pop_states(.x, bins), .id = "model") %>% 
+    map_df(~bin_pop_states(.x), .id = "model") %>% 
     #wrangle into a pretty dataframe
     separate(model, into = c("ipm", "habitat")) %>% 
     mutate(
[…]

### 2022-06-01 _targets.R

@@ -248,7 +248,8 @@ tar_plan(
       dlnm_cf = ipm_dlnm_cf,
       dlnm_ff = ipm_dlnm_ff
     ) %>%
-      plot_pop_states(bins = 7, save_path = "docs/figures/pop_states.png", height = 5, width = 7),
+      plot_pop_states(save_path = "docs/figures/pop_states.png",
+                      height = 7.5, width = 8.5),
     deployment = "main",
     format = "file"
   ),

### 2022-06-01 packages.R

@@ -16,5 +16,6 @@ suppressPackageStartupMessages({
   library(metRology)
   library(clustermq)
   library(ipmr)
+  library(patchwork)
 })
 

### 2022-06-01 docs/paper.Rmd
add figure caption and some bullet-point results

@@
 
+Figure \@ref(fig:pop-states) has some interesting things in it:
+
+-   for the deterministic IPM (and the kernel-resampled IPM?) there are slightly more of the smallest plants and the largest plants in CF compared to FF (i.e. more medium sized plants in FF).
+
+-   For the kernel-resampled IPM (random effect of year), the fluctuations are extremely similar between CF and FF
+
+-   For the parameter-resampled IPM (DLNM) the size structure of the population is a LOT more variable in FF.
+    This makes sense as we know lagged effects are more important in fragments.
+
+-   Also, the fluctuations in size structure in CF do not match the fluctuations in FF as well (can see this by the increased spread of points in Figure \@ref(fig:pop-states)B)
+
+-   Also, in the parameter-resampled IPM (and only in this one), we see a shift toward smaller plants in FF compared to CF
+
 # Discussion
@@
 
-#TODO: update table on figure once you figure out multi-line tables in lifecycle-eq-table.Rmd
 # Draft of this figure is in docs/figures/Heliconia lifecycle DAG.key
@@
 
-<!-- Just playing with something here.
-This is the population trajectories from the iterated IPMs with `normalize_pop_size=FALSE`.
-I thought maybe I'd be able to see some (lagged) similarities between the kernel sampled and DLNM IPMs because they are run with the same sequence of years.
-Nothing obvious pops out.
-Not sure this figure is worth including.
--->
+```{r pop-states}
+#| fig.cap = "Relative proportions of plant sizes in the first 250 iterations of the IPM simulations.
+Stacked area charts (A) show the relative size/stage distribution of plants in continuous forest (CF, top row) and forest fragments (FF, bottom row) in each of the three IPMs (columns).
+The proportion of each size class in CF and FF for each iteration is shown in B with the first 30 iterations removed to not include transient dynamics.
+A 1:1 line is plotted in black.
+Size categories include seedlings (a discrete category in the IPMs), pre-reproductive 1 (log(size) 0–2.5) that have low average survival (< 0.9) and a near 0 probability of flowering, pre-reproductive 2 (log(size) 2.5–4.5) that have a higher average survival probabilty (> 0.8) and a near 0 probability of flowering, reproductive 1 (log(size) 4.5–6) that have a high average survival probability (>0.95) and a lower flowering probability (< 0.25), and reproductive 2 (log(size) 6+) that have a high average survival probability (>0.95) and higher flowering proability (> 0.2)."
 
-```{r eval=FALSE}
-#| pva,
-#| fig.cap = "Population trajectories for all three IPMs in both habitats"
-tar_load(c(
-  ipm_det_cf,
-  ipm_det_ff,
-  ipm_stoch_cf,
-  ipm_stoch_ff,
-  ipm_dlnm_cf,
-  ipm_dlnm_ff
-))
-
-ipm_list <- 
-  list(det_cf = ipm_det_cf,
-       det_ff = ipm_det_ff,
-       stoch_cf = ipm_stoch_cf,
-       stoch_ff = ipm_stoch_ff,
-       dlnm_cf = ipm_dlnm_cf,
-       dlnm_ff = ipm_dlnm_ff)
-
-plot_df <- 
-  ipm_list %>% map_df(~{
-  pop_states <- pop_state(.x)
-
-    lapply(pop_states, colSums) %>%
-    as_tibble() %>% 
-    mutate(n = n_log_size + n_sdlg,
-           t = 1:n())
-}, .id = "model") %>% 
-  separate(model, into = c("ipm", "habitat"))
-
-p <- ggplot(plot_df, aes(x = t, y = n, color = ipm)) +
-  geom_line() +
-  facet_wrap(~habitat, ncol = 1)
-
-p + xlim(10, 250)
-
-p+ scale_y_log10() + labs(y = "log10(n)")
+knitr::include_graphics(tar_read(fig_pop_states))
 

### 2022-06-01 _targets.R

@@ -239,7 +239,7 @@ tar_plan(
   ## ├Plot population states --------------------------------------------------
 
   tar_target(
-    "pop_states_figure",
+    "fig_pop_states",
     list(
       det_cf = ipm_det_cf,
       det_ff = ipm_det_ff,

### 2022-06-01 docs/paper.Rmd
update manuscript to test GH pages

@@
   - Emilio M. Bruna:
+
       email: embruna@ufl.edu
@@
       correspondence: false
-
 institute:
@@
 output: 
-    bookdown::word_document2:
-      linenumbers: yes
-      number_sections: no
-      fig_caption: yes
-      reference_docx: "templates/template.docx" # Insert path for the DOCX file
-      pandoc_args:
-      - --lua-filter=templates/scholarly-metadata.lua
-      - --lua-filter=templates/author-info-blocks.lua
-      - -Fpandoc-crossref #required for cross-ref equations
     bookdown::html_document2: 
@@
       fig_caption: yes
+      pandoc_args:
+      - --lua-filter=templates/scholarly-metadata.lua
+      - --lua-filter=templates/author-info-blocks.lua
+      - -Fpandoc-crossref #required for cross-ref equations
+    bookdown::word_document2:
+      linenumbers: yes
+      number_sections: no
+      fig_caption: yes
+      reference_docx: "templates/template.docx" # Insert path for the DOCX file
       pandoc_args:
@@
   full_width = FALSE,
-  caption = "(\\#tab:lambda) Population growth rates for continuous forest (CF) and forest fragments (FF) under different kinds of IPMs with bootstrapped, bias-corrected, 95% confidence intervals."
+  caption = "(\\#tab:lambdas) Population growth rates for continuous forest (CF) and forest fragments (FF) under different kinds of IPMs with bootstrapped, bias-corrected, 95% confidence intervals."
 )
@@
 
-knitr::include_graphics(tar_read(fig_pop_states))
+knitr::include_graphics(here::here(tar_read(fig_pop_states)))
 

### 2022-06-15 docs/paper.Rmd
fix relative path issues

@@
 library(tidyverse)
+library(here)
+```
+
+```{r}
+#load targets
+withr::with_dir(here(), {
+  tar_load(c(
+    aic_tbl_df,
+    lambda_table,
+    fig_pop_states
+  ))
+})
 ```
@@
 meta_df <- 
-  tar_meta() %>% 
+  tar_meta(store = here("_targets")) %>% 
   select(name, size, bytes, time, seconds) %>% 
@@
 
-tar_read(aic_tbl_df) %>%
+aic_tbl_df %>%
   group_by(habitat, model) %>% 
@@
 ```{r lambdas, results='asis'}
-tar_read(lambda_table) %>% 
+lambda_table %>% 
   mutate(habitat = toupper(habitat)) %>% 
@@
 # Draft of this figure is in docs/figures/Heliconia lifecycle DAG.key
-knitr::include_graphics(here::here("docs", "figures", "lifecycle.pdf"))
+path <- here("docs/figures/lifecycle.png")
+knitr::include_graphics(path, rel_path = FALSE)
 ```
@@
 
-knitr::include_graphics(here::here(tar_read(fig_pop_states)))
+knitr::include_graphics(here(fig_pop_states), rel_path = FALSE)
 

### 2022-06-15 _targets.R
render paper

@@ -333,8 +333,8 @@ tar_plan(
 
 
   # Outputs --------------------------------------------------------------
-  tar_render(paper, here("docs", "paper.Rmd"), packages = "bookdown", output_format = "all"),
-  tar_render(readme, "README.Rmd")
+  tar_render(paper, here("docs", "paper.Rmd"), packages = "bookdown", output_format = "all", deployment = "main"),
+  tar_render(readme, "README.Rmd", deployment = "main")
   
   
 ) %>% 

### 2022-08-04 R/make_lambda_table.R
Update lambda table (closes #25)

@@ -13,7 +13,9 @@
 #'   det_ff = ipm_det_ff,
 #'   det_cf = ipm_det_cf,
 #'   stoch_ff = ipm_stoch_ff,
-#'   stoch_cf = ipm_stoch_cf
+#'   stoch_cf = ipm_stoch_cf,
+#'   dlnm_ff = ipm_dlnm_ff,
+#'   dlnm_cf = ipm_dlnm_cf
 #' )
 #' 
 #' #bt_list should look like this:
@@ -21,10 +23,13 @@
 #'   det_ff = lambda_bt_det_ff,
 #'   det_cf = lambda_bt_det_cf,
 #'   stoch_ff = lambda_bt_stoch_ff,
-#'   stoch_cf = lambda_bt_stoch_cf
-#'   # dlnm_ff = lambda_bt_dlnm_ff,
-#'   # dlnm_cf = lambda_bt_dlnm_cf
+#'   stoch_cf = lambda_bt_stoch_cf,
+#'   dlnm_ff = lambda_bt_dlnm_ff,
+#'   dlnm_cf = lambda_bt_dlnm_cf
 #' )
+#' 
+#' #load the above targets with tar_load(starts_with("lambda")) and
+#' tar_load(starts_with("ipm"))
 #' 
 #' make_lambda_table(ipm_list, bt_list, alpha = 0.05)
 make_lambda_table <- function(ipm_list, bt_list, alpha = 0.05) {

### 2022-08-04 docs/paper.Rmd
added bdffp tech series to ackngmnts

@@
 National Science Foundation (awards \_\_\_\_, and \_\_\_\_).
+This article is publication no. -- -- in the BDFFP Technical series.
 The authors declare no conflicts of interest.
@@
       "stoch" = "Stochastic, kernel-resampled",
-      "DLNM" = "Stochastic, parameter-resampled"
+      "dlnm" = "Stochastic, DLNM"
     )
   )) %>% 
-  select(IPM = ipm, Habitat = habitat, "$\\lambda$" = lambda) %>% 
+  select(IPM = ipm, habitat, lambda) %>% 
+  pivot_wider(names_from = "habitat", values_from = lambda) %>%
+
 pandoc.table(
@@
   keep.trailing.zeros = TRUE,
-  justify = c("left", "left", "right"),
+  justify = c("left", "right", "right"),
   full_width = FALSE,
-  caption = "(\\#tab:lambdas) Population growth rates for continuous forest (CF) and forest fragments (FF) under different kinds of IPMs with bootstrapped, bias-corrected, 95% confidence intervals."
+  caption = "(\\#tab:lambdas) Population growth rates (lambda) for continuous forest (CF) and forest fragments (FF) under different kinds of IPMs with bootstrapped, bias-corrected, 95% confidence intervals."
 )
-
+#TODO: in caption possibly explain that lambdas are calculated differently for deterministic and stochastic models
 ```

### 2023-07-18 docs/paper.Rmd
added updated packages

--- 2022-06-15 docs/paper.Rmd
@@
 output: 
+    # bookdown::pdf_document2:
+    #   floatsintext: no
+    #   linenumbers: yes
+    #   draft: yes
+    #   mask: no
+    #   figurelist: no
+    #   tablelist: no
+    #   footnotelist: no
+    #   fontsize: 12pt
+    #   toc: false
+    #   number_sections: no
+    #   fig_caption: yes
     bookdown::html_document2: 
@@
     canonical: true
+
+# header-includes:
+header-includes:
+  - \usepackage[left]{lineno}
+  - \linenumbers
+  - \usepackage{fancyhdr}
+  - \pagestyle{fancy}
+  - \fancyfoot{}
+  - \fancyhead[R]{p. \thepage}
+  - \fancyhead[L]{Draft - `r format(Sys.time(), '%d %B %Y')`}
+  - \usepackage{sectsty} \sectionfont{\centering}
+  - \DeclareUnicodeCharacter{U+2206}{$\delta$}
+  # - \usepackage{sectsty} \subsectionfont{\centering}
+  # - \AtBeginDocument{\let\maketitle\relax}
+ #  - \usepackage{float} #to keep figure in the location placed instead of end
+ #  - \floatplacement{figure}{H}
+ #  - \usepackage{booktabs}
+ #  - \usepackage{setspace} # line spacing
+ #  - \raggedbottom
+ # # - \usepackage{endfloat} #[nomarkers] excludes the {insert figure x around here] from main text.
+The others exclude the list of tables and figures.
+https://cs.brown.edu/about/system/managed/latex/doc/endfloat.pdf
+ #  - \usepackage{tabu}
+ #  - \usepackage{makecell}
+ #  - \usepackage{pdflscape}
+ #  - \usepackage{longtable}
+ #  - \newcommand{\blandscape}{\begin{landscape}}
+ #  - \newcommand{\elandscape}{\end{landscape}}
+ #  - \setlength\parindent{22pt}
 ---
@@
 
+<!-- AmNat Note Abstracts are no longer than 150 words. -->
+
 Abstract text.
@@
 # Introduction
+
+<!-- AmNat Notes: communicate concise points, using either data or theory.
+Like Major Articles, they present insights of broad general significance and interest.
+Notes, on average, should be no more than 3000 words of text (not including the Literature Cited) and have no more than three figures and/or tables in print.
+-->
 

### 2023-07-18 docs/paper.tex
added updated packages

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage[left]{lineno}
\linenumbers
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyfoot{}
\fancyhead[R]{p. \thepage}
\fancyhead[L]{Draft - 18 July 2023}
\usepackage{sectsty} \sectionfont{\centering}
\DeclareUnicodeCharacter{U+2206}{$\delta$}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Context-dependent consequences of including lagged effects in demographic models},
  pdfkeywords={integral projection models; environmental stochasticity; lagged effects},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Context-dependent consequences of including lagged effects in demographic models}
\author{true \and true}
\date{18 July, 2023}

\begin{document}
\maketitle

Keywords: integral projection models; environmental stochasticity; lagged effects

Running title: Lagged Effects IPMs

\pagebreak

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

Abstract text.

\pagebreak

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Demographic models are widely used for all kinds of stuff
\item
  It has long been recognized that there is the potential for lagged effects.
  Now there's evidence for that (Evers et al. 2021; Scott et al. 2022).
\item
  Lagged effects could be different in different habitat types (and this could be a big reason for differences between habitats)
\end{enumerate}

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{demographic-data}{%
\subsection{Demographic Data}\label{demographic-data}}

\hypertarget{ipms}{%
\subsection{IPMs}\label{ipms}}

In preliminary investigation we found that the survival and growth of plants was better explained by treating seedlings and mature plants separately.
Seedlings are physiologically different from small plants because they necessarily lack the underground reserves (of carbohydrates and meristems) that a small, mature plant may have.
Therefore, we used general IPMs to model population dynamics with seedlings treated as a separate discreet class not structured by size.
General IPMs allow for combinations of continuous and discrete states and transitions between them (Ellner et al. 2016).

We built three classes of IPMs for comparison which each required different functional forms of their underlying vital rates models.
The simplest IPM was a general, density-independent, deterministic IPM with four sub-kernels: growth and survival (\(P\), (\textbf{eq:P?})), fecundity (\(F\), i.e.~production of new seedlings, (\textbf{eq:F?})), probability of staying a seedling (always 0), and recruitment (\(R\), i.e.~seedling survival and establishment, (\textbf{eq:R?})) ((\textbf{eq:mature?}; \textbf{eq:sdlg?}), Figure \ref{fig:lifecycle}).
The probability of staying a seedling, was always equal to zero, since our definition of seedlings was first year plants only.

\[
n(z^{\prime},t+1) = R(z^\prime)n_s(t) + \int_L^U P(z^\prime,z) n(z,t)\;dz
\]\{\#eq:mature\}

\[
n_s(t+1) = \int_L^U F(z) n(z,t) \; dz
\]\{\#eq:sdlg\}

\[
R(z^\prime) = s_sG_s(z^\prime)
\]\{\#eq:R\}

\[
P(z^\prime, z) = s(z)G(z^\prime, z)
\]\{\#eq:P\}

\[
F(z) = p_f(z)f(z)g
\]\{\#eq:F\}

The number and size of mature plants in the next census is determined by seedlings entering the mature plant population (i.e.~recruitment ) and survival and growth (or regression) of mature plants (\textbf{eq:mature?}).
Seedlings survive (\(s_s\)) and grow into mature plants of a particular size (\(G_s(z^\prime)\)) (\textbf{eq:R?}).
Mature plants survive as a function of size (\(s(z)\)), and grow (or regress) to a new size as a function of their previous size (\(G(z^\prime, z)\)) (\textbf{eq:P?}).
Mature plants flower with a probability that is a function of size (\(p_f(z)\)) and produce a number of seeds as a function of size (\(f(z)\)), which germinate and establish as seedlings with probability \(g\) (\textbf{eq:F?}).

Vital rate models for growth (\(G_s(z^\prime)\) and \(G(z^\prime, z)\)) , survival (\(s_s\) and \(s(z)\)), and flowering (\(p_f(z)\)) were fit using the long term demographic dataset.
For established plants, these three vital rates were modeled as a smooth function of size in the previous census using generalized additive models (GAMs) fit with the \texttt{mgcv} package(Wood 2011) in R version 4.3.0 (2023-04-21) (R Core Team 2020) .
For consistency, seedling survival and growth were also modeled using GAMs, but without size in the previous census as a predictor (i.e.~intercept only models).
For growth models (\(G_s(z^\prime)\) and \(G(z^\prime, z)\)) a scaled t family distribution provided a better fit to the data than a gaussian fit as the residuals were leptokurtic with a simple gaussian model.

To estimate reproduction we drew on additional data sources to estimate the number of fruits per flowering plant as a function of plant size and the number of seeds per fruit (together \(f(z)\)).
Germination and establishment rates in continuous forest and forest fragments were estimated using data from \ldots.

To build the general, density-independent, stochastic, kernel-resampled IPMs, we included environmental stochasticity in all vital rate models built using the long term demographic datset by adding a random effect of year (Figure \ref{fig:lifecycle}).
The random effect of year was included using a factor--smooth interaction which allowed the relationship between plant size and vital rates to vary in functional form among transition years.
The kernel-resampling approach is to generate kernels corresponding to each transition year in the demographic dataset using the random smooths for year, and to iterate the IPM by drawing from these randomly.
This is equivalent to the matrix selection approach for matrix population models described by Caswell (2001).

For the third method, we modeled the impacts of drought on vital rates explicitly and created general, density-independent, stochastic, parameter-resampled IPMs (\emph{sensu} Metcalf et al. (2015)).
We calculated the standardized precipitation evapotranspiraton index (SPEI) for our site using a published gridded dataset based on ground measurements (Xavier et al. 2016) as described in Scott et al. (2022).
For all vital rate models fit using the long term demographic dataset, we modeled delayed effects of SPEI using distributed lag non-linear models with a maximum lag of 36 months(Scott et al. 2022) (Figure \ref{fig:lifecycle}).
To iterate these parameter-resampled IPMs, a random sequence of SPEI values was created by sampling years of the observed monthly SPEI data.
Then, 36 month lags are calculated for each year starting in February (the month of the demographic census).
These values are then used to predict fitted values from the vital rates models, generating different kernels at each iteration of the IPM.
With this method, the kernels of successive iterations are not entirely independent because the SPEI values used in calculating vital rates include values used in the previous two iterations, but they are ergodic.

All IPMs were constructed and iterated using the \texttt{ipmr} package in R (Levin et al. 2021).
The IPMs used 100 meshpoints and the midpoint rule for calculating kernels .
For each type of IPM we iterated the model for 1000 time steps, discarding the first 100 time steps to omit transient effects.
Stochastic growth rates (\(\lambda_s\)) were calculated as the average \(ln(\lambda)\) from each time step (Caswell 2001) and back-transformed to be on the same scale as deterministic lambdas for comparison.
We used the distribution of established plant sizes and proportion of seedlings from the full dataset as a starting population vector.
While other starting population vectors were possible, the choice is of little importance as it will only impact transient dynamics, which we aren't interested in for this study.

To estimate uncertainty around the per-capita growth rates (lambdas), we created 500 bootstraps of the demographic dataset by sampling individual plants with replacement within each habitat.
For each bootstrap, we then re-fit vital rates models (all except germination and establishment rate, fruits per flowering plant, and seeds per fruit, which were estimated using different datasets), constructed IPMs, and calculated a value for lambda as described above.
We then used these bootstraped estimates of lambda to calculate bias corrected 95\% confidence intervals (Ellner et al. 2016).

This workflow was managed using the \texttt{targets} R package (Landau 2021) which also allowed us to track computational time spent on each IPM for comparison.

\hypertarget{results}{%
\section{Results}\label{results}}

For all vital rates estimated using the long term demographic dataset, the DLNM model fit the best (dAIC = 0) followed by the model with a random effect of year, followed by the deterministic model (Table \ref{tab:aic}).

Population growth rates were consistently higher in continuous forest compared to forest fragments across IPM types (Table \ref{tab:lambdas}).

The time to iterate the DLNM models is much higher than than deterministic and kernel-resampled.
The greater use of computational resources is likely a result of \texttt{predict()} being much slower for GAMs with 2D smooths because the number of knots is much higher compared to the GAMs used for the vital rates models in the determinsitic and kernel-resampled IPMs.

\begin{verbatim}
#> # A tibble: 3 x 2
#>   IPM   mean_time_min
#>   <chr>         <dbl>
#> 1 det            0.02
#> 2 dlnm          87.1 
#> 3 stoch          0.07
\end{verbatim}

Figure \ref{fig:pop-states} has some interesting things in it:

\begin{itemize}
\item
  for the deterministic IPM (and the kernel-resampled IPM?) there are slightly more of the smallest plants and the largest plants in CF compared to FF (i.e.~more medium sized plants in FF).
\item
  For the kernel-resampled IPM (random effect of year), the fluctuations are extremely similar between CF and FF
\item
  For the parameter-resampled IPM (DLNM) the size structure of the population is a LOT more variable in FF.
  This makes sense as we know lagged effects are more important in fragments.
\item
  Also, the fluctuations in size structure in CF do not match the fluctuations in FF as well (can see this by the increased spread of points in Figure \ref{fig:pop-states}B)
\item
  Also, in the parameter-resampled IPM (and only in this one), we see a shift toward smaller plants in FF compared to CF
\end{itemize}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

\begin{itemize}
\tightlist
\item
  Our finding that the choice of IPM didn't change the relative ranking of CF and FF is consistent with Kaye and Pyke (2003) finding that method effected stochastic lambda, but relative ranking of populations was consistent.
\end{itemize}

\hypertarget{acknowledgments}{%
\section{Acknowledgments}\label{acknowledgments}}

We thank \textbf{,} \_, \_\_\_ and \_\_\_ anonymous reviewers for helpful discussions and comments on the manuscript.
We thank Sam Levin for his help with the \texttt{ipmr} package.
Financial support was provided by the U.S.
National Science Foundation (awards \_\_\_\_, and \_\_\_\_).
This article is publication no. -- -- in the BDFFP Technical series.
The authors declare no conflicts of interest.

\hypertarget{credit-statement}{%
\section{CRediT Statement}\label{credit-statement}}

ERS contributed to the conceptualization, methodology, formal analysis, and led the writing of the original draft.
EMB contributed to the conceptualization, methodology, and writing and also acquired funding.

\hypertarget{data-availability-statement}{%
\section{Data Availability Statement}\label{data-availability-statement}}

Data and R code used in this study are archived with Zenodo at .

\pagebreak

\hypertarget{tables}{%
\section{Tables}\label{tables}}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2778}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3333}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1111}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1111}}@{}}
\caption{\label{tab:aic} Comparison of vital rate models used to build IPM. The `Effect of Environment' column describes how environmental effects were included in models. Those with `none' were used to build deterministic IPMs; those with a random effect of year were used to build stochastic, kernel-resampled IPMs; and those with a distributed lag non-linear model (DLNM) were used to build stochastic, parameter-resampled IPMs. `edf' is the estimated degrees of freedom of the penalized GAM. ∆AIC is calculated within each habitat and vital rate combination. ∆AIC within 2 indicates models are equivalent.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Habitat
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Vital Rate
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Effect of Environment
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
edf
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
∆AIC
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Habitat
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Vital Rate
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Effect of Environment
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
edf
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
∆AIC
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
CF & Survival & Random effect of year & 43.26 & 0 \\
CF & Survival & DLNM & 19.72 & 78.92 \\
CF & Survival & None & 4.976 & 260 \\
CF & Growth & Random effect of year & 78.43 & 0 \\
CF & Growth & DLNM & 23.87 & 158.5 \\
CF & Growth & None & 7.81 & 1896 \\
CF & Flowering & DLNM & 19.59 & 0 \\
CF & Flowering & Random effect of year & 17.19 & 1.627 \\
CF & Flowering & None & 7.468 & 381.9 \\
CF & Seedling survival & None & 1 & 0 \\
CF & Seedling survival & Random effect of year & 1.817 & 1.386 \\
CF & Seedling survival & DLNM & 4.008 & 1.528 \\
CF & Seedling growth & Random effect of year & 9.475 & 0 \\
CF & Seedling growth & DLNM & 8.952 & 2.902 \\
CF & Seedling growth & None & 1 & 172.3 \\
FF & Survival & DLNM & 14.95 & 0 \\
FF & Survival & Random effect of year & 19.21 & 35.68 \\
FF & Survival & None & 4.333 & 51.25 \\
FF & Growth & DLNM & 25.18 & 0 \\
FF & Growth & Random effect of year & 37.84 & 200 \\
FF & Growth & None & 5.599 & 382.8 \\
FF & Flowering & DLNM & 20.61 & 0 \\
FF & Flowering & Random effect of year & 13.81 & 27.4 \\
FF & Flowering & None & 5.007 & 101.7 \\
FF & Seedling survival & DLNM & 5.574 & 0 \\
FF & Seedling survival & Random effect of year & 5.088 & 5.721 \\
FF & Seedling survival & None & 1 & 6.491 \\
FF & Seedling growth & Random effect of year & 6.25 & 0 \\
FF & Seedling growth & DLNM & 8.182 & 2.29 \\
FF & Seedling growth & None & 1 & 5.745 \\
\end{longtable}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4306}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1389}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3611}}@{}}
\caption{\label{tab:lambdas} Population growth rates for continuous forest (CF) and forest fragments (FF) under different kinds of IPMs with bootstrapped, bias-corrected, 95\% confidence intervals.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
IPM
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Habitat
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
\(\lambda\)
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
IPM
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Habitat
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
\(\lambda\)
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Deterministic & FF & 0.9778 (0.9736, 0.9823) \\
Deterministic & CF & 0.9897 (0.9877, 0.9920) \\
Stochastic, kernel-resampled & FF & 0.9787 (0.9735, 0.9835) \\
Stochastic, kernel-resampled & CF & 0.9913 (0.9892, 0.9939) \\
dlnm & FF & 0.9595 (0.9459, 0.9689) \\
dlnm & CF & 0.9795 (0.9752, 0.9867) \\
\end{longtable}

\hypertarget{figures}{%
\section{Figures}\label{figures}}

\begin{figure}
\centering
\includegraphics{/Users/emiliobruna/Dropbox (UFL)/Research/Heliconia/Scott_etal_AmNat/lagged-ipms-ms/docs/figures/lifecycle.png}
\caption{\label{fig:lifecycle}Lifecycle diagram of \emph{Heliconia acuminata}. Each transition is associated with an equation for a vital rate function. The functions shown on the diagram correspond to those used to construct a general, density-independent, deterministic IPM. The table below shows the equivalent equations for stochastic, kernel-resampled IPMs and stochastic, parameter-resampled IPMs.}
\end{figure}

\begin{figure}
\centering
\includegraphics{/Users/emiliobruna/Dropbox (UFL)/Research/Heliconia/Scott_etal_AmNat/lagged-ipms-ms/docs/figures/pop_states.png}
\caption{\label{fig:pop-states}Relative proportions of plant sizes in the first 250 iterations of the IPM simulations. Stacked area charts (A) show the relative size/stage distribution of plants in continuous forest (CF, top row) and forest fragments (FF, bottom row) in each of the three IPMs (columns). The proportion of each size class in CF and FF for each iteration is shown in B with the first 30 iterations removed to not include transient dynamics. A 1:1 line is plotted in black. Size categories include seedlings (a discrete category in the IPMs), pre-reproductive 1 (log(size) 0--2.5) that have low average survival (\textless{} 0.9) and a near 0 probability of flowering, pre-reproductive 2 (log(size) 2.5--4.5) that have a higher average survival probabilty (\textgreater{} 0.8) and a near 0 probability of flowering, reproductive 1 (log(size) 4.5--6) that have a high average survival probability (\textgreater0.95) and a lower flowering probability (\textless{} 0.25), and reproductive 2 (log(size) 6+) that have a high average survival probability (\textgreater0.95) and higher flowering proability (\textgreater{} 0.2).}
\end{figure}

\hypertarget{references}{%
\section{References}\label{references}}

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-caswell2001}{}}%
Caswell, H. 2001. Matrix population models: Construction, analysis, and interpretation. Sinauer Associates, Sunderland.

\leavevmode\vadjust pre{\hypertarget{ref-ellnerDatadrivenModellingStructured2016}{}}%
Ellner, S. P., D. Z. Childs, and M. Rees. 2016. Data-driven modelling of structured populations: A practical guide to the integral projection model. {Springer Science+Business Media}, {New York, NY}.

\leavevmode\vadjust pre{\hypertarget{ref-eversLaggedDormantSeason2021}{}}%
Evers, S. M., T. M. Knight, D. W. Inouye, T. E. X. Miller, R. Salguero-Gómez, A. M. Iler, and A. Compagnoni. 2021. \href{https://doi.org/10.1111/gcb.15519}{Lagged and dormant season climate better predict plant vital rates than climate during the growing season}. Global Change Biology 27:1927--1941.

\leavevmode\vadjust pre{\hypertarget{ref-kayeEffectStochasticTechnique2003}{}}%
Kaye, T. N., and D. A. Pyke. 2003. \href{https://doi.org/10.1890/0012-9658(2003)084\%5B1464:TEOSTO\%5D2.0.CO;2}{The effect of stochastic technique on estimates of population viability from transition matrix models}. Ecology 84:1464--1476.

\leavevmode\vadjust pre{\hypertarget{ref-targets}{}}%
Landau, W. M. 2021. \href{https://doi.org/10.21105/joss.02959}{The targets r package: A dynamic make-like function-oriented pipeline toolkit for reproducibility and high-performance computing} 6:2959.

\leavevmode\vadjust pre{\hypertarget{ref-levinIpmrFlexibleImplementation2021}{}}%
Levin, S. C., D. Z. Childs, A. Compagnoni, S. Evers, T. M. Knight, and R. Salguero-Gómez. 2021. \href{https://doi.org/10.1111/2041-210X.13683}{Ipmr: {Flexible} implementation of {Integral Projection Models} in {R}}. Methods in Ecology and Evolution 12:1826--1834.

\leavevmode\vadjust pre{\hypertarget{ref-metcalfStatisticalModellingAnnual2015}{}}%
Metcalf, C. J. E., S. P. Ellner, D. Z. Childs, R. Salguero-Gómez, C. Merow, S. M. McMahon, E. Jongejans, et al. 2015. \href{https://doi.org/10.1111/2041-210X.12405}{Statistical modelling of annual variation for inference on stochastic population dynamics using {Integral Projection Models}}. Methods in Ecology and Evolution 6:1007--1017.

\leavevmode\vadjust pre{\hypertarget{ref-rcoreteam2020}{}}%
R Core Team. 2020. \emph{R: {A} language and environment for statistical computing}. {Vienna, Austria}.

\leavevmode\vadjust pre{\hypertarget{ref-scottDelayedEffectsClimate2022}{}}%
Scott, E. R., M. Uriarte, and E. M. Bruna. 2022. \href{https://doi.org/10.1111/gcb.15900}{Delayed effects of climate on vital rates lead to demographic divergence in Amazonian forest fragments}. Global Change Biology 28:463--479.

\leavevmode\vadjust pre{\hypertarget{ref-mgcv}{}}%
Wood, S. N. 2011. Fast stable restricted maximum likelihood and marginal likelihood estimation of semiparametric generalized linear models 73:3--36.

\leavevmode\vadjust pre{\hypertarget{ref-xavierDailyGriddedMeteorological2016}{}}%
Xavier, A. C., C. W. King, and B. R. Scanlon. 2016. \href{https://doi.org/10.1002/joc.4518}{Daily gridded meteorological variables in {Brazil} (1980\textendash 2013)}. International Journal of Climatology 36:2644--2659.

\end{CSLReferences}

\end{document}

### 2023-07-19 docs/paper.Rmd
removed Renv on startup, fixed render to pdf

@@
 output: 
-    # bookdown::pdf_document2:
-    #   floatsintext: no
-    #   linenumbers: yes
-    #   draft: yes
-    #   mask: no
-    #   figurelist: no
-    #   tablelist: no
-    #   footnotelist: no
-    #   fontsize: 12pt
-    #   toc: false
+    # bookdown::html_document2:
+    #   theme:
+    #     version: 4
+    #     bootswatch: journal
     #   number_sections: no
     #   fig_caption: yes
-    bookdown::html_document2: 
-      theme: 
-        version: 4
-        bootswatch: journal
+    #   pandoc_args:
+    #   - --lua-filter=templates/scholarly-metadata.lua
+    #   - --lua-filter=templates/author-info-blocks.lua
+      # - -Fpandoc-crossref #required for cross-ref equations
+    bookdown::pdf_document2:
+      floatsintext: no
+      linenumbers: yes
+      draft: yes
+      mask: no
+      figurelist: no
+      tablelist: no
+      footnotelist: no
+      fontsize: 12pt
+      toc: false
       number_sections: no
@@
       - --lua-filter=templates/author-info-blocks.lua
-      - -Fpandoc-crossref #required for cross-ref equations
-    bookdown::word_document2:
-      linenumbers: yes
-      number_sections: no
-      fig_caption: yes
-      reference_docx: "templates/template.docx" # Insert path for the DOCX file
-      pandoc_args:
-      - --lua-filter=templates/scholarly-metadata.lua
-      - --lua-filter=templates/author-info-blocks.lua
-      - -Fpandoc-crossref #required for cross-ref equations
+    # bookdown::word_document2:
+    #   linenumbers: yes
+    #   number_sections: no
+    #   fig_caption: yes
+    #   reference_docx: "templates/template.docx" # Insert path for the DOCX file
+    #   pandoc_args:
+    #   - --lua-filter=templates/scholarly-metadata.lua
+    #   - --lua-filter=templates/author-info-blocks.lua
+    #   - -Fpandoc-crossref #required for cross-ref equations
 linkcolor: blue
@@
   - \usepackage{sectsty} \sectionfont{\centering}
-  - \DeclareUnicodeCharacter{U+2206}{$\delta$}
+  - \usepackage[utf8]{inputenc} # This loads package to convert unicode 
+  - \DeclareUnicodeCharacter{394}{$\Delta$} # convert unicode \u0394 to Delta
   # - \usepackage{sectsty} \subsectionfont{\centering}
@@
 
-1.  Demographic models are widely used for all kinds of stuff
+1.  Demographic models, such as IPMs and matrix models, are widely used for all kinds of stuff.
 
 2.  It has long been recognized that there is the potential for lagged effects.
-    Now there's evidence for that [@eversLaggedDormantSeason2021; @scottDelayedEffectsClimate2022].
+    Including lagged effects in models has been a major technical challenge, but there are now multiple approaches for doing so.
+    The studies assessing the potential for lagged effects on vital rates find that they indeed appear to be prevalent [@eversLaggedDormantSeason2021; @scottDelayedEffectsClimate2022].
 
@@
 
-# Tables
+<!-- # Tables -->
 
@@
     "edf" = "df",
-    "∆AIC" = "dAIC"
+    "\u0394AIC" = "dAIC"
   ) %>%
@@
 'edf' is the estimated degrees of freedom of the penalized GAM.
-∆AIC is calculated within each habitat and vital rate combination.
-∆AIC within 2 indicates models are equivalent."
+\u0394AIC is calculated within each habitat and vital rate combination.
+\u0394AIC within 2 indicates models are equivalent."
 )
 ```
+
+\newpage
 
@@
 
-# Figures
+\newpage
 
-```{r lifecycle}
+<!-- # Figures -->
+
+```{r lifecycle,  echo=FALSE, out.width="85%",fig.align='center'}
 #| fig.cap = "Lifecycle diagram of *Heliconia acuminata*.
@@
 
-```{r pop-states}
+\newpage
+
+```{r pop-states,  echo=FALSE, out.width="85%",fig.align='center'}
 #| fig.cap = "Relative proportions of plant sizes in the first 250 iterations of the IPM simulations.
@@
 
+\newpage
+
 # References

### 2023-07-20 docs/ipm_comparison_ms.qmd
adding quarto version of ms

--- 2023-07-19 docs/paper.Rmd
@@
 shorttitle: "Lagged Effects IPMs"  
-author:
-  - Eric R. Scott:
+authors:
+  - name: Eric R. Scott
     email: scottericr@gmail.com
-      institute: [UF_WEC]
-      correspondence: true
-  - Emilio M. Bruna:
-
+    affil-id: 1
+    corresponding: true
+  - name: Emilio M. Bruna
     email: embruna@ufl.edu
-      institute: [UF_WEC, UF_LAS, BDFFP]
-      correspondence: false
-institute:
-  - UF_WEC: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
-  - UF_LAS: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530  
-  - BDFFP: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
-date: "`r format(Sys.time(), '%d %B, %Y')`"
-output: 
-    # bookdown::html_document2:
-    #   theme:
-    #     version: 4
-    #     bootswatch: journal
-    #   number_sections: no
-    #   fig_caption: yes
-    #   pandoc_args:
-    #   - --lua-filter=templates/scholarly-metadata.lua
-    #   - --lua-filter=templates/author-info-blocks.lua
-      # - -Fpandoc-crossref #required for cross-ref equations
-    bookdown::pdf_document2:
-      floatsintext: no
-      linenumbers: yes
-      draft: yes
-      mask: no
-      figurelist: no
-      tablelist: no
-      footnotelist: no
-      fontsize: 12pt
-      toc: false
-      number_sections: no
-      fig_caption: yes
-      pandoc_args:
-      - --lua-filter=templates/scholarly-metadata.lua
-      - --lua-filter=templates/author-info-blocks.lua
-    # bookdown::word_document2:
-    #   linenumbers: yes
-    #   number_sections: no
-    #   fig_caption: yes
-    #   reference_docx: "templates/template.docx" # Insert path for the DOCX file
-    #   pandoc_args:
-    #   - --lua-filter=templates/scholarly-metadata.lua
-    #   - --lua-filter=templates/author-info-blocks.lua
-    #   - -Fpandoc-crossref #required for cross-ref equations
+    affil-id: 1,2,3
+affiliations:
+  - id: "1"
+    name: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
+  - id: "2"
+    name: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530
+  - id: "3"
+    name: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
+date: 'last-modified'
+date-format: '([draft:] D MMMM YYYY)'
 linkcolor: blue
@@
 csl: "templates/the-american-naturalist.csl" # Insert path for the bib-style
-keywords: |
-  integral projection models; environmental stochasticity; lagged effects
-highlights: |
-  These are the highlights
+# 
+# highlights: |
+#   These are the highlights
+format: 
+  pdf:
+    number-sections: false
+    colorlinks: true
+    keeptex: true
+    # the following line is to render the aaffiliations in pdf, which is not default in quarto, see 
+    # https://stackoverflow.com/questions/75040607/why-do-affiliations-not-show-up-anywhere-in-the-pdf-output-of-quarto
+    template-partials:  
+      - title.tex
+    include-in-header:
+      text: |
+        \usepackage[noblocks]{authblk}
+        \renewcommand*{\Authsep}{, }
+        \renewcommand*{\Authand}{, }
+        \renewcommand*{\Authands}{, }
+        \renewcommand\Affilfont{\normalsize}
+        \usepackage{lineno}\linenumbers
+        \usepackage{sectsty}\sectionfont{\centering}
+        \usepackage{sectsty}\subsectionfont{\centering}
 
-editor_options:
-  markdown:
-    wrap: sentence
-    canonical: true
+# When ready to submit, these will change to double space and add a 
+# paragraph indent
+# \usepackage{setspace}\doublespacing
+# \setlength\parindent{22pt}
 
-# header-includes:
-header-includes:
-  - \usepackage[left]{lineno}
-  - \linenumbers
-  - \usepackage{fancyhdr}
-  - \pagestyle{fancy}
-  - \fancyfoot{}
-  - \fancyhead[R]{p. \thepage}
-  - \fancyhead[L]{Draft - `r format(Sys.time(), '%d %B %Y')`}
-  - \usepackage{sectsty} \sectionfont{\centering}
-  - \usepackage[utf8]{inputenc} # This loads package to convert unicode 
-  - \DeclareUnicodeCharacter{394}{$\Delta$} # convert unicode \u0394 to Delta
-  # - \usepackage{sectsty} \subsectionfont{\centering}
-  # - \AtBeginDocument{\let\maketitle\relax}
- #  - \usepackage{float} #to keep figure in the location placed instead of end
- #  - \floatplacement{figure}{H}
- #  - \usepackage{booktabs}
- #  - \usepackage{setspace} # line spacing
- #  - \raggedbottom
- # # - \usepackage{endfloat} #[nomarkers] excludes the {insert figure x around here] from main text.
-The others exclude the list of tables and figures.
-https://cs.brown.edu/about/system/managed/latex/doc/endfloat.pdf
- #  - \usepackage{tabu}
- #  - \usepackage{makecell}
- #  - \usepackage{pdflscape}
- #  - \usepackage{longtable}
- #  - \newcommand{\blandscape}{\begin{landscape}}
- #  - \newcommand{\elandscape}{\end{landscape}}
- #  - \setlength\parindent{22pt}
+fig-cap-location: bottom
+tbl-cap-location: top
+geometry:
+  - top=1in
+  - left=1in
+  - bottom=1in
+  - right=1in
+keywords: 
+  - integral projection models
+  - environmental stochasticity
+  - lagged effects
 ---
@@
 
-Keywords: `r rmarkdown::metadata$keywords`
+<!-- Keywords: `r rmarkdown::metadata$keywords` -->
 
-Running title: `r rmarkdown::metadata$shorttitle`
+<!-- Running title: `r rmarkdown::metadata$shorttitle` -->
 
@@
 library(tidyverse)
+library(kableExtra)
 library(here)
@@
 
-Abstract text.
+Text of 150 words max summarizing this amazing paper.
 
@@
 
-# Introduction
+## Introduction
 
@@
 
-1.  Demographic models, such as IPMs and matrix models, are widely used for all kinds of stuff.
+### **Paragraph 1**
 
-2.  It has long been recognized that there is the potential for lagged effects.
-    Including lagged effects in models has been a major technical challenge, but there are now multiple approaches for doing so.
-    The studies assessing the potential for lagged effects on vital rates find that they indeed appear to be prevalent [@eversLaggedDormantSeason2021; @scottDelayedEffectsClimate2022].
+1.  Structured demographic models, such as Integral Projection Models (IPMs) and matrix models, have emerged as invaluable tools in ecology, evolution, and conservation biology.
 
-3.  Lagged effects could be different in different habitat types (and this could be a big reason for differences between habitats)
+They provide insights into \_\_\_\_, the evolution of life-history strategies, and life-history stages and ecological processes at which conservation intervention is likely to have the greatest impact on population growth rate.
 
-# Methods
+Their flexibility hasmade them increasingly applicable to species with differtent life histories.
 
-## Demographic Data
+however, there are still a number of challenges to address, namely \_\_\_\_ , \_\_\_\_, and \_\_\_\_.
+
+### **Paragraph 2**
+
+1.  It has long been recognized that there is the potential for lagged effects.
+2.  Including lagged effects in models has been a major technical challenge, but there are now multiple approaches for doing so.
+3.  The studies assessing the potential for lagged effects on vital rates find that they indeed appear to be prevalent [@eversLaggedDormantSeason2021; @scottDelayedEffectsClimate2022].
+
+### **Paragraph 3**
+
+1.  Lagged effects could be different in different habitat types.
+2.  This could be a big reason for differences between habitats.
+
+### **Paragraph 4**
+
+1.  Here we...
+2.  We conducted these analyses with both deterministic and stochastic IPMs.
+
+## Methods
+
+### *Demographic methods and data*
 
@@
 
-## IPMs
+Overview of the heliconia project and data
+
+<!-- Add info about additional datasets (Emilio) -->
+
+### *Construction of Integral Projection Models*
 
@@
 
--   Add info about additional datasets (Emilio)
 -->
@@
 
-# Results
+### *Statistical analyses*
 
-For all vital rates estimated using the long term demographic dataset, the DLNM model fit the best (dAIC = 0) followed by the model with a random effect of year, followed by the deterministic model (Table \@ref(tab:aic)).
+All about the stats.
+
+## Results
+
+For all vital rates estimated using the long term demographic dataset, the DLNM model fit the best (dAIC = 0) followed by the model with a random effect of year, followed by the deterministic model (Table \@ref(tab:aic)).\
 <!--# No longer true if random effect = different smooth for each year -->
@@
 The time to iterate the DLNM models is much higher than than deterministic and kernel-resampled.
+
 The greater use of computational resources is likely a result of `predict()` being much slower for GAMs with 2D smooths because the number of knots is much higher compared to the GAMs used for the vital rates models in the determinsitic and kernel-resampled IPMs.
@@
 -->
-
-```{r}
-meta_df <- 
-  tar_meta(store = here("_targets")) %>% 
-  select(name, size, bytes, time, seconds) %>% 
-  filter(str_detect(name, "^ipm_\\w+_[cf]{2}$")) %>% 
-  mutate(minutes = seconds / 60,
-         hours = minutes / 60)
-
-meta_df %>% 
-  separate(name, into = c("trash", "IPM", "Habitat")) %>% 
-  select(-trash) %>% 
-  select(IPM, Habitat, minutes) %>% 
-  group_by(IPM) %>% 
-  summarize(mean_time_min = round(mean(minutes), 2))
-```
 
@@
 
-# Discussion
+## Discussion
 
@@
 
-# Acknowledgments
+## Acknowledgments
 
@@
 
-# CRediT Statement
+## CRediT Statement
 
@@
 
-# Data Availability Statement
+## Data Availability Statement
 
@@
 
+```{r time, results='asis'}
+meta_df <- 
+  tar_meta(store = here("_targets")) %>% 
+  select(name, size, bytes, time, seconds) %>% 
+  filter(str_detect(name, "^ipm_\\w+_[cf]{2}$")) %>% 
+  mutate(minutes = seconds / 60,
+         hours = minutes / 60)
+
+meta_df_tbl<-meta_df %>% 
+  separate(name, into = c("trash", "IPM", "Habitat")) %>% 
+  select(-trash) %>% 
+  select(IPM, Habitat, minutes) %>% 
+  group_by(IPM) %>% 
+mutate(IPM = str_replace_all(
+    IPM,
+    c(
+      "det" = "Deterministic",
+      "stoch" = "Stochastic, kernel-resampled",
+      "dlnm" = "Stochastic, parameter-resampled"
+    ))) %>%
+  summarize(mean_time_min = round(mean(minutes), 2)) %>%
+  rename("IPM Type" = IPM) %>% 
+  rename("mean time (min.)" = mean_time_min)
+
+kbl(meta_df_tbl, 
+    booktabs = T,
+    align = "lc",
+    caption = "table caption") %>% 
+  kable_styling(full_width = F,
+                position = "center")
+
+```
+
+```{=tex}
+\newpage
+\newpage
+```
 ```{r lambdas, results='asis'}
@@
 
-# References
+## Literature Cited
 

### 2023-07-20 docs/ipm_comparison_ms.tex
adding quarto version of ms

--- 2023-07-18 docs/paper.tex
@@
 ]{article}
+
 \usepackage{amsmath,amssymb}
+\usepackage{lmodern}
 \usepackage{iftex}
@@
 \else % if luatex or xetex
-  \usepackage{unicode-math} % this also loads fontspec
+  \usepackage{unicode-math}
   \defaultfontfeatures{Scale=MatchLowercase}
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-\usepackage{lmodern}
-\ifPDFTeX\else
-  % xetex/luatex font selection
 \fi
@@
 \usepackage{xcolor}
-\usepackage[margin=1in]{geometry}
-\usepackage{longtable,booktabs,array}
+\usepackage[top=1in,left=1in,bottom=1in,right=1in]{geometry}
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+% Make \paragraph and \subparagraph free-standing
+\ifx\paragraph\undefined\else
+  \let\oldparagraph\paragraph
+  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+  \let\oldsubparagraph\subparagraph
+  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
 \usepackage{calc} % for calculating minipage widths
@@
 \makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 \newlength{\cslhangindent}
@@
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-\usepackage[left]{lineno}
-\linenumbers
-\usepackage{fancyhdr}
-\pagestyle{fancy}
-\fancyfoot{}
-\fancyhead[R]{p. \thepage}
-\fancyhead[L]{Draft - 18 July 2023}
-\usepackage{sectsty} \sectionfont{\centering}
-\DeclareUnicodeCharacter{U+2206}{$\delta$}
+
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
+\usepackage[noblocks]{authblk}
+\renewcommand*{\Authsep}{, }
+\renewcommand*{\Authand}{, }
+\renewcommand*{\Authands}{, }
+\renewcommand\Affilfont{\normalsize}
+\usepackage{lineno}\linenumbers
+\usepackage{sectsty}\sectionfont{\centering}
+\usepackage{sectsty}\subsectionfont{\centering}
+\makeatletter
+\makeatother
+\makeatletter
+\makeatother
+\makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\AtBeginDocument{%
+\ifdefined\contentsname
+  \renewcommand*\contentsname{Table of contents}
+\else
+  \newcommand\contentsname{Table of contents}
+\fi
+\ifdefined\listfigurename
+  \renewcommand*\listfigurename{List of Figures}
+\else
+  \newcommand\listfigurename{List of Figures}
+\fi
+\ifdefined\listtablename
+  \renewcommand*\listtablename{List of Tables}
+\else
+  \newcommand\listtablename{List of Tables}
+\fi
+\ifdefined\figurename
+  \renewcommand*\figurename{Figure}
+\else
+  \newcommand\figurename{Figure}
+\fi
+\ifdefined\tablename
+  \renewcommand*\tablename{Table}
+\else
+  \newcommand\tablename{Table}
+\fi
+}
+\@ifpackageloaded{float}{}{\usepackage{float}}
+\floatstyle{ruled}
+\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
+\floatname{codelisting}{Listing}
+\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
+\makeatother
+\makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
+\makeatother
+\makeatletter
+\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
+\makeatother
+\makeatletter
+\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
+\makeatother
+\makeatletter
+\makeatother
 \ifLuaTeX
@@
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same}
+\urlstyle{same} % disable monospaced font for URLs
 \hypersetup{
   pdftitle={Context-dependent consequences of including lagged effects in demographic models},
-  pdfkeywords={integral projection models; environmental stochasticity; lagged effects},
+  pdfkeywords={integral projection models, environmental
+stochasticity, lagged effects},
   colorlinks=true,
@@
 
-\title{Context-dependent consequences of including lagged effects in demographic models}
-\author{true \and true}
-\date{18 July, 2023}
+\title{Context-dependent consequences of including lagged effects in
+demographic models}
 
+\author[1]{Eric R. Scott}
+\author[1,2,3]{Emilio M. Bruna}
+
+\affil[1]{Department of Wildlife Ecology and Conservation, University of
+Florida, Gainesville, Florida 32611-0430}
+\affil[2]{Center for Latin American Studies, University of Florida,
+Gainesville, Florida 32611-5530}
+\affil[3]{Biological Dynamics of Forest Fragments Project, INPA-PDBFF,
+CP 478, Manaus, Amazonas 69011-970 Brazil}
+
+\date{(draft: 20 July 2023)}
 \begin{document}
 \maketitle
-
-Keywords: integral projection models; environmental stochasticity; lagged effects
-
-Running title: Lagged Effects IPMs
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners, frame hidden, breakable]}{\end{tcolorbox}}\fi
 
@@
 \hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\subsection{Introduction}\label{introduction}}
+
+\hypertarget{paragraph-1}{%
+\subsubsection{\texorpdfstring{\textbf{Paragraph
+1}}{Paragraph 1}}\label{paragraph-1}}
 
@@
 \def\labelenumi{\arabic{enumi}.}
+\tightlist
 \item
-  Demographic models are widely used for all kinds of stuff
+  Structured demographic models, such as Integral Projection Models
+  (IPMs) and matrix models, have emerged as invaluable tools in ecology,
+  evolution, and conservation biology.
+\end{enumerate}
+
+They provide insights into \_\_\_\_, the evolution of life-history
+strategies, and life-history stages and ecological processes at which
+conservation intervention is likely to have the greatest impact on
+population growth rate.
+
+Their flexibility hasmade them increasingly applicable to species with
+differtent life histories.
+
+however, there are still a number of challenges to address, namely
+\_\_\_\_ , \_\_\_\_, and \_\_\_\_.
+
+\hypertarget{paragraph-2}{%
+\subsubsection{\texorpdfstring{\textbf{Paragraph
+2}}{Paragraph 2}}\label{paragraph-2}}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
 \item
-  It has long been recognized that there is the potential for lagged effects.
-  Now there's evidence for that (Evers et al. 2021; Scott et al. 2022).
+  It has long been recognized that there is the potential for lagged
+  effects.
 \item
-  Lagged effects could be different in different habitat types (and this could be a big reason for differences between habitats)
+  Including lagged effects in models has been a major technical
+  challenge, but there are now multiple approaches for doing so.
+\item
+  The studies assessing the potential for lagged effects on vital rates
+  find that they indeed appear to be prevalent (Evers et al. 2021; Scott
+  et al. 2022).
+\end{enumerate}
+
+\hypertarget{paragraph-3}{%
+\subsubsection{\texorpdfstring{\textbf{Paragraph
+3}}{Paragraph 3}}\label{paragraph-3}}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Lagged effects could be different in different habitat types.
+\item
+  This could be a big reason for differences between habitats.
+\end{enumerate}
+
+\hypertarget{paragraph-4}{%
+\subsubsection{\texorpdfstring{\textbf{Paragraph
+4}}{Paragraph 4}}\label{paragraph-4}}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Here we\ldots{}
+\item
+  We conducted these analyses with both deterministic and stochastic
+  IPMs.
 \end{enumerate}
@@
 \hypertarget{methods}{%
-\section{Methods}\label{methods}}
+\subsection{Methods}\label{methods}}
 
-\hypertarget{demographic-data}{%
-\subsection{Demographic Data}\label{demographic-data}}
+\hypertarget{demographic-methods-and-data}{%
+\subsubsection{\texorpdfstring{\emph{Demographic methods and
+data}}{Demographic methods and data}}\label{demographic-methods-and-data}}
 
-\hypertarget{ipms}{%
-\subsection{IPMs}\label{ipms}}
+Overview of the heliconia project and data
 
-In preliminary investigation we found that the survival and growth of plants was better explained by treating seedlings and mature plants separately.
-Seedlings are physiologically different from small plants because they necessarily lack the underground reserves (of carbohydrates and meristems) that a small, mature plant may have.
-Therefore, we used general IPMs to model population dynamics with seedlings treated as a separate discreet class not structured by size.
-General IPMs allow for combinations of continuous and discrete states and transitions between them (Ellner et al. 2016).
+\hypertarget{construction-of-integral-projection-models}{%
+\subsubsection{\texorpdfstring{\emph{Construction of Integral Projection
+Models}}{Construction of Integral Projection Models}}\label{construction-of-integral-projection-models}}
 
-We built three classes of IPMs for comparison which each required different functional forms of their underlying vital rates models.
-The simplest IPM was a general, density-independent, deterministic IPM with four sub-kernels: growth and survival (\(P\), (\textbf{eq:P?})), fecundity (\(F\), i.e.~production of new seedlings, (\textbf{eq:F?})), probability of staying a seedling (always 0), and recruitment (\(R\), i.e.~seedling survival and establishment, (\textbf{eq:R?})) ((\textbf{eq:mature?}; \textbf{eq:sdlg?}), Figure \ref{fig:lifecycle}).
-The probability of staying a seedling, was always equal to zero, since our definition of seedlings was first year plants only.
+In preliminary investigation we found that the survival and growth of
+plants was better explained by treating seedlings and mature plants
+separately. Seedlings are physiologically different from small plants
+because they necessarily lack the underground reserves (of carbohydrates
+and meristems) that a small, mature plant may have. Therefore, we used
+general IPMs to model population dynamics with seedlings treated as a
+separate discreet class not structured by size. General IPMs allow for
+combinations of continuous and discrete states and transitions between
+them (Ellner et al. 2016).
+
+We built three classes of IPMs for comparison which each required
+different functional forms of their underlying vital rates models. The
+simplest IPM was a general, density-independent, deterministic IPM with
+four sub-kernels: growth and survival (\(P\), (\textbf{eq:P?})),
+fecundity (\(F\), i.e.~production of new seedlings, (\textbf{eq:F?})),
+probability of staying a seedling (always 0), and recruitment (\(R\),
+i.e.~seedling survival and establishment, (\textbf{eq:R?}))
+((\textbf{eq:mature?}; \textbf{eq:sdlg?}), Figure @ref(fig:lifecycle)).
+The probability of staying a seedling, was always equal to zero, since
+our definition of seedlings was first year plants only.
 
@@
 
-The number and size of mature plants in the next census is determined by seedlings entering the mature plant population (i.e.~recruitment ) and survival and growth (or regression) of mature plants (\textbf{eq:mature?}).
-Seedlings survive (\(s_s\)) and grow into mature plants of a particular size (\(G_s(z^\prime)\)) (\textbf{eq:R?}).
-Mature plants survive as a function of size (\(s(z)\)), and grow (or regress) to a new size as a function of their previous size (\(G(z^\prime, z)\)) (\textbf{eq:P?}).
-Mature plants flower with a probability that is a function of size (\(p_f(z)\)) and produce a number of seeds as a function of size (\(f(z)\)), which germinate and establish as seedlings with probability \(g\) (\textbf{eq:F?}).
+The number and size of mature plants in the next census is determined by
+seedlings entering the mature plant population (i.e.~recruitment ) and
+survival and growth (or regression) of mature plants
+(\textbf{eq:mature?}). Seedlings survive (\(s_s\)) and grow into mature
+plants of a particular size (\(G_s(z^\prime)\)) (\textbf{eq:R?}). Mature
+plants survive as a function of size (\(s(z)\)), and grow (or regress)
+to a new size as a function of their previous size (\(G(z^\prime, z)\))
+(\textbf{eq:P?}). Mature plants flower with a probability that is a
+function of size (\(p_f(z)\)) and produce a number of seeds as a
+function of size (\(f(z)\)), which germinate and establish as seedlings
+with probability \(g\) (\textbf{eq:F?}).
 
-Vital rate models for growth (\(G_s(z^\prime)\) and \(G(z^\prime, z)\)) , survival (\(s_s\) and \(s(z)\)), and flowering (\(p_f(z)\)) were fit using the long term demographic dataset.
-For established plants, these three vital rates were modeled as a smooth function of size in the previous census using generalized additive models (GAMs) fit with the \texttt{mgcv} package(Wood 2011) in R version 4.3.0 (2023-04-21) (R Core Team 2020) .
-For consistency, seedling survival and growth were also modeled using GAMs, but without size in the previous census as a predictor (i.e.~intercept only models).
-For growth models (\(G_s(z^\prime)\) and \(G(z^\prime, z)\)) a scaled t family distribution provided a better fit to the data than a gaussian fit as the residuals were leptokurtic with a simple gaussian model.
+Vital rate models for growth (\(G_s(z^\prime)\) and \(G(z^\prime, z)\))
+, survival (\(s_s\) and \(s(z)\)), and flowering (\(p_f(z)\)) were fit
+using the long term demographic dataset. For established plants, these
+three vital rates were modeled as a smooth function of size in the
+previous census using generalized additive models (GAMs) fit with the
+\texttt{mgcv} package(Wood 2011) in R version 4.3.0 (2023-04-21) (R Core
+Team 2020) . For consistency, seedling survival and growth were also
+modeled using GAMs, but without size in the previous census as a
+predictor (i.e.~intercept only models). For growth models
+(\(G_s(z^\prime)\) and \(G(z^\prime, z)\)) a scaled t family
+distribution provided a better fit to the data than a gaussian fit as
+the residuals were leptokurtic with a simple gaussian model.
 
-To estimate reproduction we drew on additional data sources to estimate the number of fruits per flowering plant as a function of plant size and the number of seeds per fruit (together \(f(z)\)).
-Germination and establishment rates in continuous forest and forest fragments were estimated using data from \ldots.
+To estimate reproduction we drew on additional data sources to estimate
+the number of fruits per flowering plant as a function of plant size and
+the number of seeds per fruit (together \(f(z)\)). Germination and
+establishment rates in continuous forest and forest fragments were
+estimated using data from \ldots.
 
-To build the general, density-independent, stochastic, kernel-resampled IPMs, we included environmental stochasticity in all vital rate models built using the long term demographic datset by adding a random effect of year (Figure \ref{fig:lifecycle}).
-The random effect of year was included using a factor--smooth interaction which allowed the relationship between plant size and vital rates to vary in functional form among transition years.
-The kernel-resampling approach is to generate kernels corresponding to each transition year in the demographic dataset using the random smooths for year, and to iterate the IPM by drawing from these randomly.
-This is equivalent to the matrix selection approach for matrix population models described by Caswell (2001).
+To build the general, density-independent, stochastic, kernel-resampled
+IPMs, we included environmental stochasticity in all vital rate models
+built using the long term demographic datset by adding a random effect
+of year (Figure @ref(fig:lifecycle)). The random effect of year was
+included using a factor--smooth interaction which allowed the
+relationship between plant size and vital rates to vary in functional
+form among transition years. The kernel-resampling approach is to
+generate kernels corresponding to each transition year in the
+demographic dataset using the random smooths for year, and to iterate
+the IPM by drawing from these randomly. This is equivalent to the matrix
+selection approach for matrix population models described by Caswell
+(2001).
 
-For the third method, we modeled the impacts of drought on vital rates explicitly and created general, density-independent, stochastic, parameter-resampled IPMs (\emph{sensu} Metcalf et al. (2015)).
-We calculated the standardized precipitation evapotranspiraton index (SPEI) for our site using a published gridded dataset based on ground measurements (Xavier et al.
-2016) as described in Scott et al.
-(2022).
-For all vital rate models fit using the long term demographic dataset, we modeled delayed effects of SPEI using distributed lag non-linear models with a maximum lag of 36 months(Scott et al.
-2022) (Figure \ref{fig:lifecycle}).
-To iterate these parameter-resampled IPMs, a random sequence of SPEI values was created by sampling years of the observed monthly SPEI data.
-Then, 36 month lags are calculated for each year starting in February (the month of the demographic census).
-These values are then used to predict fitted values from the vital rates models, generating different kernels at each iteration of the IPM.
-With this method, the kernels of successive iterations are not entirely independent because the SPEI values used in calculating vital rates include values used in the previous two iterations, but they are ergodic.
+For the third method, we modeled the impacts of drought on vital rates
+explicitly and created general, density-independent, stochastic,
+parameter-resampled IPMs (\emph{sensu} Metcalf et al. (2015)). We
+calculated the standardized precipitation evapotranspiraton index (SPEI)
+for our site using a published gridded dataset based on ground
+measurements (Xavier et al. 2016) as described in Scott et al. (2022).
+For all vital rate models fit using the long term demographic dataset,
+we modeled delayed effects of SPEI using distributed lag non-linear
+models with a maximum lag of 36 months(Scott et al. 2022) (Figure
+@ref(fig:lifecycle)). To iterate these parameter-resampled IPMs, a
+random sequence of SPEI values was created by sampling years of the
+observed monthly SPEI data. Then, 36 month lags are calculated for each
+year starting in February (the month of the demographic census). These
+values are then used to predict fitted values from the vital rates
+models, generating different kernels at each iteration of the IPM. With
+this method, the kernels of successive iterations are not entirely
+independent because the SPEI values used in calculating vital rates
+include values used in the previous two iterations, but they are
+ergodic.
 
-All IPMs were constructed and iterated using the \texttt{ipmr} package in R (Levin et al. 2021).
-The IPMs used 100 meshpoints and the midpoint rule for calculating kernels .
-For each type of IPM we iterated the model for 1000 time steps, discarding the first 100 time steps to omit transient effects.
-Stochastic growth rates (\(\lambda_s\)) were calculated as the average \(ln(\lambda)\) from each time step (Caswell 2001) and back-transformed to be on the same scale as deterministic lambdas for comparison.
-We used the distribution of established plant sizes and proportion of seedlings from the full dataset as a starting population vector.
-While other starting population vectors were possible, the choice is of little importance as it will only impact transient dynamics, which we aren't interested in for this study.
+All IPMs were constructed and iterated using the \texttt{ipmr} package
+in R (Levin et al. 2021). The IPMs used 100 meshpoints and the midpoint
+rule for calculating kernels . For each type of IPM we iterated the
+model for 1000 time steps, discarding the first 100 time steps to omit
+transient effects. Stochastic growth rates (\(\lambda_s\)) were
+calculated as the average \(ln(\lambda)\) from each time step (Caswell
+2001) and back-transformed to be on the same scale as deterministic
+lambdas for comparison. We used the distribution of established plant
+sizes and proportion of seedlings from the full dataset as a starting
+population vector. While other starting population vectors were
+possible, the choice is of little importance as it will only impact
+transient dynamics, which we aren't interested in for this study.
 
-To estimate uncertainty around the per-capita growth rates (lambdas), we created 500 bootstraps of the demographic dataset by sampling individual plants with replacement within each habitat.
-For each bootstrap, we then re-fit vital rates models (all except germination and establishment rate, fruits per flowering plant, and seeds per fruit, which were estimated using different datasets), constructed IPMs, and calculated a value for lambda as described above.
-We then used these bootstraped estimates of lambda to calculate bias corrected 95\% confidence intervals (Ellner et al. 2016).
+To estimate uncertainty around the per-capita growth rates (lambdas), we
+created 500 bootstraps of the demographic dataset by sampling individual
+plants with replacement within each habitat. For each bootstrap, we then
+re-fit vital rates models (all except germination and establishment
+rate, fruits per flowering plant, and seeds per fruit, which were
+estimated using different datasets), constructed IPMs, and calculated a
+value for lambda as described above. We then used these bootstraped
+estimates of lambda to calculate bias corrected 95\% confidence
+intervals (Ellner et al. 2016).
 
-This workflow was managed using the \texttt{targets} R package (Landau 2021) which also allowed us to track computational time spent on each IPM for comparison.
+This workflow was managed using the \texttt{targets} R package (Landau
+2021) which also allowed us to track computational time spent on each
+IPM for comparison.
+
+\hypertarget{statistical-analyses}{%
+\subsubsection{\texorpdfstring{\emph{Statistical
+analyses}}{Statistical analyses}}\label{statistical-analyses}}
+
+All about the stats.
 
 \hypertarget{results}{%
-\section{Results}\label{results}}
+\subsection{Results}\label{results}}
 
-For all vital rates estimated using the long term demographic dataset, the DLNM model fit the best (dAIC = 0) followed by the model with a random effect of year, followed by the deterministic model (Table \ref{tab:aic}).
+For all vital rates estimated using the long term demographic dataset,
+the DLNM model fit the best (dAIC = 0) followed by the model with a
+random effect of year, followed by the deterministic model (Table
+@ref(tab:aic)).\\
 
-Population growth rates were consistently higher in continuous forest compared to forest fragments across IPM types (Table \ref{tab:lambdas}).
+Population growth rates were consistently higher in continuous forest
+compared to forest fragments across IPM types (Table @ref(tab:lambdas)).
 
-The time to iterate the DLNM models is much higher than than deterministic and kernel-resampled.
-The greater use of computational resources is likely a result of \texttt{predict()} being much slower for GAMs with 2D smooths because the number of knots is much higher compared to the GAMs used for the vital rates models in the determinsitic and kernel-resampled IPMs.
+The time to iterate the DLNM models is much higher than than
+deterministic and kernel-resampled.
 
-\begin{verbatim}
-#> # A tibble: 3 x 2
-#>   IPM   mean_time_min
-#>   <chr>         <dbl>
-#> 1 det            0.02
-#> 2 dlnm          87.1 
-#> 3 stoch          0.07
-\end{verbatim}
+The greater use of computational resources is likely a result of
+\texttt{predict()} being much slower for GAMs with 2D smooths because
+the number of knots is much higher compared to the GAMs used for the
+vital rates models in the determinsitic and kernel-resampled IPMs.
 
-Figure \ref{fig:pop-states} has some interesting things in it:
+Figure @ref(fig:pop-states) has some interesting things in it:
 
@@
 \item
-  for the deterministic IPM (and the kernel-resampled IPM?) there are slightly more of the smallest plants and the largest plants in CF compared to FF (i.e.~more medium sized plants in FF).
+  for the deterministic IPM (and the kernel-resampled IPM?) there are
+  slightly more of the smallest plants and the largest plants in CF
+  compared to FF (i.e.~more medium sized plants in FF).
 \item
-  For the kernel-resampled IPM (random effect of year), the fluctuations are extremely similar between CF and FF
+  For the kernel-resampled IPM (random effect of year), the fluctuations
+  are extremely similar between CF and FF
 \item
-  For the parameter-resampled IPM (DLNM) the size structure of the population is a LOT more variable in FF.
-  This makes sense as we know lagged effects are more important in fragments.
+  For the parameter-resampled IPM (DLNM) the size structure of the
+  population is a LOT more variable in FF. This makes sense as we know
+  lagged effects are more important in fragments.
 \item
-  Also, the fluctuations in size structure in CF do not match the fluctuations in FF as well (can see this by the increased spread of points in Figure \ref{fig:pop-states}B)
+  Also, the fluctuations in size structure in CF do not match the
+  fluctuations in FF as well (can see this by the increased spread of
+  points in Figure @ref(fig:pop-states)B)
 \item
-  Also, in the parameter-resampled IPM (and only in this one), we see a shift toward smaller plants in FF compared to CF
+  Also, in the parameter-resampled IPM (and only in this one), we see a
+  shift toward smaller plants in FF compared to CF
 \end{itemize}
@@
 \hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
+\subsection{Discussion}\label{discussion}}
 
@@
 \item
-  Our finding that the choice of IPM didn't change the relative ranking of CF and FF is consistent with Kaye and Pyke (2003) finding that method effected stochastic lambda, but relative ranking of populations was consistent.
+  Our finding that the choice of IPM didn't change the relative ranking
+  of CF and FF is consistent with Kaye and Pyke (2003) finding that
+  method effected stochastic lambda, but relative ranking of populations
+  was consistent.
 \end{itemize}
@@
 \hypertarget{acknowledgments}{%
-\section{Acknowledgments}\label{acknowledgments}}
+\subsection{Acknowledgments}\label{acknowledgments}}
 
-We thank \textbf{,} \_, \_\_\_ and \_\_\_ anonymous reviewers for helpful discussions and comments on the manuscript.
-We thank Sam Levin for his help with the \texttt{ipmr} package.
-Financial support was provided by the U.S.
-National Science Foundation (awards \_\_\_\_, and \_\_\_\_).
-This article is publication no. -- -- in the BDFFP Technical series.
-The authors declare no conflicts of interest.
+We thank \textbf{,} \_, \_\_\_ and \_\_\_ anonymous reviewers for
+helpful discussions and comments on the manuscript. We thank Sam Levin
+for his help with the \texttt{ipmr} package. Financial support was
+provided by the U.S. National Science Foundation (awards \_\_\_\_, and
+\_\_\_\_). This article is publication no. -- -- in the BDFFP Technical
+series. The authors declare no conflicts of interest.
 
 \hypertarget{credit-statement}{%
-\section{CRediT Statement}\label{credit-statement}}
+\subsection{CRediT Statement}\label{credit-statement}}
 
-ERS contributed to the conceptualization, methodology, formal analysis, and led the writing of the original draft.
-EMB contributed to the conceptualization, methodology, and writing and also acquired funding.
+ERS contributed to the conceptualization, methodology, formal analysis,
+and led the writing of the original draft. EMB contributed to the
+conceptualization, methodology, and writing and also acquired funding.
 
 \hypertarget{data-availability-statement}{%
-\section{Data Availability Statement}\label{data-availability-statement}}
+\subsection{Data Availability
+Statement}\label{data-availability-statement}}
 
@@
 \pagebreak
-
-\hypertarget{tables}{%
-\section{Tables}\label{tables}}
 
@@
   >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1111}}@{}}
-\caption{\label{tab:aic} Comparison of vital rate models used to build IPM.
-The `Effect of Environment' column describes how environmental effects were included in models.
-Those with `none' were used to build deterministic IPMs; those with a random effect of year were used to build stochastic, kernel-resampled IPMs; and those with a distributed lag non-linear model (DLNM) were used to build stochastic, parameter-resampled IPMs.
-`edf' is the estimated degrees of freedom of the penalized GAM.
-∆AIC is calculated within each habitat and vital rate combination.
-∆AIC within 2 indicates models are equivalent.}\tabularnewline
-\toprule\noalign{}
+\caption{(\#tab:aic) Comparison of vital rate models used to build IPM.
+The `Effect of Environment' column describes how environmental effects
+were included in models. Those with `none' were used to build
+deterministic IPMs; those with a random effect of year were used to
+build stochastic, kernel-resampled IPMs; and those with a distributed
+lag non-linear model (DLNM) were used to build stochastic,
+parameter-resampled IPMs. `edf' is the estimated degrees of freedom of
+the penalized GAM. ΔAIC is calculated within each habitat and vital rate
+combination. ΔAIC within 2 indicates models are
+equivalent.}\tabularnewline
+\toprule()
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-∆AIC
+ΔAIC
 \end{minipage} \\
-\midrule\noalign{}
+\midrule()
 \endfirsthead
-\toprule\noalign{}
+\toprule()
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-∆AIC
+ΔAIC
 \end{minipage} \\
-\midrule\noalign{}
+\midrule()
 \endhead
-\bottomrule\noalign{}
-\endlastfoot
 CF & Survival & Random effect of year & 43.26 & 0 \\
@@
 FF & Seedling growth & None & 1 & 5.745 \\
+\bottomrule()
 \end{longtable}
+
+\newpage
+
+\begin{table}
+
+\caption{table caption}
+\centering
+\begin{tabular}[t]{lc}
+\toprule
+IPM Type & mean time (min.)\\
+\midrule
+Deterministic & 0.02\\
+Stochastic, kernel-resampled & 0.07\\
+Stochastic, parameter-resampled & 87.12\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\newpage
+\newpage
 
@@
   >{\raggedleft\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3611}}@{}}
-\caption{\label{tab:lambdas} Population growth rates for continuous forest (CF) and forest fragments (FF) under different kinds of IPMs with bootstrapped, bias-corrected, 95\% confidence intervals.}\tabularnewline
-\toprule\noalign{}
+\caption{(\#tab:lambdas) Population growth rates for continuous forest
+(CF) and forest fragments (FF) under different kinds of IPMs with
+bootstrapped, bias-corrected, 95\% confidence intervals.}\tabularnewline
+\toprule()
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} \\
-\midrule\noalign{}
+\midrule()
 \endfirsthead
-\toprule\noalign{}
+\toprule()
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} \\
-\midrule\noalign{}
+\midrule()
 \endhead
-\bottomrule\noalign{}
-\endlastfoot
 Deterministic & FF & 0.9778 (0.9736, 0.9823) \\
@@
 dlnm & CF & 0.9795 (0.9752, 0.9867) \\
+\bottomrule()
 \end{longtable}
 
-\hypertarget{figures}{%
-\section{Figures}\label{figures}}
+\newpage
 
 \begin{figure}
-\centering
-\includegraphics{/Users/emiliobruna/Dropbox (UFL)/Research/Heliconia/Scott_etal_AmNat/lagged-ipms-ms/docs/figures/lifecycle.png}
-\caption{\label{fig:lifecycle}Lifecycle diagram of \emph{Heliconia acuminata}.
-Each transition is associated with an equation for a vital rate function.
-The functions shown on the diagram correspond to those used to construct a general, density-independent, deterministic IPM.
-The table below shows the equivalent equations for stochastic, kernel-resampled IPMs and stochastic, parameter-resampled IPMs.}
+
+{\centering \includegraphics[width=0.85\textwidth,height=\textheight]{/Users/emiliobruna/Dropbox (UFL)/Research/Heliconia/Scott_etal_AmNat/lagged-ipms-ms/docs/figures/lifecycle.png}
+
+}
+
+\caption{Lifecycle diagram of \emph{Heliconia acuminata}. Each
+transition is associated with an equation for a vital rate function. The
+functions shown on the diagram correspond to those used to construct a
+general, density-independent, deterministic IPM. The table below shows
+the equivalent equations for stochastic, kernel-resampled IPMs and
+stochastic, parameter-resampled IPMs.}
+
 \end{figure}
 
+\newpage
+
 \begin{figure}
-\centering
-\includegraphics{/Users/emiliobruna/Dropbox (UFL)/Research/Heliconia/Scott_etal_AmNat/lagged-ipms-ms/docs/figures/pop_states.png}
-\caption{\label{fig:pop-states}Relative proportions of plant sizes in the first 250 iterations of the IPM simulations.
-Stacked area charts (A) show the relative size/stage distribution of plants in continuous forest (CF, top row) and forest fragments (FF, bottom row) in each of the three IPMs (columns).
-The proportion of each size class in CF and FF for each iteration is shown in B with the first 30 iterations removed to not include transient dynamics.
-A 1:1 line is plotted in black.
-Size categories include seedlings (a discrete category in the IPMs), pre-reproductive 1 (log(size) 0--2.5) that have low average survival (\textless{} 0.9) and a near 0 probability of flowering, pre-reproductive 2 (log(size) 2.5--4.5) that have a higher average survival probabilty (\textgreater{} 0.8) and a near 0 probability of flowering, reproductive 1 (log(size) 4.5--6) that have a high average survival probability (\textgreater0.95) and a lower flowering probability (\textless{} 0.25), and reproductive 2 (log(size) 6+) that have a high average survival probability (\textgreater0.95) and higher flowering proability (\textgreater{} 0.2).}
+
+{\centering \includegraphics[width=0.85\textwidth,height=\textheight]{/Users/emiliobruna/Dropbox (UFL)/Research/Heliconia/Scott_etal_AmNat/lagged-ipms-ms/docs/figures/pop_states.png}
+
+}
+
+\caption{Relative proportions of plant sizes in the first 250 iterations
+of the IPM simulations. Stacked area charts (A) show the relative
+size/stage distribution of plants in continuous forest (CF, top row) and
+forest fragments (FF, bottom row) in each of the three IPMs (columns).
+The proportion of each size class in CF and FF for each iteration is
+shown in B with the first 30 iterations removed to not include transient
+dynamics. A 1:1 line is plotted in black. Size categories include
+seedlings (a discrete category in the IPMs), pre-reproductive 1
+(log(size) 0--2.5) that have low average survival (\textless{} 0.9) and
+a near 0 probability of flowering, pre-reproductive 2 (log(size)
+2.5--4.5) that have a higher average survival probabilty (\textgreater{}
+0.8) and a near 0 probability of flowering, reproductive 1 (log(size)
+4.5--6) that have a high average survival probability (\textgreater0.95)
+and a lower flowering probability (\textless{} 0.25), and reproductive 2
+(log(size) 6+) that have a high average survival probability
+(\textgreater0.95) and higher flowering proability (\textgreater{}
+0.2).}
+
 \end{figure}
 
-\hypertarget{references}{%
-\section{References}\label{references}}
+\newpage
+
+\hypertarget{literature-cited}{%
+\subsection{Literature Cited}\label{literature-cited}}
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-caswell2001}{}}%
-Caswell, H. 2001. Matrix population models: Construction, analysis, and interpretation. Sinauer Associates, Sunderland.
+Caswell, H. 2001. Matrix population models: Construction, analysis, and
+interpretation. Sinauer Associates, Sunderland.
 
 \leavevmode\vadjust pre{\hypertarget{ref-ellnerDatadrivenModellingStructured2016}{}}%
-Ellner, S. P., D. Z. Childs, and M. Rees. 2016. Data-driven modelling of structured populations: A practical guide to the integral projection model. {Springer Science+Business Media}, {New York, NY}.
+Ellner, S. P., D. Z. Childs, and M. Rees. 2016. Data-driven modelling of
+structured populations: A practical guide to the integral projection
+model. {Springer Science+Business Media}, {New York, NY}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-eversLaggedDormantSeason2021}{}}%
-Evers, S.
-M., T.
-M.
-Knight, D.
-W.
-Inouye, T.
-E.
-X.
-Miller, R.
-Salguero-Gómez, A.
-M.
-Iler, and A.
-Compagnoni.
-2021.
-\href{https://doi.org/10.1111/gcb.15519}{Lagged and dormant season climate better predict plant vital rates than climate during the growing season}.
-Global Change Biology 27:1927--1941.
+Evers, S. M., T. M. Knight, D. W. Inouye, T. E. X. Miller, R.
+Salguero-Gómez, A. M. Iler, and A. Compagnoni. 2021.
+\href{https://doi.org/10.1111/gcb.15519}{Lagged and dormant season
+climate better predict plant vital rates than climate during the growing
+season}. Global Change Biology 27:1927--1941.
 
 \leavevmode\vadjust pre{\hypertarget{ref-kayeEffectStochasticTechnique2003}{}}%
-Kaye, T.
-N., and D.
-A.
-Pyke.
-2003.
-\href{https://doi.org/10.1890/0012-9658(2003)084\%5B1464:TEOSTO\%5D2.0.CO;2}{The effect of stochastic technique on estimates of population viability from transition matrix models}.
-Ecology 84:1464--1476.
+Kaye, T. N., and D. A. Pyke. 2003.
+\href{https://doi.org/10.1890/0012-9658(2003)084\%5B1464:TEOSTO\%5D2.0.CO;2}{The
+effect of stochastic technique on estimates of population viability from
+transition matrix models}. Ecology 84:1464--1476.
 
 \leavevmode\vadjust pre{\hypertarget{ref-targets}{}}%
-Landau, W. M. 2021. \href{https://doi.org/10.21105/joss.02959}{The targets r package: A dynamic make-like function-oriented pipeline toolkit for reproducibility and high-performance computing} 6:2959.
+Landau, W. M. 2021. \href{https://doi.org/10.21105/joss.02959}{The
+targets r package: A dynamic make-like function-oriented pipeline
+toolkit for reproducibility and high-performance computing} 6:2959.
 
 \leavevmode\vadjust pre{\hypertarget{ref-levinIpmrFlexibleImplementation2021}{}}%
-Levin, S.
-C., D.
-Z.
-Childs, A.
-Compagnoni, S.
-Evers, T.
-M.
-Knight, and R.
-Salguero-Gómez.
-2021.
-\href{https://doi.org/10.1111/2041-210X.13683}{Ipmr: {Flexible} implementation of {Integral Projection Models} in {R}}.
-Methods in Ecology and Evolution 12:1826--1834.
+Levin, S. C., D. Z. Childs, A. Compagnoni, S. Evers, T. M. Knight, and
+R. Salguero-Gómez. 2021.
+\href{https://doi.org/10.1111/2041-210X.13683}{Ipmr: {Flexible}
+implementation of {Integral Projection Models} in {R}}. Methods in
+Ecology and Evolution 12:1826--1834.
 
 \leavevmode\vadjust pre{\hypertarget{ref-metcalfStatisticalModellingAnnual2015}{}}%
-Metcalf, C.
-J.
-E., S.
-P.
-Ellner, D.
-Z.
-Childs, R.
-Salguero-Gómez, C.
-Merow, S.
-M.
-McMahon, E.
-Jongejans, et al.
-2015.
-\href{https://doi.org/10.1111/2041-210X.12405}{Statistical modelling of annual variation for inference on stochastic population dynamics using {Integral Projection Models}}.
-Methods in Ecology and Evolution 6:1007--1017.
+Metcalf, C. J. E., S. P. Ellner, D. Z. Childs, R. Salguero-Gómez, C.
+Merow, S. M. McMahon, E. Jongejans, et al. 2015.
+\href{https://doi.org/10.1111/2041-210X.12405}{Statistical modelling of
+annual variation for inference on stochastic population dynamics using
+{Integral Projection Models}}. Methods in Ecology and Evolution
+6:1007--1017.
 
 \leavevmode\vadjust pre{\hypertarget{ref-rcoreteam2020}{}}%
-R Core Team. 2020. \emph{R: {A} language and environment for statistical computing}. {Vienna, Austria}.
+R Core Team. 2020. \emph{R: {A} language and environment for statistical
+computing}. {Vienna, Austria}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-scottDelayedEffectsClimate2022}{}}%
-Scott, E.
-R., M.
-Uriarte, and E.
-M.
-Bruna.
-2022.
-\href{https://doi.org/10.1111/gcb.15900}{Delayed effects of climate on vital rates lead to demographic divergence in Amazonian forest fragments}.
-Global Change Biology 28:463--479.
+Scott, E. R., M. Uriarte, and E. M. Bruna. 2022.
+\href{https://doi.org/10.1111/gcb.15900}{Delayed effects of climate on
+vital rates lead to demographic divergence in Amazonian forest
+fragments}. Global Change Biology 28:463--479.
 
 \leavevmode\vadjust pre{\hypertarget{ref-mgcv}{}}%
-Wood, S. N. 2011. Fast stable restricted maximum likelihood and marginal likelihood estimation of semiparametric generalized linear models 73:3--36.
+Wood, S. N. 2011. Fast stable restricted maximum likelihood and marginal
+likelihood estimation of semiparametric generalized linear models
+73:3--36.
 
 \leavevmode\vadjust pre{\hypertarget{ref-xavierDailyGriddedMeteorological2016}{}}%
-Xavier, A.
-C., C.
-W.
-King, and B.
-R.
-Scanlon.
-2016.
-\href{https://doi.org/10.1002/joc.4518}{Daily gridded meteorological variables in {Brazil} (1980\textendash 2013)}.
-International Journal of Climatology 36:2644--2659.
+Xavier, A. C., C. W. King, and B. R. Scanlon. 2016.
+\href{https://doi.org/10.1002/joc.4518}{Daily gridded meteorological
+variables in {Brazil} (1980\textendash 2013)}. International Journal of
+Climatology 36:2644--2659.
 

### 2023-07-20 docs/rmd_version/paper.Rmd
adding quarto version of ms

--- 2023-07-19 docs/paper.Rmd
@@
   - \DeclareUnicodeCharacter{394}{$\Delta$} # convert unicode \u0394 to Delta
+  # more info # https://www.ctan.org/pkg/sectsty
+  - \usepackage{sectsty} \sectionfont{\centering}
+  # to center to sub-sub-headings (###) and change color of font
+  - \usepackage{sectsty} \subsectionfont{\centering}
+  # - \usepackage{sectsty} \subsubsectionfont{\centering}
   # - \usepackage{sectsty} \subsectionfont{\centering}
@@
 
+1.  **Paragraph 1**
+
     1.  Demographic models, such as IPMs and matrix models, are widely used for all kinds of stuff.
 
-2.  It has long been recognized that there is the potential for lagged effects.
-    Including lagged effects in models has been a major technical challenge, but there are now multiple approaches for doing so.
-    The studies assessing the potential for lagged effects on vital rates find that they indeed appear to be prevalent [@eversLaggedDormantSeason2021; @scottDelayedEffectsClimate2022].
+2.  **Paragraph 2**
 
-3.  Lagged effects could be different in different habitat types (and this could be a big reason for differences between habitats)
+    1.  It has long been recognized that there is the potential for lagged effects.\
+    2.  Including lagged effects in models has been a major technical challenge, but there are now multiple approaches for doing so.\
+    3.  The studies assessing the potential for lagged effects on vital rates find that they indeed appear to be prevalent [@eversLaggedDormantSeason2021; @scottDelayedEffectsClimate2022].
 
-# Methods
+3.  **Paragraph 3**
 
-## Demographic Data
+    1.  Lagged effects could be different in different habitat types.\
+    2.  This could be a big reason for differences between habitats.
 
-<!--# Emilio will fill this section -->
+4.  **Paragraph 4**
 
-## IPMs
+    1.  Here we...
+    2.  We conducted these analyses with both deterministic and stochastic IPMs.
+
+## Methods
+
+### Demographic methods and data
+
+Overviw of <!--# Emilio will fill this section -->
+
+<!-- Add info about additional datasets (Emilio) -->
+
+### Construction of Integral Projection Models
 
@@
 
-# Results
+### Statistical analyses
 
-For all vital rates estimated using the long term demographic dataset, the DLNM model fit the best (dAIC = 0) followed by the model with a random effect of year, followed by the deterministic model (Table \@ref(tab:aic)).
+All about the stats.
+
+## Results
+
+For all vital rates estimated using the long term demographic dataset, the DLNM model fit the best (dAIC = 0) followed by the model with a random effect of year, followed by the deterministic model (Table \@ref(tab:aic)).\
 <!--# No longer true if random effect = different smooth for each year -->
@@
 
+```{r time, results='asis'}
+meta_df <- 
+  tar_meta(store = here("_targets")) %>% 
+  select(name, size, bytes, time, seconds) %>% 
+  filter(str_detect(name, "^ipm_\\w+_[cf]{2}$")) %>% 
+  mutate(minutes = seconds / 60,
+         hours = minutes / 60)
+
+meta_df_tbl<-meta_df %>% 
+  separate(name, into = c("trash", "IPM", "Habitat")) %>% 
+  select(-trash) %>% 
+  select(IPM, Habitat, minutes) %>% 
+  group_by(IPM) %>% 
+mutate(IPM = str_replace_all(
+    IPM,
+    c(
+      "det" = "Deterministic",
+      "stoch" = "Stochastic, kernel-resampled",
+      "dlnm" = "Stochastic, parameter-resampled"
+    ))) %>%
+  summarize(mean_time_min = round(mean(minutes), 2)) %>%
+  rename("IPM Type" = IPM) %>% 
+  rename("mean time (min.)" = mean_time_min)
+
+kbl(meta_df_tbl, 
+    booktabs = T,
+    align = "lc",
+    caption = "table caption") %>% 
+  kable_styling(full_width = F,
+                position = "center")
+
+```
+
+\newpage
+
 ```{r lambdas, results='asis'}

### 2023-07-21 docs/ipm_comparison_ms.qmd
update MS formatting and cross refs

@@
     corresponding: true
+  - name: María Uriarte
+    email: mu2126@columbia.edu
+    institute: 2
   - name: Emilio M. Bruna
     email: embruna@ufl.edu
-    affil-id: 1,2,3
+    affil-id: 1,3,5
 affiliations:
   - id: "1"
-    name: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
+    name: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430 USA
   - id: "2" 
-    name: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530
+    name: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
   - id: "3"
+    name: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530 USA
+  - id: "4"
     name: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
@@
     keeptex: true
+    fontsize: 12pt
+    documentclass: article
     # the following line is to render the aaffiliations in pdf, which is not default in quarto, see 
@@
         \usepackage{sectsty}\subsectionfont{\centering}
+        \usepackage{fontspec}
+        \usepackage{titlesec}
+
+#\sectionfont{\color{black}\fontsize{12}{16.8}\selectfont}
+#\subsectionfont{\color{black}\fontsize{12}{16.8}\selectfont}
+
+# number after font size is the interline spacing. It should about 20-25 % more 
+# than the font size (for a 10 pt document, you can measure it is equal to 12 pt).
+# \titleformat{\subsection}
+# \titleformat*{\subsection}{\fontsize{12{16.8}\selectfont
+        #     {\sffamily\Large\bfseries}{\thesection}{1em}{}[{\titlerule[0.8pt]}]
 
@@
 We built three classes of IPMs for comparison which each required different functional forms of their underlying vital rates models.
-The simplest IPM was a general, density-independent, deterministic IPM with four sub-kernels: growth and survival ($P$, @eq:P), fecundity ($F$, i.e.
-production of new seedlings, @eq:F), probability of staying a seedling (always 0), and recruitment ($R$, i.e.
-seedling survival and establishment, @eq:R) ([@eq:mature; @eq:sdlg], Figure \@ref(fig:lifecycle)).
+The simplest IPM was a general, density-independent, deterministic IPM with four sub-kernels: growth and survival ($P$, @eq-P), fecundity ($F$, i.e.
+production of new seedlings, @eq-F), probability of staying a seedling (always 0), and recruitment ($R$, i.e.
+seedling survival and establishment, @eq-R) ([@eq-mature; @eq-sdlg], @fig-lifecycle).
 The probability of staying a seedling, was always equal to zero, since our definition of seedlings was first year plants only.
@@
 n(z^{\prime},t+1) = R(z^\prime)n_s(t) + \int_L^U P(z^\prime,z) n(z,t)\;dz
-$${#eq:mature}
+$$ {#eq-mature}
 
@@
 n_s(t+1) = \int_L^U F(z) n(z,t) \; dz
-$${#eq:sdlg}
+$$ {#eq-sdlg}
 
@@
 R(z^\prime) = s_sG_s(z^\prime)
-$${#eq:R}
+$$ {#eq-R}
 
@@
 P(z^\prime, z) = s(z)G(z^\prime, z)
-$${#eq:P}
+$$ {#eq-P}
 
@@
 F(z) = p_f(z)f(z)g
-$${#eq:F}
+$$ {#eq-F}
 
@@
 recruitment <!--# *Is* this recruitment?
--->) and survival and growth (or regression) of mature plants [@eq:mature].
-Seedlings survive ($s_s$) and grow into mature plants of a particular size ($G_s(z^\prime)$) [@eq:R].
-Mature plants survive as a function of size ($s(z)$), and grow (or regress) to a new size as a function of their previous size ($G(z^\prime, z)$) [@eq:P].
-Mature plants flower with a probability that is a function of size ($p_f(z)$) and produce a number of seeds as a function of size ($f(z)$), which germinate and establish as seedlings with probability $g$ [@eq:F].
+-->) and survival and growth (or regression) of mature plants ([@eq-mature]).
+Seedlings survive ($s_s$) and grow into mature plants of a particular size ($G_s(z^\prime)$) ([@eq-R]).
+Mature plants survive as a function of size ($s(z)$), and grow (or regress) to a new size as a function of their previous size ($G(z^\prime, z)$) ([@eq-P]).
+Mature plants flower with a probability that is a function of size ($p_f(z)$) and produce a number of seeds as a function of size ($f(z)$), which germinate and establish as seedlings with probability $g$ ([@eq-F]).
 
@@
 
-To build the general, density-independent, stochastic, kernel-resampled IPMs, we included environmental stochasticity in all vital rate models built using the long term demographic datset by adding a random effect of year (Figure \@ref(fig:lifecycle)).
+To build the general, density-independent, stochastic, kernel-resampled IPMs, we included environmental stochasticity in all vital rate models built using the long term demographic datset by adding a random effect of year (@fig-lifecycle).
 The random effect of year was included using a factor--smooth interaction which allowed the relationship between plant size and vital rates to vary in functional form among transition years.
@@
 We calculated the standardized precipitation evapotranspiraton index (SPEI) for our site using a published gridded dataset based on ground measurements [@xavierDailyGriddedMeteorological2016] as described in @scottDelayedEffectsClimate2022.
-For all vital rate models fit using the long term demographic dataset, we modeled delayed effects of SPEI using distributed lag non-linear models with a maximum lag of 36 months[@scottDelayedEffectsClimate2022] (Figure \@ref(fig:lifecycle)).
+For all vital rate models fit using the long term demographic dataset, we modeled delayed effects of SPEI using distributed lag non-linear models with a maximum lag of 36 months [@scottDelayedEffectsClimate2022] (@fig-lifecycle).
 To iterate these parameter-resampled IPMs, a random sequence of SPEI values was created by sampling years of the observed monthly SPEI data.
@@
 
-For all vital rates estimated using the long term demographic dataset, the DLNM model fit the best (dAIC = 0) followed by the model with a random effect of year, followed by the deterministic model (Table \@ref(tab:aic)).\
+For all vital rates estimated using the long term demographic dataset, the DLNM model fit the best (dAIC = 0) followed by the model with a random effect of year, followed by the deterministic model (@tbl-aic).\
 <!--# No longer true if random effect = different smooth for each year -->
 
-Population growth rates were consistently higher in continuous forest compared to forest fragments across IPM types (Table \@ref(tab:lambdas)).
+Population growth rates were consistently higher in continuous forest compared to forest fragments across IPM types (@tbl-lambdas).
 
@@
 
-Figure \@ref(fig:pop-states) has some interesting things in it:
+@fig-pop-states has some interesting things in it:
 
@@
 
--   Also, the fluctuations in size structure in CF do not match the fluctuations in FF as well (can see this by the increased spread of points in Figure \@ref(fig:pop-states)B)
+-   Also, the fluctuations in size structure in CF do not match the fluctuations in FF as well (can see this by the increased spread of points in @fig-pop-states (B)
 
@@
 
+## Literature Cited
+
+<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
+
+::: {#refs}
+:::
+
 \pagebreak
@@
 ```{r aic, results='asis'}
+#| label: tbl-aic
+#| tbl-cap: "Comparison of vital rate models used to build IPM.
+The 'Effect of Environment' column describes how environmental effects were included in models.
+Those with 'none' were used to build deterministic IPMs; those with a random effect of year were used to build stochastic, kernel-resampled IPMs; and those with a distributed lag non-linear model (DLNM) were used to build stochastic, parameter-resampled IPMs.
+'edf' is the estimated degrees of freedom of the penalized GAM.
+\u0394AIC is calculated within each habitat and vital rate combination.
+\u0394AIC within 2 indicates models are equivalent."
+#| 
 #TODO: only show first instance of habitat and vital rate columns, not on every row.  Alternatively do something fancy with flextable or gt.
@@
   justify = c( "left", "left","left", "right",  "right"),
-  full_width = FALSE,
-  caption = "(\\#tab:aic) Comparison of vital rate models used to build IPM.
+  full_width = FALSE
+  # ,
+  # caption = Comparison of vital rate models used to build IPM.
 The 'Effect of Environment' column describes how environmental effects were included in models.
@@
 
-```{r time, results='asis'}
+```{r tbl-time, results='asis'}
+#| label: tbl-time
+#| tbl-cap: "Figure caption to be written"
 meta_df <- 
@@
     booktabs = T,
-    align = "lc",
-    caption = "table caption") %>% 
+    align = "lc") %>% 
   kable_styling(full_width = F,
@@
 
-```{=tex}
 \newpage
-\newpage
-```
-```{r lambdas, results='asis'}
+
+```{r tbl-lambdas, results='asis'}
+#| label: tbl-lambdas
+#| tbl-cap: "Population growth rates for continuous forest (CF) and forest fragments (FF) under different kinds of IPMs with bootstrapped, bias-corrected, 95% confidence intervals."
+
 lambda_table %>% 
@@
   justify = c("left", "left", "right"),
-  full_width = FALSE,
-  caption = "(\\#tab:lambdas) Population growth rates for continuous forest (CF) and forest fragments (FF) under different kinds of IPMs with bootstrapped, bias-corrected, 95% confidence intervals."
+  full_width = FALSE
+  # ,
+  # caption = "(\\#tab-lambdas) Population growth rates for continuous forest (CF) and forest fragments (FF) under different kinds of IPMs with bootstrapped, bias-corrected, 95% confidence intervals."
 )
@@
 
-```{r lifecycle,  echo=FALSE, out.width="85%",fig.align='center'}
+```{r #fig-lifecycle,  echo=FALSE, out.width="85%",fig.align='center'}
 #| fig.cap = "Lifecycle diagram of *Heliconia acuminata*.
@@
 
-```{r pop-states,  echo=FALSE, out.width="85%",fig.align='center'}
+```{r #fig-pop-states,  echo=FALSE, out.width="85%",fig.align='center'}
 #| fig.cap = "Relative proportions of plant sizes in the first 250 iterations of the IPM simulations.
@@
 knitr::include_graphics(here(fig_pop_states), rel_path = FALSE)
-
 ```
-
-\newpage
-
-## Literature Cited
-
-<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
-
-::: {#refs}
-:::

### 2023-07-21 docs/ipm_comparison_ms.tex
update MS formatting and cross refs

@@
 \documentclass[
+  12pt,
 ]{article}
@@
 \usepackage{sectsty}\subsectionfont{\centering}
+\usepackage{fontspec}
+\usepackage{titlesec}
 \makeatletter
@@
 \author[1]{Eric R. Scott}
-\author[1,2,3]{Emilio M. Bruna}
+\author{María Uriarte}
+\author[1,3,5]{Emilio M. Bruna}
 
 \affil[1]{Department of Wildlife Ecology and Conservation, University of
-Florida, Gainesville, Florida 32611-0430}
-\affil[2]{Center for Latin American Studies, University of Florida,
-Gainesville, Florida 32611-5530}
-\affil[3]{Biological Dynamics of Forest Fragments Project, INPA-PDBFF,
+Florida, Gainesville, Florida 32611-0430 USA}
+\affil[2]{Department of Ecology, Evolution and Environmental Biology,
+Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA}
+\affil[3]{Center for Latin American Studies, University of Florida,
+Gainesville, Florida 32611-5530 USA}
+\affil[4]{Biological Dynamics of Forest Fragments Project, INPA-PDBFF,
 CP 478, Manaus, Amazonas 69011-970 Brazil}
 
-\date{(draft: 20 July 2023)}
+\date{(draft: 21 July 2023)}
 \begin{document}
 \maketitle
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners, frame hidden, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable, sharp corners, boxrule=0pt, interior hidden, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 simplest IPM was a general, density-independent, deterministic IPM with
-four sub-kernels: growth and survival (\(P\), (\textbf{eq:P?})),
-fecundity (\(F\), i.e.~production of new seedlings, (\textbf{eq:F?})),
-probability of staying a seedling (always 0), and recruitment (\(R\),
-i.e.~seedling survival and establishment, (\textbf{eq:R?}))
-((\textbf{eq:mature?}; \textbf{eq:sdlg?}), Figure @ref(fig:lifecycle)).
-The probability of staying a seedling, was always equal to zero, since
-our definition of seedlings was first year plants only.
+four sub-kernels: growth and survival (\(P\), Equation~\ref{eq-P}),
+fecundity (\(F\), i.e.~production of new seedlings,
+Equation~\ref{eq-F}), probability of staying a seedling (always 0), and
+recruitment (\(R\), i.e.~seedling survival and establishment,
+Equation~\ref{eq-R}) (Equation~\ref{eq-mature}, Equation~\ref{eq-sdlg},
+Figure~\ref{fig-lifecycle}). The probability of staying a seedling, was
+always equal to zero, since our definition of seedlings was first year
+plants only.
 
-\[
+\begin{equation}\protect\hypertarget{eq-mature}{}{
 n(z^{\prime},t+1) = R(z^\prime)n_s(t) + \int_L^U P(z^\prime,z) n(z,t)\;dz
-\]\{\#eq:mature\}
+}\label{eq-mature}\end{equation}
 
-\[
+\begin{equation}\protect\hypertarget{eq-sdlg}{}{
 n_s(t+1) = \int_L^U F(z) n(z,t) \; dz
-\]\{\#eq:sdlg\}
+}\label{eq-sdlg}\end{equation}
 
-\[
+\begin{equation}\protect\hypertarget{eq-R}{}{
 R(z^\prime) = s_sG_s(z^\prime)
-\]\{\#eq:R\}
+}\label{eq-R}\end{equation}
 
-\[
+\begin{equation}\protect\hypertarget{eq-P}{}{
 P(z^\prime, z) = s(z)G(z^\prime, z)
-\]\{\#eq:P\}
+}\label{eq-P}\end{equation}
 
-\[
+\begin{equation}\protect\hypertarget{eq-F}{}{
 F(z) = p_f(z)f(z)g
-\]\{\#eq:F\}
+}\label{eq-F}\end{equation}
 
@@
 survival and growth (or regression) of mature plants
-(\textbf{eq:mature?}). Seedlings survive (\(s_s\)) and grow into mature
-plants of a particular size (\(G_s(z^\prime)\)) (\textbf{eq:R?}). Mature
-plants survive as a function of size (\(s(z)\)), and grow (or regress)
-to a new size as a function of their previous size (\(G(z^\prime, z)\))
-(\textbf{eq:P?}). Mature plants flower with a probability that is a
-function of size (\(p_f(z)\)) and produce a number of seeds as a
-function of size (\(f(z)\)), which germinate and establish as seedlings
-with probability \(g\) (\textbf{eq:F?}).
+(Equation~\ref{eq-mature}). Seedlings survive (\(s_s\)) and grow into
+mature plants of a particular size (\(G_s(z^\prime)\))
+(Equation~\ref{eq-R}). Mature plants survive as a function of size
+(\(s(z)\)), and grow (or regress) to a new size as a function of their
+previous size (\(G(z^\prime, z)\)) (Equation~\ref{eq-P}). Mature plants
+flower with a probability that is a function of size (\(p_f(z)\)) and
+produce a number of seeds as a function of size (\(f(z)\)), which
+germinate and establish as seedlings with probability \(g\)
+(Equation~\ref{eq-F}).
 
@@
 built using the long term demographic datset by adding a random effect
-of year (Figure @ref(fig:lifecycle)). The random effect of year was
+of year (Figure~\ref{fig-lifecycle}). The random effect of year was
 included using a factor--smooth interaction which allowed the
@@
 we modeled delayed effects of SPEI using distributed lag non-linear
-models with a maximum lag of 36 months(Scott et al. 2022) (Figure
-@ref(fig:lifecycle)). To iterate these parameter-resampled IPMs, a
-random sequence of SPEI values was created by sampling years of the
+models with a maximum lag of 36 months (Scott et al. 2022)
+(Figure~\ref{fig-lifecycle}). To iterate these parameter-resampled IPMs,
+a random sequence of SPEI values was created by sampling years of the
 observed monthly SPEI data. Then, 36 month lags are calculated for each
@@
 the DLNM model fit the best (dAIC = 0) followed by the model with a
-random effect of year, followed by the deterministic model (Table
-@ref(tab:aic)).\\
+random effect of year, followed by the deterministic model
+(Table~\ref{tbl-aic}).\\
 
 Population growth rates were consistently higher in continuous forest
-compared to forest fragments across IPM types (Table @ref(tab:lambdas)).
+compared to forest fragments across IPM types (Table~\ref{tbl-lambdas}).
 
@@
 
-Figure @ref(fig:pop-states) has some interesting things in it:
+Figure~\ref{fig-pop-states} has some interesting things in it:
 
@@
   fluctuations in FF as well (can see this by the increased spread of
-  points in Figure @ref(fig:pop-states)B)
+  points in Figure~\ref{fig-pop-states} (B)
 \item
@@
 Data and R code used in this study are archived with Zenodo at .
-
-\pagebreak
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2778}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3333}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1111}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1111}}@{}}
-\caption{(\#tab:aic) Comparison of vital rate models used to build IPM.
-The `Effect of Environment' column describes how environmental effects
-were included in models. Those with `none' were used to build
-deterministic IPMs; those with a random effect of year were used to
-build stochastic, kernel-resampled IPMs; and those with a distributed
-lag non-linear model (DLNM) were used to build stochastic,
-parameter-resampled IPMs. `edf' is the estimated degrees of freedom of
-the penalized GAM. ΔAIC is calculated within each habitat and vital rate
-combination. ΔAIC within 2 indicates models are
-equivalent.}\tabularnewline
-\toprule()
-\begin{minipage}[b]{\linewidth}\raggedright
-Habitat
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Vital Rate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Effect of Environment
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-edf
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-ΔAIC
-\end{minipage} \\
-\midrule()
-\endfirsthead
-\toprule()
-\begin{minipage}[b]{\linewidth}\raggedright
-Habitat
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Vital Rate
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Effect of Environment
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-edf
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-ΔAIC
-\end{minipage} \\
-\midrule()
-\endhead
-CF & Survival & Random effect of year & 43.26 & 0 \\
-CF & Survival & DLNM & 19.72 & 78.92 \\
-CF & Survival & None & 4.976 & 260 \\
-CF & Growth & Random effect of year & 78.43 & 0 \\
-CF & Growth & DLNM & 23.87 & 158.5 \\
-CF & Growth & None & 7.81 & 1896 \\
-CF & Flowering & DLNM & 19.59 & 0 \\
-CF & Flowering & Random effect of year & 17.19 & 1.627 \\
-CF & Flowering & None & 7.468 & 381.9 \\
-CF & Seedling survival & None & 1 & 0 \\
-CF & Seedling survival & Random effect of year & 1.817 & 1.386 \\
-CF & Seedling survival & DLNM & 4.008 & 1.528 \\
-CF & Seedling growth & Random effect of year & 9.475 & 0 \\
-CF & Seedling growth & DLNM & 8.952 & 2.902 \\
-CF & Seedling growth & None & 1 & 172.3 \\
-FF & Survival & DLNM & 14.95 & 0 \\
-FF & Survival & Random effect of year & 19.21 & 35.68 \\
-FF & Survival & None & 4.333 & 51.25 \\
-FF & Growth & DLNM & 25.18 & 0 \\
-FF & Growth & Random effect of year & 37.84 & 200 \\
-FF & Growth & None & 5.599 & 382.8 \\
-FF & Flowering & DLNM & 20.61 & 0 \\
-FF & Flowering & Random effect of year & 13.81 & 27.4 \\
-FF & Flowering & None & 5.007 & 101.7 \\
-FF & Seedling survival & DLNM & 5.574 & 0 \\
-FF & Seedling survival & Random effect of year & 5.088 & 5.721 \\
-FF & Seedling survival & None & 1 & 6.491 \\
-FF & Seedling growth & Random effect of year & 6.25 & 0 \\
-FF & Seedling growth & DLNM & 8.182 & 2.29 \\
-FF & Seedling growth & None & 1 & 5.745 \\
-\bottomrule()
-\end{longtable}
-
-\newpage
-
-\begin{table}
-
-\caption{table caption}
-\centering
-\begin{tabular}[t]{lc}
-\toprule
-IPM Type & mean time (min.)\\
-\midrule
-Deterministic & 0.02\\
-Stochastic, kernel-resampled & 0.07\\
-Stochastic, parameter-resampled & 87.12\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\newpage
-\newpage
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4306}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1389}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3611}}@{}}
-\caption{(\#tab:lambdas) Population growth rates for continuous forest
-(CF) and forest fragments (FF) under different kinds of IPMs with
-bootstrapped, bias-corrected, 95\% confidence intervals.}\tabularnewline
-\toprule()
-\begin{minipage}[b]{\linewidth}\raggedright
-IPM
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Habitat
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-\(\lambda\)
-\end{minipage} \\
-\midrule()
-\endfirsthead
-\toprule()
-\begin{minipage}[b]{\linewidth}\raggedright
-IPM
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Habitat
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-\(\lambda\)
-\end{minipage} \\
-\midrule()
-\endhead
-Deterministic & FF & 0.9778 (0.9736, 0.9823) \\
-Deterministic & CF & 0.9897 (0.9877, 0.9920) \\
-Stochastic, kernel-resampled & FF & 0.9787 (0.9735, 0.9835) \\
-Stochastic, kernel-resampled & CF & 0.9913 (0.9892, 0.9939) \\
-dlnm & FF & 0.9595 (0.9459, 0.9689) \\
-dlnm & CF & 0.9795 (0.9752, 0.9867) \\
-\bottomrule()
-\end{longtable}
-
-\newpage
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.85\textwidth,height=\textheight]{/Users/emiliobruna/Dropbox (UFL)/Research/Heliconia/Scott_etal_AmNat/lagged-ipms-ms/docs/figures/lifecycle.png}
-
-}
-
-\caption{Lifecycle diagram of \emph{Heliconia acuminata}. Each
-transition is associated with an equation for a vital rate function. The
-functions shown on the diagram correspond to those used to construct a
-general, density-independent, deterministic IPM. The table below shows
-the equivalent equations for stochastic, kernel-resampled IPMs and
-stochastic, parameter-resampled IPMs.}
-
-\end{figure}
-
-\newpage
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.85\textwidth,height=\textheight]{/Users/emiliobruna/Dropbox (UFL)/Research/Heliconia/Scott_etal_AmNat/lagged-ipms-ms/docs/figures/pop_states.png}
-
-}
-
-\caption{Relative proportions of plant sizes in the first 250 iterations
-of the IPM simulations. Stacked area charts (A) show the relative
-size/stage distribution of plants in continuous forest (CF, top row) and
-forest fragments (FF, bottom row) in each of the three IPMs (columns).
-The proportion of each size class in CF and FF for each iteration is
-shown in B with the first 30 iterations removed to not include transient
-dynamics. A 1:1 line is plotted in black. Size categories include
-seedlings (a discrete category in the IPMs), pre-reproductive 1
-(log(size) 0--2.5) that have low average survival (\textless{} 0.9) and
-a near 0 probability of flowering, pre-reproductive 2 (log(size)
-2.5--4.5) that have a higher average survival probabilty (\textgreater{}
-0.8) and a near 0 probability of flowering, reproductive 1 (log(size)
-4.5--6) that have a high average survival probability (\textgreater0.95)
-and a lower flowering probability (\textless{} 0.25), and reproductive 2
-(log(size) 6+) that have a high average survival probability
-(\textgreater0.95) and higher flowering proability (\textgreater{}
-0.2).}
-
-\end{figure}
-
-\newpage
 
@@
 
+\pagebreak
+
+\hypertarget{tbl-aic}{}
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2778}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3333}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1111}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1111}}@{}}
+\caption{\label{tbl-aic}Comparison of vital rate models used to build
+IPM. The `Effect of Environment' column describes how environmental
+effects were included in models. Those with `none' were used to build
+deterministic IPMs; those with a random effect of year were used to
+build stochastic, kernel-resampled IPMs; and those with a distributed
+lag non-linear model (DLNM) were used to build stochastic,
+parameter-resampled IPMs. `edf' is the estimated degrees of freedom of
+the penalized GAM. ΔAIC is calculated within each habitat and vital rate
+combination. ΔAIC within 2 indicates models are
+equivalent.}\tabularnewline
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+Habitat
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Vital Rate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Effect of Environment
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+edf
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+ΔAIC
+\end{minipage} \\
+\midrule()
+\endfirsthead
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+Habitat
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Vital Rate
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Effect of Environment
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+edf
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+ΔAIC
+\end{minipage} \\
+\midrule()
+\endhead
+CF & Survival & Random effect of year & 43.26 & 0 \\
+CF & Survival & DLNM & 19.72 & 78.92 \\
+CF & Survival & None & 4.976 & 260 \\
+CF & Growth & Random effect of year & 78.43 & 0 \\
+CF & Growth & DLNM & 23.87 & 158.5 \\
+CF & Growth & None & 7.81 & 1896 \\
+CF & Flowering & DLNM & 19.59 & 0 \\
+CF & Flowering & Random effect of year & 17.19 & 1.627 \\
+CF & Flowering & None & 7.468 & 381.9 \\
+CF & Seedling survival & None & 1 & 0 \\
+CF & Seedling survival & Random effect of year & 1.817 & 1.386 \\
+CF & Seedling survival & DLNM & 4.008 & 1.528 \\
+CF & Seedling growth & Random effect of year & 9.475 & 0 \\
+CF & Seedling growth & DLNM & 8.952 & 2.902 \\
+CF & Seedling growth & None & 1 & 172.3 \\
+FF & Survival & DLNM & 14.95 & 0 \\
+FF & Survival & Random effect of year & 19.21 & 35.68 \\
+FF & Survival & None & 4.333 & 51.25 \\
+FF & Growth & DLNM & 25.18 & 0 \\
+FF & Growth & Random effect of year & 37.84 & 200 \\
+FF & Growth & None & 5.599 & 382.8 \\
+FF & Flowering & DLNM & 20.61 & 0 \\
+FF & Flowering & Random effect of year & 13.81 & 27.4 \\
+FF & Flowering & None & 5.007 & 101.7 \\
+FF & Seedling survival & DLNM & 5.574 & 0 \\
+FF & Seedling survival & Random effect of year & 5.088 & 5.721 \\
+FF & Seedling survival & None & 1 & 6.491 \\
+FF & Seedling growth & Random effect of year & 6.25 & 0 \\
+FF & Seedling growth & DLNM & 8.182 & 2.29 \\
+FF & Seedling growth & None & 1 & 5.745 \\
+\bottomrule()
+\end{longtable}
+
+\newpage
+
+\hypertarget{tbl-time}{}
+\begin{table}
+\caption{\label{tbl-time}Figure caption to be written }\tabularnewline
+
+\centering
+\begin{tabular}[t]{lc}
+\toprule
+IPM Type & mean time (min.)\\
+\midrule
+Deterministic & 0.02\\
+Stochastic, kernel-resampled & 0.07\\
+Stochastic, parameter-resampled & 87.12\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\newpage
+
+\hypertarget{tbl-lambdas}{}
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4306}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1389}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3611}}@{}}
+\caption{\label{tbl-lambdas}Population growth rates for continuous
+forest (CF) and forest fragments (FF) under different kinds of IPMs with
+bootstrapped, bias-corrected, 95\% confidence intervals.}\tabularnewline
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+IPM
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Habitat
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+\(\lambda\)
+\end{minipage} \\
+\midrule()
+\endfirsthead
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+IPM
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Habitat
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+\(\lambda\)
+\end{minipage} \\
+\midrule()
+\endhead
+Deterministic & FF & 0.9778 (0.9736, 0.9823) \\
+Deterministic & CF & 0.9897 (0.9877, 0.9920) \\
+Stochastic, kernel-resampled & FF & 0.9787 (0.9735, 0.9835) \\
+Stochastic, kernel-resampled & CF & 0.9913 (0.9892, 0.9939) \\
+dlnm & FF & 0.9595 (0.9459, 0.9689) \\
+dlnm & CF & 0.9795 (0.9752, 0.9867) \\
+\bottomrule()
+\end{longtable}
+
+\newpage
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.85\textwidth,height=\textheight]{/Users/emiliobruna/Dropbox (UFL)/Research/Heliconia/Scott_etal_AmNat/lagged-ipms-ms/docs/figures/lifecycle.png}
+
+}
+
+\caption{\label{fig-lifecycle}Lifecycle diagram of \emph{Heliconia
+acuminata}. Each transition is associated with an equation for a vital
+rate function. The functions shown on the diagram correspond to those
+used to construct a general, density-independent, deterministic IPM. The
+table below shows the equivalent equations for stochastic,
+kernel-resampled IPMs and stochastic, parameter-resampled IPMs.}
+
+\end{figure}
+
+\newpage
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.85\textwidth,height=\textheight]{/Users/emiliobruna/Dropbox (UFL)/Research/Heliconia/Scott_etal_AmNat/lagged-ipms-ms/docs/figures/pop_states.png}
+
+}
+
+\caption{\label{fig-pop-states}Relative proportions of plant sizes in
+the first 250 iterations of the IPM simulations. Stacked area charts (A)
+show the relative size/stage distribution of plants in continuous forest
+(CF, top row) and forest fragments (FF, bottom row) in each of the three
+IPMs (columns). The proportion of each size class in CF and FF for each
+iteration is shown in B with the first 30 iterations removed to not
+include transient dynamics. A 1:1 line is plotted in black. Size
+categories include seedlings (a discrete category in the IPMs),
+pre-reproductive 1 (log(size) 0--2.5) that have low average survival
+(\textless{} 0.9) and a near 0 probability of flowering,
+pre-reproductive 2 (log(size) 2.5--4.5) that have a higher average
+survival probabilty (\textgreater{} 0.8) and a near 0 probability of
+flowering, reproductive 1 (log(size) 4.5--6) that have a high average
+survival probability (\textgreater0.95) and a lower flowering
+probability (\textless{} 0.25), and reproductive 2 (log(size) 6+) that
+have a high average survival probability (\textgreater0.95) and higher
+flowering proability (\textgreater{} 0.2).}
+
+\end{figure}
+
 \end{document}

### 2024-09-03 docs/ipm_comparison_ms.qmd
back to it

@@
 ---
-title: "Context-dependent consequences of including lagged effects in demographic models"
+title: "Context-dependent consequences of lagged effects in demographic models"
 shorttitle: "Lagged Effects IPMs"  
@@
     email: scottericr@gmail.com
-    affil-id: 1
+    affiliations:
+      - ref: 1
     corresponding: true
@@
     email: mu2126@columbia.edu
-    institute: 2
+    affiliations:
+      - ref: 2
   - name: Emilio M. Bruna
     email: embruna@ufl.edu
-    affil-id: 1,3,5
+    affiliations:
+      - ref: 1
+      - ref: 3
+      - ref: 4
 affiliations:
@@
 linkcolor: blue
-bibliography: references.bib
+bibliography: lagged-ipms-ms.bib
 csl: "templates/the-american-naturalist.csl" # Insert path for the bib-style
@@
     keeptex: true
-    fontsize: 12pt
+    fontsize: 11pt
     documentclass: article
@@
         \usepackage{titlesec}
+        \usepackage[sc]{mathpazo} %Like Palatino with extensive math support
+        \usepackage[utf8]{inputenc}
 
@@
   - integral projection models
+  - demography
+  - population dynamics
   - environmental stochasticity
   - lagged effects
+
+Notes: >
+  Manuscript elements: Figure~1, figure~2, table~1, appendix~A.  
+  Manuscript type: e-note
+
+# abstract: >
+#  The abstract should have 150 words. 
 ---
-
-<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->
-
-<!-- With the following code you can access and display values from the yml header above. -->
-
-<!-- Keywords: `r rmarkdown::metadata$keywords` -->
-
-<!-- Running title: `r rmarkdown::metadata$shorttitle` -->
-
-<!-- Highlights: `r rmarkdown::metadata$highlights` -->
-
-<!-- The following code chunk defines some general settings how code chunks should behave. -->
 
@@
 library(targets)
+library(grateful)
 library(tidyverse)
@@
 
-### **Paragraph 1**
+### Paragraph 1: MM and IPM broadly used, but technical challenges remain
 
-1.  Structured demographic models, such as Integral Projection Models (IPMs) and matrix models, have emerged as invaluable tools in ecology, evolution, and conservation biology.
+1.
+Two types of structured population models - matrix models [@leslie1945use; @caswell2001] and integral projection models [@ellnerIntegralProjectionModels2006] - are fundamental frameworks used to study demography and population dynamics.\
+2.
+Their flexibility, in concert with a rapidly growing suite of software, data, and other resources [@ellnerDatadrivenModellingStructured2016; @levinIpmrFlexibleImplementation2021; @salguero-gomezCOMPADREPlantMatrix2015], have facilitated their use to study a wide range of topics in ecology, evolution, and conservation [@ellnerDatadrivenModellingStructured2016; @morrisQuantitativeConservationBiology2002; @croneHowPlantEcologists2011].
+3.
+While recent mathematical and statistical developments have expanded the scope of these applications [e.g., @brooksStatisticalModelingPatterns2019; @reesEvolvingIntegralProjection2016], a number of technical challenges remain.
+4.
+While many of these relate to general issues surrounding model development and validation [e.g., @williams2012avoiding], others stem from the fundamental challenge of using mathematical models - albeit highly advanced ones - to describe and analyze complex biological systems.
 
-They provide insights into \_\_\_\_, the evolution of life-history strategies, and life-history stages and ecological processes at which conservation intervention is likely to have the greatest impact on population growth rate.
+### Paragraph 2: Challenge of lagged effects
 
-Their flexibility hasmade them increasingly applicable to species with differtent life histories.
-
-however, there are still a number of challenges to address, namely \_\_\_\_ , \_\_\_\_, and \_\_\_\_.
-
-### **Paragraph 2**
-
-1.  It has long been recognized that there is the potential for lagged effects.
+1.  Once such biological challenge is the implementation of alagged effects. It has long been recognized that there is the potential for lagged effects.
 2.  Including lagged effects in models has been a major technical challenge, but there are now multiple approaches for doing so.
@@
 
-### **Paragraph 3**
+### Paragraph 3: Why does it matter when modeling?
 
@@
 2.  This could be a big reason for differences between habitats.
-
-### **Paragraph 4**
-
-1.  Here we...
-2.  We conducted these analyses with both deterministic and stochastic IPMs.
+3.  Could also be important for stochastic vs. deterministic models.
+4.  Here we...
+5.  We conducted these analyses with both deterministic and stochastic IPMs.
 
@@
 
-## Results
+## Results & Discussion
 
@@
 Population growth rates were consistently higher in continuous forest compared to forest fragments across IPM types (@tbl-lambdas).
+
+We found that the the choice of Integral Projection Model didn't change the relative ranking of lambda in Continous Forest and Fragments.
 
@@
 
-## Discussion
-
--   Our finding that the choice of IPM didn't change the relative ranking of CF and FF is consistent with @kayeEffectStochasticTechnique2003 finding that method effected stochastic lambda, but relative ranking of populations was consistent.
+These results are consistent with those of @kayeEffectStochasticTechnique2003, who found who found that the method effected stochastic lambda but relative ranking of populations was consistent.
 
@@
 Data and R code used in this study are archived with Zenodo at <doi url>.
+
+<!-- ```{r } -->
+
+<!-- pkgs <- cite_packages(output = "table", out.dir = ".") -->
+
+<!-- knitr::kable(pkgs) -->
+
+<!-- ``` -->
 
@@
 ```
+
+\`\`\`\`

### 2024-09-03 docs/ipm_comparison_ms.qmd
formatting

@@
 1.
-Two types of structured population models - matrix models [@leslie1945use; @caswell2001] and integral projection models [@ellnerIntegralProjectionModels2006] - are fundamental frameworks used to study demography and population dynamics.\
+Two types of structured population models - matrix models [@leslie1945use; @caswell2001] and integral projection models [@ellnerIntegralProjectionModels2006] - are fundamental frameworks used to study demography and population dynamics.
 2.

### 2024-11-07 docs/AmNat_MS_template.tex
update to ms

\documentclass[11pt]{scrartcl}
\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
\usepackage[authoryear,sectionbib,sort]{natbib}
\linespread{1.7}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\usepackage{titlesec}
\titleformat{\section}[block]{\Large\bfseries\filcenter}{\thesection}{1em}{}
\titleformat{\subsection}[block]{\Large\itshape\filcenter}{\thesubsection}{1em}{}
\titleformat{\subsubsection}[block]{\large\itshape}{\thesubsubsection}{1em}{}
\titleformat{\paragraph}[runin]{\itshape}{\theparagraph}{1em}{}[. ]\renewcommand{\refname}{Literature Cited}

%%%%%%%%%%%%%%%%%%%%%
% Line numbering
%%%%%%%%%%%%%%%%%%%%%
%
% Please use line numbering with your initial submission and
% subsequent revisions. After acceptance, please turn line numbering
% off by adding percent signs to the lines %\usepackage{lineno} and
% to %\linenumbers{} and %\modulolinenumbers[1] below.
%
% To avoid line numbering being thrown off around math environments,
% the math environments have to be wrapped using
% \begin{linenomath*} and \end{linenomath*}
%
% (Thanks to Vlastimil Krivan for pointing this out to us!)

\title{Template and Guidelines for Using \LaTeX{} in \textit{The~American~Naturalist} }

% This version of the LaTeX template was last updated on
% July 16, 2024.

%%%%%%%%%%%%%%%%%%%%%
% Authorship
%%%%%%%%%%%%%%%%%%%%%
% Please commet out authorship information while your paper is under review. 
% You will need to add this information back in to your final files after
% acceptance.

\author{Owen E. Cook$^${1,\ast}$ \\ 
Generic H. Collaborator$^{2,\dag}$ \\ 
Additional Q. Expert$^{3}$}

\date{}

\begin{document}

\maketitle

\noindent{} 1. University of Chicago, Chicago, Illinois 60637;

\noindent{} 2. University of Toronto, Toronto, Ontario M5S 1A5, Canada;

\noindent{} 3. Middle Eastern Technical University, Çankaya, Ankara 06800, Turkey.

\noindent{} $\ast$ Corresponding author; e-mail: amnat@uchicago.edu.

\noindent{} $\dag$ Deceased.

\bigskip

\textit{Manuscript elements}: Figure~1, figure~2, table~1, appendix~A (for print; including figure~A1, figure~A2, and table~A1), supplemental PDF. Figure~2 is to print in color.

\bigskip

\textit{Keywords}: Examples, model, template, guidelines.

\bigskip

\textit{Manuscript type}: Article. %Or note, natural history miscellany note, comment, reply, invited symposium, featured topic, or historical perspective.

\bigskip

\noindent{\footnotesize Prepared using the suggested \LaTeX{} template for \textit{Am.\ Nat.}}

\linenumbers{}
\modulolinenumbers[1]

\newpage{}

\section*{Abstract}

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Sed non risus. Suspendisse lectus tortor, dignissim sit amet, adipiscing nec, ultricies sed, dolor. Cras elementum ultrices diam. Praesent quis dolor in dolor molestie cursus et ac nisi. Vestibulum ante purus, semper eget est vitae, vehicula ornare nisl. Morbi efficitur euismod enim, nec feugiat tellus cursus eget.

\newpage{}

\section*{Introduction}

% The journal does not have numbered sections in the main portion of
% articles. Please refrain from using section references (à la
% section~\ref{section:CountingOwlEggs}), and refer to sections by name
% (e.g. section ``Counting Owl Eggs'').

The quick red fox jumps over the lazy brown dog. Furthermore, the quick brown fox jumps over the lazy red dog. In addition, the quick R\"{u}ppell's fox (\textit{Vulpes rueppellii}) jumps over the lazy golden retriever.

\section*{Methods}

Lorem ipsum ut velit mauris, egestas sed, gravida nec, ornare ut, mi. Aenean ut orci vel massa suscipit pulvinar. Nulla sollicitudin. Fusce varius, ligula non tempus aliquam, nunc turpis ullamcorper nibh, in tempus sapien eros vitae ligula (\citealt{OriginSpec}).

\subsection*{Second-order heading}

Lorem ipsum nulla facilisi, pace \citet{LemKapEx07}. Etiam semper, orci sit amet facilisis interdum, tellus nunc consequat erat, quis viverra nisi diam ut metus. Pellentesque cursus, sapien malesuada euismod iaculis, mauris purus interdum diam, vel vestibulum justo enim vitae tellus. Nunc interdum lorem sit amet diam volutpat tristique. Quisque pulvinar ac metus commodo lacinia (\citealt{Ing11,Xiao2015}).

\subsubsection*{Third-order heading}

Usually two or three levels of heading will be all you need. Journal style even permits a fourth level in case you need it.

\paragraph*{Fourth-order heading}
The quick red fox jumps over the lazy brown dog in this paragraph as well. Donec mauris nibh, volutpat vehicula viverra at, iaculis congue sem. Praesent eget erat rhoncus erat sollicitudin volutpat.

\begin{equation}
{ \frac{1}{N_k-1} \sum \limits_{t=1}^{N_k} (M_{tjk} - \bar{M}_{jk})^2}
\end{equation}

Praesent quis dolor in dolor molestie cursus et ac nisi. Vestibulum ante purus, semper eget est vitae, vehicula ornare nisl. Morbi efficitur euismod enim, nec feugiat tellus cursus eget. Donec mauris nibh, volutpat vehicula viverra at, iaculis congue sem.

\subsection*{Another second-order heading}

As \citet{Xiao2015} argued, phasellus porttitor eros et ante condimentum, eget facilisis orci condimentum. Nulla facilisi. Proin placerat elit blandit, euismod dolor nec, dapibus diam. Mauris posuere malesuada lacus, at elementum lacus auctor eu (fig~\ref{Fig:Jumps}A).

\section*{Results}

Lorem ipsum dolor sit amet. Aenean pulvinar malesuada commodo (see \citealt{DavisEtAl2011}; table~\ref{Table:Founders}). Sed aliquet mauris odio, in tristique dui egestas a. Etiam eu malesuada quam. Suspendisse tincidunt eu erat sit amet vulputate. Duis at arcu et nisl dictum mattis. Maecenas vel cursus ante. Cras eleifend elit nec velit sollicitudin fermentum in ac mauris. Pellentesque rutrum magna vel elit maximus hendrerit.

\subsection*{The height of the jump}

Aenean eu pellentesque quam (fig.~\ref{Fig:OkapiHorn}). Nam pellentesque augue eu finibus lacinia. Nullam nec justo vitae odio imperdiet rhoncus vitae vitae quam. Pellentesque porttitor metus et lectus ornare, ac cursus urna efficitur (fig~\ref{Fig:Jumps}B).

\subsection*{The laziness of the dog}

Example paragraph with embedded references (video~\ref{VideoExample}, fig.~\ref{Fig:AnotherFigure}). 
If you have deposited data to Dryad, it is advisable to cite them somewhere in the main text (usually in the Methods or Results sections). A sentence like the following will do: All data are available in the Dryad Digital Repository (\citealt{CookEtAl2015}).

\section*{Discussion}

Nam pulvinar lorem at lorem ultrices, vel accumsan massa feugiat (\citealt{Ing11}). Proin tristique velit eget lacus iaculis, in pellentesque nulla varius. Phasellus sodales est odio, eu pulvinar magna pellentesque eu. Sed ut lobortis eros. Aliquam eget metus turpis. Sed et convallis lectus, id tincidunt enim. In porta nibh ut lacus feugiat, non consequat orci rhoncus. Morbi blandit at augue nec tempor. Sed fringilla ipsum ut justo viverra, ut euismod nisi gravida.

Curabitur non posuere augue, id suscipit orci. Nunc luctus accumsan aliquam. Cras egestas turpis vitae nisl vulputate interdum. Donec pellentesque libero egestas tortor pharetra laoreet. Phasellus facilisis auctor ligula, eu sollicitudin mi sagittis non.

\section*{Conclusion}

Duis pharetra enim at libero cursus, eu commodo mi vestibulum. Nullam eget velit nec lectus viverra sodales. Suspendisse egestas, eros at dictum tincidunt, mi orci laoreet libero, eget rutrum sapien arcu blandit odio.

%%%%%%%%%%%%%%%%%%%%%
% Acknowledgments
%%%%%%%%%%%%%%%%%%%%%
% You may wish to remove the Acknowledgments section while your paper 
% is under review as the Acknowledgments may reveal your identity.
% If you remove this section, you will need to add it back in to your
% final files after acceptance.

\section*{Acknowledgments}

OEC would like to thank Madlen Wilmes, Gyuri Barab\'{a}s, Flo D\'{e}barre, Vlastimil K\v{r}ivan, and Greg Dwyer for their comments and suggestions on this template.

%%%%%%%%%%%%%%%%%%%%%
% Statement of Authorship
%%%%%%%%%%%%%%%%%%%%%
% This section should also be commented out while your MS is undergoing
% double-blind review. The specifics should of course be adapted to
% your paper, but the paragraph below gives some hints of possible
% contributions.

\section*{Statement of Authorship}

OEC conceived the experiments, collected the data, and wrote the original draft.
GHC provided specimens and analyzed the model.
AQE oversaw data analysis and developed the code. 
All authors reviewed and edited the writing at all stages of revision.

\section*{Data and Code Availability}

On initial submission, you may use this section to provide a URL for editors and reviewers that is `private for peer review'. After acceptance, this section must be updated with correct, working DOIs for data and code deposits (such as in Zenodo, Dryad, or DataVerse). An example statement could resemble the following: All data and code for this work are available from the Dryad Digital Repository, \citealt{CookEtAl2015}).

\section*{Appendix A: Additional Methods and Parameters}

% In most cases, authors should typeset supplementary material in a separate,
% author-supplied PDF. For author-supplied PDFs, please consult the
% AmNat_supp_template.tex document, available from
% https://www.journals.uchicago.edu/journals/an/instruct 
%
% If you want to use cross-references within the supplemental TeX file,
% it may be helpful to \usepackage{xr} and include the command 
% \externaldocument{name-of-supplemental-tex-file}
%
% The Appendix instructions below apply to cases in which
% a brief appendix is to appear in print after the main body of the article.
% That notably includes descriptions of methods, tables defining parameters,
% and other material necessary for reproducing the MS's results.
%
% Please reset counters for the appendix (thus normally figure A1, 
% figure A2, table A1, etc.).
%
% Most AmNat articles have no more than one print appendix. If your article
% has more than one, counters for each appendix should match the letter of
% that appendix. For example, tables in Appendix B should be numbered table B1, % table C2, etc. This applies to tables, equations, and figures.
%
% It's better not to use the \appendix command, because we have some
% formatting peculiarities that \appendix conflicts with.

\renewcommand{\theequation}{A\arabic{equation}}
% redefine the command that creates the equation number.
\renewcommand{\thetable}{A\arabic{table}}
\setcounter{equation}{0}  % reset counter 
\setcounter{figure}{0}
\setcounter{table}{0}

\subsection*{Fox--dog encounters through the ages}

The quick red fox jumps over the lazy brown dog. The quick red fox has always jumped over the lazy brown dog. The quick red fox began jumping over the lazy brown dog in the 19th century and has never ceased from so jumping, as we shall see in figure~\ref{Fig:Jumps}. But there can be surprises (figure~\ref{Fig:JumpsOk}).

If the order and location of figures is not otherwise clear, feel free to include explanatory dummy text like this:

[Figure A1 goes here.]

[Figure A2 goes here.]

\subsection*{Further insights}

Tables in the appendices can appear in the appendix text (see table~\ref{Table:Rivers} for an example), unlike appendix figure legends which should be grouped at the end of the document together with the other figure legends.

\begin{table}[h]
\caption{Various rivers, cities, and animals}
\label{Table:Rivers}
\centering
\begin{tabular}{lll}\hline
River        & City        & Animal            \\ \hline
Chicago      & Chicago     & Raccoon           \\
Des Plaines  & Joliet      & Coyote            \\
Illinois     & Peoria      & Cardinal          \\
Kankakee     & Bourbonnais & White-tailed deer \\
Mississippi  & Galena      & Bald eagle        \\ \hline
\end{tabular}
\bigskip{}
\\
{\footnotesize Note: See table~\ref{Table:Founders} below for further table formatting hints.}
\end{table}

Lorem ipsum dolor sit amet, as we have seen in figures~\ref{Fig:Jumps} and \ref{Fig:JumpsOk}.

%%%%%%%%%%%%%%%%%%%%%
% Bibliography
%%%%%%%%%%%%%%%%%%%%%
% You can either type your references following the examples below, or
% compile your BiBTeX database and paste the contents of your .bbl file
% here. The amnatnat.bst style file should work for this---but please
% let us know if you run into any hitches with it!
%
% If you upload a .bib file with your submission, please upload the .bbl
% file as well; this will be required for typesetting.
%
% The list below includes sample journal articles, book chapters, and
% Dryad references.

\begin{thebibliography}{}

\bibitem[{Cook et~al.(2015)Cook, Collaborator, and Expert}]{CookEtAl2015}
Cook, O.~E., G.~H. Collaborator, and A.~Q. Expert. 2015.
\newblock Data from: Template and guidelines for using \LaTeX{} in \textit{The American Naturalist}.
\newblock American Naturalist, Dryad Digital Repository, https://dx.doi.org/10.5061/dryad.XYZAB123.

\bibitem[{Darwin(1859)Darwin}]{OriginSpec}
Darwin, C. 1859.
\newblock On the origin of species by means of natural selection, or the preservation of favoured races in the struggle for life.
\newblock J.~Murray, London.

\bibitem[{Davis et~al.(2011)Davis, Brakora, and Lee}]{DavisEtAl2011}
Davis, E.~B., K.~A. Brakora, and A.~H. Lee. 2011.
\newblock Evolution of ruminant headgear: a review.
\newblock Proceedings of the Royal Society~B 278:2857--2865.

\bibitem[{Inglis et~al.(2011)Inglis, Roberts, Gardner, and Buckling}]{Ing11}
Inglis, R.~F., P.~G. Roberts, A.~Gardner, and A.~Buckling. 2011.
\newblock Spite and the scale of competition in \textit{Pseudomonas
  aeruginosa}.
\newblock American Naturalist 178:276--285.

\bibitem[{Fastovsky(2009)Fastovsky}]{LemKapEx07}
Fastovsky, D.~E. 2009.
\newblock Ideas in dinosaur paleontology: resonating to social and political context.
\newblock Pages 239--253 \emph{in} D. Sepkoski and M. Ruse, eds. The Paleobiological Revolution. University of Chicago Press, Chicago~IL.

\bibitem[{Xiao et~al.(2015)Xiao, McGlinn, and White}]{Xiao2015}
Xiao, X., D.~J. McGlinn, and E.~P. White. 2015.
\newblock A strong test of the maximum entropy theory of ecology.
\newblock American Naturalist 185:E705--E80.

\section*{References Cited Only in the Online Enhancements}

% This section is for works cited in your supplemental PDF but
% not in the main text.

\bibitem[{Tytler(1759)Tytler}]{tytler-mqos}
Tytler, W. 1759.
\newblock The Inquiry, Historical and Critical, into the Evidence against Mary Queen of Scots, and an Examination of the Histories of Dr. Robertson and David Hume with respect to that Evidence.
\newblock W.~Creech, Edinburgh.

\end{thebibliography}

\newpage{}

\section*{Tables}
\renewcommand{\thetable}{\arabic{table}}
\setcounter{table}{0}

\begin{table}[h]
\caption{Founders of \textit{The~American Naturalist}}
\label{Table:Founders}
\centering
\begin{tabular}{lll}\hline
Early editor            & Years with the journal \\ \hline
Alpheus S. Packard Jr.  & 1867--1886 \\
Frederick W. Putnam     & 1867--1874 \\ 
Edward S. Morse         & 1867--1871 \\ 
Alpheus Hyatt           & 1867--1871 \\
Edward Drinker Cope$^a$ & 1878--1897 \\
J.~S. Kingsley          & 1887--1896 \\ \hline 
\end{tabular}
\bigskip{}
\\
{\footnotesize Note: Table titles should be short. Further details should go in a `notes' area after the tabular environment, like this. $^a$ Published the first description of \textit{Dimetrodon}.}
\end{table}

\newpage{}

\section*{Figure legends}

\begin{figure}[h!]
%\includegraphics{horn-of-okapi}
\caption{Figure legends can be longer than the titles of tables. However, they should not be excessively long---in most cases, they should be no more than 100 words each.}
\label{Fig:OkapiHorn}
\end{figure}

\begin{figure}[h!]
%\includegraphics{elegance}
\caption{In this way, figure legends can be listed at the end of the document, with references that work, even though the graphic itself should be included for final files after acceptance. Instead, upload the relevant figure files separately to Editorial Manager; Editorial Manager should insert them at the end of the PDF automatically.}
\label{Fig:AnotherFigure}
\end{figure}

% Figure legends for any print appendices you might have.

\renewcommand{\thefigure}{A\arabic{figure}}
\setcounter{figure}{0}

\begin{figure}[h!]
%\includegraphics{jumps20m}
\caption{\textit{A}, the quick red fox proceeding to jump 20~m straight into the air over not one, but several lazy dogs. \textit{B}, the quick red fox landing gracefully despite the skepticism of naysayers.}
\label{Fig:Jumps}
\end{figure}

\begin{figure}[h!]
%\includegraphics{jumps20m}
\caption{The quicker the red fox jumps, the likelier it is to land near an okapi. For further details, see \citet{LemKapEx07}.}
\label{Fig:JumpsOk}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%
% Videos
%%%%%%%%%%%%%%%%%%%%%
% If you have videos, journal style for them is generally similar to that for
% figures.

%%%%% Include the text below if you have videos

\renewcommand{\figurename}{Video} 
\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}

\begin{figure}[h!]
%\includegraphics{VideoScreengrab.jpg}
\caption{Video legends can follow the same principles as figure legends. Counters should be set and reset so that videos and figures are enumerated separately.}
\label{VideoExample}
\end{figure}

\renewcommand{\figurename}{Figure}
\setcounter{figure}{1}

%%%%% Include the above if you have videos

\end{document}

### 2024-11-07 docs/ipm_comparison_ms.qmd
update to ms

@@
   pdf:
+    # template: AmNat_MS_template.tex
+    # documentclass: scrarticle
+    documentclass: article
     number-sections: false
@@
     keeptex: true
-    fontsize: 11pt
-    documentclass: article
+    fontsize: 12pt
     # the following line is to render the aaffiliations in pdf, which is not default in quarto, see 
@@
         \usepackage[utf8]{inputenc}
+        \usepackage{parskip}
+        \setlength{\parskip}{1pt} % 1ex plus 0.5ex minus 0.2ex}
+        \setlength{\parindent}{0pt}
 
@@
 
+\bigskip
+
+**Keywords:** demography, environmental stochasticity, integral projection models, lagged effects, structured population models, population dynamics
+
+\bigskip
+
+**Manuscript elements:** Figure\~1, figure\~2, table\~1, appendix\~A..
+
+\bigskip
+
+**Manuscript type:** e-note
+
 \pagebreak
@@
 3.
-While recent mathematical and statistical developments have expanded the scope of these applications [e.g., @brooksStatisticalModelingPatterns2019; @reesEvolvingIntegralProjection2016], a number of technical challenges remain.
+Mathematical and statistical advances have rapidly expand the scope of questions that can be addressed with these models [e.g., @brooksStatisticalModelingPatterns2019; @reesEvolvingIntegralProjection2016].
 4.
-While many of these relate to general issues surrounding model development and validation [e.g., @williams2012avoiding], others stem from the fundamental challenge of using mathematical models - albeit highly advanced ones - to describe and analyze complex biological systems.
+However, there are still several fundamental questions for which these models are not yet readily applied, or for which it is unclear how potential alternative approaches will shape model results [e.g., @williams2012avoiding].
+
+<!-- Not quite right yet...maybe:  Others, however, reflect the difficulty of describing complex biological processes with the mathematical and statistical frameworks used for demographic models;  models may be extremely difficult to parameterize, limited in  -->
+
+<!-- recent advances greatly expanded the scope of questions that can be addressed. -->
 
@@
 
-1.  Once such biological challenge is the implementation of alagged effects. It has long been recognized that there is the potential for lagged effects.
-2.  Including lagged effects in models has been a major technical challenge, but there are now multiple approaches for doing so.
-3.  The studies assessing the potential for lagged effects on vital rates find that they indeed appear to be prevalent [@eversLaggedDormantSeason2021; @scottDelayedEffectsClimate2022].
+1.  One example of such a process is *Delayed Life-history Events* (i.e., DLHEs), also known as *Lagged Effects*.
+2.  Lagged effects are those in which the demographic vital rates observed in a given year are influenced -- or even determined by -- past environmental conditions.
+3.  For instance, environmental conditions during juvenile development can shape the expression of traits (e.g., defensive spikes on *Daphnia*) that determine adult survival.
+4.
+Alternatively, the physiological mechanisms responsible for a vital rate can take an extended period of time to complete (e.g., bud formation in \_\_\_\_\_ can begin 1-2 years before flowers appear).
+5.
+Vital rates can even be influenced by environmental conditions during the parental life-cycle or historical trade-offs between vital rates (e.g., delayed costs of reproduction, competition-colonization trade-offs).
+6.
+Although these lagged effects could potentially have major consequences for population dynamics [@beckermanPopulationDynamicConsequences2002], their impacts remain poorly understood for two primary reasons.
+7.  The first is the empirical challenge of designing studies to quantify lagged responses and evaluate their potential drivers [@kussEvolutionaryDemographyLonglived2008a].
+8.
+The second is the technical challenge of incorporating these complex or poorly understood biological processes in demographic models; doing often renders them less parameterizable, tractable, or broadly applicable.
 
@@
 
-1.  Lagged effects could be different in different habitat types.
-2.  This could be a big reason for differences between habitats.
-3.  Could also be important for stochastic vs. deterministic models.
-4.  Here we...
-5.  We conducted these analyses with both deterministic and stochastic IPMs.
+1.  Recent years have seen a surge in new approaches for doing so, including ***,*** \_, and \_\_\_\_.
+2.  Studies assessing potential for lagged effects on vital rates find that they indeed appear to be prevalent [@eversLaggedDormantSeason2021; @scottDelayedEffectsClimate2022].
+3.  Lagged effects could be different in different habitat types.
+4.  This could be a big reason for differences between habitats.
+5.  Could also be important for stochastic vs. deterministic models.
+6.  Here we...
+7.  We conducted these analyses with both deterministic and stochastic IPMs.
 
@@
 ```
-
-\`\`\`\`

### 2024-11-25 docs/ipm_comparison_ms.qmd
updates - MS

@@
 
-## Introduction
-
 <!-- AmNat Notes: communicate concise points, using either data or theory.
@@
 
-### Paragraph 1: MM and IPM broadly used, but technical challenges remain
+## Introduction
 
-1.
 Two types of structured population models - matrix models [@leslie1945use; @caswell2001] and integral projection models [@ellnerIntegralProjectionModels2006] - are fundamental frameworks used to study demography and population dynamics.
-2.
 Their flexibility, in concert with a rapidly growing suite of software, data, and other resources [@ellnerDatadrivenModellingStructured2016; @levinIpmrFlexibleImplementation2021; @salguero-gomezCOMPADREPlantMatrix2015], have facilitated their use to study a wide range of topics in ecology, evolution, and conservation [@ellnerDatadrivenModellingStructured2016; @morrisQuantitativeConservationBiology2002; @croneHowPlantEcologists2011].
-3.
-Mathematical and statistical advances have rapidly expand the scope of questions that can be addressed with these models [e.g., @brooksStatisticalModelingPatterns2019; @reesEvolvingIntegralProjection2016].
-4.
-However, there are still several fundamental questions for which these models are not yet readily applied, or for which it is unclear how potential alternative approaches will shape model results [e.g., @williams2012avoiding].
+Mathematical and statistical advances [e.g., @brooksStatisticalModelingPatterns2019; @williams2012avoiding]have rapidly expand the scope of questions and biological processes that can be investigated with these models [e.g., @reesEvolvingIntegralProjection2016; @ellnerDatadrivenModellingStructured2016].
+Despite this progress, however, several important biological processes have proven challenging to incorporate in structured population models.
+In some cases this is because theoretical and analytical methods for doing so remain underdeveloped; in others this is because data with which to parametertize and assess alternative model structures are lacking [@ellnerDatadrivenModellingStructured2016; @metcalfStatisticalModellingAnnual2015].\
 
-<!-- Not quite right yet...maybe:  Others, however, reflect the difficulty of describing complex biological processes with the mathematical and statistical frameworks used for demographic models;  models may be extremely difficult to parameterize, limited in  -->
+One of these biological processes is *Delayed Life-history Events* (i.e., DLHEs), also known as *Lagged Effects* [@beckermanPopulationDynamicConsequences2002].Lagged effects are those in which the demographic vital rates observed in a given year are influenced -- or even determined by -- past environmental conditions.
+For instance, environmental conditions during juvenile development can shape the expression of traits (e.g., defensive spikes on *Daphnia*) that determine adult survival.
+Alternatively, the physiological mechanisms responsible for a vital rate can take an extended period of time to complete [@eversLaggedDormantSeason2021]; for example, flowering bud formation may be initiated several months before flowers appear [@crileyYearProductionHigh1994].
+Vital rates can even be influenced by environmental conditions during the parental life-cycle or historical trade-offs between vital rates (e.g., delayed costs of reproduction, competition-colonization trade-offs).
+Although these lagged effects could potentially have major consequences for population dynamics [@beckermanPopulationDynamicConsequences2002], their impacts remain poorly understood for two primary reasons.
+The first is limited data - parametrizing models requires long-term data on lagged effects and their potential drivers [@metcalfStatisticalModellingAnnual2015], and these studies can be challenging to design and maintain [@kussEvolutionaryDemographyLonglived2008a].
+The second is challenge is technical - incorporating complex biological processes in demographic models can render them less tractable.\
 
-<!-- recent advances greatly expanded the scope of questions that can be addressed. -->
+Despite these challenges, several recent studies have found there can be large delayed effects of environmental conditions (e.g., climate) on demographic vital rates [@eversLaggedDormantSeason2021; @scottDelayedEffectsClimate2022].
+It remains unclear, however, if including such lagged effects will significantly alter the results of demographic models.
+Using a decade of survey and climate data, we assessed the effects of precipitation extremes on the demographic vital rates of the Amazonian understory herb *Heliconia acuminata* (Heliconiaceae).
+We found that the effects of climate on vital rates could be delayed up to 36 months, with the presence and duration of these effects differing by vital rate and habitat type (i.e., continuous forest, forest fragments).
+Here we used these data to parameterize three different classes of Intgral Projection Models (IPMs): a deterministic IPM, a stochastic IPM, and a stochastic IPM with lagged effects of SPEI on vital rates.
+We then evaluated how model choice influenced projections of population growth rate (i.e., $\lambda$) and structure?
+Based on previous studies [@brunaArePlantPopulations2003; @brunaDemographicEffectsHabitat2005; @brunaExperimentalAssessmentHeliconia2002; @brunaHabitatFragmentationDemographic2002] and demographic theory [@caswell2001; @tuljapurkarPopulationDynamicsVariable1990] we predicted that:\
 
-### Paragraph 2: Challenge of lagged effects
-
-1.  One example of such a process is *Delayed Life-history Events* (i.e., DLHEs), also known as *Lagged Effects*.
-2.  Lagged effects are those in which the demographic vital rates observed in a given year are influenced -- or even determined by -- past environmental conditions.
-3.  For instance, environmental conditions during juvenile development can shape the expression of traits (e.g., defensive spikes on *Daphnia*) that determine adult survival.
-4.
-Alternatively, the physiological mechanisms responsible for a vital rate can take an extended period of time to complete (e.g., bud formation in \_\_\_\_\_ can begin 1-2 years before flowers appear).
-5.
-Vital rates can even be influenced by environmental conditions during the parental life-cycle or historical trade-offs between vital rates (e.g., delayed costs of reproduction, competition-colonization trade-offs).
-6.
-Although these lagged effects could potentially have major consequences for population dynamics [@beckermanPopulationDynamicConsequences2002], their impacts remain poorly understood for two primary reasons.
-7.  The first is the empirical challenge of designing studies to quantify lagged responses and evaluate their potential drivers [@kussEvolutionaryDemographyLonglived2008a].
-8.
-The second is the technical challenge of incorporating these complex or poorly understood biological processes in demographic models; doing often renders them less parameterizable, tractable, or broadly applicable.
-
-### Paragraph 3: Why does it matter when modeling?
-
-1.  Recent years have seen a surge in new approaches for doing so, including ***,*** \_, and \_\_\_\_.
-2.  Studies assessing potential for lagged effects on vital rates find that they indeed appear to be prevalent [@eversLaggedDormantSeason2021; @scottDelayedEffectsClimate2022].
-3.  Lagged effects could be different in different habitat types.
-4.  This could be a big reason for differences between habitats.
-5.  Could also be important for stochastic vs. deterministic models.
-6.  Here we...
-7.  We conducted these analyses with both deterministic and stochastic IPMs.
+(i) $\lambda$ would be higher in continuous forest forest fragments regardless of model type,
+(ii) that projections of $\lambda$ from deterministic models would be higher than those of stochastic models,
+(iii) that $\lambda$ would be lowest for models including lagged effects, and
+(iv) populations would be more skewed towards pre-reproductive size classes in fragments that forest, regadless of wether models included stochaticity or lagged effects.
 
@@
 
-### *Demographic methods and data*
+### *Study System and Demographic Data*
 
-<!--# Emilio will fill this section -->
+<!--# Emilio will fill this section with info about Ha and demog data -->
 
-Overview of the heliconia project and data
-
-<!-- Add info about additional datasets (Emilio) -->
+> Overview of the *Heliconia* project
 
@@
 ```
-In preliminary investigation we found that the survival and growth of plants was better explained by treating seedlings and mature plants separately.
+
+In preliminary investigation, we found that the survival and growth of plants was better explained by treating seedlings and mature plants separately.
 Seedlings are physiologically different from small plants because they necessarily lack the underground reserves (of carbohydrates and meristems) that a small, mature plant may have.
 Therefore, we used general IPMs to model population dynamics with seedlings treated as a separate discreet class not structured by size.
-General IPMs allow for combinations of continuous and discrete states and transitions between them [@ellnerDatadrivenModellingStructured2016].
+General IPMs allow for combinations of continuous and discrete states and transitions between them [@ellnerDatadrivenModellingStructured2016].\
 
@@
 seedling survival and establishment, @eq-R) ([@eq-mature; @eq-sdlg], @fig-lifecycle).
-The probability of staying a seedling, was always equal to zero, since our definition of seedlings was first year plants only.
+The probability of staying a seedling, was always equal to zero, since our definition of seedlings was first year plants only.\
 
@@
 -->) and survival and growth (or regression) of mature plants ([@eq-mature]).
-Seedlings survive ($s_s$) and grow into mature plants of a particular size ($G_s(z^\prime)$) ([@eq-R]).
+Seedlings survive ($s_s$) and grow into mature plants of a particular size ($G_s(z^\prime)$) ([@eq-R]).\
+
 Mature plants survive as a function of size ($s(z)$), and grow (or regress) to a new size as a function of their previous size ($G(z^\prime, z)$) ([@eq-P]).
-Mature plants flower with a probability that is a function of size ($p_f(z)$) and produce a number of seeds as a function of size ($f(z)$), which germinate and establish as seedlings with probability $g$ ([@eq-F]).
+Mature plants flower with a probability that is a function of size ($p_f(z)$) and produce a number of seeds as a function of size ($f(z)$), which germinate and establish as seedlings with probability $g$ ([@eq-F]).\
 
@@
 -->.
-For consistency, seedling survival and growth were also modeled using GAMs, but without size in the previous census as a predictor (i.e. intercept only models).
-For growth models ($G_s(z^\prime)$ and $G(z^\prime, z)$) a scaled t family distribution provided a better fit to the data than a gaussian fit as the residuals were leptokurtic with a simple gaussian model.
+For consistency, seedling survival and growth were also modeled using GAMs, but without size in the previous census as a predictor (i.e.
+intercept only models).
+For growth models ($G_s(z^\prime)$ and $G(z^\prime, z)$) a scaled t family distribution provided a better fit to the data than a gaussian fit as the residuals were leptokurtic with a simple Gaussian model.\
 
@@
 Germination and establishment rates in continuous forest and forest fragments were estimated using data from ....
-<!--# Emilio: add more info here -->
+<!--# Emilio: add more info here -->.\
 
@@
 The kernel-resampling approach is to generate kernels corresponding to each transition year in the demographic dataset using the random smooths for year, and to iterate the IPM by drawing from these randomly.
-This is equivalent to the matrix selection approach for matrix population models described by @caswell2001.
+This is equivalent to the matrix selection approach for matrix population models described by @caswell2001.\
 
@@
 These values are then used to predict fitted values from the vital rates models, generating different kernels at each iteration of the IPM.
-With this method, the kernels of successive iterations are not entirely independent because the SPEI values used in calculating vital rates include values used in the previous two iterations, but they are ergodic.
+With this method, the kernels of successive iterations are not entirely independent because the SPEI values used in calculating vital rates include values used in the previous two iterations, but they are ergodic.\
+
 <!--# ergodicity is somethign discussed in the Caswell book.
@@
 All IPMs were constructed and iterated <!--# optimized in notes/optimizing-meshpoints.Rmd which could be turned into supplemental -->using the `ipmr` package in R [@levinIpmrFlexibleImplementation2021].
-The IPMs used 100 meshpoints and the midpoint rule for calculating kernels <!--# citation? -->.
+The IPMs used 100 meshpoints and the midpoint rule for calculating kernels <!--# citation?
+-->.
 For each type of IPM we iterated the model for 1000 time steps, discarding the first 100 time steps to omit transient effects.
@@
 We used the distribution of established plant sizes and proportion of seedlings from the full dataset as a starting population vector.
-While other starting population vectors were possible, the choice is of little importance as it will only impact transient dynamics, which we aren't interested in for this study.
+While other starting population vectors were possible, the choice is of little importance as it will only impact transient dynamics, which we aren't interested in for this study.\
 
@@
 For each bootstrap, we then re-fit vital rates models (all except germination and establishment rate, fruits per flowering plant, and seeds per fruit, which were estimated using different datasets), constructed IPMs, and calculated a value for lambda as described above.
-We then used these bootstraped estimates of lambda to calculate bias corrected 95% confidence intervals [@ellnerDatadrivenModellingStructured2016].
+We then used these bootstraped estimates of lambda to calculate bias corrected 95% confidence intervals [@ellnerDatadrivenModellingStructured2016].\
 
-This workflow was managed using the `targets` R package [@targets] which also allowed us to track computational time spent on each IPM for comparison.
+This workflow was managed using the `targets` R package [@targets] which also allowed us to track computational time spent on each IPM for comparison.\
 
@@
 
-All about the stats.
+> All about the stats.
 
@@
 
+1.
 For all vital rates estimated using the long term demographic dataset, the DLNM model fit the best (dAIC = 0) followed by the model with a random effect of year, followed by the deterministic model (@tbl-aic).\
@@
 
-Population growth rates were consistently higher in continuous forest compared to forest fragments across IPM types (@tbl-lambdas).
+2.  Population growth rates were consistently higher in continuous forest compared to forest fragments across IPM types (@tbl-lambdas).\
 
-We found that the the choice of Integral Projection Model didn't change the relative ranking of lambda in Continous Forest and Fragments.
+3.  We found that the the choice of Integral Projection Model didn't change the relative ranking of lambda in Continous Forest and Fragments.\
 
-The time to iterate the DLNM models is much higher than than deterministic and kernel-resampled.
+4.  The time to iterate the DLNM models is much higher than than deterministic and kernel-resampled.\
 
-The greater use of computational resources is likely a result of `predict()` being much slower for GAMs with 2D smooths because the number of knots is much higher compared to the GAMs used for the vital rates models in the determinsitic and kernel-resampled IPMs.
+5.
+The greater use of computational resources is likely a result of `predict()` being much slower for GAMs with 2D smooths because the number of knots is much higher compared to the GAMs used for the vital rates models in the determinsitic and kernel-resampled IPMs.\
+
 <!--# also note that this can't really be paralellized because each iteration depends on the previous one.
@@
 
-@fig-pop-states has some interesting things in it:
+6.  @fig-pop-states has some interesting things in it:
 
--   for the deterministic IPM (and the kernel-resampled IPM?) there are slightly more of the smallest plants and the largest plants in CF compared to FF (i.e. more medium sized plants in FF).
+> -   For the deterministic IPM (and the kernel-resampled IPM?) there are slightly more of the smallest plants and the largest plants in CF compared to FF (i.e. more medium sized plants in FF).
 
--   For the kernel-resampled IPM (random effect of year), the fluctuations are extremely similar between CF and FF
+> -   For the kernel-resampled IPM (random effect of year), the fluctuations are extremely similar between CF and FF
 
--   For the parameter-resampled IPM (DLNM) the size structure of the population is a LOT more variable in FF.
-    This makes sense as we know lagged effects are more important in fragments.
+> -   For the parameter-resampled IPM (DLNM) the size structure of the population is a LOT more variable in FF. This makes sense as we know lagged effects are more important in fragments.
 
--   Also, the fluctuations in size structure in CF do not match the fluctuations in FF as well (can see this by the increased spread of points in @fig-pop-states (B)
+> -   Also, the fluctuations in size structure in CF do not match the fluctuations in FF as well (can see this by the increased spread of points in @fig-pop-states (B)
 
--   Also, in the parameter-resampled IPM (and only in this one), we see a shift toward smaller plants in FF compared to CF
+> -   Also, in the parameter-resampled IPM (and only in this one), we see a shift toward smaller plants in FF compared to CF
 
-These results are consistent with those of @kayeEffectStochasticTechnique2003, who found who found that the method effected stochastic lambda but relative ranking of populations was consistent.
+> -   These results are consistent with those of @kayeEffectStochasticTechnique2003, who found who found that the method effected stochastic lambda but relative ranking of populations was consistent.
 
@@
 National Science Foundation (awards \_\_\_\_, and \_\_\_\_).
-This article is publication no. -- -- in the BDFFP Technical series.
+This article is publication no.
+-- -- in the BDFFP Technical series.
 The authors declare no conflicts of interest.
@@
 ```{r #fig-pop-states,  echo=FALSE, out.width="85%",fig.align='center'}
+
 #| fig.cap = "Relative proportions of plant sizes in the first 250 iterations of the IPM simulations.

### 2024-12-02 docs/ipm_comparison_ms.qmd
added methods

@@
 
-Two types of structured population models - matrix models [@leslie1945use; @caswell2001] and integral projection models [@ellnerIntegralProjectionModels2006] - are fundamental frameworks used to study demography and population dynamics.
+Integral Projection Models (i.e., IPMs) are an important and widely used tool for studying demography and population dynamics [@ellnerIntegralProjectionModels2006; @reesBuildingIntegralProjection2014; @reesIntegralProjectionModels2009].
 Their flexibility, in concert with a rapidly growing suite of software, data, and other resources [@ellnerDatadrivenModellingStructured2016; @levinIpmrFlexibleImplementation2021; @salguero-gomezCOMPADREPlantMatrix2015], have facilitated their use to study a wide range of topics in ecology, evolution, and conservation [@ellnerDatadrivenModellingStructured2016; @morrisQuantitativeConservationBiology2002; @croneHowPlantEcologists2011].
@@
 
-One of these biological processes is *Delayed Life-history Events* (i.e., DLHEs), also known as *Lagged Effects* [@beckermanPopulationDynamicConsequences2002].Lagged effects are those in which the demographic vital rates observed in a given year are influenced -- or even determined by -- past environmental conditions.
+One of these biological processes is *Delayed Life-history Events* (i.e., DLHEs), also known as *Lagged Effects* [@beckermanPopulationDynamicConsequences2002].
+Lagged effects are those in which the demographic vital rates observed in a given year are influenced -- or even determined by -- past environmental conditions.
 For instance, environmental conditions during juvenile development can shape the expression of traits (e.g., defensive spikes on *Daphnia*) that determine adult survival.
 Alternatively, the physiological mechanisms responsible for a vital rate can take an extended period of time to complete [@eversLaggedDormantSeason2021]; for example, flowering bud formation may be initiated several months before flowers appear [@crileyYearProductionHigh1994].
-Vital rates can even be influenced by environmental conditions during the parental life-cycle or historical trade-offs between vital rates (e.g., delayed costs of reproduction, competition-colonization trade-offs).
-Although these lagged effects could potentially have major consequences for population dynamics [@beckermanPopulationDynamicConsequences2002], their impacts remain poorly understood for two primary reasons.
-The first is limited data - parametrizing models requires long-term data on lagged effects and their potential drivers [@metcalfStatisticalModellingAnnual2015], and these studies can be challenging to design and maintain [@kussEvolutionaryDemographyLonglived2008a].
-The second is challenge is technical - incorporating complex biological processes in demographic models can render them less tractable.\
+Vital rates can even be influenced by environmental conditions during the parental life-cycle or historical trade-offs between vital rates (e.g., delayed costs of reproduction, competition-colonization trade-offs).\
 
-Despite these challenges, several recent studies have found there can be large delayed effects of environmental conditions (e.g., climate) on demographic vital rates [@eversLaggedDormantSeason2021; @scottDelayedEffectsClimate2022].
-It remains unclear, however, if including such lagged effects will significantly alter the results of demographic models.
-Using a decade of survey and climate data, we assessed the effects of precipitation extremes on the demographic vital rates of the Amazonian understory herb *Heliconia acuminata* (Heliconiaceae).
-We found that the effects of climate on vital rates could be delayed up to 36 months, with the presence and duration of these effects differing by vital rate and habitat type (i.e., continuous forest, forest fragments).
-Here we used these data to parameterize three different classes of Intgral Projection Models (IPMs): a deterministic IPM, a stochastic IPM, and a stochastic IPM with lagged effects of SPEI on vital rates.
-We then evaluated how model choice influenced projections of population growth rate (i.e., $\lambda$) and structure?
+Although such lagged effects could potentially have major consequences for population dynamics [@beckermanPopulationDynamicConsequences2002], their their demographic impacts remain poorly understood [but see @molowny-horasChangesNaturalDynamics2017; @tenhumbergTimelaggedEffectsWeather2018; @williamsLifeHistoryEvolution2015].
+There are two primary reasons for this limited understanding.
+The first is a lack of empirical data [sensu @eversLaggedDormantSeason2021].
+Detecting lagged effects requires long-term data on both the putative lagged effects and their potential drivers [@metcalfStatisticalModellingAnnual2015], and studies to detect them can be challenging to design and maintain [@kussEvolutionaryDemographyLonglived2008a].
+The second is challenge is technical - incorporating complex biological processes such as lagged effects in demographic models can render the models less tractable.
+<!-- would be good to close with a sentence stating that if model outputs are incorrect, thenm failure to do so could lead to major problems...
+-->\
+
+We used a decade of survey and climate data to assess the effects of precipitation extremes on the demographic vital rates of the Amazonian understory herb *Heliconia acuminata* [@scottDelayedEffectsClimate2022].
+Our analyses revealed that the effects of climate on vital rates could be delayed up to 36 months, and that the presence and duration of these effects could differ by vital rate and habitat type (i.e., continuous forest vs.
+forest fragments).
+Here we investigate how including lagged effects in Integral Projection Models influences projections of population growth rate (i.e., $\lambda$) and structure.
+To do so we parameterized and compared three different classes of Integral Projection Models: a deterministic IPM, a stochastic IPM, and a stochastic IPM with lagged effects of SPEI on vital rates.
 Based on previous studies [@brunaArePlantPopulations2003; @brunaDemographicEffectsHabitat2005; @brunaExperimentalAssessmentHeliconia2002; @brunaHabitatFragmentationDemographic2002] and demographic theory [@caswell2001; @tuljapurkarPopulationDynamicsVariable1990] we predicted that:\
 
-(i) $\lambda$ would be higher in continuous forest forest fragments regardless of model type,
+(i) $\lambda$ would be higher in forest than fragments regardless of model type,
 (ii) that projections of $\lambda$ from deterministic models would be higher than those of stochastic models,
 (iii) that $\lambda$ would be lowest for models including lagged effects, and
-(iv) populations would be more skewed towards pre-reproductive size classes in fragments that forest, regadless of wether models included stochaticity or lagged effects.
+(iv) populations would be more skewed towards pre-reproductive size classes in fragments that forest, regardless of whether models included stochasticity or lagged effects.
 
@@
 
-<!--# Emilio will fill this section with info about Ha and demog data -->
+*Heliconia acuminata* (Heliconiaceae) is a perennial, self-incompatible monocot [@kressDiversityDistributionHeliconia1990] that is distributed throughout much of the Amazon basin [@kressDiversityDistributionHeliconia1990].
+While some *Heliconia* species grow in large aggregations on roadsides, gaps, and in other disturbed habitats, others - including *H.
+acuminata* - grow primarily in the forest understory [@kressSelfincompatibilitySystemsCentral1983; @ribeiroInfluencePostclearingTreatment2010].
+Understory *Heliconia* species typically produce fewer flowers and are pollinated by traplining hummingbirds [@stoufferForestFragmentationSeasonal1996; @brunaHeliconiaAcuminataReproductive2004].\
 
-> Overview of the *Heliconia* project
+The models and analyses here are based on 11 years (1998-2009) of demographic data collected on \>8500 *H.
+acuminata* found at Brazil's Biological Dynamics of Forest Fragments Project (BDFFP), located \~70 km north of Manaus, Brazil.
+The BDFFP reserves include both continuous forest and forest fragments that range in size from 1-100 ha.
+These fragment reserves were originally isolated in the early 1980’s by the creation of cattle pastures, with the secondary growth surrounding them periodically cleared to ensure their continued isolation.
+The habitat in all sites is non-flooded lowland rain forest with rugged topography.
+A complete summary of the BDFFP and its history can be found in @bierregaardLessonsAmazoniaEcology2001.\
+
+In 1997--1998 a series of 5000 m^2^ plots were established in the BDFFP's Continuous Forest Reserves (N=6 plots) and 1-ha Fragments (N=4 plots) in which all of the *Heliconia acuminata* were marked and measured.
+The plots were censused annually, at which time a team maked marked and measured new seedlings, identified any previously marked plants that died, and recorded the size of surviving individuals.
+Each plot was also surveyed 4-5 times during the flowering season to identify reproductive plants; in our site *H.
+acuminata* begin flowering early in the rainy season (e.g., January) and most reproductive plants produce a single infloresence (range = 1--7) with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
+Fruits mature April-May and have 1--3 seeds per fruit ($\bar{x}=2$) that are dispersed by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011].
+Dispersed seeds germinate approximately 6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999].
+On average olots in CF also had more than twice as many plants as the plots in 1-ha fragments (CF median = 788, range = \[201, 1549\]; 1-ha median = 339, range = \[297, 400\]).
+A complete description of the demographic methods, data, and analyses to date can be found in [@brunaDemographyUnderstoryHerb2023].\
 
 ### *Construction of Integral Projection Models*
-
-```{=html}
-<!-- TODO:
-
--   Add paragraph about why we used IPMs or why a General IPM---depends on what goes in introduction maybe?
-
--->
-```
 
@@
 
+```{r }
+meta_df <- 
+  tar_meta(store = here("_targets")) %>% 
+  select(name, size, bytes, time, seconds) %>% 
+  filter(str_detect(name, "^ipm_\\w+_[cf]{2}$")) %>% 
+  mutate(minutes = seconds / 60,
+         hours = minutes / 60)
+
+meta_df_tbl<-meta_df %>% 
+  separate(name, into = c("trash", "IPM", "Habitat")) %>% 
+  select(-trash) %>% 
+  select(IPM, Habitat, minutes) %>% 
+  group_by(IPM) %>% 
+mutate(IPM = str_replace_all(
+    IPM,
+    c(
+      "det" = "det",  # Deterministic",
+      "stoch" = "sk", # Stochastic, kernel-resampled
+      "dlnm" = "sp" # Stochastic, parameter-resampled
+    ))) %>%
+  summarize(mean_time_min = round(mean(minutes), 2)) %>%
+  rename("IPM Type" = IPM) %>% 
+  rename("mean time (min.)" = mean_time_min)
+
+det_time<-meta_df_tbl %>% filter(`IPM Type`=="det")
+det_time<-as.numeric(det_time[1,2])
+sker_time<-meta_df_tbl %>% filter(`IPM Type`=="sk")
+sker_time<-as.numeric(sker_time[1,2])
+spar_time<-meta_df_tbl %>% filter(`IPM Type`=="sp")
+spar_time<-as.numeric(spar_time[1,2])
+```
+
 1.
@@
 
-4.  The time to iterate the DLNM models is much higher than than deterministic and kernel-resampled.\
+4.
+DLNM models take much, much longer to iterate: while the Deterministic and Kernel-resampled stochastic models took \~`r det_time` and \~`r sker_time` min to iterate (respectively), the Parameter-resampled stochastic models with lagged effects took \~`r spar_time` min.\
 
@@
 
-```{r tbl-time, results='asis'}
-#| label: tbl-time
-#| tbl-cap: "Figure caption to be written"
-meta_df <- 
-  tar_meta(store = here("_targets")) %>% 
-  select(name, size, bytes, time, seconds) %>% 
-  filter(str_detect(name, "^ipm_\\w+_[cf]{2}$")) %>% 
-  mutate(minutes = seconds / 60,
-         hours = minutes / 60)
-
-meta_df_tbl<-meta_df %>% 
-  separate(name, into = c("trash", "IPM", "Habitat")) %>% 
-  select(-trash) %>% 
-  select(IPM, Habitat, minutes) %>% 
-  group_by(IPM) %>% 
-mutate(IPM = str_replace_all(
-    IPM,
-    c(
-      "det" = "Deterministic",
-      "stoch" = "Stochastic, kernel-resampled",
-      "dlnm" = "Stochastic, parameter-resampled"
-    ))) %>%
-  summarize(mean_time_min = round(mean(minutes), 2)) %>%
-  rename("IPM Type" = IPM) %>% 
-  rename("mean time (min.)" = mean_time_min)
-
-kbl(meta_df_tbl, 
-    booktabs = T,
-    align = "lc") %>% 
-  kable_styling(full_width = F,
-                position = "center")
-
-```
-
-\newpage
-
 ```{r tbl-lambdas, results='asis'}

### 2025-04-09 docs/ipm_comparison_ms.qmd
update MS

@@
         \usepackage{titlesec}
-        \usepackage[sc]{mathpazo} %Like Palatino with extensive math support
-        \usepackage[utf8]{inputenc}
+        <!-- \usepackage[sc]{mathpazo} %Like Palatino with extensive math support -->
+        <!-- \usepackage[utf8]{inputenc} -->
         \usepackage{parskip}
@@
 
-Integral Projection Models (i.e., IPMs) are an important and widely used tool for studying demography and population dynamics [@ellnerIntegralProjectionModels2006; @reesBuildingIntegralProjection2014; @reesIntegralProjectionModels2009].
-Their flexibility, in concert with a rapidly growing suite of software, data, and other resources [@ellnerDatadrivenModellingStructured2016; @levinIpmrFlexibleImplementation2021; @salguero-gomezCOMPADREPlantMatrix2015], have facilitated their use to study a wide range of topics in ecology, evolution, and conservation [@ellnerDatadrivenModellingStructured2016; @morrisQuantitativeConservationBiology2002; @croneHowPlantEcologists2011].
-Mathematical and statistical advances [e.g., @brooksStatisticalModelingPatterns2019; @williams2012avoiding]have rapidly expand the scope of questions and biological processes that can be investigated with these models [e.g., @reesEvolvingIntegralProjection2016; @ellnerDatadrivenModellingStructured2016].
-Despite this progress, however, several important biological processes have proven challenging to incorporate in structured population models.
-In some cases this is because theoretical and analytical methods for doing so remain underdeveloped; in others this is because data with which to parametertize and assess alternative model structures are lacking [@ellnerDatadrivenModellingStructured2016; @metcalfStatisticalModellingAnnual2015].\
+There is increasing evidence that an organism's current likelihood of growth, survival, or reproduction can be strongly influenced by previous environmental conditions.
+These *Lagged Effects*, also known as *Delayed Life-history Events* (i.e., DLHEs) [@beckerman_population_2002], can simultaneously affect an entire cohort (e.g., juveniles hatching during a period of scarcity will all have delayed maturation and lower lifetime fecundity) or only a subset of the population (e.g., cold temperatures in one year lead to reduced flowering by potentially reproductive individuals in the next).
+In addition, the temporal delay between an environmental event and changes in demographic vital rates depends on both the intensity of the event and its timing relative to the underlying physiological processes [@eversLaggedDormantSeason2021; @criley_year_1994].
+A drought during the early stages of gestation or floral bud formation, for example, might have a much larger impact on the number of fruits or offspring produced than one several months later whose timing coincides with birth or flowering.
+The delay or intensity of lagged effects can also depend on local ecological conditions, with individuals in some habitats buffered against -- or able to recover more quickly from - the delayed effects of environmental variation.\
 
-One of these biological processes is *Delayed Life-history Events* (i.e., DLHEs), also known as *Lagged Effects* [@beckermanPopulationDynamicConsequences2002].
-Lagged effects are those in which the demographic vital rates observed in a given year are influenced -- or even determined by -- past environmental conditions.
-For instance, environmental conditions during juvenile development can shape the expression of traits (e.g., defensive spikes on *Daphnia*) that determine adult survival.
-Alternatively, the physiological mechanisms responsible for a vital rate can take an extended period of time to complete [@eversLaggedDormantSeason2021]; for example, flowering bud formation may be initiated several months before flowers appear [@crileyYearProductionHigh1994].
-Vital rates can even be influenced by environmental conditions during the parental life-cycle or historical trade-offs between vital rates (e.g., delayed costs of reproduction, competition-colonization trade-offs).\
+Because Lagged Effects are often directly linked to reproduction and survival, it is thought they could have major consequences for population dynamics [@beckerman_population_2002].
+Although there is emerging evidence that this is indeed the case [e.g., @molowny-horas_changes_2017; @tenhumberg_time-lagged_2018; @williams_life_2015], broader efforts to test this hypothesis have been hampered by two primary factors: First, detecting lagged effects requires long-term data on both the putative lagged effect (i.e., probability of flowering) and its potential environmental drivers [@metcalfStatisticalModellingAnnual2015].
+These coupled data sets are rare [sensu @eversLaggedDormantSeason2021], in part because studies to disentangle lagged effects can be challenging to design and maintain [@kuss_evolutionary_2008].
+Second, the methods for identifying lagged effects and modeling their demographic impacts can be challenging to implement.
+Many of the statistical methods have stringent data requirements [@metcalfStatisticalModellingAnnual2015] and assumptions, while the including complex biological processes in demographic models can render them less tractable.
+Addressing these obstacles is a major undertaking; the value of doing so will depend on the effort required vs.
+the potential consequences of failing to consider lagged effects - consequences that range from overestimating projections of population growth rate (i.e., $\lambda$) in a conservation setting to drawing invalid conclusions regarding support for the predictions of ecological or evolutionary theory.\
 
-Although such lagged effects could potentially have major consequences for population dynamics [@beckermanPopulationDynamicConsequences2002], their their demographic impacts remain poorly understood [but see @molowny-horasChangesNaturalDynamics2017; @tenhumbergTimelaggedEffectsWeather2018; @williamsLifeHistoryEvolution2015].
-There are two primary reasons for this limited understanding.
-The first is a lack of empirical data [sensu @eversLaggedDormantSeason2021].
-Detecting lagged effects requires long-term data on both the putative lagged effects and their potential drivers [@metcalfStatisticalModellingAnnual2015], and studies to detect them can be challenging to design and maintain [@kussEvolutionaryDemographyLonglived2008a].
-The second is challenge is technical - incorporating complex biological processes such as lagged effects in demographic models can render the models less tractable.
-<!-- would be good to close with a sentence stating that if model outputs are incorrect, thenm failure to do so could lead to major problems...
--->\
+Integral Projection Models (i.e., IPMs) are an important and widely used tool for studying demography and population dynamics [@ellner_integral_2006; @rees_building_2014; @rees_integral_2009].
+Their flexibility, in concert with a rapidly growing suite of software, data, and other resources [@ellnerDatadrivenModellingStructured2016; @levinIpmrFlexibleImplementation2021; @salguerogomez_compadre_2015], have facilitated their use to study a wide range of topics in ecology, evolution, and conservation [@ellnerDatadrivenModellingStructured2016; @morris_quantitative_2002 and @crone_how_2011].
+Mathematical and statistical advances [e.g., @williams2012avoiding; @brooks_statistical_2019] have rapidly expand the scope of questions and biological processes that can be investigated with these models [e.g., @metcalfStatisticalModellingAnnual2015; @rees_evolving_2016; @ellnerDatadrivenModellingStructured2016].
+Here we investigate how including lagged effects in Integral Projection Models influences projections of $\lambda$ and population structure.\
 
-We used a decade of survey and climate data to assess the effects of precipitation extremes on the demographic vital rates of the Amazonian understory herb *Heliconia acuminata* [@scottDelayedEffectsClimate2022].
-Our analyses revealed that the effects of climate on vital rates could be delayed up to 36 months, and that the presence and duration of these effects could differ by vital rate and habitat type (i.e., continuous forest vs.
+We have previously shown that the effects of precipitation extremes on the demographic vital rates of an Amazonian understory herb (*Heliconia acuminata*, Heliconiaceae) can be delayed up to 36 months [@scottDelayedEffectsClimate2022], with the presence and duration of these lagged effects varying by vital rate and habitat.
+We parameterized three classes of Integral Projection Models - a deterministic IPM, a stochastic IPM, and a stochastic IPM with lagged effects of precipitation on vital rates - for populations in two habitat type (i.e., continuous forest vs.
 forest fragments).
-Here we investigate how including lagged effects in Integral Projection Models influences projections of population growth rate (i.e., $\lambda$) and structure.
-To do so we parameterized and compared three different classes of Integral Projection Models: a deterministic IPM, a stochastic IPM, and a stochastic IPM with lagged effects of SPEI on vital rates.
-Based on previous studies [@brunaArePlantPopulations2003; @brunaDemographicEffectsHabitat2005; @brunaExperimentalAssessmentHeliconia2002; @brunaHabitatFragmentationDemographic2002] and demographic theory [@caswell2001; @tuljapurkarPopulationDynamicsVariable1990] we predicted that:\
+Based on previous studies [@bruna_are_2003; @bruna_demographic_2005; @bruna_experimental_2002; @bruna_habitat_2002] and demographic theory [@caswell2001; @tuljapurkar_population_1990] we predicted that:\
 
-(i) $\lambda$ would be higher in forest than fragments regardless of model type,
-(ii) that projections of $\lambda$ from deterministic models would be higher than those of stochastic models,
-(iii) that $\lambda$ would be lowest for models including lagged effects, and
-(iv) populations would be more skewed towards pre-reproductive size classes in fragments that forest, regardless of whether models included stochasticity or lagged effects.
+(i) projections of $\lambda$ would be lower in forest fragments than in continuous forest for all model types,
+(ii) projections of $\lambda$ from deterministic models would be higher than those of stochastic models,
+(iii) projections of $\lambda$ would be lowest for models with lagged effects, and
+(iv) populations would be more skewed towards pre-reproductive size classes in fragments that forest, regardless of model structure
 
@@
 
-*Heliconia acuminata* (Heliconiaceae) is a perennial, self-incompatible monocot [@kressDiversityDistributionHeliconia1990] that is distributed throughout much of the Amazon basin [@kressDiversityDistributionHeliconia1990].
+*Heliconia acuminata* (Heliconiaceae) is a perennial, self-incompatible monocot that is distributed throughout much of the Amazon basin [@kress_diversity_1990].
 While some *Heliconia* species grow in large aggregations on roadsides, gaps, and in other disturbed habitats, others - including *H.
-acuminata* - grow primarily in the forest understory [@kressSelfincompatibilitySystemsCentral1983; @ribeiroInfluencePostclearingTreatment2010].
-Understory *Heliconia* species typically produce fewer flowers and are pollinated by traplining hummingbirds [@stoufferForestFragmentationSeasonal1996; @brunaHeliconiaAcuminataReproductive2004].\
-
+acuminata* - grow primarily in the forest understory [@kress_self-incompatibility_1983; @ribeiro_influence_2010].
+Understory *Heliconia* species produce fewer flowers and are pollinated by traplining hummingbirds [@bruna_heliconia_2004; @stouffer_forest_1996].
 The models and analyses here are based on 11 years (1998-2009) of demographic data collected on \>8500 *H.
@@
 The habitat in all sites is non-flooded lowland rain forest with rugged topography.
-A complete summary of the BDFFP and its history can be found in @bierregaardLessonsAmazoniaEcology2001.\
+A complete summary of the BDFFP and its history can be found in @bierregaard_lessons_2001.\
 
-In 1997--1998 a series of 5000 m^2^ plots were established in the BDFFP's Continuous Forest Reserves (N=6 plots) and 1-ha Fragments (N=4 plots) in which all of the *Heliconia acuminata* were marked and measured.
-The plots were censused annually, at which time a team maked marked and measured new seedlings, identified any previously marked plants that died, and recorded the size of surviving individuals.
+A complete description of the demographic methods, data, and analyses to date can be found in @bruna_demography_2023.
+Briefly, in 1997--1998 a series of 5000 m^2^ plots were established in the BDFFP reserves: N=6 in Continuous Forest and N=4 in 1-ha Forest Fragments.
+All of the *Heliconia acuminata* in these plots were marked and measured; the plots were censused annually, at which time a team recorded the size of surviving individuals, marked and measured new seedlings, and identified any previously marked plants that died.
 Each plot was also surveyed 4-5 times during the flowering season to identify reproductive plants; in our site *H.
-acuminata* begin flowering early in the rainy season (e.g., January) and most reproductive plants produce a single infloresence (range = 1--7) with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
-Fruits mature April-May and have 1--3 seeds per fruit ($\bar{x}=2$) that are dispersed by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011].
-Dispersed seeds germinate approximately 6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999].
-On average olots in CF also had more than twice as many plants as the plots in 1-ha fragments (CF median = 788, range = \[201, 1549\]; 1-ha median = 339, range = \[297, 400\]).
-A complete description of the demographic methods, data, and analyses to date can be found in [@brunaDemographyUnderstoryHerb2023].\
+acuminata* begin flowering early in the rainy season (e.g., January) and most reproductive plants produce a single infloresence (range = 1--7) with 20--25 flowers [@bruna_habitat_2002].
+Fruits mature April-May and have 1--3 seeds per fruit ($\bar{x}=2$) that are dispersed by a thrush and several species of manakin [@uriarte_disentangling_2011].
+Dispersed seeds germinate approximately 6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@bruna_habitat_2002; @bruna_seed_1999].
+On average olots in CF also had more than twice as many plants as the plots in 1-ha fragments (CF median = 788, range = (201-1549); 1-ha median = 339, range = (297-400)).\
 
@@
 
-In preliminary investigation, we found that the survival and growth of plants was better explained by treating seedlings and mature plants separately.
-Seedlings are physiologically different from small plants because they necessarily lack the underground reserves (of carbohydrates and meristems) that a small, mature plant may have.
-Therefore, we used general IPMs to model population dynamics with seedlings treated as a separate discreet class not structured by size.
-General IPMs allow for combinations of continuous and discrete states and transitions between them [@ellnerDatadrivenModellingStructured2016].\
+We compared three classes of IPMs - Deterministic, Stochastic, and Stochastic with lagged effects of precipitation - each of which required different functional forms of their underlying vital rates models (@fig-lifecycle).
+All of the models were density-independent.
 
-We built three classes of IPMs for comparison which each required different functional forms of their underlying vital rates models.
-The simplest IPM was a general, density-independent, deterministic IPM with four sub-kernels: growth and survival ($P$, @eq-P), fecundity ($F$, i.e.
-production of new seedlings, @eq-F), probability of staying a seedling (always 0), and recruitment ($R$, i.e.
-seedling survival and establishment, @eq-R) ([@eq-mature; @eq-sdlg], @fig-lifecycle).
-The probability of staying a seedling, was always equal to zero, since our definition of seedlings was first year plants only.\
+**(1) Deterministic IPM:** In this model, which served as the foundation for more complex models, the number and size of mature plants in year $t+1$ is determined by sub-kernels describing the survival and growth (or regression) of mature plants from year $t$ to $t+1$, the number of seedlings establishing in year $t$ that survive to $t+1$ (@eq-mature), and the number of new seedlings that enter the population (@eq-sdlg).
 
@@
 
-$$
-R(z^\prime) = s_sG_s(z^\prime)
-$$ {#eq-R}
+In @eq-mature the sub-kernel $P(z^\prime, z)$ describes the size-dependent survival and growth or regression ($s(z)$ and $G(z^\prime, z)$, respectively) of mature plants from one year to the next ([@eq-P]).
+Sub-kernel $R(z^\prime)$ describes the survival and growth of seedlings that established in year $t$ to year $t+1$ ([@eq-R]); note that these values are not size-dependent - in all our IPMs 'seedlings' are a discrete, size-independent category with a probability $s_s$ of survival for one year following germination and establishment, with the size at which they transition into the 'mature' plant category given by $G(z^\prime, z)$.
+IPMs can include transitions between continuous and discrete states [@ellnerDatadrivenModellingStructured2016]); we treat seedlings as a distinct category because they have lower survival and growth in their first year than comparably sized plants [@bruna_are_2003; @scottDelayedEffectsClimate2022].
 
@@
 $$
+R(z^\prime) = s_sG_s(z^\prime)
+$$ {#eq-R}
+
+The number of new seedlings entering the population, $n_s(t+1)$ ([@eq-sdlg]), is described by the fecundity sub-kernel ([@eq-F]).
+Both the probability that a mature plant will flower, $p_f(z)$, and the number of seeds they produce, $f(z)$, are size dependent.
+These seeds germinate and establish as seedlings with probability $g$.
+
+$$
 F(z) = p_f(z)f(z)g
@@
 
-The number and size of mature plants in the next census is determined by seedlings entering the mature plant population (i.e.
-recruitment <!--# *Is* this recruitment?
--->) and survival and growth (or regression) of mature plants ([@eq-mature]).
-Seedlings survive ($s_s$) and grow into mature plants of a particular size ($G_s(z^\prime)$) ([@eq-R]).\
-
-Mature plants survive as a function of size ($s(z)$), and grow (or regress) to a new size as a function of their previous size ($G(z^\prime, z)$) ([@eq-P]).
-Mature plants flower with a probability that is a function of size ($p_f(z)$) and produce a number of seeds as a function of size ($f(z)$), which germinate and establish as seedlings with probability $g$ ([@eq-F]).\
-
-Vital rate models for growth ($G_s(z^\prime)$ and $G(z^\prime, z)$) , survival ($s_s$ and $s(z)$), and flowering ($p_f(z)$) were fit using the long term demographic dataset.
-For established plants, these three vital rates were modeled as a smooth function of size in the previous census using generalized additive models (GAMs) fit with the `mgcv` package[@mgcv] in `r version$version.string` [@rcoreteam2020] <!--# probably include a supplement with detail on things like basis, number of knots, etc.
+Vital rate models for growth ($G_s(z^\prime)$ and $G(z^\prime, z)$), survival ($s_s$ and $s(z)$), and flowering ($p_f(z)$) were fit using the long-term demographic data [@bruna_are_2003].
+For established plants, these three vital rates were modeled as a smooth function of size in the previous census using generalized additive models (GAMs) fit with the `mgcv` package [@mgcv] in `r version$version.string` [@rcoreteam2020]<!--# probably include a supplement with detail on things like basis, number of knots, etc.
 -->.
@@
 intercept only models).
-For growth models ($G_s(z^\prime)$ and $G(z^\prime, z)$) a scaled t family distribution provided a better fit to the data than a gaussian fit as the residuals were leptokurtic with a simple Gaussian model.\
+For growth models ($G_s(z^\prime)$ and $G(z^\prime, z)$) a scaled t family distribution provided a better fit to the data than a gaussian fit as the residuals were leptokurtic with a simple Gaussian model.
+Size-specific fecundity was based on surveys of fruits per flowering plant [@brunaLeafNumberLeaf2021] and seeds per fruit [@brunaDatasetHeliconiaAcuminata2014] (together $f(z)$) and experimentally derived estimates of germination and establishment in forest and fragments [@bruna_effects_2002; @bruna_seed_1999] <!--# Emilio: add more info here -->.\
 
-To estimate reproduction we drew on additional data sources to estimate the number of fruits per flowering plant as a function of plant size and the number of seeds per fruit (together $f(z)$).
-Germination and establishment rates in continuous forest and forest fragments were estimated using data from ....
-<!--# Emilio: add more info here -->.\
-
-To build the general, density-independent, stochastic, kernel-resampled IPMs, we included environmental stochasticity in all vital rate models built using the long term demographic datset by adding a random effect of year (@fig-lifecycle).
+**(2) Stochastic IPM:** To build the stochastic, kernel-resampled IPMs we included environmental stochasticity in all vital rate models built using the long term demographic datset by adding a random effect of year (@fig-lifecycle).
 The random effect of year was included using a factor--smooth interaction which allowed the relationship between plant size and vital rates to vary in functional form among transition years.
@@
 
-For the third method, we modeled the impacts of drought on vital rates explicitly and created general, density-independent, stochastic, parameter-resampled IPMs (*sensu* @metcalfStatisticalModellingAnnual2015).
+**(3) Stochastic IPM with lagged effects of precipitation on vital rates:** In our The stochastic, parameter-resampled IPMs the impacts of precipitation extremes on vital rates were modeled explicitly (*sensu* @metcalfStatisticalModellingAnnual2015).
 We calculated the standardized precipitation evapotranspiraton index (SPEI) for our site using a published gridded dataset based on ground measurements [@xavierDailyGriddedMeteorological2016] as described in @scottDelayedEffectsClimate2022.
@@
 
+#TODO: need to add something about the comparison of population structure (figure)
+
 This workflow was managed using the `targets` R package [@targets] which also allowed us to track computational time spent on each IPM for comparison.\
@@
 
-> All about the stats.
+Comparison of CF vs FF
+
+-   For det need to bootstrap
+
+-   for within stoch and lag: glm/t-test.
 
@@
 
-2.  Population growth rates were consistently higher in continuous forest compared to forest fragments across IPM types (@tbl-lambdas).\
+2.  Population growth rates in CF were significantly higher than in forest fragments for all IPM types (@tbl-lambdas).\
 
-3.  We found that the the choice of Integral Projection Model didn't change the relative ranking of lambda in Continous Forest and Fragments.\
+3.  Deterministic and stocuhastic-kernal similar means.
 
-4.
-DLNM models take much, much longer to iterate: while the Deterministic and Kernel-resampled stochastic models took \~`r det_time` and \~`r sker_time` min to iterate (respectively), the Parameter-resampled stochastic models with lagged effects took \~`r spar_time` min.\
+4.  These results are consistent with those of @kayeEffectStochasticTechnique2003, who found who found that the method effected stochastic lambda but relative ranking of populations was consistent.
 
 5.
-The greater use of computational resources is likely a result of `predict()` being much slower for GAMs with 2D smooths because the number of knots is much higher compared to the GAMs used for the vital rates models in the determinsitic and kernel-resampled IPMs.\
+BUT lag dropped lambdas considerably.
+In fact, took both well below zero.
+This is a big, big difference.
+If you are doing conservation and management, a growth rate lower by 5-6 % would really rock your world.
 
-<!--# also note that this can't really be paralellized because each iteration depends on the previous one.
-There's *maybe* some potential speed gains in using multiple cores with predict.bam(), but unlikely to be substantial.
--->
+### Population Structure
 
@@
 
-> -   These results are consistent with those of @kayeEffectStochasticTechnique2003, who found who found that the method effected stochastic lambda but relative ranking of populations was consistent.
+### Time and Effort
+
+4.
+DLNM models take much, much longer to iterate: while the Deterministic and Kernel-resampled stochastic models took \~`r det_time` and \~`r sker_time` min to iterate (respectively), the Parameter-resampled stochastic models with lagged effects took \~`r spar_time` min.\
+
+5.
+The greater use of computational resources is likely a result of `predict()` being much slower for GAMs with 2D smooths because the number of knots is much higher compared to the GAMs used for the vital rates models in the determinsitic and kernel-resampled IPMs.\
+
+<!--# also note that this can't really be paralellized because each iteration depends on the previous one.
+There's *maybe* some potential speed gains in using multiple cores with predict.bam(), but unlikely to be substantial.
+-->
 
@@
 
-Data and R code used in this study are archived with Zenodo at <doi url>.
+Data and R code used in this study are archived with Zenodo at *(doi and url to be added on acceptance)*.
 
@@
 <!-- ``` -->
+
+\newpage
 

### 2025-04-23 docs/ipm_comparison_ms.qmd
updating plots and ms

@@
 forest fragments).
-Based on previous studies [@bruna_are_2003; @bruna_demographic_2005; @bruna_experimental_2002; @bruna_habitat_2002] and demographic theory [@caswell2001; @tuljapurkar_population_1990] we predicted that:\
-
-(i) projections of $\lambda$ would be lower in forest fragments than in continuous forest for all model types,
-(ii) projections of $\lambda$ from deterministic models would be higher than those of stochastic models,
-(iii) projections of $\lambda$ would be lowest for models with lagged effects, and
-(iv) populations would be more skewed towards pre-reproductive size classes in fragments that forest, regardless of model structure
+Based on previous studies [@bruna_are_2003; @bruna_demographic_2005; @bruna_experimental_2002; @bruna_habitat_2002] and demographic theory [@caswell2001; @tuljapurkar_population_1990] we predicted that: (i) projections of $\lambda$ from deterministic models would be higher than those of stochastic models, and that (ii) for all IPMs the projections of $\lambda$ forest fragments would be lower than for continuous forest, but that the difference would be proportionately greater for IPM with lagged effects.\
 
@@
 Dispersed seeds germinate approximately 6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@bruna_habitat_2002; @bruna_seed_1999].
-On average olots in CF also had more than twice as many plants as the plots in 1-ha fragments (CF median = 788, range = (201-1549); 1-ha median = 339, range = (297-400)).\
+On average olots in CF also had more than twice as many plants as the plots in 1-ha fragments (CF median = 788, range = (201-1549); 1-ha median = 339, range = (297-400)).
 
@@
 
-We compared three classes of IPMs - Deterministic, Stochastic, and Stochastic with lagged effects of precipitation - each of which required different functional forms of their underlying vital rates models (@fig-lifecycle).
-All of the models were density-independent.
+We projected the growth rate and structure of *Heliconia acuminata* populations in Continuous Forest and Forest Fragments with three classes of IPMs - Deterministic, Stochastic, and Stochastic with Lagged Environmental Effects.
+Each of these IPMs required different functional forms of the underlying vital rate functions used to describe the *H.
+acuminata* life cycle (@fig-lifecycle).
+All models were density-independent, with the deterministic model serving as the foundation for the more complex models.\
 
-**(1) Deterministic IPM:** In this model, which served as the foundation for more complex models, the number and size of mature plants in year $t+1$ is determined by sub-kernels describing the survival and growth (or regression) of mature plants from year $t$ to $t+1$, the number of seedlings establishing in year $t$ that survive to $t+1$ (@eq-mature), and the number of new seedlings that enter the population (@eq-sdlg).
+**(1) Deterministic IPM:** In this model the size and structure of a population in year $t+1$ is determined by the survival and growth of plants alive in year $t$ (@eq-mature) plus the number of new seedlings that entered the population (@eq-sdlg).
 
@@
 
-In @eq-mature the sub-kernel $P(z^\prime, z)$ describes the size-dependent survival and growth or regression ($s(z)$ and $G(z^\prime, z)$, respectively) of mature plants from one year to the next ([@eq-P]).
-Sub-kernel $R(z^\prime)$ describes the survival and growth of seedlings that established in year $t$ to year $t+1$ ([@eq-R]); note that these values are not size-dependent - in all our IPMs 'seedlings' are a discrete, size-independent category with a probability $s_s$ of survival for one year following germination and establishment, with the size at which they transition into the 'mature' plant category given by $G(z^\prime, z)$.
-IPMs can include transitions between continuous and discrete states [@ellnerDatadrivenModellingStructured2016]); we treat seedlings as a distinct category because they have lower survival and growth in their first year than comparably sized plants [@bruna_are_2003; @scottDelayedEffectsClimate2022].
+<!-- The number and size of mature plants in year $t+1$ is determined by sub-kernels describing the survival and growth (or regression) of mature plants from year $t$ to $t+1$, the number of seedlings establishing in year $t$ that survive to $t+1$ (@eq-mature), and the number of new seedlings that enter the population (@eq-sdlg).\ -->
+
+@eq-mature has two components. The first is the sub-kernel $P(z^\prime, z)$, which describes the size-dependent survival and growth/regression of mature plants (@eq-P):
 
@@
 
+The second is sub-kernel $R(z^\prime)$, which describes the survival of seedlings established in year $t$ and their size when entering the mature plant population in year $t+1$ (@eq-R):
+
 $$
@@
 
-The number of new seedlings entering the population, $n_s(t+1)$ ([@eq-sdlg]), is described by the fecundity sub-kernel ([@eq-F]).
-Both the probability that a mature plant will flower, $p_f(z)$, and the number of seeds they produce, $f(z)$, are size dependent.
-These seeds germinate and establish as seedlings with probability $g$.
+Note that in @eq-R both the probability that new seedlings survive their first year, $s_s$, and their size at the end of this year, $G(z^\prime, z)$, are size-independent.
+IPMs can include transitions between individuals from a discrete state to a continuous one [i.e., from 'seedling' to 'mature plant of size $z'$', @ellnerDatadrivenModellingStructured2016]; we treat seedlings as a distinct and discrete category because they have lower survival and growth in their first year than comparably sized plants [@bruna_are_2003; @scottDelayedEffectsClimate2022].\
 
-$$
+The number of new seedlings entering the population in year $t+1$ is a function of the number of mature plants in year $t$ and a sub-kernel describing the size-dependent fecundity of these individuals (@eq-F): $$
 F(z) = p_f(z)f(z)g
@@
 
-Vital rate models for growth ($G_s(z^\prime)$ and $G(z^\prime, z)$), survival ($s_s$ and $s(z)$), and flowering ($p_f(z)$) were fit using the long-term demographic data [@bruna_are_2003].
-For established plants, these three vital rates were modeled as a smooth function of size in the previous census using generalized additive models (GAMs) fit with the `mgcv` package [@mgcv] in `r version$version.string` [@rcoreteam2020]<!--# probably include a supplement with detail on things like basis, number of knots, etc.
--->.
-For consistency, seedling survival and growth were also modeled using GAMs, but without size in the previous census as a predictor (i.e.
-intercept only models).
-For growth models ($G_s(z^\prime)$ and $G(z^\prime, z)$) a scaled t family distribution provided a better fit to the data than a gaussian fit as the residuals were leptokurtic with a simple Gaussian model.
-Size-specific fecundity was based on surveys of fruits per flowering plant [@brunaLeafNumberLeaf2021] and seeds per fruit [@brunaDatasetHeliconiaAcuminata2014] (together $f(z)$) and experimentally derived estimates of germination and establishment in forest and fragments [@bruna_effects_2002; @bruna_seed_1999] <!--# Emilio: add more info here -->.\
+Both the probability that a mature plant will flower, $p_f(z)$, and the number of seeds a flowering plant will produce, $f(z)$, are size-dependent.
+All seeds germinate and establish as seedlings with probability $g$.\
+<!-- `r version$version.string`  -->
 
-**(2) Stochastic IPM:** To build the stochastic, kernel-resampled IPMs we included environmental stochasticity in all vital rate models built using the long term demographic datset by adding a random effect of year (@fig-lifecycle).
-The random effect of year was included using a factor--smooth interaction which allowed the relationship between plant size and vital rates to vary in functional form among transition years.
-The kernel-resampling approach is to generate kernels corresponding to each transition year in the demographic dataset using the random smooths for year, and to iterate the IPM by drawing from these randomly.
-This is equivalent to the matrix selection approach for matrix population models described by @caswell2001.\
+We used the annual census data [@bruna_are_2003] to fit the deterministic vital rate functions for growth, survival, and flowering in each habitat type (i.e., Fragments, Continuous Forest; the data from all plots within a habitat class were pooled to create a single 'summary population' [@bruna_are_2003].
+For established plants these were modeled as a smooth function of size in the previous census with generalized additive models (GAMs) fit with the `mgcv` library [@mgcv] for the R statistical programming language [@rcoreteam2020].<!--`r version$version.string`---> <!--# probably include a supplement with detail on things like basis, number of knots, etc.
+--> For consistency, seedling survival and growth were also modeled using GAMs, but without size in the previous census as a predictor (i.e.
+'intercept-only' models).
+For growth models a scaled t family distribution provided a better fit to the data than a Gaussian fit, as the residuals with a simple Gaussian model were Leptokurtic.
+To model size-specific fecundity we used data on the number of fruits per flowering plant [@brunaLeafNumberLeaf2021] and seeds per fruit [@brunaDatasetHeliconiaAcuminata2014] that were collected outside of the demographic plots (to avoid altering within-plot recruitment) and experimentally-derived estimates of seed germination and seedling establishment [@bruna_effects_2002; @bruna_seed_1999] <!--# Emilio: add more info here -->.\
 
-**(3) Stochastic IPM with lagged effects of precipitation on vital rates:** In our The stochastic, parameter-resampled IPMs the impacts of precipitation extremes on vital rates were modeled explicitly (*sensu* @metcalfStatisticalModellingAnnual2015).
-We calculated the standardized precipitation evapotranspiraton index (SPEI) for our site using a published gridded dataset based on ground measurements [@xavierDailyGriddedMeteorological2016] as described in @scottDelayedEffectsClimate2022.
-For all vital rate models fit using the long term demographic dataset, we modeled delayed effects of SPEI using distributed lag non-linear models with a maximum lag of 36 months [@scottDelayedEffectsClimate2022] (@fig-lifecycle).
-To iterate these parameter-resampled IPMs, a random sequence of SPEI values was created by sampling years of the observed monthly SPEI data.
-Then, 36 month lags are calculated for each year starting in February (the month of the demographic census).
-These values are then used to predict fitted values from the vital rates models, generating different kernels at each iteration of the IPM.
-With this method, the kernels of successive iterations are not entirely independent because the SPEI values used in calculating vital rates include values used in the previous two iterations, but they are ergodic.\
+**(2) Stochastic IPMs:** To include temporal stochasticity in our IPMs we included a random effect of year.
+This was done using a factor--smooth interaction that allowed the functional form of the relationship between plant size and vital rates to vary among transition years (@fig-lifecycle).
+We generated kernels for every transition year using the long-term survey data [@bruna_demography_2023] and random smooths for year.
+We then randomly selected one of these sets of kernels to use in each iteration of the IPM.
+This procedure is equivalent to *'kernel resampling'* [*sensu* @metcalfStatisticalModellingAnnual2015] or matrix selection for matrix population models [@caswell2001].\
+
+**(3) Stochastic IPMs with lagged effects of precipitation on vital rates:** We explicit modeled the lagged effects of precipitation extremes on vital rates using the procedure described in @scottDelayedEffectsClimate2022.
+Briefly, we first calculated the Standardized Precipitation Evapotranspiraton Index (i.e., SPEI) for our study site using a published gridded dataset based on ground measurements [@xavierDailyGriddedMeteorological2016].
+After we fit vital rate models using the long-term survey data (@fig-lifecycle), we modeled delayed effects of SPEI with Distributed Lag Non-linear Models (i.e., DLNMs) with a maximum lag of 36 months [@scottDelayedEffectsClimate2022].
+To iterate these parameter-resampled IPMs [*sensu* @metcalfStatisticalModellingAnnual2015] we first created a random sequence of SPEI values by sampling years of the observed (monthly) SPEI data.
+For every year we then calculated a lag of 36 months from the month in which that year's census was completed.
+These values were then used to predict the fitted values from vital rate models, which generated different kernels for each iteration of the IPM.
+The kernels of successive iterations are not entirely independent -- the SPEI values used to calculate vital rates include values used in the previous two iterations -- they are ergodic.\
 
@@
 
-All IPMs were constructed and iterated <!--# optimized in notes/optimizing-meshpoints.Rmd which could be turned into supplemental -->using the `ipmr` package in R [@levinIpmrFlexibleImplementation2021].
+All IPMs were constructed and iterated <!--# optimized in notes/optimizing-meshpoints.Rmd which could be turned into supplemental -->using the `ipmr` package [@levinIpmrFlexibleImplementation2021] for the R statistical programming language [@rcoreteam2020].
 The IPMs used 100 meshpoints and the midpoint rule for calculating kernels <!--# citation?
 -->.
-For each type of IPM we iterated the model for 1000 time steps, discarding the first 100 time steps to omit transient effects.
-Stochastic growth rates ($\lambda_s$) were calculated as the average $ln(\lambda)$ from each time step [@caswell2001] and back-transformed to be on the same scale as deterministic lambdas for comparison.
-We used the distribution of established plant sizes and proportion of seedlings from the full dataset as a starting population vector.
-While other starting population vectors were possible, the choice is of little importance as it will only impact transient dynamics, which we aren't interested in for this study.\
-
-To estimate uncertainty around the per-capita growth rates (lambdas), we created 500 bootstraps of the demographic dataset by sampling individual plants with replacement within each habitat.
-For each bootstrap, we then re-fit vital rates models (all except germination and establishment rate, fruits per flowering plant, and seeds per fruit, which were estimated using different datasets), constructed IPMs, and calculated a value for lambda as described above.
-We then used these bootstraped estimates of lambda to calculate bias corrected 95% confidence intervals [@ellnerDatadrivenModellingStructured2016].\
-
-#TODO: need to add something about the comparison of population structure (figure)
-
-This workflow was managed using the `targets` R package [@targets] which also allowed us to track computational time spent on each IPM for comparison.\
+For each type of IPM we iterated the model for 1000 time steps, but discarded the first 100 time steps to omit transient effects.
+Stochastic growth rates ($\lambda_s$) were calculated as the average $ln(\lambda)$ from each time step [@caswell2001] and then back-transformed to allow for direct comparison with projections of $\lambda$ from deterministic models.
+The initial starting vector primarily influences a population's transient dynamics; we therefore used the distribution of established plant sizes and proportion of seedlings in the full demographic data set as the initial population vector for all models.
+Finally, we estimated the 95% confidence intervals for each IPMs projections of $\lambda$.
+To do so we first created 500 populations for each habitat type by sampling individual plants with replacement (i.e., bootstrapping) until the population size of each matched that of the initial population vector.
+We then re-fit the vital rate models for growth, survival, and flowering for each of these bootstrapped population and constructed new IPMs for each population as above (the models for germination and establishment rate, fruits per flowering plant, and seeds per fruit were not refit because these were estimated using different data sets).
+The projections of $\lambda$ for the new populations were then used to estimate the upper and lower 95% bias-corrected percentile intervals [@manly2018randomization; @caswell2001].
+This workflow was managed using the `targets` R package [@targets], which also allowed us to track computational time spent processing and analyzing each class of IPM.\
 
@@
 
-Comparison of CF vs FF
+##### Comparison of CF vs FF
 
@@
 -   for within stoch and lag: glm/t-test.
+
+-   Comparison of pop structure in figure (test for increasing variance with variance / F or permutation test).
+Villase∼nor, J.
+A., & González-Estrada, E.
+(2024).
+A nonparametric test for homogeneity of variances.
+Communications in Statistics - Theory and Methods, 54(2), 646–654.
+https://doi.org/10.1080/03610926.2024.2316274 OR https://www.datanovia.com/en/lessons/homogeneity-of-variance-test-in-r/#google_vignette
 
@@
 
+### Lambda Comparison
+
+1.  We found that including Delayed Life-History Events in Intergral Projection Models led to significantly lower projections of $\lambda$
+2.
+It is reassuring that all IPM models projected higher values of $\lambda$ in continuous forest than forest fragments (@tbl-lambdas); this is consistent with the results of previous studies comparing the relative rankings of projections from stochastic and deterministic models [e.g., @kayeEffectStochasticTechnique2003.]
+3.
+However, the underestimating $\lambda$ by 2-3% difference in $\lambda$ resulting projections from the different IPMs could have major implications for many of the contexts in which IPMs are brought to bear.
+4.  If you are doing conservation and management, a growth rate lower by 5-6 % would really rock your world.
+5.  For instance, underestimating $\lambda$ by failing to include lagged effects could be especially severe in studies of climate change, given many DLHE are climate driven.\
+    <!-- CF: deterministic is > than 70% of dnlm --> <!-- FF: deterministic is > than 76% of dnlm -->
+
+### Pop Structure Comparison
+
 1.
-For all vital rates estimated using the long term demographic dataset, the DLNM model fit the best (dAIC = 0) followed by the model with a random effect of year, followed by the deterministic model (@tbl-aic).\
+Including lagged effects in IPMs also resulted in more variable projections of future population structure (@fig-pop-states a), especially in forest fragments where we have previously shown lagged effects have very large impacts on growth and survival [@scottDelayedEffectsClimate2022].
+2.  Consequently, projections of population structure were also far less less consistent across habitats (@fig-pop-states b).
+3.  Finally, the choice of IPM also resulted in some subtle but notable differences in projected population structure.
+4.  For example, the deterministic IPM projected the population in Continuous Forest had slightly more of the smallest and largest plants than the population in Forest Fragments.
+5.  In other words, populations in forest fragments had proportionatly less recruitment and fewer individuals growing into the larger and reproductivce size classes [@bruna_demographic_2005].
+6.  The shift towards intermediately sized, pre-reproductive plants in forest fragments is even more dramatic when using the parameter-resampled stochastic IPM (@fig-pop-states b).\
+
+### So...should we care?
+
+1.  Including delayed life-history events in demographic models has important implications both projections of population growth rate and population structure.
+2.
+These projections may be more accurrate - the underlying vital rate models for IPMS with Lagged effects all had the best fit to the survey data (dAIC = 0, @tbl-aic).
 <!--# No longer true if random effect = different smooth for each year -->
-
-2.  Population growth rates in CF were significantly higher than in forest fragments for all IPM types (@tbl-lambdas).\
-
-3.  Deterministic and stocuhastic-kernal similar means.
-
-4.  These results are consistent with those of @kayeEffectStochasticTechnique2003, who found who found that the method effected stochastic lambda but relative ranking of populations was consistent.
-
+3.  While this may not be critical in situations where relative differences allow for hypothesis testing, it may be especially important in conservation or manaagement settings.
+4.  Including lagged effects comes at a cost, however.
 5.
-BUT lag dropped lambdas considerably.
-In fact, took both well below zero.
-This is a big, big difference.
-If you are doing conservation and management, a growth rate lower by 5-6 % would really rock your world.
-
-### Population Structure
-
-6.  @fig-pop-states has some interesting things in it:
-
-> -   For the deterministic IPM (and the kernel-resampled IPM?) there are slightly more of the smallest plants and the largest plants in CF compared to FF (i.e. more medium sized plants in FF).
-
-> -   For the kernel-resampled IPM (random effect of year), the fluctuations are extremely similar between CF and FF
-
-> -   For the parameter-resampled IPM (DLNM) the size structure of the population is a LOT more variable in FF. This makes sense as we know lagged effects are more important in fragments.
-
-> -   Also, the fluctuations in size structure in CF do not match the fluctuations in FF as well (can see this by the increased spread of points in @fig-pop-states (B)
-
-> -   Also, in the parameter-resampled IPM (and only in this one), we see a shift toward smaller plants in FF compared to CF
-
-### Time and Effort
-
-4.
-DLNM models take much, much longer to iterate: while the Deterministic and Kernel-resampled stochastic models took \~`r det_time` and \~`r sker_time` min to iterate (respectively), the Parameter-resampled stochastic models with lagged effects took \~`r spar_time` min.\
-
-5.
-The greater use of computational resources is likely a result of `predict()` being much slower for GAMs with 2D smooths because the number of knots is much higher compared to the GAMs used for the vital rates models in the determinsitic and kernel-resampled IPMs.\
+While Deterministic and Kernel-resampled Stochastic IPMs took only \~`r det_time` and \~`r sker_time` min to iterate (respectively), the Stochastic IPMs with using parameter-resampled kernels and lagged effects took \~`r spar_time` min.
+6.
+This is largely due to the required computational resources and algorithms (i.e., `predict()`), which are much slower forGeneral Additive Models (i.e., GAMs) with 2D smooths because of the much higher number of knots than for the GAMs used in Determinsitic and Kernel-resampled IPMs.
+7.  Advances in computational power and access to high-performance computing resources could lower this cost.\
+8.  However, computational power cannot compensate for limited data.
+9.
+Detecting lagged effects and evaluating their consequences requires long-term demographic data - data that are only available for a relatively small number of species, few of which are in the tropics.
+10.
+The increasing evidence that lagged effects are ubiquitous, and that they can have major demographic impacts, underscores the need to support the collection of such long-term data, the complementary development of experimental and statistical approaches to disentangling lagged effects, and community driven efforts to identify priority or model systems for in-depth investigation.\
 
@@
 
-We thank **,** \_, \_\_\_ and \_\_\_ anonymous reviewers for helpful discussions and comments on the manuscript.
-We thank Sam Levin for his help with the `ipmr` package.
-Financial support was provided by the U.S.
-National Science Foundation (awards \_\_\_\_, and \_\_\_\_).
+We thank \_\_\_, \_\_\_, and \_\_\_ anonymous reviewers for helpful discussions and comments on the manuscript.
+We also thank Sam Levin for his help with the `ipmr` package.
+Financial support for collecting the demographic data used in this study was provided by the U.S.
+National Science Foundation (awards INT 98-06351, DEB-0309819, DEB-0614149,DEB-0614339, and DEB-1948607).
 This article is publication no.
--- -- in the BDFFP Technical series.
+\_\_\_ in the BDFFP Technical series.
 The authors declare no conflicts of interest.
@@
 ERS contributed to the conceptualization, methodology, formal analysis, and led the writing of the original draft.
-EMB contributed to the conceptualization, methodology, and writing and also acquired funding.
+EMB contributed to the conceptualization, methodology, writing, and, acquired funding.
 
@@
 
+<!-- # Figures -->
+
+\pagebreak
+
+```{r #fig-lifecycle,  echo=FALSE, out.width="85%",fig.align='center'}
+#| fig.cap = "Life cycle diagram of *Heliconia acuminata*.
+Each transition is associated with an equation for a vital rate function.
+The functions shown on the diagram correspond to those used to construct a general, density-independent, deterministic IPM.
+The table below shows the equivalent equations for stochastic, kernel-resampled IPMs and stochastic, parameter-resampled IPMs."
+
+# Draft of this figure is in docs/figures/Heliconia lifecycle DAG.key
+path <- here("docs/figures/lifecycle.png")
+knitr::include_graphics(path, rel_path = FALSE)
+```
+
+\pagebreak
+
+```{r #fig-pop-states,  echo=FALSE, out.width="85%",fig.align='center'}
+#| fig.cap = "**(A)** The proportion of _Heliconia acuminata_ populations in different size/stage classes when simulating population dyammics with three different Integral Projection Models.
+Results are shown for the first 250 iterations of populations in both in Continuous Forest (CF) and Forest Fragments (FF); for the criteria used to define the size categories see @tbl-plantcats.
+**(B)** For each interation, the relative proportion of the population in each size class (FF vs CF).
+Note that this excludes transient dynamics (iterations 1-30).
+Values on the 1-1 line indicate an iteration where CF and FF have the same proportion of the population in a given size class."
+# note flowering prob for rp1 <0.25 but for rp2>0.2???'
+# include _seedlings_ (a discrete category in the IPMs), _pre-reproductive 1_ (log(size) = 0–2.5, average survival < 0.9, probability of flowering \\approx 0), _pre-reproductive 2_ (log(size) = 2.5–4.5, average survival probabilty > 0.8, probability of flowering \\approx 0), _reproductive 1_ (log(size) = 4.5–6, average survival probability >0.95, flowering probability < 0.25), and _reproductive 2_ (log(size) = 6+, average survival probability >0.95, flowering probability > 0.2)
+knitr::include_graphics(here(fig_pop_states), rel_path = FALSE)
+```
+
+\pagebreak
+
+```{r #fig-lambdas,  echo=FALSE, out.width="85%",fig.align='center'}
+#| fig.cap = "Distribution of 900 values of \u03BB projected with (A) Stochastic, kernel-resampled IPMs and (B) Stochastic, paramenter-resampled IPMs.
+IPMs were used to project \u03BB for both Continuous Forest (green) and Forest Fragments (gray).
+The solid line indicates the mean value, the dashed line indicates the value of \u03BB $\\lambda$ in that habitat projected with Deterministic IPMs."
+
+# The figures is currently generated with "plotting_lambdas.R, need to move to targets_r 
+path <- here("docs/figures/lambdas_fig.png")
+knitr::include_graphics(path, rel_path = FALSE)
+```
+
+\pagebreak
+
+```{r #fig-var,  echo=FALSE, out.width="85%",fig.align='center'}
+#| fig.cap = "------"
+
+# The figures is currently generated with "plotting_lambdas.R, need to move to targets_r 
+path <- here("docs/figures/relative_variances_fig.png")
+knitr::include_graphics(path, rel_path = FALSE)
+```
+
+\pagebreak
+
+```{r #fig-marginal,  echo=FALSE, out.width="85%",fig.align='center'}
+#| fig.cap = "------"
+
+# The figures is currently generated with "plotting_lambdas.R, need to move to targets_r 
+path <- here("docs/figures/marginal_stage_fig.png")
+knitr::include_graphics(path, rel_path = FALSE)
+```
+
 \pagebreak
@@
 #| 
-#TODO: only show first instance of habitat and vital rate columns, not on every row.  Alternatively do something fancy with flextable or gt.
-
 aic_tbl_df %>%
@@
   arrange(habitat, model) %>% 
+  mutate(habitat=c(rep("Continuous Forest",15),rep("Forest Fragments",15))) %>% 
+  mutate(model=(rep(c("Survival","","","Growth","","","Flowering","","","Seedling survival","","","Seedling growth","",""),2)))  %>%
   select(
@@
   )  %>% 
+  mutate(Habitat = str_replace_all(
+    Habitat,
+    c(
+      "Continuous Forest" = "",
+      "Forest Fragments" = ""
+      )
+  )) %>% 
+  kable(digits = 2,
+  format = "latex",
+  align = "cllcc",
+  escape = FALSE,
+  row.names = FALSE,
+  booktabs = T,
+  linesep = ""
+) %>%
+  kable_styling(
+    bootstrap_options = c("hover"),
+    # full_width = F,
+    latex_options = c("scale_down"),
+    # latex_options = c("hold_position"),
+    font_size = 12,
+    position = "center"
+  ) %>% 
+# %>% 
+  # collapse_rows(columns = 1:2, valign = "top")
+  column_spec(1, width = "10em") %>% 
+  pack_rows("Continuous Forest", 1, 15) %>%
+  pack_rows("Forest Fragments", 16, 30)
 
-pandoc.table(
-  justify = c( "left", "left","left", "right",  "right"),
-  full_width = FALSE
-  # ,
-  # caption = Comparison of vital rate models used to build IPM.
+# pandoc.table(
+#   justify = c( "left", "left","left", "right",  "right"),
+#   full_width = FALSE
+#   # ,
+#   # caption = Comparison of vital rate models used to build IPM.
 The 'Effect of Environment' column describes how environmental effects were included in models.
@@
 \u0394AIC within 2 indicates models are equivalent."
-)
+# )
+
 ```
 
-\newpage
+\pagebreak
 
@@
       "stoch" = "Stochastic, kernel-resampled",
-      "DLNM" = "Stochastic, parameter-resampled"
+      "dlnm" = "Stochastic, parameter-resampled"
     )
@@
   select(IPM = ipm, Habitat = habitat, "$\\lambda$" = lambda) %>% 
-pandoc.table(
-  digits = 5,
-  keep.trailing.zeros = TRUE,
-  justify = c("left", "left", "right"),
-  full_width = FALSE
+  mutate(IPM = str_replace_all(
+    IPM,
+    c(
+      "Deterministic" = "",
+      "Stochastic, kernel-resampled" = "",
+      "Stochastic, parameter-resampled" = ""
+      )
+  )) %>% 
+  kable(digits = 2,
+  format = "latex",
+  align = "lcc",
+  escape = FALSE,
+  row.names = FALSE,
+  booktabs = T,
+  linesep = ""
+) %>% 
+  kable_styling(
+    bootstrap_options = c("hover"),
+    # full_width = F,
+    latex_options = c("scale_down"),
+    # latex_options = c("hold_position"),
+    font_size = 12,
+    position = "center"
+  ) %>% 
+  # collapse_rows(columns = 1:2, valign = "top")
+  column_spec(1, width = "15em") %>% 
+  pack_rows("Deterministic", 1, 2) %>%
+  pack_rows("Stochastic, kernel resampled", 3, 4) %>% 
+  pack_rows("Stochastic, parameter-resampled", 5, 5)
+
+# 
+# pandoc.table(
+#   digits = 5,
+#   keep.trailing.zeros = TRUE,
+#   justify = c("left", "center", "center"),
+#   full_width = FALSE
   # ,
   # caption = "(\\#tab-lambdas) Population growth rates for continuous forest (CF) and forest fragments (FF) under different kinds of IPMs with bootstrapped, bias-corrected, 95% confidence intervals."
-)
+# )
 
@@
 
-\newpage
+\pagebreak
 
-<!-- # Figures -->
+```{r tbl-plantcats, results='asis'}
+#| label: tbl-plantcats
+#| tbl-cap: "Size and stage categories used for comparing _Heliconia acuminata_ population structure.
+Note that seedlings are a discrete size class not based on size (see _Methods_ for additional details)."
 
-```{r #fig-lifecycle,  echo=FALSE, out.width="85%",fig.align='center'}
-#| fig.cap = "Lifecycle diagram of *Heliconia acuminata*.
-Each transition is associated with an equation for a vital rate function.
-The functions shown on the diagram correspond to those used to construct a general, density-independent, deterministic IPM.
-The table below shows the equivalent equations for stochastic, kernel-resampled IPMs and stochastic, parameter-resampled IPMs."
+category<-c("seedlings", "pre-reproductive 1", "pre-reproductive 2", "reproductive 1", "reproductive 2")
+size<-c(" - ","0–2.5","2.5-4.5","4.5-6","$\\geq$ 6")
+survival<-c("-","$\\leq$ 0.9","$\\geq$ 0.8","$\\geq$ 0.95","$\\geq$ 0.95")
+flowering<-c("-","$\\approx$ 0","$\\approx$ 0","$\\leq$ 0.25","$\\geq$ 0.2")
+cat_table<-bind_cols(category=category,
+                     `log(size)`=size, 
+                     `avg. prob. survival` = survival,
+                     `prob. flowering` = flowering)
+cat_table %>% 
+kable(digits = 2,
+  format = "latex",
+  align = "lccc",
+  escape = FALSE,
+  row.names = FALSE,
+  booktabs = T,
+  linesep = ""
+) %>% 
+  kable_styling(
+    bootstrap_options = c("hover"),
+    # full_width = F,
+    latex_options = c("scale_down"),
+    # latex_options = c("hold_position"),
+    font_size = 12,
+    position = "center"
+  ) 
+# %>% 
+#   column_spec(1, width = "50em") %>%
 
-# Draft of this figure is in docs/figures/Heliconia lifecycle DAG.key
-path <- here("docs/figures/lifecycle.png")
-knitr::include_graphics(path, rel_path = FALSE)
+# cat_table %>% 
+#   pandoc.table(
+#   digits = 5,
+#   keep.trailing.zeros = TRUE,
+#   justify = c("left", "center", "center", "center"),
+#   full_width = FALSE
+# )
+
 ```
-
-\newpage
-
-```{r #fig-pop-states,  echo=FALSE, out.width="85%",fig.align='center'}
-
-#| fig.cap = "Relative proportions of plant sizes in the first 250 iterations of the IPM simulations.
-Stacked area charts (A) show the relative size/stage distribution of plants in continuous forest (CF, top row) and forest fragments (FF, bottom row) in each of the three IPMs (columns).
-The proportion of each size class in CF and FF for each iteration is shown in B with the first 30 iterations removed to not include transient dynamics.
-A 1:1 line is plotted in black.
-Size categories include seedlings (a discrete category in the IPMs), pre-reproductive 1 (log(size) 0–2.5) that have low average survival (< 0.9) and a near 0 probability of flowering, pre-reproductive 2 (log(size) 2.5–4.5) that have a higher average survival probabilty (> 0.8) and a near 0 probability of flowering, reproductive 1 (log(size) 4.5–6) that have a high average survival probability (>0.95) and a lower flowering probability (< 0.25), and reproductive 2 (log(size) 6+) that have a high average survival probability (>0.95) and higher flowering proability (> 0.2)."
-
-knitr::include_graphics(here(fig_pop_states), rel_path = FALSE)
-```

### 2025-04-23 R/plotting_lambdas.R

@@ -1,0 +1,859 @@
+# extract values of lambda fom targets to make figures
+
+# load libraries ----------------------------------------------------------
+
+
+library(targets)
+library(tidyverse)
+library(RColorBrewer)  
+
+
+# extract data ------------------------------------------------------------
+
+# Deterministic CF
+ipm_det_cf<-tar_read(ipm_det_cf)
+l_det_cf<-ipm_det_cf$pop_state$lambda
+l_det_cf<-as.vector(l_det_cf)
+l_det_cf<-as_tibble(l_det_cf) %>% 
+  slice_tail(n=900) %>% 
+  rename(det_cf=value)
+
+# Deterministic FF
+ipm_det_ff<-tar_read(ipm_det_ff)
+l_det_ff<-ipm_det_ff$pop_state$lambda
+l_det_ff<-as.vector(l_det_ff)
+l_det_ff<-as_tibble(l_det_ff) %>% 
+  slice_tail(n=900) %>% 
+  rename(det_ff=value)
+
+# lagged CF
+ipm_dlnm_cf<-tar_read(ipm_dlnm_cf)
+l_lag_cf<-ipm_dlnm_cf$pop_state$lambda
+l_lag_cf<-as.vector(l_lag_cf)
+l_lag_cf<-as_tibble(l_lag_cf) %>% 
+  slice_tail(n=900) %>% 
+  rename(lag_cf=value)
+
+
+
+# lagged FF
+ipm_dlnm_ff<-tar_read(ipm_dlnm_ff)
+l_lag_ff<-ipm_dlnm_ff$pop_state$lambda
+l_lag_ff<-as.vector(l_lag_ff)
+l_lag_ff<-as_tibble(l_lag_ff) %>% 
+  slice_tail(n=900) %>% 
+  rename(lag_ff=value)
+
+# stochastic CF
+ipm_stoch_cf<-tar_read(ipm_stoch_cf)
+l_stoch_cf<-ipm_stoch_cf$pop_state$lambda
+l_stoch_cf<-as.vector(l_stoch_cf)
+l_stoch_cf<-as_tibble(l_stoch_cf) %>% 
+  slice_tail(n=900) %>% 
+  rename(stoch_cf=value)
+
+# stochastic FF
+ipm_stoch_ff<-tar_read(ipm_stoch_ff)
+l_stoch_ff<-ipm_stoch_ff$pop_state$lambda
+l_stoch_ff<-as.vector(l_stoch_ff)
[…]

### 2025-08-27 docs/ipm_comparison_ms.qmd
ready for review by ES

@@
         \setlength{\parindent}{0pt}
+        \usepackage{caption}
+        \captionsetup[figure]{labelfont={bf},name={fig.},labelsep=period}
+        \raggedright
+        \usepackage{setspace}\doublespacing
+        \setlength\parindent{22pt}
 
@@
 library(here)
+# library(vartest)
+library(purrr)
+library(broom)
 ```
@@
 })
+
 ```
@@
 
-\pagebreak
+\newpage
 
@@
 
+<!-- There is increasing evidence for delays of months or even years between current environmental conditions and changes in the growth, survival, or reproduction of long-lived organisms.  -->
+
 ## Introduction
 
-There is increasing evidence that an organism's current likelihood of growth, survival, or reproduction can be strongly influenced by previous environmental conditions.
-These *Lagged Effects*, also known as *Delayed Life-history Events* (i.e., DLHEs) [@beckerman_population_2002], can simultaneously affect an entire cohort (e.g., juveniles hatching during a period of scarcity will all have delayed maturation and lower lifetime fecundity) or only a subset of the population (e.g., cold temperatures in one year lead to reduced flowering by potentially reproductive individuals in the next).
-In addition, the temporal delay between an environmental event and changes in demographic vital rates depends on both the intensity of the event and its timing relative to the underlying physiological processes [@eversLaggedDormantSeason2021; @criley_year_1994].
-A drought during the early stages of gestation or floral bud formation, for example, might have a much larger impact on the number of fruits or offspring produced than one several months later whose timing coincides with birth or flowering.
-The delay or intensity of lagged effects can also depend on local ecological conditions, with individuals in some habitats buffered against -- or able to recover more quickly from - the delayed effects of environmental variation.\
+Current environmental conditions can have large and immediate effects on the growth, survival, or reproduction of long-lived organisms.
+There is also mounting evidence, however, for intervals that range from months to years between environmental conditions and the resulting changes in demographic vital rates.
+These *Lagged Effects*, also known as *Delayed Life-history Events* (i.e., DLHEs) [@beckermanPopulationDynamicConsequences2002], can simultaneously affect an entire cohort (e.g., juveniles hatching during a period of scarcity will all have delayed maturation and lower lifetime fecundity) or only a subset of the population (e.g., cold temperatures in one year lead to reduced flowering by potentially reproductive individuals in the next).
+In addition, the temporal delay between an environmental event and changes in demographic vital rates depends on both the intensity of the event and its timing relative to the underlying physiological processes [@eversLaggedDormantSeason2021; @crileyYearProductionHigh1994].
+A drought during the early stages of gestation or floral bud formation, for example, might have a much larger impact on the number of fruits or offspring produced than one at a later stage of the reproductive process.
+The delay or intensity of lagged effects can also be location- or habitat-specific, with individuals in some sites or habitats buffered against -- or able to recover more quickly from - the delayed effects of environmental variation.
 
-Because Lagged Effects are often directly linked to reproduction and survival, it is thought they could have major consequences for population dynamics [@beckerman_population_2002].
-Although there is emerging evidence that this is indeed the case [e.g., @molowny-horas_changes_2017; @tenhumberg_time-lagged_2018; @williams_life_2015], broader efforts to test this hypothesis have been hampered by two primary factors: First, detecting lagged effects requires long-term data on both the putative lagged effect (i.e., probability of flowering) and its potential environmental drivers [@metcalfStatisticalModellingAnnual2015].
-These coupled data sets are rare [sensu @eversLaggedDormantSeason2021], in part because studies to disentangle lagged effects can be challenging to design and maintain [@kuss_evolutionary_2008].
-Second, the methods for identifying lagged effects and modeling their demographic impacts can be challenging to implement.
-Many of the statistical methods have stringent data requirements [@metcalfStatisticalModellingAnnual2015] and assumptions, while the including complex biological processes in demographic models can render them less tractable.
-Addressing these obstacles is a major undertaking; the value of doing so will depend on the effort required vs.
-the potential consequences of failing to consider lagged effects - consequences that range from overestimating projections of population growth rate (i.e., $\lambda$) in a conservation setting to drawing invalid conclusions regarding support for the predictions of ecological or evolutionary theory.\
+Because Lagged Effects are often directly linked to reproduction and survival, it has been argued they could have major but underappreciated consequences for population dynamics [@beckermanPopulationDynamicConsequences2002].
+While recent studies suggest this could indeed the case [e.g., @molowny-horasChangesNaturalDynamics2017; @tenhumbergTimelaggedEffectsWeather2018; @williamsLifeHistoryEvolution2015], broader efforts to test this hypothesis have face a number of obstacles [@metcalfStatisticalModellingAnnual2015].
+Empirical tests of putative lagged effects because these are challenging to design, implement, and maintain [@kussEvolutionaryDemographyLonglived2008a].
+While one could use conceivably use observational data to detect lagged effects, doing so requires long-term data on both the putative lagged effect (e.g., reduced probability of flowering) and its potential environmental drivers (e.g., drought during early floral development), and such coupled data sets are rare [sensu @eversLaggedDormantSeason2021].
+In addition, the methods for both identifying lagged effects and modeling their demographic impacts can be challenging to implement.
+For example, many of the potentially applicable statistical methods have stringent assumptions and data requirements rarely met by ecological data [@metcalfStatisticalModellingAnnual2015], while the including complex biological processes in demographic models can render them less tractable or amenable to analysis using currently available tools.
+Addressing these challenges is a major undertaking; the value of doing so will depend on the effort required vs.
+the potential consequences of failing to consider lagged effects - consequences that range from overestimating projections of population growth rate (i.e., $\lambda$) in a conservation setting to drawing invalid conclusions regarding support for the predictions of ecological or evolutionary theory.
 
-Integral Projection Models (i.e., IPMs) are an important and widely used tool for studying demography and population dynamics [@ellner_integral_2006; @rees_building_2014; @rees_integral_2009].
-Their flexibility, in concert with a rapidly growing suite of software, data, and other resources [@ellnerDatadrivenModellingStructured2016; @levinIpmrFlexibleImplementation2021; @salguerogomez_compadre_2015], have facilitated their use to study a wide range of topics in ecology, evolution, and conservation [@ellnerDatadrivenModellingStructured2016; @morris_quantitative_2002 and @crone_how_2011].
-Mathematical and statistical advances [e.g., @williams2012avoiding; @brooks_statistical_2019] have rapidly expand the scope of questions and biological processes that can be investigated with these models [e.g., @metcalfStatisticalModellingAnnual2015; @rees_evolving_2016; @ellnerDatadrivenModellingStructured2016].
-Here we investigate how including lagged effects in Integral Projection Models influences projections of $\lambda$ and population structure.\
+Integral Projection Models (i.e., IPMs) are an important and widely used tool for studying demography and population dynamics [@ellnerIntegralProjectionModels2006; @reesBuildingIntegralProjection2014; @reesIntegralProjectionModels2009].
+Their flexibility, in concert with a rapidly growing suite of software, data, and other resources [@ellnerDatadrivenModellingStructured2016; @levinIpmrFlexibleImplementation2021; @salguero-gomezCOMPADREPlantMatrix2015], have facilitated their use to study a wide range of topics in ecology, evolution, and conservation [@ellnerDatadrivenModellingStructured2016; @morrisQuantitativeConservationBiology2002 and @croneHowPlantEcologists2011].
+Mathematical and statistical advances [e.g., @williams2012avoiding; @brooksStatisticalModelingPatterns2019] have rapidly expand the scope of questions and biological processes that can be investigated with these models [e.g., @metcalfStatisticalModellingAnnual2015; @reesEvolvingIntegralProjection2016; @ellnerDatadrivenModellingStructured2016].
+Here we investigate how including lagged effects in Integral Projection Models influences projections of $\lambda$ and population structure.
 
@@
 forest fragments).
-Based on previous studies [@bruna_are_2003; @bruna_demographic_2005; @bruna_experimental_2002; @bruna_habitat_2002] and demographic theory [@caswell2001; @tuljapurkar_population_1990] we predicted that: (i) projections of $\lambda$ from deterministic models would be higher than those of stochastic models, and that (ii) for all IPMs the projections of $\lambda$ forest fragments would be lower than for continuous forest, but that the difference would be proportionately greater for IPM with lagged effects.\
+Based on previous studies [@brunaArePlantPopulations2003; @brunaDemographicEffectsHabitat2005; @brunaHabitatFragmentationDemographic2002] and demographic theory [@caswell2001; @tuljapurkarPopulationDynamicsVariable1990] we predicted that: (i) projections of $\lambda$ from deterministic models would be higher than those of stochastic models, and that (ii) $\lambda$ is lower for Forest Fragment than Continuous Forest populations regardless of model, but the difference between the two habitats is greatest for the IPM with lagged effects.
 
@@
 
-*Heliconia acuminata* (Heliconiaceae) is a perennial, self-incompatible monocot that is distributed throughout much of the Amazon basin [@kress_diversity_1990].
+*Heliconia acuminata* (Heliconiaceae) is a perennial, self-incompatible monocot that is distributed throughout much of the Amazon basin [@kressDiversityDistributionHeliconia1990].
 While some *Heliconia* species grow in large aggregations on roadsides, gaps, and in other disturbed habitats, others - including *H.
-acuminata* - grow primarily in the forest understory [@kress_self-incompatibility_1983; @ribeiro_influence_2010].
-Understory *Heliconia* species produce fewer flowers and are pollinated by traplining hummingbirds [@bruna_heliconia_2004; @stouffer_forest_1996].
+acuminata* - grow primarily in the forest understory [@kressSelfincompatibilitySystemsCentral1983; @ribeiroInfluencePostclearingTreatment2010].
+Understory *Heliconia* species produce fewer flowers and are pollinated by traplining hummingbirds [@brunaHeliconiaAcuminataReproductive2004; @stoufferForestFragmentationSeasonal1996].
 The models and analyses here are based on 11 years (1998-2009) of demographic data collected on \>8500 *H.
@@
 The habitat in all sites is non-flooded lowland rain forest with rugged topography.
-A complete summary of the BDFFP and its history can be found in @bierregaard_lessons_2001.\
+A complete summary of the BDFFP and its history can be found in @bierregaardLessonsAmazoniaEcology2001.\
 
-A complete description of the demographic methods, data, and analyses to date can be found in @bruna_demography_2023.
-Briefly, in 1997--1998 a series of 5000 m^2^ plots were established in the BDFFP reserves: N=6 in Continuous Forest and N=4 in 1-ha Forest Fragments.
+A complete description of our demographic methods, data, and analyses to date can be found in @brunaDemographyUnderstoryHerb2023.
+Briefly, in 1997--1998 a series of 5000 m^2^ plots were established in the BDFFP reserves: N=6 in Continuous Forest and N=4 in 1-ha Forest Fragments (i.e., CF and FF, respectively).
 All of the *Heliconia acuminata* in these plots were marked and measured; the plots were censused annually, at which time a team recorded the size of surviving individuals, marked and measured new seedlings, and identified any previously marked plants that died.
-Each plot was also surveyed 4-5 times during the flowering season to identify reproductive plants; in our site *H.
-acuminata* begin flowering early in the rainy season (e.g., January) and most reproductive plants produce a single infloresence (range = 1--7) with 20--25 flowers [@bruna_habitat_2002].
-Fruits mature April-May and have 1--3 seeds per fruit ($\bar{x}=2$) that are dispersed by a thrush and several species of manakin [@uriarte_disentangling_2011].
-Dispersed seeds germinate approximately 6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@bruna_habitat_2002; @bruna_seed_1999].
-On average olots in CF also had more than twice as many plants as the plots in 1-ha fragments (CF median = 788, range = (201-1549); 1-ha median = 339, range = (297-400)).
+Each plot was also surveyed 4-5 times during the flowering season to identify reproductive plants.
+These surveys were complemented by data on the number of fruits per flowering plant [@brunaLeafNumberLeaf2021] and seeds per fruit [@brunaDatasetHeliconiaAcuminata2014] that were collected outside of the demographic plots to avoid altering within-plot recruitment.
+We also conducted experiments to quantify the probability of seed germination and seedling establishment in both forest fragments and continuous forest [@brunaEffectsForestFragmentation2002; @brunaSeedGerminationRainforest1999].
+
+*Heliconia acuminata* in our site begin flowering early in the rainy season (e.g., January) and most reproductive plants produce a single infloresence (range = 1--7) with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
+Fruits mature April-May and have 1--3 seeds per fruit ($\bar{x}=2$) that are dispersed by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011].
+Dispersed seeds germinate approximately 6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@brunaEffectsForestFragmentation2002; @brunaSeedGerminationRainforest1999].
+On average plots in CF also had more than twice as many plants as the plots in 1-ha fragments (CF median = 788, range = (201-1549); 1-ha median = 339, range = (297-400)).
 
@@
 
-We projected the growth rate and structure of *Heliconia acuminata* populations in Continuous Forest and Forest Fragments with three classes of IPMs - Deterministic, Stochastic, and Stochastic with Lagged Environmental Effects.
+We projected the growth rate and structure of *Heliconia acuminata* with three classes of IPMs - Deterministic, Stochastic, and Stochastic with Lagged Environmental Effects.
 Each of these IPMs required different functional forms of the underlying vital rate functions used to describe the *H.
 acuminata* life cycle (@fig-lifecycle).
-All models were density-independent, with the deterministic model serving as the foundation for the more complex models.\
+All models were density-independent, with the deterministic model serving as the foundation for the more complex models.
+Our modeling workflows were managed using the `targets` R package [@targets]; in adddition to ensuring reproducibility this allowed us to track computational time spent processing and analyzing each class of IPM.\
 
@@
 Note that in @eq-R both the probability that new seedlings survive their first year, $s_s$, and their size at the end of this year, $G(z^\prime, z)$, are size-independent.
-IPMs can include transitions between individuals from a discrete state to a continuous one [i.e., from 'seedling' to 'mature plant of size $z'$', @ellnerDatadrivenModellingStructured2016]; we treat seedlings as a distinct and discrete category because they have lower survival and growth in their first year than comparably sized plants [@bruna_are_2003; @scottDelayedEffectsClimate2022].\
+IPMs can include transitions between individuals from a discrete state to a continuous one [i.e., from 'seedling' to 'mature plant of size $z'$', @ellnerDatadrivenModellingStructured2016]; we treat seedlings as a distinct and discrete category because they have lower survival and growth in their first year than comparably sized plants [@brunaArePlantPopulations2003; @scottDelayedEffectsClimate2022].\
 
@@
 
-We used the annual census data [@bruna_are_2003] to fit the deterministic vital rate functions for growth, survival, and flowering in each habitat type (i.e., Fragments, Continuous Forest; the data from all plots within a habitat class were pooled to create a single 'summary population' [@bruna_are_2003].
+We used the annual census data [@brunaArePlantPopulations2003] to fit the deterministic vital rate functions for growth, survival, and flowering in each habitat type (i.e., Fragments, Continuous Forest; the data from all plots within a habitat class were pooled to create a single 'summary population' for the CF and FF habitats [@brunaArePlantPopulations2003].
 For established plants these were modeled as a smooth function of size in the previous census with generalized additive models (GAMs) fit with the `mgcv` library [@mgcv] for the R statistical programming language [@rcoreteam2020].<!--`r version$version.string`---> <!--# probably include a supplement with detail on things like basis, number of knots, etc.
@@
 For growth models a scaled t family distribution provided a better fit to the data than a Gaussian fit, as the residuals with a simple Gaussian model were Leptokurtic.
-To model size-specific fecundity we used data on the number of fruits per flowering plant [@brunaLeafNumberLeaf2021] and seeds per fruit [@brunaDatasetHeliconiaAcuminata2014] that were collected outside of the demographic plots (to avoid altering within-plot recruitment) and experimentally-derived estimates of seed germination and seedling establishment [@bruna_effects_2002; @bruna_seed_1999] <!--# Emilio: add more info here -->.\
+To model size-specific fecundity we used the data on fruits per flowering plant [@brunaLeafNumberLeaf2021], seeds per fruit [@brunaDatasetHeliconiaAcuminata2014], and the experimentally-derived estimates of seed germination and seedling establishment [@brunaEffectsForestFragmentation2002; @brunaSeedGerminationRainforest1999].
+<!--# Emilio: add more info here -->\
 
@@
 This was done using a factor--smooth interaction that allowed the functional form of the relationship between plant size and vital rates to vary among transition years (@fig-lifecycle).
-We generated kernels for every transition year using the long-term survey data [@bruna_demography_2023] and random smooths for year.
+We generated kernels for every transition year using the long-term survey data [@brunaDemographyUnderstoryHerb2023] and random smooths for year.
 We then randomly selected one of these sets of kernels to use in each iteration of the IPM.
@@
 These values were then used to predict the fitted values from vital rate models, which generated different kernels for each iteration of the IPM.
-The kernels of successive iterations are not entirely independent -- the SPEI values used to calculate vital rates include values used in the previous two iterations -- they are ergodic.\
+Though the kernels of successive iterations are not entirely independent -- the SPEI values used to calculate vital rates include values used in the previous two iterations -- they are ergodic.\
 
@@
 All IPMs were constructed and iterated <!--# optimized in notes/optimizing-meshpoints.Rmd which could be turned into supplemental -->using the `ipmr` package [@levinIpmrFlexibleImplementation2021] for the R statistical programming language [@rcoreteam2020].
-The IPMs used 100 meshpoints and the midpoint rule for calculating kernels <!--# citation?
+The IPMs used 100 meshpoints and the midpoint rule for calculating kernels<!--# citation?
 -->.
@@
 The initial starting vector primarily influences a population's transient dynamics; we therefore used the distribution of established plant sizes and proportion of seedlings in the full demographic data set as the initial population vector for all models.
-Finally, we estimated the 95% confidence intervals for each IPMs projections of $\lambda$.
+
+Finally, We estimated the 95% confidence intervals for each IPMs projections of $\lambda$.
 To do so we first created 500 populations for each habitat type by sampling individual plants with replacement (i.e., bootstrapping) until the population size of each matched that of the initial population vector.
@@
 The projections of $\lambda$ for the new populations were then used to estimate the upper and lower 95% bias-corrected percentile intervals [@manly2018randomization; @caswell2001].
-This workflow was managed using the `targets` R package [@targets], which also allowed us to track computational time spent processing and analyzing each class of IPM.\
 
@@
 
-##### Comparison of CF vs FF
+##### $\lambda$ in Continuous Forest vs. Forest Fragments.
 
--   For det need to bootstrap
+To determine if the deterministic projections of $\lambda$ for Continuous Forest and Forest Fragments were significantly different between habitats we used the randomization test procedure described by @caswell2001.
+Briefly, we randomly assigned plants with their demographic history among two populations, $R_{1}$ and $R_{2}$, that were equal in size to the original CF and FF populations.
+We then calculated $\lambda^{R_1}$, $\lambda^{R_2}$, and the absolute value of the difference between the two (i.e., $\theta$).
+This was repeated N = 1000 times, after which we determined the proportion of simulations in which $\theta$ was greater than the observed difference between $\lambda^{CF}$ and $\lambda^{FF}$ (i.e., $P[\theta \geq \theta_{obs}]$).
 
--   for within stoch and lag: glm/t-test.
+<!-- -   for within stoch and lag: glm/t-test. -->
 
--   Comparison of pop structure in figure (test for increasing variance with variance / F or permutation test).
-Villase∼nor, J.
-A., & González-Estrada, E.
-(2024).
-A nonparametric test for homogeneity of variances.
-Communications in Statistics - Theory and Methods, 54(2), 646–654.
-https://doi.org/10.1080/03610926.2024.2316274 OR https://www.datanovia.com/en/lessons/homogeneity-of-variance-test-in-r/#google_vignette
+<!-- see https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12577 for suggestions on text ffor describing and reporting results -->
+
+We used a Generalized Linear Model to compare the projections of $\lambda$ from the two stochastic IPMs.
+We modelled the lambda as a function of IPM type (kernel-resampling vs parameter-resampling) and Habitat (continuous forest vs.
+fragments), and the interaction of IPM type and Habitat.
+Because the response variable ($\lambda$) was continuous we used the Gaussian distribution with the identity function in our model.
+We verified the model assumptions by plotting residuals versus fitted values.\
+
+```{r glm-lambda}
+
+path <- here("docs/figures/stoch_lambdas.csv")
+stoch_lambdas<-read_csv(path)
+lambda_glm <- glm(lambda ~ habitat*ipm_type,
+             family = gaussian(link = "identity"), data = stoch_lambdas)
+
+# summary(lambda_glm)
+# (over)dispersaion
+# pearson's residuals
+# sum(resid(lambda_glm, type="pearson")^2)/lambda_glm$df.resid
+#deviance
+# lambda_glm$deviance/model$df.residual
+# plot(lambda_glm, which = 1)
+# plot(lambda_glm, which = 2)
+# https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12577
+# acf(residuals(lambda_glm)) ## check autocorrelation
+```
+
+##### Population Structure.
+
+Finally, we compared the structure of populations through the first 250 time steps in each habitat by each of the stochastic IPMs (i.e, kernel- vs.
+parameter-resampling).
+At each time step, we assigned the individuals to one of four stage classes based on plant size, probability of survival, and probability of reproduction (see Table @tbl-plantcats).
+We then calculated the proportion of the population that was in each stage class in each habitat for each IPM type.
+Comparing the means and variances of these proportions will allow us to determine (1) if the two IPMs project different population structures over time, (2) if the population structure for a given IPM is consistent across habitat type.
+Because Shapiro-Wilk tests indicated that the distributions of proportions were not normally distributed, we used the non-parametric Ansari-Bradley Test to compare the distribution of proportions in each habitat x stage combination.
+These analyses were conducted using the R libraries `rstatix` and `vartest`, respectively [@kassambaraRstatixPipefriendlyFramework2023; @cosarVartestTestsVariance2024].
 
@@
 spar_time<-as.numeric(spar_time[1,2])
+
+path_det_stats <- here("docs","figures","det_randomization_test.csv")
+det_stats<-read_csv(path_det_stats) %>% 
+  mutate(p_val=100-perc_below)
 ```
 
-### Lambda Comparison
+Deterministic projections of $\lambda$ for Continuous Forest were significantly greater than for Forest Fragments ($P[\theta \geq \theta_{obs}]$ = `r det_stats$p_val`, N = 1000 randomization).
+We also found found significant effects of IPM Type, Habitat, and their interaction on projections of $\lambda$ (F(df1, df2) = F-value, p = p-value, η² = effect size).
+Generally speaking, the effects of model choice on $\lambda$ were consistent across habitats - both the deterministic $\lambda$ and the average $\lambda$ from stochastic IPMs were higher for Continuous Forest than for Forest Fragments (@tbl-lambdas).
+However, model choice did influence the absolute value of projections within a habitat.
+Deterministic $\lambda$ was nearly identical to the average $\lambda$ from kernel-resampled IPMs, which was unexpected given that stochasticity is predicted to reduce population growth rates [@doakCorrectlyEstimatingHow2005a; @tuljapurkarManyGrowthRates2003; @metcalfStatisticalModellingAnnual2015].
+However the average projection of $\lambda$ was 1.5-2% lower when including lagged effects, with 70-76% of $\lambda$ projections below deterministic $\lambda$.
+These IPMs also appear to be more accurate - the underlying vital rate models for IPMs with lagged effects all had the best fit to the survey data (dAIC = 0, @tbl-aic).
+<!--# No longer true if random effect = different smooth for each year -->
 
-1.  We found that including Delayed Life-History Events in Intergral Projection Models led to significantly lower projections of $\lambda$
-2.
-It is reassuring that all IPM models projected higher values of $\lambda$ in continuous forest than forest fragments (@tbl-lambdas); this is consistent with the results of previous studies comparing the relative rankings of projections from stochastic and deterministic models [e.g., @kayeEffectStochasticTechnique2003.]
-3.
-However, the underestimating $\lambda$ by 2-3% difference in $\lambda$ resulting projections from the different IPMs could have major implications for many of the contexts in which IPMs are brought to bear.
-4.  If you are doing conservation and management, a growth rate lower by 5-6 % would really rock your world.
-5.  For instance, underestimating $\lambda$ by failing to include lagged effects could be especially severe in studies of climate change, given many DLHE are climate driven.\
-    <!-- CF: deterministic is > than 70% of dnlm --> <!-- FF: deterministic is > than 76% of dnlm -->
+Including lagged effects in IPMs also resulted in less predictable projections of population structure, both with (@fig-pop-states a) and across habitats (@fig-pop-states b, @tbl-vartest).
+This variability is particularly notable in forest fragments, where we have previously shown lagged effects have very large impacts on growth and survival [@scottDelayedEffectsClimate2022].
+The deterministic IPM for Continuous Forest projected slightly more of the smallest and largest plants than the one for Forest Fragments.
+In other words, populations in forest fragments had proportionately less recruitment and fewer individuals growing into the larger, reproductive size classes [@brunaDemographicEffectsHabitat2005].
+This shift towards intermediately sized, pre-reproductive plants in forest fragments is even more dramatic when using the parameter-resampled stochastic IPM (@fig-pop-states b).\
 
-### Pop Structure Comparison
+What are the implications of these results for the use of IPMs to study plant demography?
+First, they suggest that failing to include lagged effects in models could result in underestimates of $\lambda$.
+While relative differences in $\lambda$ may be sufficient for some tests of theory, they may be critical in conservation or management settings because even small differences in $\lambda$ can have major consequences for population dynamics.
+They could also be critical for studies assessing the effects of climate change, given many DLHE are climate driven.
+That said, there is a cost to including lagged effects in models, even in if they are more accurate.
+While Deterministic and Kernel-resampled Stochastic IPMs took only \~`r det_time` and \~`r sker_time` min to iterate (respectively), the Stochastic IPMs with using parameter-resampled kernels and lagged effects took \~`r spar_time` min.
+This is largely due to the required computational resources and algorithms (i.e., `predict()`), which are much slower forGeneral Additive Models (i.e., GAMs) with 2D smooths because of the much higher number of knots than for the GAMs used in Determinsitic and Kernel-resampled IPMs.
+Advances in computational power and access to high-performance computing resources could lower this cost.
 
-1.
-Including lagged effects in IPMs also resulted in more variable projections of future population structure (@fig-pop-states a), especially in forest fragments where we have previously shown lagged effects have very large impacts on growth and survival [@scottDelayedEffectsClimate2022].
-2.  Consequently, projections of population structure were also far less less consistent across habitats (@fig-pop-states b).
-3.  Finally, the choice of IPM also resulted in some subtle but notable differences in projected population structure.
-4.  For example, the deterministic IPM projected the population in Continuous Forest had slightly more of the smallest and largest plants than the population in Forest Fragments.
-5.  In other words, populations in forest fragments had proportionatly less recruitment and fewer individuals growing into the larger and reproductivce size classes [@bruna_demographic_2005].
-6.  The shift towards intermediately sized, pre-reproductive plants in forest fragments is even more dramatic when using the parameter-resampled stochastic IPM (@fig-pop-states b).\
-
-### So...should we care?
-
-1.  Including delayed life-history events in demographic models has important implications both projections of population growth rate and population structure.
-2.
-These projections may be more accurrate - the underlying vital rate models for IPMS with Lagged effects all had the best fit to the survey data (dAIC = 0, @tbl-aic).
-<!--# No longer true if random effect = different smooth for each year -->
-3.  While this may not be critical in situations where relative differences allow for hypothesis testing, it may be especially important in conservation or manaagement settings.
-4.  Including lagged effects comes at a cost, however.
-5.
-While Deterministic and Kernel-resampled Stochastic IPMs took only \~`r det_time` and \~`r sker_time` min to iterate (respectively), the Stochastic IPMs with using parameter-resampled kernels and lagged effects took \~`r spar_time` min.
-6.
-This is largely due to the required computational resources and algorithms (i.e., `predict()`), which are much slower forGeneral Additive Models (i.e., GAMs) with 2D smooths because of the much higher number of knots than for the GAMs used in Determinsitic and Kernel-resampled IPMs.
-7.  Advances in computational power and access to high-performance computing resources could lower this cost.\
-8.  However, computational power cannot compensate for limited data.
-9.
+Computational power cannot compensate for limited data, however.
 Detecting lagged effects and evaluating their consequences requires long-term demographic data - data that are only available for a relatively small number of species, few of which are in the tropics.
-10.
 The increasing evidence that lagged effects are ubiquitous, and that they can have major demographic impacts, underscores the need to support the collection of such long-term data, the complementary development of experimental and statistical approaches to disentangling lagged effects, and community driven efforts to identify priority or model systems for in-depth investigation.\
@@
 # Draft of this figure is in docs/figures/Heliconia lifecycle DAG.key
-path <- here("docs/figures/lifecycle.png")
-knitr::include_graphics(path, rel_path = FALSE)
+path <- here("docs","figures","lifecycle.png")
+knitr::include_graphics(path, rel_path = TRUE)
+
 ```
 
-\pagebreak
+\newpage
 
 ```{r #fig-pop-states,  echo=FALSE, out.width="85%",fig.align='center'}
-#| fig.cap = "**(A)** The proportion of _Heliconia acuminata_ populations in different size/stage classes when simulating population dyammics with three different Integral Projection Models.
-Results are shown for the first 250 iterations of populations in both in Continuous Forest (CF) and Forest Fragments (FF); for the criteria used to define the size categories see @tbl-plantcats.
-**(B)** For each interation, the relative proportion of the population in each size class (FF vs CF).
+#| fig.cap = "**(A)** The change over time in the the proportion of _Heliconia acuminata_ populations in different size/stage classes when simulating population dyammics in Continuous Forest (CF) or Forest Fragment (FF) with three different Integral Projection Models.
+Results are shown for the first 250 iterations of populations; for the criteria used to define the size categories see @tbl-plantcats.
+**(B)** The relative proportion of the population in each size class (FF vs.
+CF) for 250 iterations of each IPM model.
 Note that this excludes transient dynamics (iterations 1-30).
@@
 # include _seedlings_ (a discrete category in the IPMs), _pre-reproductive 1_ (log(size) = 0–2.5, average survival < 0.9, probability of flowering \\approx 0), _pre-reproductive 2_ (log(size) = 2.5–4.5, average survival probabilty > 0.8, probability of flowering \\approx 0), _reproductive 1_ (log(size) = 4.5–6, average survival probability >0.95, flowering probability < 0.25), and _reproductive 2_ (log(size) = 6+, average survival probability >0.95, flowering probability > 0.2)
-knitr::include_graphics(here(fig_pop_states), rel_path = FALSE)
+knitr::include_graphics(here(fig_pop_states), rel_path = TRUE)
 ```
 
-\pagebreak
+\newpage
 
 ```{r #fig-lambdas,  echo=FALSE, out.width="85%",fig.align='center'}
-#| fig.cap = "Distribution of 900 values of \u03BB projected with (A) Stochastic, kernel-resampled IPMs and (B) Stochastic, paramenter-resampled IPMs.
-IPMs were used to project \u03BB for both Continuous Forest (green) and Forest Fragments (gray).
-The solid line indicates the mean value, the dashed line indicates the value of \u03BB $\\lambda$ in that habitat projected with Deterministic IPMs."
+#| fig.cap = "Distribution of 900 values of $\\lambda$ projected with (A) Stochastic, kernel-resampled IPMs and (B) Stochastic, paramenter-resampled IPMs.
+IPMs were used to project $\\lambda$ for both Continuous Forest (above, in green) and Forest Fragments (below, in gray).
+The solid line indicates the mean value of $\\lambda$, the dashed line indicates the value of $\\lambda$ in that habitat projected with Deterministic IPMs."
 
 # The figures is currently generated with "plotting_lambdas.R, need to move to targets_r
-path <- here("docs/figures/lambdas_fig.png")
-knitr::include_graphics(path, rel_path = FALSE)
+
+path <- here("docs","figures","lambdas_fig.png")
+knitr::include_graphics(path, rel_path = TRUE)
 ```
 
-\pagebreak
+<!-- \newpage -->
 
-```{r #fig-var,  echo=FALSE, out.width="85%",fig.align='center'}
-#| fig.cap = "------"
+<!-- ```{r #fig-var,  echo=FALSE, out.width="85%",fig.align='center'} -->
+
+<!-- #| fig.cap = "------" -->
+
+<!-- # The figures is currently generated with "plotting_lambdas.R, need to move to targets_r  -->
+
+<!-- path <- here("docs","figures","relative_variances_fig.png") -->
+
+<!-- knitr::include_graphics(path, rel_path = TRUE) -->
+
+<!-- ``` -->
+
+<!-- \newpage -->
+
+<!-- ```{r #fig-marginal-long,  echo=FALSE, out.width="85%",fig.align='center'} -->
+
+<!-- #| fig.cap = "The relative proportion of the population in each size class (FF vs.
+CF) for each of 250 iterations of the kernel-resampled (dark shading) and parameter-resampled IPM models (light shading).
+Values on the 1-1 line indicate an iteration where CF and FF have the same proportion of the population in a given size class; the marginal plots indicate the distribution of these relative proportions for each class of IPM in each habitat.
+Note that both the scatterplots and marginal plots exclude transient dynamics (iterations 1-30)." -->
+
+<!-- # The figures is currently generated with "plotting_lambdas.R, need to move to targets_r  -->
+
+<!-- path <- here("docs","figures","marginal_stage_fig_long.png") -->
+
+<!-- knitr::include_graphics(path, rel_path = TRUE) -->
+
+<!-- ``` -->
+
+\newpage
+
+```{r #fig-marginal,  echo=FALSE, out.width="85%",fig.align='center'}
+#| fig.cap = "The relative proportion of the population in each size class (FF vs.
+CF) for each of 250 iterations of the kernel-resampled (dark shading) and parameter-resampled IPM models (light shading).
+Values on the 1-1 line indicate an iteration where CF and FF have the same proportion of the population in a given size class; the marginal plots indicate the distribution of these relative proportions for each class of IPM in each habitat.
+Note that both the scatterplots and marginal plots exclude transient dynamics (iterations 1-30)."
 
 # The figures is currently generated with "plotting_lambdas.R, need to move to targets_r
-path <- here("docs/figures/relative_variances_fig.png")
-knitr::include_graphics(path, rel_path = FALSE)
+
+path <- here("docs","figures","marginal_stage_fig.png")
+knitr::include_graphics(path, rel_path = TRUE)
 ```
 
-\pagebreak
-
-```{r #fig-marginal,  echo=FALSE, out.width="85%",fig.align='center'}
-#| fig.cap = "------"
-
-# The figures is currently generated with "plotting_lambdas.R, need to move to targets_r 
-path <- here("docs/figures/marginal_stage_fig.png")
-knitr::include_graphics(path, rel_path = FALSE)
-```
-
-\pagebreak
+\newpage
 
@@
 \u0394AIC within 2 indicates models are equivalent."
+#| tbl-pos: H
 #| 
@@
          )
-  )%>% 
+  ) %>% 
   mutate(
@@
     # latex_options = c("hold_position"),
-    font_size = 12,
+    font_size = 10,
     position = "center"
@@
 #| tbl-cap: "Population growth rates for continuous forest (CF) and forest fragments (FF) under different kinds of IPMs with bootstrapped, bias-corrected, 95% confidence intervals."
+#| tbl-pos: H
 
@@
   mutate(across(c(est, lower, upper), ~format(round(.,4)))) %>% 
-  mutate(lambda = glue::glue("{est} ({lower}, {upper})")) %>%
+  # mutate(lambda = glue::glue("{est} ({lower}, {upper})")) %>%
+  mutate(CI = glue::glue("({lower}, {upper})")) %>%
   mutate(ipm = str_replace_all(
@@
   )) %>% 
-  select(IPM = ipm, Habitat = habitat, "$\\lambda$" = lambda) %>% 
+  # separate(lambda,c("lambda","CI"),sep=" [(]") %>% 
+  select(IPM = ipm, 
+         Habitat = habitat, 
+         # "$\\lambda$" = lambda) %>% 
+         "$\\lambda$" = est,
+         "(Lower, Upper 95\\% CI)"=CI) %>%
   mutate(IPM = str_replace_all(
@@
   format = "latex",
-  align = "lcc",
+  align = "lccc",
   escape = FALSE,
@@
 Note that seedlings are a discrete size class not based on size (see _Methods_ for additional details)."
+#| tbl-pos: H
 
-category<-c("seedlings", "pre-reproductive 1", "pre-reproductive 2", "reproductive 1", "reproductive 2")
+category<-c("Seedlings", "Pre-reproductive 1", "Pre-reproductive 2", "Reproductive 1", "Reproductive 2")
 size<-c(" - ","0–2.5","2.5-4.5","4.5-6","$\\geq$ 6")
@@
 flowering<-c("-","$\\approx$ 0","$\\approx$ 0","$\\leq$ 0.25","$\\geq$ 0.2")
-cat_table<-bind_cols(category=category,
-                     `log(size)`=size, 
-                     `avg. prob. survival` = survival,
-                     `prob. flowering` = flowering)
+cat_table<-bind_cols(Category=category,
+                     `Log(size)`=size, 
+                     `Avg. prob. survival` = survival,
+                     `Prob. flowering` = flowering)
 cat_table %>% 
@@
 ```
+
+\newpage
+
+```{r tbl-vartest, results='asis'}
+#| label: tbl-vartest
+#| tbl-cap: "Results of statistical tests comparing the variance in the proportion of each habitat's population projected to be in each stage class by kernel-resampled vs.
+parameter resampled IPMs (N = 220 projections per stage class).
+The variances for each habitat x stage class x IPM combination can be found in @tbl-meanvars.
+Comparisons where the p-value of the test was < 0.05 are indicated with an asterisk."
+#| tbl-pos: H
+
+# source(here("R","variance_stats.R"))
+# variance_stats()
+path <- here("docs","figures","varstats_table.csv")
+stats_table<-read_csv(path)
+
+stats_table %>%
+  # select(-`Vital Rate`) %>%
+  rename(`Stage`=`Vital Rate`) %>% 
+  mutate(`Stage`= " ") %>% 
+  # mutate(Comparison="KR v PR") %>% 
+  # relocate(Comparison,.before=1) %>% 
+kable(digits = 5,
+  format = "latex",
+  align = "lcrrcr",
+  escape = FALSE,
+  row.names = FALSE,
+  booktabs = T,
+  # caption = "Test comparing variances of vital rates in stochastic ipms.",
+  linesep = ""
+) %>%
+  kable_styling(
+    bootstrap_options = c("hover"),
+    # full_width = F,
+    latex_options = c("scale_down"),
+    # latex_options = c("hold_position"),
+    font_size = 12,
+    position = "center"
+  ) %>%
+  # collapse_rows(columns = 1:2, valign = "top")
+  column_spec(1, width = "8em") %>%
+  # collapse_rows(columns = 1:2,valign="top")
+  pack_rows("Seedling", 1, 2) %>%
+  pack_rows("Pre-reproductive 1", 3, 4) %>%
+  pack_rows("Pre-reproductive 2", 5, 6) %>%
+  pack_rows("Reproductive 1", 7, 8) %>%
+  pack_rows("Reproductive 2", 9, 10)
+
+```
+
+\newpage
+
+```{r tbl-meanvars, results='asis'}
+#| label: tbl-meanvars
+#| tbl-cap: "Summary statistics (median, mean, and variance) describing the proportion of populations projected to be in each of five life-history stages by kernel- and parameter-resampled IPMs (N = 220 projections for each IPM class x habitat combination.)"
+#| tbl-pos: H
+
+# source(here("R","variance_stats.R"))
+# variance_stats()
+
+path <- here("docs","figures","meanvar_table.csv")
+meavvar_table<-read_csv(path) %>%
+  select(-n) %>% 
+  mutate(meanCF=round(meanCF,4)) %>% 
+  mutate(meanFF=round(meanCF,4)) %>% 
+  mutate(medianCF=round(medianCF,4)) %>% 
+  mutate(medianFF=round(medianFF,4)) %>% 
+  mutate(varCF=round(varCF,6)) %>% 
+  mutate(varFF=round(varFF,6)) %>% 
+  rename(Stage=log_size_bin) %>% 
+  mutate(Stage=" ")
+
+names(meavvar_table)<-c("Stage","IPM","CF","FF","CF","FF","CF","FF")
+# %>% 
+  # mutate(CF=paste(meanCF,"±",varCF,sep=" "),.keep="unused") %>% 
+  # mutate(FF=paste(meanFF,"±",varFF,sep=" "),.keep="unused")
+# pandoc.table(meavvar_table,
+#   digits = 5,
+#   keep.trailing.zeros = TRUE,
+#   justify = c("left", "center", "center","center","center"),
+#   full_width = FALSE,
+# caption = "(\\#tab-meanvars) caption."
+# )
+meavvar_table %>%
+kable(
+  # digits = 2,
+  # format = "latex",
+  align = "llcccccc",
+  escape = FALSE,
+  row.names = FALSE,
+  booktabs = T,
+  # caption="Summary statistics (median, mean, and variance) describing the proportion of projected populations in each of five life-history stages (N = 220 projections for each class of stochastic IPM).",
+  linesep = ""
+) %>%
+  kable_styling(
+    bootstrap_options = c("hover"),
+    # full_width = F,
+    latex_options = c("scale_down"),
+    # latex_options = c("hold_position"),
+    font_size = 10,
+    position = "center"
+  )%>%
+  # collapse_rows(columns = 1:2, valign = "top")
+  # column_spec(1, width = "15em") %>%
+  add_header_above(c(" " = 2, "Median" = 2, "Mean" = 2, "Variance" = 2)) %>%
+  column_spec(1, width = "8em") %>%
+  pack_rows("Seedling", 1, 2) %>%
+  pack_rows("Pre-reproductive 1", 3, 4) %>%
+  pack_rows("Pre-reproductive 2", 5, 6) %>%
+  pack_rows("Reproductive 1", 7, 8) %>%
+  pack_rows("Reproductive 2", 9, 10)
+```
+
+\newpage
+
+```{r table-glm, results="asis"}
+#| label: tbl-glm
+#| tbl-cap: "Estimated parameters, standard errors, t-values and P-values for the GLM of the effect of Habitat and IPM Type on projections of lambda."
+#| tbl-pos: H
+
+glm_table<-tidy(lambda_glm) %>% 
+  mutate(estimate=round(estimate,3),
+  std.error=round(std.error,4),
+  statistic=round(statistic,3),
+  p.value=round(p.value,4)) %>% 
+  mutate(p.value=as.character(p.value)) %>% 
+  # mutate(p.value=if_else(p.value< 0.01,"0.001",p.value)) %>% 
+  rename(Term=term,
+         Estimate=estimate,
+         SE=std.error,
+         "z value"=statistic,
+         "P value"=p.value) %>% 
+  mutate(Term=gsub("[:]"," : ",Term)) %>%
+  mutate(Term=gsub("habitatff","$Habitat_{FF}$",Term)) %>%
+  mutate(Term=gsub("ipm_typestoch","$IPM_{Stoch}$",Term))
+
+# %>%
+#   mutate(Term=gsub(":"," ",Term)) %>% 
+#   mutate(Term=gsub("_"," ",Term))
+
+glm_table %>% 
+  kable(
+  digits = 2,
+  format = "latex",
+  align = "lccccc",
+  escape = FALSE,
+  row.names = FALSE,
+  booktabs = T,
+  linesep = ""
+) %>%
+  kable_styling(
+    bootstrap_options = c("hover"),
+    # full_width = F,
+    latex_options = c("scale_down"),
+    # latex_options = c("hold_position"),
+    font_size = 12,
+    position = "center"
+  )
+
+```

### 2025-08-27 R/bootstrap_cf_v_ff.R

@@ -1,4 +1,4 @@
-# extract values of lambda fom targets to make figures
+# compare det lambda in CF and FF using bootstrapping 
 
 # load libraries ----------------------------------------------------------
 
@@ -6,6 +6,97 @@
 library(targets)
 library(tidyverse)
 library(RColorBrewer)  
+library(patchwork)
+library(ggExtra)
+
+
+tar_read(data_full)
+tar_read(demog)
+tar_read(vit_list_det_cf)
+tar_read(vit_list_det_ff)
+tar_read(pop_vec_cf)
+tar_read(data_cf)
+
+# load data ---------------------------------------------------------------
+ipm_det_cf_1<-tibble(lambda_cf=tar_read(lambda_bt_det_cf_6387ee2c))
+ipm_det_cf_2<-tibble(lambda_cf=tar_read(lambda_bt_det_cf_e8cfe512))
+ipm_det_cf_3<-tibble(lambda_cf=tar_read(lambda_bt_det_cf_1f90f462))
+ipm_det_cf_4<-tibble(lambda_cf=tar_read(lambda_bt_det_cf_832d921b))
+ipm_det_cf_5<-tibble(lambda_cf=tar_read(lambda_bt_det_cf_ac9b0ec2))
+det_cf<-bind_rows(
+  ipm_det_cf_1,
+  ipm_det_cf_2,
+  ipm_det_cf_3,
+  ipm_det_cf_4,
+  ipm_det_cf_5
+) %>% 
+  mutate(run=row_number())
+
+ipm_det_ff_1<-tibble(lambda_ff=tar_read(lambda_bt_det_ff_3305ceac))
+ipm_det_ff_2<-tibble(lambda_ff=tar_read(lambda_bt_det_ff_7944832e))
+ipm_det_ff_3<-tibble(lambda_ff=tar_read(lambda_bt_det_ff_2db550e8))
+ipm_det_ff_4<-tibble(lambda_ff=tar_read(lambda_bt_det_ff_2258367a))
+ipm_det_ff_5<-tibble(lambda_ff=tar_read(lambda_bt_det_ff_aad0a33a))
+
+
+det_ff<-bind_rows(
+  ipm_det_ff_1,
+  ipm_det_ff_2,
+  ipm_det_ff_3,
+  ipm_det_ff_4,
+  ipm_det_ff_5
+)%>% 
+  mutate(run=row_number())
+
+bootstrap_lambdas<-left_join(det_ff,det_cf,by="run") %>% 
+  relocate(run,.before=1) %>% 
[…]

### 2025-08-27 R/ipm_boot_det_sig_testing.R

@@ -1,0 +1,291 @@
+
+#'
+#' This samples individuals with replacement, fits vital rate
+#' models, builds an IPM, iterates it, then returns lambda. No intermediate
+#' steps are saved, so this is intended for use in bootstrapping confidence
+#' intervals around lambda.
+#' 
+#' 
+## Load your packages, e.g. library(targets).
+source("./packages.R")
+
+## Load your R files
+# lapply(list.files("./R", full.names = TRUE), source)
+
+conflicts_prefer(dplyr::filter)
+## tar_plan supports drake-style targets and also tar_target()
+# tar_plan(
+  
+  # Load and wrangle data ---------------------------------------------------
+  # All of these targets are run locally (deployment = "main")
+  # Heliconia demographic datset
+  # tar_file(file_demog, here("data", "HDP_1997_2009.csv"), deployment = "main")
+  # tar_file(file_plots, here("data", "HDP_plots.csv"), deployment = "main")
+  # 
+  # # Other unpublished Heliconia datasets for fruit and seed vital rates
+  # tar_file(file_1998, here("data", "ha_size_data_1998_cor.csv"), deployment = "main")
+  # tar_file(file_2008, here("data", "Heliconia_acuminata_seedset_2008.csv"),
+  #          deployment = "main")
+  # tar_target(data_1998, read_wrangle_1998(file_1998), deployment = "main")
+  # tar_target(data_2008, read.csv(file_2008), deployment = "main")
+  # # 
+  # # # Wrangle and combine data
+  # tar_target(demog, read_wrangle_demog(file_demog, file_plots), deployment = "main")
+  # tar_target(data_full, join_data(demog, clim), deployment = "main")
+  
+  
+  # 
+  # 
+  # tar_target(data_cf, data_full %>% filter(habitat == "CF"), deployment = "main")
+  # 
+  # tar_target(data_ff, data_full %>% filter(habitat == "1-ha"), deployment = "main")
+  # 
+  
+  data_full<-tar_read(data_full)
+  data_1998<-tar_read(data_1998)
+  data_2008<-tar_read(data_2008)
+  
+  
+  n_cf<-data_full %>%
+    filter(habitat=="CF") %>% 
[…]

### 2025-08-27 R/marginal_value_plot_proportions.R

@@ -1,0 +1,461 @@
+
+
+
+# load libraries ----------------------------------------------------------
+
+
+library(targets)
+library(tidyverse)
+library(patchwork)
+library(ggExtra)
+
+
+# load data ---------------------------------------------------------------
+ipm_det_cf<-tar_read(ipm_det_cf)
+ipm_det_ff<-tar_read(ipm_det_ff)
+ipm_dlnm_cf<-tar_read(ipm_dlnm_cf)
+ipm_dlnm_ff<-tar_read(ipm_dlnm_ff)
+ipm_stoch_cf<-tar_read(ipm_stoch_cf)
+ipm_stoch_ff<-tar_read(ipm_stoch_ff)
+
+ipm_list<-
+  list(
+    det_cf = ipm_det_cf,
+    det_ff = ipm_det_ff,
+    stoch_cf = ipm_stoch_cf,
+    stoch_ff = ipm_stoch_ff,
+    dlnm_cf = ipm_dlnm_cf,
+    dlnm_ff = ipm_dlnm_ff
+  )
+
+# create df of proportions in CF and FF  ----------------------------------
+
+
+
+## FUNCTION - BIN POP STATES
+
+bin_pop_states <- function(ipm) {
+  
+  meshpts <- int_mesh(ipm, full_mesh = FALSE)$log_size_1
+  
+  pop_states <- pop_state(ipm)
+  rownames(pop_states$n_log_size) <-  meshpts
+  
+  post_sdlg_df <-
+    as.data.frame(pop_states$n_log_size) %>%
+    as_tibble(rownames = "log_size") %>%
+    mutate(log_size = as.numeric(log_size)) %>% 
+    pivot_longer(
+      -log_size,
+      names_to = "iteration",
+      names_transform = list(iteration = ~factor(as.numeric(.))),
+      values_to = "n_plants",
+      names_prefix = "V")
+  
+  #bin sizes for easier plotting
+  log_size_range <- as.numeric(ipmr::domains(ipm)$log_size[1:2])
+  
+  post_sdlg_binned <-
+    post_sdlg_df %>%
+    mutate(log_size_bin = case_when(
+      log_size <= 2.5 ~ "pre-reproductive 1 [0, 2.5]",
+      log_size > 2.5 & log_size <= 4.5 ~ "pre-reproductive 2 (2.5, 4.5]",
+      log_size > 4.5 & log_size <= 6 ~ "reproductive 1 (4.5, 6]",
[…]

### 2025-08-27 R/variance_stats.R

@@ -1,0 +1,167 @@
+variance_stats <- function() {
+  
+  path <- here("docs/figures/df_for_stats.csv")
+  df_for_stats<-read_csv(path) %>%
+    dplyr::filter(IPM!="Deterministic")
+  
+  seedling_cf <- df_for_stats %>%
+    dplyr::filter(log_size_bin== 'seedling') %>%
+    dplyr::filter(Habitat=="CF")
+  
+  seedling_ff <- df_for_stats %>%
+    dplyr::filter(log_size_bin== 'seedling') %>%
+    dplyr::filter(Habitat=="FF")
+  
+  pre1_cf <- df_for_stats %>%
+    dplyr::filter(log_size_bin== 'pre-reproductive1') %>%
+    dplyr::filter(Habitat=="CF")
+  
+  pre1_ff <- df_for_stats %>%
+    dplyr::filter(log_size_bin== 'pre-reproductive1') %>%
+    dplyr::filter(Habitat=="FF")
+  
+  pre2_cf <- df_for_stats %>%
+    dplyr::filter(log_size_bin== 'pre-reproductive2') %>%
+    dplyr::filter(Habitat=="CF")
+  
+  pre2_ff <- df_for_stats %>%
+    dplyr::filter(log_size_bin== 'pre-reproductive2') %>%
+    dplyr::filter(Habitat=="FF")
+  
+  rep1_cf <- df_for_stats %>%
+    dplyr::filter(log_size_bin== 'reproductive1') %>%
+    dplyr::filter(Habitat=="CF")
+  
+  rep1_ff <- df_for_stats %>%
+    dplyr::filter(log_size_bin== 'reproductive1') %>%
+    dplyr::filter(Habitat=="FF")
+  
+  rep2_cf <- df_for_stats %>%
+    dplyr::filter(log_size_bin== 'reproductive2') %>%
+    dplyr::filter(Habitat=="CF")
+  
+  rep2_ff <- df_for_stats %>%
+    dplyr::filter(log_size_bin== 'reproductive2') %>%
+    dplyr::filter(Habitat=="FF")
+  
+  # seedling_cf
+  # seedling_ff
+  # pre1_cf
+  # pre1_ff
+  # pre2_cf
+  # pre2_ff
+  # rep1_cf
+  # rep1_ff
+  # rep2_cf
+  # rep2_ff
+  
+  #
+  
+  #
+  
+  
+  
+  
+  # Create a list of datasets
+  
+  datasets <- list(seedling_cf,
[…]

### 2025-09-02 docs/ipm_comparison_ms.qmd
ms edits

@@
 Notes: >
-  Manuscript elements: Figure~1, figure~2, table~1, appendix~A.  
+  Manuscript elements: Figures 1-4, Tables 1-6.  
   Manuscript type: e-note
@@
 
-**Manuscript elements:** Figure\~1, figure\~2, table\~1, appendix\~A..
+**Manuscript elements:** Figures 1-4, Tables 1-6
 
@@
 
-##### $\lambda$ in Continuous Forest vs. Forest Fragments.
+##### Comparison of $\lambda$ in Continuous Forest and Forest Fragments.
 
-To determine if the deterministic projections of $\lambda$ for Continuous Forest and Forest Fragments were significantly different between habitats we used the randomization test procedure described by @caswell2001.
+To determine if the deterministic projections of $\lambda$ for Continuous Forest and Forest Fragment populations were significantly different we used the randomization test procedure described by @caswell2001.
 Briefly, we randomly assigned plants with their demographic history among two populations, $R_{1}$ and $R_{2}$, that were equal in size to the original CF and FF populations.
@@
 
-Deterministic projections of $\lambda$ for Continuous Forest were significantly greater than for Forest Fragments ($P[\theta \geq \theta_{obs}]$ = `r det_stats$p_val`, N = 1000 randomization).
-We also found found significant effects of IPM Type, Habitat, and their interaction on projections of $\lambda$ (F(df1, df2) = F-value, p = p-value, η² = effect size).
-Generally speaking, the effects of model choice on $\lambda$ were consistent across habitats - both the deterministic $\lambda$ and the average $\lambda$ from stochastic IPMs were higher for Continuous Forest than for Forest Fragments (@tbl-lambdas).
-However, model choice did influence the absolute value of projections within a habitat.
-Deterministic $\lambda$ was nearly identical to the average $\lambda$ from kernel-resampled IPMs, which was unexpected given that stochasticity is predicted to reduce population growth rates [@doakCorrectlyEstimatingHow2005a; @tuljapurkarManyGrowthRates2003; @metcalfStatisticalModellingAnnual2015].
-However the average projection of $\lambda$ was 1.5-2% lower when including lagged effects, with 70-76% of $\lambda$ projections below deterministic $\lambda$.
-These IPMs also appear to be more accurate - the underlying vital rate models for IPMs with lagged effects all had the best fit to the survey data (dAIC = 0, @tbl-aic).
+All IPMs projected higher population growth rates for *Heliconia acuminata* populations in Continuous Forest than for those in Forest Fragments (@tbl-lambdas, @tbl-glm).
+The differences between $\lambda^{CF}$ and $\lambda^{FF}$, which ranged from 1.19-2%, were significant for both deterministic ($P[\theta \geq \theta_{obs}]$ = `r det_stats$p_val`, N = 1000 randomization) and stochastic IPMs (@tbl-glm).
+There were also significant effects of IPM Type, Habitat, and their interaction on $\lambda$ (@tbl-glm).
+Deterministic $\lambda$ was nearly identical to the average $\lambda$ from kernel-resampled IPMs (@fig-lambdas), which is somewhat surpring given that stochasticity is predicted to reduce population growth rates [@doakCorrectlyEstimatingHow2005a; @tuljapurkarManyGrowthRates2003; @metcalfStatisticalModellingAnnual2015].
+However including lagged effects in IPMs resulted in projections of $\lambda$ that were on average 1.5-2% lower than those of deterministic models, with 70-76% of projections below deterministic $\lambda$ (@fig-lambdas).
+The IPMs with lagged effects also appear to be more accurate - the underlying vital rate models for IPMs with lagged effects all had the best fit to the survey data (dAIC = 0, @tbl-aic).
 <!--# No longer true if random effect = different smooth for each year -->
 
-Including lagged effects in IPMs also resulted in less predictable projections of population structure, both with (@fig-pop-states a) and across habitats (@fig-pop-states b, @tbl-vartest).
-This variability is particularly notable in forest fragments, where we have previously shown lagged effects have very large impacts on growth and survival [@scottDelayedEffectsClimate2022].
+Including lagged effects in IPMs also resulted in significantly less predictable projections of population structure, both within (@fig-pop-states a) and across habitats (@fig-pop-states b, @tbl-vartest).
+This variability is particularly notable in forest fragments (@fig-marginal, @tbl-meanvars), where we have previously shown lagged effects have very large impacts on growth and survival [@scottDelayedEffectsClimate2022].
 The deterministic IPM for Continuous Forest projected slightly more of the smallest and largest plants than the one for Forest Fragments.

### 2025-09-08 docs/ipm_comparison_ms.qmd
updates - eric comments

@@
     fontsize: 12pt
+    fig-pos: "H"
     # the following line is to render the aaffiliations in pdf, which is not default in quarto, see 
@@
 A drought during the early stages of gestation or floral bud formation, for example, might have a much larger impact on the number of fruits or offspring produced than one at a later stage of the reproductive process.
-The delay or intensity of lagged effects can also be location- or habitat-specific, with individuals in some sites or habitats buffered against -- or able to recover more quickly from - the delayed effects of environmental variation.
+The delay or intensity of lagged effects can also be location- or habitat-specific, with individuals in some sites or habitats buffered against, or able to recover more quickly from, the delayed effects of environmental variation.
 
@@
 While recent studies suggest this could indeed the case [e.g., @molowny-horasChangesNaturalDynamics2017; @tenhumbergTimelaggedEffectsWeather2018; @williamsLifeHistoryEvolution2015], broader efforts to test this hypothesis have face a number of obstacles [@metcalfStatisticalModellingAnnual2015].
-Empirical tests of putative lagged effects because these are challenging to design, implement, and maintain [@kussEvolutionaryDemographyLonglived2008a].
-While one could use conceivably use observational data to detect lagged effects, doing so requires long-term data on both the putative lagged effect (e.g., reduced probability of flowering) and its potential environmental drivers (e.g., drought during early floral development), and such coupled data sets are rare [sensu @eversLaggedDormantSeason2021].
+Empirical tests of putative lagged effects are rare because these are challenging to design, implement, and maintain [@kussEvolutionaryDemographyLonglived2008a].
+While one could conceivably use observational data to detect lagged effects, doing so requires long-term data on both the putative lagged effect (e.g., reduced probability of flowering) and its potential environmental drivers (e.g., drought during early floral development), and such coupled data sets are rare [sensu @eversLaggedDormantSeason2021].
 In addition, the methods for both identifying lagged effects and modeling their demographic impacts can be challenging to implement.
@@
 Addressing these challenges is a major undertaking; the value of doing so will depend on the effort required vs.
-the potential consequences of failing to consider lagged effects - consequences that range from overestimating projections of population growth rate (i.e., $\lambda$) in a conservation setting to drawing invalid conclusions regarding support for the predictions of ecological or evolutionary theory.
+the potential consequences of failing to consider lagged effects --- consequences that range from overestimating projections of population growth rate (i.e., $\lambda$) in a conservation setting to drawing invalid conclusions regarding support for the predictions of ecological or evolutionary theory.
 
 Integral Projection Models (i.e., IPMs) are an important and widely used tool for studying demography and population dynamics [@ellnerIntegralProjectionModels2006; @reesBuildingIntegralProjection2014; @reesIntegralProjectionModels2009].
-Their flexibility, in concert with a rapidly growing suite of software, data, and other resources [@ellnerDatadrivenModellingStructured2016; @levinIpmrFlexibleImplementation2021; @salguero-gomezCOMPADREPlantMatrix2015], have facilitated their use to study a wide range of topics in ecology, evolution, and conservation [@ellnerDatadrivenModellingStructured2016; @morrisQuantitativeConservationBiology2002 and @croneHowPlantEcologists2011].
-Mathematical and statistical advances [e.g., @williams2012avoiding; @brooksStatisticalModelingPatterns2019] have rapidly expand the scope of questions and biological processes that can be investigated with these models [e.g., @metcalfStatisticalModellingAnnual2015; @reesEvolvingIntegralProjection2016; @ellnerDatadrivenModellingStructured2016].
+Their flexibility, in concert with a rapidly growing suite of software, data, and other resources [@ellnerDatadrivenModellingStructured2016; @levinIpmrFlexibleImplementation2021; @salguero-gomezCOMPADREPlantMatrix2015], have facilitated their use to study a wide range of topics in ecology, evolution, and conservation [@ellnerDatadrivenModellingStructured2016; @morrisQuantitativeConservationBiology2002; @croneHowPlantEcologists2011].
+Mathematical and statistical advances [e.g., @williams2012avoiding; @brooksStatisticalModelingPatterns2019] have rapidly expanded the scope of questions and biological processes that can be investigated with these models [e.g., @metcalfStatisticalModellingAnnual2015; @reesEvolvingIntegralProjection2016; @ellnerDatadrivenModellingStructured2016].
 Here we investigate how including lagged effects in Integral Projection Models influences projections of $\lambda$ and population structure.
@@
 We have previously shown that the effects of precipitation extremes on the demographic vital rates of an Amazonian understory herb (*Heliconia acuminata*, Heliconiaceae) can be delayed up to 36 months [@scottDelayedEffectsClimate2022], with the presence and duration of these lagged effects varying by vital rate and habitat.
-We parameterized three classes of Integral Projection Models - a deterministic IPM, a stochastic IPM, and a stochastic IPM with lagged effects of precipitation on vital rates - for populations in two habitat type (i.e., continuous forest vs.
+We parameterized three classes of Integral Projection Models --- a deterministic IPM, a stochastic IPM, and a stochastic IPM with lagged effects of precipitation on vital rates --- for populations in two habitat types (i.e., continuous forest vs.
 forest fragments).
@@
 *Heliconia acuminata* (Heliconiaceae) is a perennial, self-incompatible monocot that is distributed throughout much of the Amazon basin [@kressDiversityDistributionHeliconia1990].
-While some *Heliconia* species grow in large aggregations on roadsides, gaps, and in other disturbed habitats, others - including *H.
-acuminata* - grow primarily in the forest understory [@kressSelfincompatibilitySystemsCentral1983; @ribeiroInfluencePostclearingTreatment2010].
+While some *Heliconia* species grow in large aggregations on roadsides, gaps, and in other disturbed habitats, others, including *H.
+acuminata*, grow primarily in the forest understory [@kressSelfincompatibilitySystemsCentral1983; @ribeiroInfluencePostclearingTreatment2010].
 Understory *Heliconia* species produce fewer flowers and are pollinated by traplining hummingbirds [@brunaHeliconiaAcuminataReproductive2004; @stoufferForestFragmentationSeasonal1996].
@@
 
-*Heliconia acuminata* in our site begin flowering early in the rainy season (e.g., January) and most reproductive plants produce a single infloresence (range = 1--7) with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
+The rainy season in our sites is typically from late December through late May.
+*Heliconia acuminata* in our site begin flowering early in the rainy season and most reproductive plants produce a single infloresence (range = 1--7) with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
 Fruits mature April-May and have 1--3 seeds per fruit ($\bar{x}=2$) that are dispersed by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011].
 Dispersed seeds germinate approximately 6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@brunaEffectsForestFragmentation2002; @brunaSeedGerminationRainforest1999].
-On average plots in CF also had more than twice as many plants as the plots in 1-ha fragments (CF median = 788, range = (201-1549); 1-ha median = 339, range = (297-400)).
+On average, plots in CF also had more than twice as many plants as the plots in 1-ha fragments (CF median = 788, range = (201-1549); 1-ha median = 339, range = (297-400)).
 
@@
 
-We projected the growth rate and structure of *Heliconia acuminata* with three classes of IPMs - Deterministic, Stochastic, and Stochastic with Lagged Environmental Effects.
+We projected the growth rate and structure of *Heliconia acuminata* with three classes of IPMs: Deterministic, Stochastic, and Stochastic with Lagged Environmental Effects.
 Each of these IPMs required different functional forms of the underlying vital rate functions used to describe the *H.
@@
 We used the annual census data [@brunaArePlantPopulations2003] to fit the deterministic vital rate functions for growth, survival, and flowering in each habitat type (i.e., Fragments, Continuous Forest; the data from all plots within a habitat class were pooled to create a single 'summary population' for the CF and FF habitats [@brunaArePlantPopulations2003].
-For established plants these were modeled as a smooth function of size in the previous census with generalized additive models (GAMs) fit with the `mgcv` library [@mgcv] for the R statistical programming language [@rcoreteam2020].<!--`r version$version.string`---> <!--# probably include a supplement with detail on things like basis, number of knots, etc.
+For established plants these were modeled as a smooth function of size in the previous census with generalized additive models (GAMs) fit with the `mgcv` package [@mgcv] for the R statistical programming language [@rcoreteam2020].<!--`r version$version.string`---> <!--# probably include a supplement with detail on things like basis, number of knots, etc.
 --> For consistency, seedling survival and growth were also modeled using GAMs, but without size in the previous census as a predictor (i.e.
@@
 These values were then used to predict the fitted values from vital rate models, which generated different kernels for each iteration of the IPM.
-Though the kernels of successive iterations are not entirely independent -- the SPEI values used to calculate vital rates include values used in the previous two iterations -- they are ergodic.\
+The kernels of successive iterations are not entirely independent --- the SPEI values used to calculate vital rates include values used in the previous two iterations --- but they are ergodic.\
 
@@
 fragments), and the interaction of IPM type and Habitat.
-Because the response variable ($\lambda$) was continuous we used the Gaussian distribution with the identity function in our model.
+Because the response variable ($\lambda$) was continuous we used the Gaussian distribution with the identity link function in our model.
 We verified the model assumptions by plotting residuals versus fitted values.\
@@
 Because Shapiro-Wilk tests indicated that the distributions of proportions were not normally distributed, we used the non-parametric Ansari-Bradley Test to compare the distribution of proportions in each habitat x stage combination.
-These analyses were conducted using the R libraries `rstatix` and `vartest`, respectively [@kassambaraRstatixPipefriendlyFramework2023; @cosarVartestTestsVariance2024].
+These analyses were conducted using the R packages `rstatix` and `vartest`, respectively [@kassambaraRstatixPipefriendlyFramework2023; @cosarVartestTestsVariance2024].
 
@@
 There were also significant effects of IPM Type, Habitat, and their interaction on $\lambda$ (@tbl-glm).
-Deterministic $\lambda$ was nearly identical to the average $\lambda$ from kernel-resampled IPMs (@fig-lambdas), which is somewhat surpring given that stochasticity is predicted to reduce population growth rates [@doakCorrectlyEstimatingHow2005a; @tuljapurkarManyGrowthRates2003; @metcalfStatisticalModellingAnnual2015].
-However including lagged effects in IPMs resulted in projections of $\lambda$ that were on average 1.5-2% lower than those of deterministic models, with 70-76% of projections below deterministic $\lambda$ (@fig-lambdas).
-The IPMs with lagged effects also appear to be more accurate - the underlying vital rate models for IPMs with lagged effects all had the best fit to the survey data (dAIC = 0, @tbl-aic).
+Because stochasticity is predicted to reduce population growth rates [@doakCorrectlyEstimatingHow2005a; @tuljapurkarManyGrowthRates2003; @metcalfStatisticalModellingAnnual2015], we were surprised to fine that deterministic $\lambda$ was nearly identical to the average $\lambda$ from kernel-resampled IPMs (@fig-lambdas).
+However, we did see reductions in population growth rate when including lagged effects in IPMs: on average the projections of $\lambda$ from these models were 1.5-2% lower, with 70-76% of the values below deterministic $\lambda$ (@fig-lambdas).
+The IPMs with lagged effects also appear to be more accurate --- the underlying vital rate models for IPMs with lagged effects all had the best fit to the survey data (dAIC = 0, @tbl-aic).
 <!--# No longer true if random effect = different smooth for each year -->
@@
 What are the implications of these results for the use of IPMs to study plant demography?
-First, they suggest that failing to include lagged effects in models could result in underestimates of $\lambda$.
-While relative differences in $\lambda$ may be sufficient for some tests of theory, they may be critical in conservation or management settings because even small differences in $\lambda$ can have major consequences for population dynamics.
-They could also be critical for studies assessing the effects of climate change, given many DLHE are climate driven.
-That said, there is a cost to including lagged effects in models, even in if they are more accurate.
+First, they suggest that failing to include lagged effects in models could significantly underestimate the growth rate of populations.
+While relative differences in projections of $\lambda$ may be sufficient for some tests of theory, the actual value of projections may be critical in conservation or management settings.
+For example, they could be used to identify which populations face the greatest risk of extinction (e.g., $\lambda=0.99$ vs.
+$\lambda=1.01$) or --- because even small differences in $\lambda$ can have major short-term consequences for population dynamics (e.g., $\lambda=1.01$ vs.
+$\lambda=1.02$) - be used to chose among management strategies.
+Moreover, including lagged effects in models could be especially critical for studies assessing the effects of climate change, given many DLHE are climate driven.
+That said, there is a cost to including lagged effects in demographic models, even when doing so is conceptually appropriate or if the outcome is potentially more accurate.
 While Deterministic and Kernel-resampled Stochastic IPMs took only \~`r det_time` and \~`r sker_time` min to iterate (respectively), the Stochastic IPMs with using parameter-resampled kernels and lagged effects took \~`r spar_time` min.
-This is largely due to the required computational resources and algorithms (i.e., `predict()`), which are much slower forGeneral Additive Models (i.e., GAMs) with 2D smooths because of the much higher number of knots than for the GAMs used in Determinsitic and Kernel-resampled IPMs.
+This is largely due to the required computational resources and algorithms (i.e., `predict()`), which are much slower for Generalized Additive Models with 2D smooths because of the much higher number of knots than for the GAMs used in Determinsitic and Kernel-resampled IPMs.
 Advances in computational power and access to high-performance computing resources could lower this cost.
 
-Computational power cannot compensate for limited data, however.
-Detecting lagged effects and evaluating their consequences requires long-term demographic data - data that are only available for a relatively small number of species, few of which are in the tropics.
+That said, computational power cannot compensate for limited data.
+Detecting lagged effects and evaluating their consequences requires long-term demographic data --- data that are only available for a relatively small number of species, few of which are in the tropics.
 The increasing evidence that lagged effects are ubiquitous, and that they can have major demographic impacts, underscores the need to support the collection of such long-term data, the complementary development of experimental and statistical approaches to disentangling lagged effects, and community driven efforts to identify priority or model systems for in-depth investigation.\
@@
 <!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
+
+<!-- Figure 2 Guides = collect not working?  -->
+
+<!-- TODO: see notes table 5 -->
 
@@
 ```
-
-<!-- \newpage -->
-
-<!-- ```{r #fig-var,  echo=FALSE, out.width="85%",fig.align='center'} -->
-
-<!-- #| fig.cap = "------" -->
-
-<!-- # The figures is currently generated with "plotting_lambdas.R, need to move to targets_r  -->
-
-<!-- path <- here("docs","figures","relative_variances_fig.png") -->
-
-<!-- knitr::include_graphics(path, rel_path = TRUE) -->
-
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r #fig-marginal-long,  echo=FALSE, out.width="85%",fig.align='center'} -->
-
-<!-- #| fig.cap = "The relative proportion of the population in each size class (FF vs.
-CF) for each of 250 iterations of the kernel-resampled (dark shading) and parameter-resampled IPM models (light shading).
-Values on the 1-1 line indicate an iteration where CF and FF have the same proportion of the population in a given size class; the marginal plots indicate the distribution of these relative proportions for each class of IPM in each habitat.
-Note that both the scatterplots and marginal plots exclude transient dynamics (iterations 1-30)." -->
-
-<!-- # The figures is currently generated with "plotting_lambdas.R, need to move to targets_r  -->
-
-<!-- path <- here("docs","figures","marginal_stage_fig_long.png") -->
-
-<!-- knitr::include_graphics(path, rel_path = TRUE) -->
-
-<!-- ``` -->
 

### 2025-09-08 R/plot_pop_states.R

@@ -1,5 +1,6 @@
 # library(targets)
 # library(ipmr)
+# library(patchwork)
 # tar_load(c(
 #   ipm_det_cf,
 #   ipm_det_ff,
@@ -8,7 +9,7 @@
 #   ipm_dlnm_cf,
 #   ipm_dlnm_ff
 # ))
-
+# 
 
 
 #' Plot population stage structure over time
@@ -32,15 +33,15 @@
 #'   Otherwise `save_path` is returned.
 #'
 #' @examples
-#' list(
-#'   det_cf = ipm_det_cf,
-#'   det_ff = ipm_det_ff,
-#'   stoch_cf = ipm_stoch_cf,
-#'   stoch_ff = ipm_stoch_ff,
-#'   dlnm_cf = ipm_dlnm_cf,
-#'   dlnm_ff = ipm_dlnm_ff
-#' ) %>%
-#'   plot_pop_states(save_path = "test.png", height = 5)
+# list(
+#   det_cf = ipm_det_cf,
+#   det_ff = ipm_det_ff,
+#   stoch_cf = ipm_stoch_cf,
+#   stoch_ff = ipm_stoch_ff,
+#   dlnm_cf = ipm_dlnm_cf,
+#   dlnm_ff = ipm_dlnm_ff
+# ) %>%
+#   plot_pop_states(save_path = "test.png", height = 5)
 #'   
 plot_pop_states <- function(ipm_list, xlim = c(0, 250), save_path = NULL, ...) {
   df <-

### 2025-09-08 R/plotting_lambdas.R

@@ -103,8 +103,8 @@ lambdas <- lambdas %>%
 
 # lambda stats ------------------------------------------------------------
 unique(lambdas$ipm)
-unique(cf_stoch$ipm)
-summary(cf_stoch)
+# unique(cf_stoch$ipm)
+# summary(cf_stoch)
 
 stoch_lambdas<-lambdas %>% 
   dplyr::filter(ipm=="stoch_cf"|
@@ -256,7 +256,7 @@ lambdas_fig <- ggplot(plot_data2, aes(x=lambda)) +
                       # alpha=0.9), 
                   fill="darkgreen",
                  color="gray20") +
-  scale_y_continuous(n.breaks = 10, limits = c(-55, 55))+
+  scale_y_continuous(labels=abs,n.breaks = 10, limits = c(-55, 55))+
   # scale_x_continuous(n.breaks = 20, limits = c(0.8, 1.06))+
   scale_x_continuous(breaks = seq(0.8,1.06,by=.02))+
   annotate("text", x = 0.87, y = 20, label = "Continuous Forest", color="darkgreen")+
@@ -413,9 +413,6 @@ geom_text(data = data.frame(x = 0.985,
           aes(x = x, y = y, label = label), 
           parse = TRUE, 
           size = 4)
-  
-  
-  
   
   
 lambdas_fig

### 2025-10-28 docs/ipm_comparison_ms.qmd
edits figs and bib

@@
 
-Current environmental conditions can have large and immediate effects on the growth, survival, or reproduction of long-lived organisms.
-There is also mounting evidence, however, for intervals that range from months to years between environmental conditions and the resulting changes in demographic vital rates.
-These *Lagged Effects*, also known as *Delayed Life-history Events* (i.e., DLHEs) [@beckermanPopulationDynamicConsequences2002], can simultaneously affect an entire cohort (e.g., juveniles hatching during a period of scarcity will all have delayed maturation and lower lifetime fecundity) or only a subset of the population (e.g., cold temperatures in one year lead to reduced flowering by potentially reproductive individuals in the next).
-In addition, the temporal delay between an environmental event and changes in demographic vital rates depends on both the intensity of the event and its timing relative to the underlying physiological processes [@eversLaggedDormantSeason2021; @crileyYearProductionHigh1994].
-A drought during the early stages of gestation or floral bud formation, for example, might have a much larger impact on the number of fruits or offspring produced than one at a later stage of the reproductive process.
+Current environmental conditions can have immediate effects on the growth, survival, or reproduction of long-lived organisms.
+However, there is mounting evidence for potential delays of months or even years between current environmental conditions and resulting changes in demographic vital rates.
+These *Lagged Effects*, also known as *Delayed Life-history Events* [i.e.
+DLHEs, @beckermanPopulationDynamicConsequences2002], can simultaneously affect an entire cohort (e.g., all juveniles hatching during a period of scarcity will have delayed maturation) or only a subset of the population (e.g., cold winter temperatures reduce flowering by small trees but not large ones).
+Moreover, the temporal delay between an environmental event and changes in demographic vital rates depends on both the intensity of the event and its timing relative to the underlying physiological processes [@eversLaggedDormantSeason2021; @crileyYearProductionHigh1994].
+For example, a drought during the early stages of gestation or floral bud formation might have a much larger impact on the number of fruits or offspring produced than one at a later stage of the reproductive process.
 The delay or intensity of lagged effects can also be location- or habitat-specific, with individuals in some sites or habitats buffered against, or able to recover more quickly from, the delayed effects of environmental variation.
@@
 Because Lagged Effects are often directly linked to reproduction and survival, it has been argued they could have major but underappreciated consequences for population dynamics [@beckermanPopulationDynamicConsequences2002].
-While recent studies suggest this could indeed the case [e.g., @molowny-horasChangesNaturalDynamics2017; @tenhumbergTimelaggedEffectsWeather2018; @williamsLifeHistoryEvolution2015], broader efforts to test this hypothesis have face a number of obstacles [@metcalfStatisticalModellingAnnual2015].
-Empirical tests of putative lagged effects are rare because these are challenging to design, implement, and maintain [@kussEvolutionaryDemographyLonglived2008a].
+While recent studies suggest this could indeed the case [e.g., @williamsLifeHistoryEvolution2015; @molowny-horasChangesNaturalDynamics2017; @tenhumbergTimelaggedEffectsWeather2018], efforts to test this hypothesis are rare.
+[@metcalfStatisticalModellingAnnual2015].
+In part this is because empirical tests of putative lagged effects are are challenging to design, implement, and maintain [@kussEvolutionaryDemographyLonglived2008a].
 While one could conceivably use observational data to detect lagged effects, doing so requires long-term data on both the putative lagged effect (e.g., reduced probability of flowering) and its potential environmental drivers (e.g., drought during early floral development), and such coupled data sets are rare [sensu @eversLaggedDormantSeason2021].
@@
 For example, many of the potentially applicable statistical methods have stringent assumptions and data requirements rarely met by ecological data [@metcalfStatisticalModellingAnnual2015], while the including complex biological processes in demographic models can render them less tractable or amenable to analysis using currently available tools.
-Addressing these challenges is a major undertaking; the value of doing so will depend on the effort required vs.
+Addressing these challenges is a major undertaking, and the value of doing so will depend on the effort required vs.
 the potential consequences of failing to consider lagged effects --- consequences that range from overestimating projections of population growth rate (i.e., $\lambda$) in a conservation setting to drawing invalid conclusions regarding support for the predictions of ecological or evolutionary theory.
@@
 Mathematical and statistical advances [e.g., @williams2012avoiding; @brooksStatisticalModelingPatterns2019] have rapidly expanded the scope of questions and biological processes that can be investigated with these models [e.g., @metcalfStatisticalModellingAnnual2015; @reesEvolvingIntegralProjection2016; @ellnerDatadrivenModellingStructured2016].
-Here we investigate how including lagged effects in Integral Projection Models influences projections of $\lambda$ and population structure.
+Here we test predictions for how including lagged effects in Integral Projection Models will influence projections of $\lambda$ and population structure.
 
-We have previously shown that the effects of precipitation extremes on the demographic vital rates of an Amazonian understory herb (*Heliconia acuminata*, Heliconiaceae) can be delayed up to 36 months [@scottDelayedEffectsClimate2022], with the presence and duration of these lagged effects varying by vital rate and habitat.
-We parameterized three classes of Integral Projection Models --- a deterministic IPM, a stochastic IPM, and a stochastic IPM with lagged effects of precipitation on vital rates --- for populations in two habitat types (i.e., continuous forest vs.
+We have previously shown that the effects of precipitation extremes on the demographic vital rates of an Amazonian understory herb (*Heliconia acuminata*, Heliconiaceae) can be delayed up to 36 months [@scottDelayedEffectsClimate2022], with the presence and duration of these lagged effects varying by vital rate and habitat (i.e., continuous forest vs.
 forest fragments).
-Based on previous studies [@brunaArePlantPopulations2003; @brunaDemographicEffectsHabitat2005; @brunaHabitatFragmentationDemographic2002] and demographic theory [@caswell2001; @tuljapurkarPopulationDynamicsVariable1990] we predicted that: (i) projections of $\lambda$ from deterministic models would be higher than those of stochastic models, and that (ii) $\lambda$ is lower for Forest Fragment than Continuous Forest populations regardless of model, but the difference between the two habitats is greatest for the IPM with lagged effects.
+We parameterized three classes of Integral Projection Models --- a deterministic IPM, a stochastic IPM, and a stochastic IPM with lagged effects of precipitation on vital rates --- for populations in these two habitat types to test the following predictions from demographic theory [@caswell2001; @tuljapurkarPopulationDynamicsVariable1990]: (i) projections of $\lambda$ from deterministic models would be higher than those of stochastic models, and that (ii) $\lambda$ is lower for Forest Fragment than Continuous Forest populations regardless of model, but the difference between the two habitats is greatest for the IPM with lagged effects.
 
@@
 Dispersed seeds germinate approximately 6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@brunaEffectsForestFragmentation2002; @brunaSeedGerminationRainforest1999].
-On average, plots in CF also had more than twice as many plants as the plots in 1-ha fragments (CF median = 788, range = (201-1549); 1-ha median = 339, range = (297-400)).
+On average, plots in CF also had more than twice as many plants as the plots in 1-ha fragments (CF: median = 788, range = 201-1549; 1-ha: median = 339, range = 297-400).
 
@@
 
-The second is sub-kernel $R(z^\prime)$, which describes the survival of seedlings established in year $t$ and their size when entering the mature plant population in year $t+1$ (@eq-R):
+\noindent{The} second is sub-kernel $R(z^\prime)$, which describes the survival of seedlings established in year $t$ and their size when entering the mature plant population in year $t+1$ (@eq-R):
 
@@
 
-Note that in @eq-R both the probability that new seedlings survive their first year, $s_s$, and their size at the end of this year, $G(z^\prime, z)$, are size-independent.
+\noindent{Note} that in @eq-R both the probability that new seedlings survive their first year, $s_s$, and their size at the end of this year, $G(z^\prime, z)$, are size-independent.
 IPMs can include transitions between individuals from a discrete state to a continuous one [i.e., from 'seedling' to 'mature plant of size $z'$', @ellnerDatadrivenModellingStructured2016]; we treat seedlings as a distinct and discrete category because they have lower survival and growth in their first year than comparably sized plants [@brunaArePlantPopulations2003; @scottDelayedEffectsClimate2022].\
@@
 
-Both the probability that a mature plant will flower, $p_f(z)$, and the number of seeds a flowering plant will produce, $f(z)$, are size-dependent.
+\noindent{Both} the probability that a mature plant will flower, $p_f(z)$, and the number of seeds a flowering plant will produce, $f(z)$, are size-dependent.
 All seeds germinate and establish as seedlings with probability $g$.\
@@
 We then randomly selected one of these sets of kernels to use in each iteration of the IPM.
-This procedure is equivalent to *'kernel resampling'* [*sensu* @metcalfStatisticalModellingAnnual2015] or matrix selection for matrix population models [@caswell2001].\
+This procedure is equivalent to *'kernel resampling'* [*sensu* @metcalfStatisticalModellingAnnual2015] or matrix selection for matrix population models [@caswell2001; @boyceDemographyIncreasinglyVariable2006a].\
 
@@
 
-##### Comparison of $\lambda$ in Continuous Forest and Forest Fragments.
+### Comparison of $\lambda$ in Continuous Forest and Forest Fragments
 
@@
 
-##### Population Structure.
+### Population Structure
 
@@
 
-Including lagged effects in IPMs also resulted in significantly less predictable projections of population structure, both within (@fig-pop-states a) and across habitats (@fig-pop-states b, @tbl-vartest).
+Including lagged effects in IPMs also resulted in significantly less predictable projections of population structure, both within (@fig-pop-states A) and across habitats (@fig-pop-states B, @tbl-vartest).
 This variability is particularly notable in forest fragments (@fig-marginal, @tbl-meanvars), where we have previously shown lagged effects have very large impacts on growth and survival [@scottDelayedEffectsClimate2022].
@@
 In other words, populations in forest fragments had proportionately less recruitment and fewer individuals growing into the larger, reproductive size classes [@brunaDemographicEffectsHabitat2005].
-This shift towards intermediately sized, pre-reproductive plants in forest fragments is even more dramatic when using the parameter-resampled stochastic IPM (@fig-pop-states b).\
+This shift towards intermediately sized, pre-reproductive plants in forest fragments is even more dramatic when using the parameter-resampled stochastic IPM (@fig-pop-states B).\
 
@@
 For example, they could be used to identify which populations face the greatest risk of extinction (e.g., $\lambda=0.99$ vs.
-$\lambda=1.01$) or --- because even small differences in $\lambda$ can have major short-term consequences for population dynamics (e.g., $\lambda=1.01$ vs.
-$\lambda=1.02$) - be used to chose among management strategies.
+$\lambda=1.01$) or, because even small differences in $\lambda$ can have major short-term consequences for population dynamics (e.g., $\lambda=1.01$ vs.
+$\lambda=1.02$), be used to chose among alternative management strategies.
 Moreover, including lagged effects in models could be especially critical for studies assessing the effects of climate change, given many DLHE are climate driven.
+
 That said, there is a cost to including lagged effects in demographic models, even when doing so is conceptually appropriate or if the outcome is potentially more accurate.
@@
 This is largely due to the required computational resources and algorithms (i.e., `predict()`), which are much slower for Generalized Additive Models with 2D smooths because of the much higher number of knots than for the GAMs used in Determinsitic and Kernel-resampled IPMs.
-Advances in computational power and access to high-performance computing resources could lower this cost.
+Advances in computational power and access to high-performance computing resources could lower this temporal and computational cost.
 
-That said, computational power cannot compensate for limited data.
-Detecting lagged effects and evaluating their consequences requires long-term demographic data --- data that are only available for a relatively small number of species, few of which are in the tropics.
+It is important to emphasize, however, that computational power cannot compensate for limited data.
+Detecting lagged effects and evaluating their consequences requires long-term demographic data --- data that are only available for a relatively small number of species, few of which are in the tropics [@brunaHabitatFragmentationPlant2009; @romerPlantDemographicKnowledge2024].
 The increasing evidence that lagged effects are ubiquitous, and that they can have major demographic impacts, underscores the need to support the collection of such long-term data, the complementary development of experimental and statistical approaches to disentangling lagged effects, and community driven efforts to identify priority or model systems for in-depth investigation.\
@@
 ```{r #fig-pop-states,  echo=FALSE, out.width="85%",fig.align='center'}
-#| fig.cap = "**(A)** The change over time in the the proportion of _Heliconia acuminata_ populations in different size/stage classes when simulating population dyammics in Continuous Forest (CF) or Forest Fragment (FF) with three different Integral Projection Models.
+#| fig.cap = "**(A)** The change over time in the the proportion of _Heliconia acuminata_ populations in different size/stage classes when simulating population dyammics in Continuous Forest (CF) or Forest Fragment (FF) with three different integral projection models (e.g., IPMs).
 Results are shown for the first 250 iterations of populations; for the criteria used to define the size categories see @tbl-plantcats.
@@
 
-```{r #fig-lambdas,  echo=FALSE, out.width="85%",fig.align='center'}
+```{r #fig-lambdas,  echo=FALSE, out.width="100%",fig.align='center'}
 #| fig.cap = "Distribution of 900 values of $\\lambda$ projected with (A) Stochastic, kernel-resampled IPMs and (B) Stochastic, paramenter-resampled IPMs.
@@
 
-```{r #fig-marginal,  echo=FALSE, out.width="85%",fig.align='center'}
+```{r #fig-marginal,  echo=FALSE, out.width="75%",fig.align='center'}
 #| fig.cap = "The relative proportion of the population in each size class (FF vs.
@@
 Those with 'none' were used to build deterministic IPMs; those with a random effect of year were used to build stochastic, kernel-resampled IPMs; and those with a distributed lag non-linear model (DLNM) were used to build stochastic, parameter-resampled IPMs.
-'edf' is the estimated degrees of freedom of the penalized GAM.
+<i>'edf'</i> is the estimated degrees of freedom of the penalized GAM.
 \u0394AIC is calculated within each habitat and vital rate combination.
@@
     #"Model" = "RHS",
-    "edf" = "df",
+    "\\textit{edf}" = "df",
     "\u0394AIC" = "dAIC"
@@
          "$\\lambda$" = est,
-         "(Lower, Upper 95\\% CI)"=CI) %>%
+         "95\\% CI (Lower, Upper)"=CI) %>%
   mutate(IPM = str_replace_all(
@@
   # select(-`Vital Rate`) %>%
-  rename(`Stage`=`Vital Rate`) %>% 
+  rename(`Stage`=`Vital Rate`,
+         "\\textit{df}"=df,
+         "\\textit{P}"=`p value`) %>%  
   mutate(`Stage`= " ") %>% 
@@
   p.value=round(p.value,4)) %>% 
-  mutate(p.value=as.character(p.value)) %>% 
+  mutate(p.value=if_else(p.value<0.001,as.character("< 0.0001"),as.character(p.value))) %>% 
   # mutate(p.value=if_else(p.value< 0.01,"0.001",p.value)) %>% 
@@
          "z value"=statistic,
-         "P value"=p.value) %>% 
+         "\\textit{P}"=p.value) %>% 
   mutate(Term=gsub("[:]"," : ",Term)) %>%

### 2025-10-28 R/plotting_lambdas.R

@@ -170,8 +170,8 @@ means<-plot_data %>%
   mutate(ipm_type = str_replace_all(
     ipm_type,
     c(
-      "lag" = "Stochastic, parameter-resampled",
-      "stoch" = "Stochastic, kernel-resampled",
+      "lag" = "B) Stochastic, parameter-resampled",
+      "stoch" = "A) Stochastic, kernel-resampled",
       "det" = "Deterministic"
     )
   )) %>%
@@ -190,35 +190,35 @@ means<-plot_data %>%
 # 
 # display.brewer.pal(n = 8, name = 'PRGn')
 # brewer.pal(n = 8, name = "PRGn")
-
-fig <- ggplot(plot_data, 
-              aes(x=lambda,
-                  fill=habitat)) +  
-  geom_histogram(binwidth=.0015,
-                 alpha=0.9,
-                 color="gray35") +
-  scale_fill_manual(values=c("#C2A5CF", "#A6DBA0")) +
-  scale_y_continuous(n.breaks = 10, limits = c(0, 250))+
-  # scale_fill_brewer(palette = "PRGn")
-  theme_classic()+
-  facet_grid(rows = vars(ipm_type))+ #,
-             # rows= vars(habitat))+
-  # facet_wrap(~ipm_type)+
-  # geom_vline(data=means, aes(xintercept=mean_l))
-  geom_segment(aes(x = mean_l, 
-                   y = 0, 
-                   xend = mean_l, 
-                   yend = 140,
-                   color=habitat
-                   ),
-               linewidth = 0.8,
-               data = means
-               )+
-  scale_color_manual(values=c("#C2A5CF", "#A6DBA0"))
-fig
-
+# 
+# fig <- ggplot(plot_data, 
+#               aes(x=lambda,
+#                   fill=habitat)) +  
+#   geom_histogram(binwidth=.0015,
+#                  alpha=0.9,
+#                  color="gray35") +
+#   scale_fill_manual(values=c("#C2A5CF", "#A6DBA0")) +
+#   scale_y_continuous(n.breaks = 10, limits = c(0, 250))+
+#   # scale_fill_brewer(palette = "PRGn")
+#   theme_classic()+
+#   facet_grid(rows = vars(ipm_type))+ #,
+#              # rows= vars(habitat))+
+#   # facet_wrap(~ipm_type)+
+#   # geom_vline(data=means, aes(xintercept=mean_l))
+#   geom_segment(aes(x = mean_l, 
+#                    y = 0, 
+#                    xend = mean_l, 
[…]

### 2025-10-28 R/plot_pop_states.R
not entirely sure...

@@ -54,10 +54,10 @@ plot_pop_states <- function(ipm_list, xlim = c(0, 250), save_path = NULL, ...) {
         ipm,
         c(
           "det" = "Deterministic",
-          "stoch" = "Kernel-resampled",
-          "dlnm" = "Parameter-resampled"
+          "stoch" = "Stochastic",
+          "dlnm" = "Lagged effects"
         )
-      ),
+      ) %>% fct_inorder(),
       habitat = toupper(habitat)
     ) %>% 
     rename(IPM = ipm, Habitat = habitat)

### 2025-10-28 notes/lambda plot ggdist.R

@@ -1,0 +1,58 @@
+library(targets)
+library(tidyverse)
+library(ggdist)
+
+tar_load(starts_with("lambda_"))
+
+df <-
+  tibble(
+   det_ff = lambda_bt_det_ff,
+   det_cf = lambda_bt_det_cf,
+   stoch_ff = lambda_bt_stoch_ff,
+   stoch_cf = lambda_bt_stoch_cf,
+   dlnm_ff = lambda_bt_dlnm_ff,
+   dlnm_cf = lambda_bt_dlnm_cf
+ ) %>% 
+    pivot_longer(everything(), 
+                 names_to = c("ipm", "habitat"),
+                 names_sep = "_", 
+                 values_to = "lambda") %>% 
+  mutate(
+    ipm = fct_relevel(ipm, "det", "stoch", "dlnm") %>%
+      fct_recode(Deterministic = "det", Stochastic = "stoch", `Lagged effects` = "dlnm")
+  ) %>% 
+  mutate(
+    habitat = toupper(habitat)
+  )
+
+
+p <-
+  ggplot(df, aes(x = lambda, fill = habitat)) +
+  stat_gradientinterval(
+    aes(slab_alpha = stat(-pmax(abs(1-2*cdf),.95))),
+    .width = c(0.83, 0.95), #multiple quantiles can be specified for the interval (black lines)
+    fill_type = "auto",
+    point_interval = "mean_qi"
+    ) +
+  geom_vline(xintercept = 1) +
+  facet_wrap(~ipm, ncol = 1) +
+  scale_x_continuous(breaks = scales::extended_breaks(n = 10)) +
+  scale_fill_manual(values = c(FF = "#E66101", CF = "#5E3C99")) +
+  theme_bw() +
+  theme(
+    text = element_text(size = 8),
+    axis.title.y = element_blank(),
+    axis.text.y = element_blank(),
+    axis.ticks.y = element_blank(),
+    panel.grid.major.y = element_blank(),
+    panel.grid.minor.y = element_blank(),
+    plot.caption.position = "plot"
+  ) +
+  labs(
+    x = "Population Growth Rate (lambda)",
+    fill = "Habitat",
+    caption = "Mean (point) ± 83% (thick) and 95% (thin) CIs\n from 500 bootstrapped simulations"
+  )
+p
+ggsave("test.png", width = 950, height = 975, units = "px")
+
